### 2024-04-02 Chapters/Chapter0.tex
Initial commit

% Chapter Template

\chapter{Chapter Title Here} % Main chapter title

\label{ChapterX} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}

%----------------------------------------------------------------------------------------
%	SECTION 1
%----------------------------------------------------------------------------------------

\section{Main Section 1}

The decision was made to construct a NN, as they are good at discerning patterns, to try and predict wind gusts. Several NNs were constructed to compare their predictive capabilities.
%-----------------------------------
%	SUBSECTION 1
%-----------------------------------
\subsection{Subsection 1}

Nunc posuere quam at lectus tristique eu ultrices augue venenatis. Vestibulum ante ipsum primis in faucibus orci luctus et ultrices posuere cubilia Curae; Aliquam erat volutpat. Vivamus sodales tortor eget quam adipiscing in vulputate ante ullamcorper. Sed eros ante, lacinia et sollicitudin et, aliquam sit amet augue. In hac habitasse platea dictumst.

%-----------------------------------
%	SUBSECTION 2
%-----------------------------------

\subsection{Subsection 2}
Morbi rutrum odio eget arcu adipiscing sodales. Aenean et purus a est pulvinar pellentesque. Cras in elit neque, quis varius elit. Phasellus fringilla, nibh eu tempus venenatis, dolor elit posuere quam, quis adipiscing urna leo nec orci. Sed nec nulla auctor odio aliquet consequat. Ut nec nulla in ante ullamcorper aliquam at sed dolor. Phasellus fermentum magna in augue gravida cursus. Cras sed pretium lorem. Pellentesque eget ornare odio. Proin accumsan, massa viverra cursus pharetra, ipsum nisi lobortis velit, a malesuada dolor lorem eu neque.

%----------------------------------------------------------------------------------------
%	SECTION 2
%----------------------------------------------------------------------------------------

\section{Main Section 2}

Sed ullamcorper quam eu nisl interdum at interdum enim egestas. Aliquam placerat justo sed lectus lobortis ut porta nisl porttitor. Vestibulum mi dolor, lacinia molestie gravida at, tempus vitae ligula. Donec eget quam sapien, in viverra eros. Donec pellentesque justo a massa fringilla non vestibulum metus vestibulum. Vestibulum in orci quis felis tempor lacinia. Vivamus ornare ultrices facilisis. Ut hendrerit volutpat vulputate. Morbi condimentum venenatis augue, id porta ipsum vulputate in. Curabitur luctus tempus justo. Vestibulum risus lectus, adipiscing nec condimentum quis, condimentum nec nisl. Aliquam dictum sagittis velit sed iaculis. Morbi tristique augue sit amet nulla pulvinar id facilisis ligula mollis. Nam elit libero, tincidunt ut aliquam at, molestie in quam. Aenean rhoncus vehicula hendrerit.

### 2024-04-02 Chapters/Chapter1.tex
Initial commit

% Chapter 1

\chapter{Introduction} % Main chapter title

\label{Chapter1} % For referencing the chapter elsewhere, use \ref{Chapter1}

%----------------------------------------------------------------------------------------

% Define some commands to keep the formatting separated from the content 
\newcommand{\keyword}[1]{\textbf{#1}}
\newcommand{\tabhead}[1]{\textbf{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\file}[1]{\texttt{\bfseries#1}}
\newcommand{\option}[1]{\texttt{\itshape#1}}

%----------------------------------------------------------------------------------------
Wind gusts are brief increase in wind speed (lasting seconds) as compared to mean wind speed. The gust factor is defined as the peak gust divided by the mean wind speed over some defined time period. The peak wind gust is often defined as the highest 3 second rolling average measured wind speed over a period of 10 minutes, while the mean wind is the average of all measurements in the 10 minute interval. This varies, with the US using a 1 minute interval, leading to 14\% higher results \cite{why_wind_gusts}. As the Navier Stokes Equation (\ref{eq:navierstokes}) shows the variability of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth, and resistance. It states that the change in wind is described by the pressure gradient, the oscillation, the acceleration due to gravity and the resistance from the landscape.

\begin{equation}
    \label{eq:navierstokes}
    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{pressure} -\overbrace{ f\mathbf{k}x\mathbf{V}}^{oscillation} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{resistance}
\end{equation}

Equation \ref{eq:navierstokes} relates the change in wind w.r.t. time ($\frac{\delta \boldsymbol{V}}{\delta t}$) and change in wind w.r.t. position ($\nabla\boldsymbol{V}$) to pressure change w.r.t. position ($\nabla P$). The wind is also influenced by the oscillation, gravity and resistance but the main drive of wind is pressure gradient\cite{uncertainties_in_numerical_weather_predictions}.

Traditionally, numerical weather prediction (NWP) systems are used to forecast and analyze weather patterns\cite{medium_range_3d_weather_forecasting_NN}. These models describe the transition between discretized packages of atmospheric states using partial differential equations based on physical reality. These results are usually published for 3-6 hour intervals.They describe the state over the period and so do not necessarily grasp fluctuations well. These fluctuations would include fluctuations in the wind speed, wind gusts\cite{canNNBeatNWPw}.

This thesis looks at how best to predict gust factor based on various factors, using several different data sources, including NWP and observations. NWP forecasts or outputs analysis, traditionally, at intervals of around 3-6 hours and thus, alone, do not capture the variability that is the essence of wind gusts. This is why another model that uses the information produced by the NWP along with observations and landscape information, is needed to try to accurately predict wind gusts. Being able to accurately predict the wind gust is important as it is often the peak wind gusts that will cause failures in structures. A problem that will become increasingly prevalent in the near future\cite{nasa_extreme_weather}.

%This thesis looks at how best to predict gust factor based on wind speed, wind direction, pressure and temperature at different heights above ground, as well as the elevation of the landscape over an area in the direction the wind is coming from. A neural network (NN) was constructed to predict the gust factor.

\section{Background}
 The history of numerical weather predictions goes all the way back to the 1920's when Lewis Fry Richardson pioneered the field and tried to produce forecasts. The results were flawed and impractical because of the time it took to calculate was much longer than the forecast period. In the 1950's, with the advent of computers we get the first operational forecasts. In September of 1954, Rossby and his Stockholm based team produced the first real-time barotropic forecasts. The next year the Joint Numerical Weather Prediction Unit (JNWPU), based in Princeton New Jersey, released their first forecasts. These forecasts were for 36 hours at 400, 700 and 900 mb. The results were inferior to subjective human-based forecasts but showed that such forecasts were feasible and promoted further development in the area \cite{historyNWP}. The field underwent significant transformation over the next 50 years, and predictions got significantly better.

Currently there has been another transformation in the field of weather prediction driven by artificial intelligence. Interest in AI has come in waves. Some progress is made, then interest dwindles. The interest has been increasing steadily since 2010. Notable work that has driven this wave of interest include increase in computational abilities due to parallel processing in graphical processing units (GPU), convolutional neural networks (CNN), which allowed much faster processing of massive (image) datasets and the availability of large datasets online. It is to be noted that images are grid data with some number of channels. Using CNNs could work on any gridded data where there is some spatial features\cite{canNNBeatNWP}. Since 2018, there has been significant work done in the weather prediction field using AI. In 2018, Dueben and Bauer showed that you can build a NN that can outperform a simple persistence forecast and is competitive with very coarse-resolution atmosphere models of similar complexity for short lead times\cite{dueben2018}. Also in 2018, Scher created a deep convolutional neural network (CNN) to emulate a general circulation model (GCM, a numerical model representing the physical processes), training on the GCM which allows it to emulate the dynamics of the model and maintain stability for much longer than Dueben\cite{scher2018}. These two papers were more proof of concept rather than production ready models to replace NWP. They showed that models based on deep learning might, with further development, compete with standard models in the field.

In the last two years there have been even more developments with the emergence of Large AI Weather forecast Models (LWM). In 2024 Ling et al. tried to standardize the definition of LWM in meteorology and came up with 3 rules that need to be met to count as LWM.

\begin{enumerate}[label = Rule \arabic*:]
    \item Large Parameter Count. The number of parameters can vary wildly but a general range might be from tens of millions to billions of parameters
    \item Large Number of Predictands: predicting on different levels (such as pressure levels or height levels) and offering detailed information on the atmospheric vertical structure and surface conditions
    \item Scalability and downstream applicability. This might crystallize in predicting cyclones. Often, the teams responsible for creating these models try to show their applicability to predict cyclones when not trained specifically on cyclone data (e.g.  \href{https://www.youtube.com/watch?v=PD1v5PCJs_o&ab_channel=GregBronevetsky}{GraphCast})\cite{SecondRevolution}. This is done to show the versability of the models.
\end{enumerate}

Before 2022, LWM had been shown to be able to compete with traditional NWP for some specifc cases as well as making predictions quicker, after training. No model had shown that it could in any way completely replace the traditional systems. In early 2022, Pathak et al. presented FourCastNet. FourCastNet uses an Adaptive Fourier Neural Operator model that leverages transfomer architecture rather than the popular convolutional model architechture. FourCastNet matches the performance of standard forecasting techniques at short lead times for large-scale variables and outperforms for smaller variables. It generates a week-long forecast in less than 2 seconds, orders of magnitude faster than standard physical methods\cite{FourCastNet}. In 2022, machine learning methods were presented that made predictions much faster than traditional NWP, after a one time training (or at least training that wouldn't have to be redone often). These were in some cases performing better than NWP.  In 2023, Remi Lam and the GraphCast team at Google introduced GraphCast. This model was able to outperform the industry standard High Resolution Forecast (HRES) produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This model as the name suggests leverages graphing connections rather than traditional grid like data structure. The base data is given in latitude and longitude degrees at a resolution of 0.25 degrees. This means points are closer to each other at the poles. Using the graphing structure is supposed to help with bias incurred as a result of this\cite{GraphCast}.

There has been a lot of progress made over the last 6 years (since 2018) and especially in the last 2 years (since 2022). The progression from machine learning methods being an interesting idea in the field of numerical weather predictions, to outperforming the standard NWP has been remarkably quick. Two years ago, machine learning methods were able to predict quickly and in some niche cases outperform traditional models. They were not generally competitive with standard weather models. Now they are competitive. It will be very interesting to watch what the next few years will have in store for the development of machine learning in weather predictions.

\section{Methodology and related work}

This study looks at data from three sources and an attempt is made to predict the gust factor in a given place in Iceland. It uses reanalysis data, along with elevation data to predict the gust factor. It looks at the data at the point of interest and does not look at the data as a time series. This thesis tries to improve on the baseline model of always predicting the mean and show that some structure can be learned from reanalysis data about gusts. To do this a neural network was created. The goal of this neural network was to predict the gust factor and improve upon a baseline of always guessing the mean gust factor. This baseline model gave an average error of around 10\%. Any significant improvement on this would indicate that the final model has something to contribute.

In 2004, H.Ágústsson and H. Ólafsson looked at the variability of gust factor in complex landscapes. They looked at data from automatic weather stations that measure wind at 10 meters above ground. A common definition of wind gust is the average 1 second measured wind over three seconds. This is a rolling average. A common definition of average wind is the average of the measured wind over 10 minutes. These are the definitions used both by H. Ágústsson and H. Ólafsson in 2004 as well as this study. The data that was studied in 2004 comes from the same source as used in this study, but limits itself to a smaller section. They only looked at the years 1999-2001. They looked at three factors and how these effected the gust factor. These were $d_m, D, H$, that is direction of wind blowing off a mountain, distance to the mountain and the height of the mountain above the weather station. Their main results were these, gust factor is inversely correlated to the distance from a mountain and correlated to the height of the mountain. The study in 2004 looked at the effect of a dominant point upwind. It did not look at the effects of the landscape more broadly. In this study, landscape upwind is looked at along with landscape downwind. Both the upwind and downwind landscape can influence the gustiness at a certain point \cite{GNP_vidtal}.

To be able to capture the patterns in the data we construct a neural network. The choice to select a neural network is one that is made because they are known to capture well patterns in complex data and can be applied to varied problems. This comes in handy when training on different types of data. It is also easy to construct different types of neural networks and see how they fit well with parts of the dataset. To measure the performance of these models, both to train and test, mean absolute percentage error was used. This measures the average percentage error. This was chosen because the target is the gust factor (the wind gust over the average wind). If the target would have been the wind gust rather than the gust factor then something like mean absolute error might be more appropriate.

### 2024-04-02 Chapters/Chapter2.tex
Initial commit

% Chapter Template

\chapter{Data gathering and preprocessing} % Main chapter title

\label{Chapter3} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}

%----------------------------------------------------------------------------------------
%	SECTION 1
%----------------------------------------------------------------------------------------

Data was sourced from several streams. Veðurstofan provided recordings from weather stations all around the Iceland. Data was downloaded from Copernicus Arctic Regional Reanalysis dataset (CARRA). A land elevation model was also provided by Veðurstofan.

\section{Automatic Weather Station Data}

Veðurstofan provided recordings from weather stations all around Iceland. These automatic weather stations (AWS) are at a height of 10 meters above ground level. The information that is provided by these AWS is presented in two different type of documents, hourly and 10 minute documents. The hourly documents are summations of the 10 minute documents, with the exception that any errors (so called nails) still in the 10 minute documents should have been removed from the hourly documents. Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another document), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction. These features for each of the stations go back at least two decades, but do not have the same start date. This paper does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.

\section{Carra Data}
The CARRA dataset goes back to September 1990 and is currently updated monthly, with a latency of 2-3 months. The oldest Vedurstofu data example that fulfills given criteria is from \startDateVedur. This is covered by Carra. A few of the newer points from AWS were not available from Carra when data was gathered. The Carra dataset is available for two regions, west and east. Each of these covers a vastly larger area than the area of interest. It is possible to request a subsection of the area, a rectangular subsection of the area. This leads to having to store a large amount of data, because we cannot simply ask for the specific points of interest. To get the data one has two options. Their web interface or using their API client. Using the API client is the only realistic option here, as we are making thousands of requests for different times.\\

Each request is made by going through all instances of the Veðurstofu data were able to fulfill the requirements (of mean wind speed reaching \averageWindSpeedLimit 
m/s over a 10 minute interval). For each such observation in the Veðurstofa gögn, we need to make two API (calls excluding redundancy). One for the 3 hour interval before and one for the 3 hour interval after. That is if the observation was at 1 pm, we would need Carra data for noon and 3 pm. Date and times were generated automatically from the AWS and requested by calling the client. Using these datapoints, we should be able to bridge and get an estimation for the point of the given weather station. The Carra data contains several types of layers. These are single levels, model levels, height levels, pressure levels. The data for this observation was downloaded from height levels. That is, data was requested at heights of 15, 150, 250 and 500 meters above ground.\\

For each point 4 features were requested, wind speed, wind direction, pressure and temperature. Each of these features needed to be bridged to create data for model to be trained on.

After using this method to request terabytes of data, it was discovered that it is apparently possible to query a specific area. This decreased the size of each file created after a request from around 50 MB to around 2 MB. Downloading and processing this data would take a lot less time. The largest bottleneck in terms of retrieving the data is the request time. That is the time the request is queued and the time it is running before it begins downloading. This could range from almost immediate to 15 minutes or more. This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during the winter and spring, which increased the wait time and sometimes resulted in queries not being responded to. This meant that the time it would take to retrieve the remaining information went from around a day, when the requests were at their quickest, to many months, something that would not be possible given the time frame of the project.

\section{Elevation data}
Veðurstofa provided a tif file containing the elevation of Iceland on a 20 meter by 20 meter grid. This file encompasses Iceland and is around 685 Mb. At first I thought I might have to split the file into subsections and read for each one but that doesn't seem to be the necessary. A good amount of time was spent finding a data structure that was best for lookup when trying to find points within a certain area. The country was divided into 10 parts (as only around 13\% of the file was able to be read into memory at each time as a part of dictionary object) with boundary boxes. A quadtree was constructed for each of the sections, so as to simplify the lookup. A quadtree allows for efficient lookup by searching for points that intersect a given boundary. A quadtree, like the name indicates, is a tree structure. To initialize it you give it a boundary box. This boundary box is then the parent of exactly four nodes and so on, dividing the area in four at each level and allowing quick lookup. This was not necessary as the rasterio package allowed for quick lookup with it's index and the affine transform.

### 2024-04-02 Chapters/Chapter3.tex
Initial commit

% Chapter Template
\chapter{Data processing and structure} % Main chapter title

\label{Chapter3} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}

%----------------------------------------------------------------------------------------
%	SECTION 1
%----------------------------------------------------------------------------------------

\section{Combining data sources}

This project used three main data sources, which need to be queried, filtered and combined to prepare the data for use in the models. When working with hundred of thousands of rows, the efficiency of the code is very important. Iterating through those rows might be necessary at times but will increase the time exponentially as compared to using vectorizing methods were possible. The three data sources were all in a different format. Measurement data from Veðurstofa was in text files, elevation data was in GeoTiff and reanalysis data from Carra was in a GRIB format. To use the data to train, these three data sources needed to be combined into one file. This was done based on the measurement data from the Veðurstofa. A limit was set on the average wind speed and it was used to select measurement points. Along with the average wind speed having to be above a certain limit, to make sure that we were not essentially getting duplicates, we would select only the top wind speed in any given 48 hour period. That is, for a data point to be included it must have been the highest wind speed in the previous and following 24 hours, so as to not describe the same weather multiple times. The data from Veðurstofan was supplied for 10 minute increments, while Carra data is in 3 hour intervals. This means that to use the Carra data to predict the measured values from Veðurstofa, temporal bridging would need to be done. Along with the temporal bridging, we note that the Carra data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the Veðurstofa is given at specific locations. The elevation information was given by a 20 by 20 rectangular grid that covers Iceland. When combining these data sources a selection of bridging needs to occur. The bridging was linear, both temporally and spatially. This might influence the results but was not considered in this study.

The procedure of combining these sources was as follows. We start with the measured data from the AWS, we filter this by using a limit on the average wind speed. The gust factor generally drops with increased wind speed (although not always dependent on the factors such as the landscape \cite{GNP_vidtal}). Even so being able to predict the gust factor is more important for higher average wind speed as there we have the highest wind gusts. After this stripped dataset over every AWS has been created it is used to query the Carra data by using their API. When querying the Carra API we will have to query in such a way that we query for given hours, days, months, years and a given area. That is, if we ask for a given hour, we get that hour for every day that we ask for. Similarly if we ask for a given day, we get that day for every month we ask for. In light of these restraints, it was decided to query month by month. Querying only the days needed (both the days included in the 10 minute measurements and the days needed for bridging if close to midnight) but every hour of the day (midnight, 3 AM, 6 AM, 9 AM, midday, 3 PM, 6 PM and 9 PM) and then take these values and bridge for the 10 minute values. After querying and downloading the data for the height levels and variables requested, bridging is done for the points of interest and values stored in a pandas dataframe. After this is done the downloaded data is discarded and we look to the next month. This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (goes from several terabytes to less than a gigabyte).

The elevation data comes in a GeoTif file that covers Iceland. It is a rectangular grid of resolution 20 meters. For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved. For each point retrieved bridging needs to be done. This is done in a similar manner to the briding of the Carra data. Weights are assigned to each of the four points bounding the point of interest. These weights are then used to calculate the weighted average that represents the bridged value. This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.

\section{Data structure}

Once data has been retrieved for all three sources and processed, including bridging values, it needs to be made ready to use by the model, for both training, validation and test. We start with a dataframe that contains the measured information from AWS. This includes the average wind, the wind gust, wind direction along with the station number and coordinates. When selecting the Carra data we choose certain height levels. That is at which heights we want the reanalysis data. These present as separate lines in the Carra dataframe. We need to combine them so that each line indicates a single observation. When this is done we can combine the AWS Veðurstofu data and Carra reanalysis data on the location and time columns. Lastly we look at the elevation data. A couple of different sections of land around the weather stations have been looked at. We tried a sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point. In any case the points, that represent these sections, was selected like as shown in code listing \ref{code:sectorElevation}. That is we look at the wind direction d and define a range of angles around that direction at some distance from the given point. This means that we get some number of points (equal to the length of angleRange) at a distance from given weather station as defined by length\_rng. This then gives us some sectors at some distances away from the weather station.

\begin{lstlisting}[language = Python, caption = {Sector elevation points generated}, label = code:sectorElevation]
angles = [(angle + (90 - d)) * pi/180 for angle in angleRange]
length_rng = [(exp(i * log(n + 1)/ k) - 1) * 1000 
                for i in range(1, k + 1)]
points = np.array([[(X + l * cos(angle), Y + l * sin(angle))
                    for angle in angles] for l in length_rng])   
\end{lstlisting}

We end up with a dataframe that has measured data from AWS, which gives us our target, reanalysis data from Carra, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data. An example of what the data looks like can be seen in table \ref{table:trainDataExample}.

\begin{table}[h]
    \resizebox{\textwidth}{!}{
    \centering
    \begin{tabular}{c|c|c|c|c|c|c|c}
         Ri\_01 & Ri\_12 & N\_01 & N\_12 & station\_elevation & relative\_corner & PC1 & ...\\\hline
         -1.18e+00 &  2.67e+04 & -8.57e-06 & 6.78e-05 & 3.34e+01 & 2.73e+00 & 1.36e+01 & ... 
    \end{tabular}
    }
    \caption{An example of data structure used with model}
    \label{table:trainDataExample}
\end{table}

Looking at table \ref{table:trainDataExample} we note that the last \nPCA columns represent the principle component analysis of the elevation data. We also note that the first four columns represent two variables that describe the stability of the air. These are the Richardson number (Ri) and Brunt–Väisälä frequency (N). They are calculated using equations \ref{eqn:Ri} and \ref{eqn:N}. They are calculated using reanalysis data about the the weather at two different height levels. Thus $Ri_{01}$ refers to the Richardson number calculated between height levels 0 and 1. Exactly the same notation is used with the Brunt–Väisälä frequency. The height levels used were 15, 250 and 500 meters above the ground.

\begin{equation}
    \label{eqn:Ri}
    Ri = \frac{g \cdot dT \cdot dz}{T_{ave} \cdot dU^2}
\end{equation}

\begin{equation}
    \label{eqn:N}
    N = \sqrt{\frac{g \cdot dT }{T_{ave} \cdot dz}}
\end{equation}

Here, g is the acceleration due to gravity, dT is the temperature difference between the two height levels, dz is the elevation difference, $T_{ave}$ is the average temperature (that is the average of the two temperatures in the height levels) and dU is the wind speed difference between the two height levels. Both of these numbers tell us something about the stability of the air. These are derived factors from the reanalysis data and as such there shouldn't be a significant information gain in calculating the as opposed to having the raw data. Including these factors instead of the every variable might speed up training as well as making the model more easily explainable with the use of Shapley values or other tools for explainability.

### 2024-04-02 Chapters/Chapter4.tex
Initial commit

% Chapter Template

\chapter{Model architecture} % Main chapter title

\label{Chapter3} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}

%----------------------------------------------------------------------------------------
%	SECTION 1
%----------------------------------------------------------------------------------------

\section{Model structure}
The structure of the neural network is such that it contains some number n of  fully connected layers and batch normalization for each layer, along with regularization. All of these layers have the same number of units. The last layer has a dropout of 50\%. In addition to these layers there is one more output layer. This is simply a dense layer with 1 unit. A grid search was performed to figure out the best hyper parameters. These hyperparameters include number of units in each layer, number of epochs to train for, number of layers, batch size, optimizer and penalty to enforce in the regularization. The possible combinations tried can be seen in table \ref{table:gridSearchHyperparamters}.

\begin{table}[h]
    \centering
    \begin{tabular}{c|c}
        Layers &  4, \textcolor{red}{5}, 6, 7, 8, 9\\\hline
        Units &  128, 256, 512\\\hline
        Epochs & 100, 200, 500\\\hline
        Batch size & 32, 64, 128, 256\\\hline
        Optimizers & Adam, RMSprop, Adamax\\\hline
        Penalties & 1, 0.1, 0.01, 0\\\hline
    \end{tabular}
    \caption{Hyperparamter search with best performing combination colored red}
    \label{table:gridSearchHyperparamters}
\end{table}

\section{Feature selection and importance}
Using the three datasources, we train a model. The data from Veðurstofa only represents the ground truth and is not used as part of the training data. We use the gust ($f_g$) and wind speed ($f$) to calculate the target gust factor. From the Carra reanalysis data we get all of our weather information, with variables wind speed, wind direction, temperature, pressure. All of these variables are queried in three height levels (15, 250, 500 meters above ground) at the location of a given weather station. In addition we have the elevation above sea level over Iceland. We select the elevation in a couple of different shapes. A sector upwind, a sector upwind and downwind as well as a circle surrounding a weather station.

To add to these observed values and reanalysis, we calculate two derived variables, the Richardson number and the Brunt-Väisälä frequency. To begin with we simply use all the features we have to train our model along with the derived variables to see. We do this to be able to then use tools such as Shapley to see which features are impacting our predictions. Using Shapley values to see which features, will then allow us to exclude features that don't affect our model prediction and simplify the training data. We thus start with much higher dimensionality of our training data and reduce it as much as we can without impacting our results. This crystallizes in the landscape points. We start with n points that describe the elevation of the landscape around our weather station. This n might be in the hundreds. If we look at concentric circles outward from the weather station, we might have 10 circles (largest with radius 20 km), each with 72 points (5°spacing) and end up with 720 points for the landscape. This might slow down training. In addition, we might imagine that we are more interested in certain important landscape features. There might be a large mountain up- or downwind of our weather station, that might be the most important feature, in addition to some hills and such. To address this we use principal component analysis (PCA) to try to reduce the dimensionality of our landscape data. Another thing that might serve the same purpose is to select some number of m points that have the most increase or decrease in elevation from the weather station along with the coordinates of those points in relation to the weather station and the direction of the wind.

Using all the variables from Carra, along with the two derived factors (Ri, N) and two sectors (upwind and downwind) downscaled to 10 features using PCA we can see the feature importance in figure \ref{fig:ShapleyWaterfallFirstTry}. Looking at the waterfall graph we can see that Richardson number for higher levels is most important.

\begin{figure}[h]
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_bar_nn_256_example100.png}
    \caption{Feature importance of a neural network as described with 5 hidden layers and 256 units in each}
    \label{fig:ShapleyWaterfallFirstTry}
\end{figure}

### 2024-04-02 MSc.tex
Initial commit

\documentclass[a4paper,12pt,twoside,BCOR=10mm]{scrbook}

% UoI MSc thesis template (English) V2.0.4 25.4.2022

% The license of this template is not fully clear, but: This template
% version is based on an earlier LaTeX template that was once available
% on a SENS UGLA page where the author and license are unknown.  As
% that version has obviously been put once online with the intend to be
% used by students, providing and using it should not be problem.

% Helmut Neukirchen https://uni.hi.is/helmut updated that earlier
% template using a tikzpicture-based approach for the title page
% created by Þór Arnar Curtis.

% The included logos are very likely copyrighted by the University of
% Iceland, but
% https://zeroheight.com/1b323cfd9/p/20f1e6-theses/b/29a61a
% states that the document is intended to make it easier for
% students, staff, and print-shops to format the thesis and ensure a
% standardised appearance. So providing them here and using them
% should be legal.

% Search below for "MODIFY THESE LINES ONLY": there you can enter your name, thesis title, etc.

% BibLaTeX is assumed for references. While Overleaf does this
% automatically for you, if you run it locally on the commandline,
% then you need to run first pdflatex MSc.tex, then biber MSc (without
% the .tex extension), and then again pdflatex MSc.tex

% Packages
\usepackage[utf8]{inputenc}
\usepackage[icelandic, english]{babel}
\usepackage{t1enc}
\usepackage{graphicx}
\usepackage[intoc]{nomencl}
\usepackage{enumerate,color}
\usepackage{url}
\usepackage[pdfborder={0 0 0}]{hyperref}
\BeforeTOCHead[toc]{\cleardoublepage\pdfbookmark{\contentsname}{toc}} % Add Table of Contents to PDF "bookmark" table of contents
\usepackage{appendix}
\usepackage{eso-pic}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[sf,normalsize]{subfigure}
\usepackage[format=plain,labelformat=simple,labelsep=colon]{caption}
\usepackage{placeins}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{listings}

% Packages used for title page layout
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{positioning}

\input{variables.tex}

% Blue color according to HÍ corporate design
\convertcolorspec{RGB}{16,9,159}{rgb}\tmphiblue
\definecolor{hiblue}{rgb}\tmphiblue

\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0cm}
\raggedbottom

\setkomafont{captionlabel}{\itshape}
\setkomafont{caption}{\itshape}
\setkomafont{section}{\FloatBarrier\Large}
\setcapwidth{\textwidth}
%\setcapwidth[l]{\textwidth} % The original template had the [l] which leads to a warning that it gets ignored, so to reduce warnings, removed it.
\setcapindent{1em}

\usepackage{lmodern} % Use Latin Modern (instead of the default Computer Modern that is rendered using a bitmap font).
\usepackage{fixcmex} % To fix that Latin Modern large symbol math fonts has by default only one size: https://tex.stackexchange.com/a/621536
% Times new roman font instead of the standard LaTeX fonts: has not been test -- try this on your own risk
%\usepackage[T1]{fontenc}
%\usepackage{mathptmx}

%%%%%%%%%%%%%%%%% Configurations (Useful defaults, but OK to change %%%%%%%%%%%%%%%%%%%
\graphicspath{{figs/}} % Figures in directory figs

% Bibliography

% \usepackage[sort&compress,authoryear]{natbib} % Uncoment if you want to used NatBib instead of BibLaTeX (and comment the bitlatex line below)

\usepackage{biblatex}  % BibLaTeX used for references. 
\usepackage{csquotes} % BibLaTex wants to have context sensitive quotes
\addbibresource{references.bib} %  Name of *.bib file containing references

%%%%%%%%%%% MODIFY THESE LINES ONLY %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Some note on advisor(s) vs. thesis committee: At the School of
% Engineering and Natural Sciences, according to Regulation
% no. 994-2017
% https://english.hi.is/university_school_of_engineering_and_natural_sciences/regulation_no_994_2017_on_masters_study_at
% Article 5., a student has an administrative supervisor who is
% typically also the academic supervisor, i.e. supervising the thesis.
% If someone from outside is supervising, than the person becomes the
% academic supervisor and you have in addition the administrative
% supervisor from within HÍ.
% In addition, according to Article 7., there is a Master's degree
% committee that includes at least two persons, one of which shall be
% the student’s administrative supervisor.
% Despite these formally defined roles, it is somewhat a matter of
% taste whether you list all persons as supervisors or all as just
% thesis committee or list the one single academic supervisor as
% supervisor and then all persons (repeating the supervisor's name)
% as thesis committee.
% In the settings, that follow here, you can set this and many other settings.

\def\thesisyear{2024}       					% Year thesis submitted
\def\thesismonth{June}					% Month thesis submitted, e.g. "December"
\def\thesisauthor{Brynjar Geir Sigurðsson}				% Thesis authoreiningaraðferðinni
\def\thesistitle{A NN approach to predicting gust factors in complex landscape} % Title of thesis
%\def\thesissubtitle{XXSubtitle is rarely usedXX}		% Subtitle of thesis (optional)
\def\thesisshorttitle{} 	% Optional: if title of thesis is longer than 50 characters, it would not fit on the spine (kjölur) and in this case you, need to provide a short title here for the print shop. Otherwhise: make it empty
\def\thesiscredits{60 ECTS} 						% Credits awarded for the project
\def\thesissubject{Mechanical Engineering}
\def\thesiskind{M.Sc.}					% E.g. M.Sc. or Ph.D. thesis
\def\thesiskindformal{CCMagister Scientiarum}			% E.g. Magister Scientiarum
\def\thesisschool{School of Engineering and {Natural Sciences}}		% School
\def\thesisfaculty{Industrial Engineering, Mechanical Engineering and Computer Science}% Faculty name without "Faculty of" (gets added) 
\def\thesisaddress{Dunhagi 5}			        % Faculty or school office address street
\def\thesispostalcode{107}			                % Faculty or school office zip code
\def\thesistelephone{525 4000}					% Office telephone
%\def\thesispublisher{XX}					% Publisher (not used)
\def\thesissupervisors{Kristján Jónasson // Ólafur Pétur Pálsson}					% Names of supervisors (split by \\ if more than one).
\def\thesisnrofsupervisors{2}					% Number of supervisors (to use "Supervisor" vs. "Supervisors")
\def\thesiscommittee{Kristján Jónasson \\ Ólafur Pétur Pálsson}			% Thesis commitee must include the supervisor(s)
\def\thesisexaminer{XXNN3XX}				        % Examiner (prófdómari)
\def\thesisISBN{}           					% Thesis ISBN number (keep empty: not used anymore)
\def\thesisprinting{}						% Name of printsthop (keep empty if thesis is never printed)
%\def\thesisprinting{Háskólaprent, Fálkagata 2, 107 Reykjavík}	% Name of printsthop (keep empty if thesis is never printed)
\def\thesislicense{This thesis may not be copied in any form without author permission.} % Set license here (could also be some Creative Commons license)
%\def\thesiskeywords{Keyword1, Keyword2, Keyword3}		% Keywords (not used anywhere, hence commented out)

%%%%%%%%%%% STOP MODIFYING HERE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%% Next modifications: search for "START MODIFYING HERE AGAIN" below %%%%%%%%%%

% We need this command if someone used \\ in the thesis title
\newcommand{\removelinebreaks}[1]{%
  \begingroup\def\\{}#1\endgroup}

\begin{document}
\hypersetup{pageanchor=false}
\pagenumbering{Alph} % To prevent page numer "1" to be used multiple times, used "A", "B", etc. for the first pages
\begin{titlepage} % This titlepage environment spans in fact multiple pages

% This is the cover title page. If you go to a print shop, they will
% ignore it and create their own cover page, i.e. this cover page here
% is only used by the PDF version that gets electronically archived.

  \thispagestyle{empty}

  % The banner at top and bottom (using a tikz overlay)
  \begin{tikzpicture}[remember picture,overlay]
    \node[anchor=north west, inner sep=0pt] at (current page.north west)
        {\includegraphics[width=\paperwidth]{banner}}; % The top banner (as a PNG) % TODO: A vector graphic would be better

    \node(bottom)[shape=rectangle, fill=hiblue, minimum height=10mm, minimum width=\paperwidth, anchor=south west] at (current page.south west) {}; % The bottom banner (a filled rectangle)

    \node[above=0.4cm of bottom] {
        \begin{tabular}{c} 
          \sffamily \small \textcolor{hiblue}{\textbf{\MakeUppercase{Faculty of \thesisfaculty{}}}}
        \end{tabular}
    };
  \end{tikzpicture}

  \enlargethispage{3cm}
  \vspace*{3.5cm} % Here starts the white space below the top banner

  % The centering used below is with respect to the page margins which
  % are not the same on left and right which prevents proper centering
  % with respect to the tikz centering (and the title page in
  % general).  Instead, we use a minipage that we shift horizontally
  % by 2.6 cm.  But minipage sets \parskip to 0, so we need to save
  % and restore it.  To be able to use vfill/stretch in a minipage,
  % the height needs to be specified: 20.0 cm.
  \newlength{\currentparskip}
  \setlength{\currentparskip}{\parskip}
  \hspace*{-2.6cm}
  \begin{minipage}[t][20.0cm]{1.0\paperwidth}
    \setlength{\parskip}{\currentparskip}
    \begin{center}
      \vspace*{ \stretch{1.5} }
      \huge \sffamily \bfseries \thesistitle{}

      %\normalfont \LARGE \sffamily \thesissubtitle{}

      \vspace{ \stretch{1.0} }
      \normalfont \Large \sffamily \thesisauthor{}

      \vspace*{ \stretch{2.75} }

      \thesismonth{}~\thesisyear{} % E.g. "March 2022"

      \vspace{ \stretch{0.5} }

      \normalfont \Large \sffamily {\thesiskind{}~thesis \\
      in \thesissubject{}}

      \vspace*{ \stretch{1.0} }
    \end{center}
  \end{minipage}

  \newpage

  \thispagestyle{empty} \mbox{} % This is the inside page of the cover (cover verso) which remains empty

  \newpage

  \thispagestyle{empty} % This is the inner title page
  \begin{center}
    \vspace*{ \stretch{0.5} }

    \Large \sffamily \bfseries \thesistitle{}

    %\normalfont \large \sffamily \thesissubtitle{}

    \vspace*{ \stretch{1.0} }

    \sffamily{\thesisauthor{}}

    \vspace*{ \stretch{1.0} }
    \normalsize \thesiscredits{}~ECTS thesis submitted in partial fulfillment of a \\
    \textit{\thesiskindformal{}} degree in \thesissubject{}
    \large

    \ifx\thesissupervisors\empty % Only print supervisor part if supervisor names are not empty
    \else
      \vspace*{ \stretch{1.0} }
      \ifnum\thesisnrofsupervisors>1 Supervisors \\
      \else Supervisor \\
      \fi
      \thesissupervisors{}
    \fi

    \ifx\thesiscommittee\empty % Only print thesis committee part if committee names are not empty
    \else
      \vspace*{ \stretch{0.25} }
      \thesiskind{}~Committee\\
      \thesiscommittee{}
    \fi

    \ifx\thesisexaminer\empty % Only print examiner part if thesisexaminer is not empty
    \else
      \vspace*{ \stretch{0.25} }
      Examiner \\
      \thesisexaminer
    \fi

    \vspace*{ \stretch{1.0} }

    Faculty of \thesisfaculty \\
    \thesisschool \\
    University of Iceland \\
    Reykjavik, \thesismonth~\thesisyear

    \vspace*{ \stretch{0.5} }
  \end{center}

  \newpage

  \thispagestyle{empty} % This is the title verso (colophon), i.e. imprint/copyright page
  \vspace*{\fill}
  % \setcounter{page}{0} \renewcommand{\baselinestretch}{1.5}\normalsize
  \sffamily{\removelinebreaks{\thesistitle}} \\
  \ifx\thesisshorttitle\empty % Show only if shorttitle is provided
  \else
  (\sffamily{\thesisshorttitle{}}) \\
  \fi
  %\sffamily{\removelinebreaks{\thesissubtitle{}}} \\

  \thesiscredits ~ECTS thesis submitted in partial fulfillment of a \thesiskind{}~degree in \thesissubject
\\ \\
  Faculty of \thesisfaculty \\
  \thesisschool \\
  University of Iceland \\
  \thesisaddress \\ 
  \thesispostalcode, Reykjavik 
  Iceland

  Telephone: \thesistelephone \\ \\ 
  \vspace*{\lineskip}

  Bibliographic information: \\
  \thesisauthor{} (\thesisyear{}) \emph{\removelinebreaks{\thesistitle{}}}, \thesiskind{}~thesis, Faculty of \thesisfaculty, University of Iceland.\\

  Copyright \textcopyright~\thesisyear~ \thesisauthor \\
  \thesislicense{}\\

  \ifx\thesisISBN\empty % Show only if ISBN is provided
  \else
  ISBN~\thesisISBN
  \fi

  \ifx\thesisprinting\empty % Show only if print shop is provided
  \else
  Printing: \thesisprinting \\
  \fi

  Reykjavik, Iceland, \thesismonth~\thesisyear \\

%%%%%%%%%%% START MODIFYING HERE AGAIN %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

  \newpage % Dedication page: remove completely if you have no dedication

  \thispagestyle{empty} \mbox{}

  \vfill

  \begin{center}
    \textit{
      To all the students who made the wise decision to use \LaTeX. % Replace this by your dedication
    }
  \end{center} \vspace*{5cm}

  \vfill

%%%%%%%%%%%%%%%%%%%%%%%%% If you have no dedication: remove until here %%%%%%%%%%%%%%%%%%%%

\end{titlepage}

\cleardoublepage

\pagenumbering{roman} % Abstract page

\setcounter{page}{5}

\setkomafont{section}{\huge} % The title "Abstract" and "Útfdráttur" should look like the chapters, i.e. use \huge (\chapter cannot be used as this would create a new page)
\section*{Abstract}
English abstract (ca. 250 words).

\vfill \vspace*{1cm}

\section*{Útdráttur}
Hér kemur útdráttur á íslensku sem er að hámarki 250 orð.
\vfill

\newpage % Empty page
\setkomafont{section}{\FloatBarrier\Large}
% The first "chapter" which will start on a new page
% Table of contents starts automatically on a right-hand side page ("recto").
% Table of contents, list of figures and tables are automatically generated by the commands below
\hypersetup{pageanchor=true}
\tableofcontents
\listoffigures
\listoftables

\chapter*{Abbreviations}
\addcontentsline{toc}{chapter}{Abbreviations}
\markboth{Abbreviations}{Abbreviations}

Í þessum kafla mega koma fram listar yfir skammstafanir og/eða breytuheiti. Gefið kaflanum nafn við hæfi, t.d.\ Skammstafanir eða Breytuheiti. Þessum kafla má sleppa ef hans er ekki þörf. \\

The section could be titled: Glossary, Variable Names, etc.

If you use acronyms, it is strongly recommended to use a LaTeX
acronyms package. For example, to use the package \emph{acronym}, add
in the preamble:
\begin{verbatim}
\usepackage{acronym}
\end{verbatim}
and then here in this chapter (to create the list of all acronyms), use:
\begin{verbatim}
\begin{acronym}[SENS]
  \acro{SENS}{School of Engineering and Natural Sciences}
  \acro{UoI}{University of Iceland}
\end{acronym}
\end{verbatim}
In the square bracket above, you need to put the longest acronym, so
that the list gets proper indentation based on the longest acronym.
Also note that you need to manually sort here that list.

Then, whereever you use the acronym in your text: \verb|\ac{UoI}|:
that will automatically expand the acronym at the first use, but use
only the short version after the first use.  (Use \verb|\acp| if need
the plural form. \verb|\acl| if you explicitly want to have the long
form only, \verb|\acf| if you explicitly want to have the full long
and short form, i.e.\ like first use of \verb|\ac|.)

Use \verb|\acresetall| to forget about earlier usage (=expansion) of
acronyms, e.g.\ if despite already used in Abstract or Introduction,
but you want to expand them later once again (starting from, e.g., in
Foundations chapter): add \verb|\acresetall| at start of Introduction
chapter (and maybe also again at start of Foundations chapter).

\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}
Í þessum kafla koma fram þakkir til þeirra sem hafa styrkt rannsóknina með fjárframlögum, aðstöðu eða vinnu. t.d.\ styrktarsjóðir, fyrirtæki, leiðbeinendur, og aðrir aðilar sem hafa á einhvern hátt aðstoðað við gerð verkefnisins, þ.m.t.\ vinir og fjölskylda ef við á. Þakkir byrja á oddatölusíðu (hægri síðu).

\pagenumbering{arabic}
\setcounter{page}{1}

\include{Chapters/Chapter1}
\include{Chapters/Chapter2}
\include{Chapters/Chapter3}
\include{Chapters/Chapter4}

\end{document}

### 2024-04-02 Chapters/Chapter1.tex
Some updates after review from OPP

@@
 These results are usually published for 3-6 hour intervals.They describe the state over the period and so do not necessarily grasp fluctuations well.
-These fluctuations would include fluctuations in the wind speed, wind gusts\cite{canNNBeatNWPw}.
+These fluctuations would include fluctuations in the wind speed, wind gusts\cite{canNNBeatNWP}.
 
@@
 It looks at the data at the point of interest and does not look at the data as a time series.
-This thesis tries to improve on the baseline model of always predicting the mean and show that some structure can be learned from reanalysis data about gusts.
+This thesis aims to improve on the baseline model of always predicting the mean gust factor and show that some structure can be learned from reanalysis data about gusts.
 To do this a neural network was created.
-The goal of this neural network was to predict the gust factor and improve upon a baseline of always guessing the mean gust factor.
-This baseline model gave an average error of around 10\%.
-Any significant improvement on this would indicate that the final model has something to contribute.
+Any significant improvement on a base model, that always guesses the mean gust factor, would indicate that the final model has something to contribute.
 
@@
 They looked at data from automatic weather stations that measure wind at 10 meters above ground.
-A common definition of wind gust is the average 1 second measured wind over three seconds.
-This is a rolling average.
-A common definition of average wind is the average of the measured wind over 10 minutes.
+A common definition of wind gust is the maximum average 1 second measured wind speed over a rolling 3 second interval using all measurements over a ten minute span.
+A common definition of average wind is the average of the measured wind speed over 10 minutes.
 These are the definitions used both by H.
 Ágústsson and H.
-Ólafsson in 2004 as well as this study.
-The data that was studied in 2004 comes from the same source as used in this study, but limits itself to a smaller section.
+Ólafsson in 2004 as well as this thesis.
+The data that was studied in 2004 comes from the same source as used in this thesis, but limits itself to a smaller section.
 They only looked at the years 1999-2001.
-They looked at three factors and how these effected the gust factor.
+They looked at three factors and how these three parameters effected the gust factor.
 These were $d_m, D, H$, that is direction of wind blowing off a mountain, distance to the mountain and the height of the mountain above the weather station.
@@
 
-To be able to capture the patterns in the data we construct a neural network.
-The choice to select a neural network is one that is made because they are known to capture well patterns in complex data and can be applied to varied problems.
+To be able to capture the patterns in the data a neural network was constructed.
+A NN architechture was chosen as they are known to be able to capture patterns well in complex data and handle high parameter counts.
 This comes in handy when training on different types of data.
 It is also easy to construct different types of neural networks and see how they fit well with parts of the dataset.
-To measure the performance of these models, both to train and test, mean absolute percentage error was used.
-This measures the average percentage error.
+To measure the performance of these models, both to train and test, mean absolute percentage error (MAPE, \ref{eq:mape})was used.
+\begin{equation}
+    \label{eq:mape}
+    MAPE = \frac{}{}
+\end{equation}
+
 This was chosen because the target is the gust factor (the wind gust over the average wind).

### 2024-04-02 Chapters/Chapter5.tex
Some updates after review from OPP

% Chapter Template

\chapter{Results} % Main chapter title

\label{Chapter5} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}

%----------------------------------------------------------------------------------------
%	SECTION 1
%----------------------------------------------------------------------------------------

\section{Model structure}
This baseline model gave an average error of around 10\%.

### 2024-04-03 Chapters/Chapter1.tex
Uppfært eftir ábendingar frá ÓPP

@@
 It states that the change in wind is described by the pressure gradient, the oscillation, the acceleration due to gravity and the resistance from the landscape.
-
 \begin{equation}
@@
 \end{equation}
-
 Equation \ref{eq:navierstokes} relates the change in wind w.r.t.
@@
 position ($\nabla\boldsymbol{V}$) to pressure change w.r.t.
-position ($\nabla P$).
+position ($\nabla\boldsymbol{P}$).
 The wind is also influenced by the oscillation, gravity and resistance but the main drive of wind is pressure gradient\cite{uncertainties_in_numerical_weather_predictions}.
@@
 A problem that will become increasingly prevalent in the near future\cite{nasa_extreme_weather}.
-
-%This thesis looks at how best to predict gust factor based on wind speed, wind direction, pressure and temperature at different heights above ground, as well as the elevation of the landscape over an area in the direction the wind is coming from.
-A neural network (NN) was constructed to predict the gust factor.
 
@@
 \end{enumerate}
-
 Before 2022, LWM had been shown to be able to compete with traditional NWP for some specifc cases as well as making predictions quicker, after training.
@@
 \section{Methodology and related work}
-
 This study looks at data from three sources and an attempt is made to predict the gust factor in a given place in Iceland.
@@
 It is also easy to construct different types of neural networks and see how they fit well with parts of the dataset.
-To measure the performance of these models, both to train and test, mean absolute percentage error (MAPE, \ref{eq:mape})was used.
+To measure the performance of these models, both to train and test, mean absolute percentage error (MAPE) as defined in Equation (\ref{eq:mape}) was used.
 \begin{equation}
     \label{eq:mape}
-    MAPE = \frac{}{}
+    MAPE = \frac{1}{n}\Sigma_{i=1}^n\frac{|y_{predict} - y_{true}|}{y_{predict}}
 \end{equation}
-
 This was chosen because the target is the gust factor (the wind gust over the average wind).

### 2024-04-03 Chapters/Chapter2.tex
Uppfært eftir ábendingar frá ÓPP

@@
 
-\label{Chapter3} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}
+\label{Chapter2} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}
 
@@
 These features for each of the stations go back at least two decades, but do not have the same start date.
-This paper does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.
+This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.
 
@@
 It is possible to request a subsection of the area, a rectangular subsection of the area.
-This leads to having to store a large amount of data, because we cannot simply ask for the specific points of interest.
+This leads to having to store a large amount of data, because it is not possible to ask for specific points of interest.
 To get the data one has two options.
 Their web interface or using their API client.
-Using the API client is the only realistic option here, as we are making thousands of requests for different times.\\
+Using the API client is the only realistic option here, as there were thousands of requests made for different times.
 
-Each request is made by going through all instances of the Veðurstofu data were able to fulfill the requirements (of mean wind speed reaching \averageWindSpeedLimit 
-m/s over a 10 minute interval).
-For each such observation in the Veðurstofa gögn, we need to make two API (calls excluding redundancy).
-One for the 3 hour interval before and one for the 3 hour interval after.
-That is if the observation was at 1 pm, we would need Carra data for noon and 3 pm.
+Each request is made by going through all instances of the Veðurstofu data were able to fulfill the requirements (of mean wind speed reaching \averageWindSpeedLimit m/s over a 10 minute interval).
+For each such observation in the Veðurstofa data, two API calls excluding redundancy are needed.
+One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly one of the three hour interval times.
+That is if the observation was at 13:00, we would need Carra data for noon and 15:00.
 Date and times were generated automatically from the AWS and requested by calling the client.
-Using these datapoints, we should be able to bridge and get an estimation for the point of the given weather station.
+Using these datapoints, interpolation was used to get an estimation for the point of the given weather station.
 The Carra data contains several types of layers.
@@
 The data for this observation was downloaded from height levels.
-That is, data was requested at heights of 15, 150, 250 and 500 meters above ground.\\
-
-For each point 4 features were requested, wind speed, wind direction, pressure and temperature. Each of these features needed to be bridged to create data for model to be trained on.
+That is, data was requested at heights of 15, 150, 250 and 500 meters above ground.
+For each point 4 parameters were requested, wind speed, wind direction, pressure and temperature.
+Each of these features needed to be interpolated to create data for model to be trained on.
 
@@
 Downloading and processing this data would take a lot less time.
-The largest bottleneck in terms of retrieving the data is the request time.
-That is the time the request is queued and the time it is running before it begins downloading.
-This could range from almost immediate to 15 minutes or more.
-This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during the winter and spring, which increased the wait time and sometimes resulted in queries not being responded to.
+The bottleneck for retrieving the data is the request time.
+That is the time the request is queued and running (server preparing the data) before it begins downloading.
+This could range from almost immediate to 30 minutes.
+This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during the winter and spring of 2023/2024, which increased the wait time and sometimes resulted in queries not being responded to.
 This meant that the time it would take to retrieve the remaining information went from around a day, when the requests were at their quickest, to many months, something that would not be possible given the time frame of the project.
+Fortunately these problems were only particularly time consuming during mid winter.
 
@@
 This file encompasses Iceland and is around 685 Mb.
-At first I thought I might have to split the file into subsections and read for each one but that doesn't seem to be the necessary.
 A good amount of time was spent finding a data structure that was best for lookup when trying to find points within a certain area.
@@
 A quadtree, like the name indicates, is a tree structure.
-To initialize it you give it a boundary box.
+To initialize it a boundary box is given.
 This boundary box is then the parent of exactly four nodes and so on, dividing the area in four at each level and allowing quick lookup.
-This was not necessary as the rasterio package allowed for quick lookup with it's index and the affine transform.
+This was not necessary as the Python package Rasterio allowed for quick lookup with it's index and the affine transform.
+Using this package it is possible to quickly look up elevation given coordinates using matrix calculations.

### 2024-04-03 Chapters/Chapter3.tex
Uppfært eftir ábendingar frá ÓPP

@@
 The data from Veðurstofan was supplied for 10 minute increments, while Carra data is in 3 hour intervals.
-This means that to use the Carra data to predict the measured values from Veðurstofa, temporal bridging would need to be done.
-Along with the temporal bridging, we note that the Carra data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the Veðurstofa is given at specific locations.
+This means that to use the Carra data to predict the measured values from Veðurstofa, temporal interpolation would need to be done.
+Along with the temporal interpolation, note that the Carra data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the Veðurstofa is given at specific locations.
 The elevation information was given by a 20 by 20 rectangular grid that covers Iceland.
-When combining these data sources a selection of bridging needs to occur.
-The bridging was linear, both temporally and spatially.
+When combining these data sources a selection of interpolation method needs to occur.
+The interpolation was linear, both temporally and spatially.
 This might influence the results but was not considered in this study.
@@
 The procedure of combining these sources was as follows.
-We start with the measured data from the AWS, we filter this by using a limit on the average wind speed.
-The gust factor generally drops with increased wind speed (although not always dependent on the factors such as the landscape \cite{GNP_vidtal}).
+The measured data from the AWS is filtered by using a limit on the average wind speed.
+The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}).
 Even so being able to predict the gust factor is more important for higher average wind speed as there we have the highest wind gusts.
 After this stripped dataset over every AWS has been created it is used to query the Carra data by using their API.
-When querying the Carra API we will have to query in such a way that we query for given hours, days, months, years and a given area.
-That is, if we ask for a given hour, we get that hour for every day that we ask for.
-Similarly if we ask for a given day, we get that day for every month we ask for.
+The Carra API needs to be queried for given hours, days, months, years and a given area.
+That is, if queried for a given hour, it returns that hour for every day that is queried.
+Similarly if queried for a given day, it returns that day for every month.
 In light of these restraints, it was decided to query month by month.
-Querying only the days needed (both the days included in the 10 minute measurements and the days needed for bridging if close to midnight) but every hour of the day (midnight, 3 AM, 6 AM, 9 AM, midday, 3 PM, 6 PM and 9 PM) and then take these values and bridge for the 10 minute values.
+Querying only the days needed (both the days included in the 10 minute measurements and the days needed for bridging if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00 PM) and then take these values and bridge for the 10 minute values.
 After querying and downloading the data for the height levels and variables requested, bridging is done for the points of interest and values stored in a pandas dataframe.
-After this is done the downloaded data is discarded and we look to the next month.
+After this is done the downloaded data is discarded and the next month is queried.
 This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (goes from several terabytes to less than a gigabyte).
@@
 For each point retrieved bridging needs to be done.
-This is done in a similar manner to the briding of the Carra data.
+This is done in a similar manner to the interpolation of the Carra data.
 Weights are assigned to each of the four points bounding the point of interest.
@@
 Once data has been retrieved for all three sources and processed, including bridging values, it needs to be made ready to use by the model, for both training, validation and test.
-We start with a dataframe that contains the measured information from AWS.
+Starting with a dataframe that contains measured information from AWS.
 This includes the average wind, the wind gust, wind direction along with the station number and coordinates.
-When selecting the Carra data we choose certain height levels.
-That is at which heights we want the reanalysis data.
+When selecting the Carra data certain height levels are chosen.
+That is at which heights above ground we want the reanalysis parameters that are requested.
 These present as separate lines in the Carra dataframe.
 We need to combine them so that each line indicates a single observation.
-When this is done we can combine the AWS Veðurstofu data and Carra reanalysis data on the location and time columns.
-Lastly we look at the elevation data.
+When this is done it is possible to combine the AWS Veðurstofu data and Carra reanalysis data on the location and time columns.
+The last data source is the elevation.
 A couple of different sections of land around the weather stations have been looked at.
-We tried a sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point.
-In any case the points, that represent these sections, was selected like as shown in code listing \ref{code:sectorElevation}.
-That is we look at the wind direction d and define a range of angles around that direction at some distance from the given point.
-This means that we get some number of points (equal to the length of angleRange) at a distance from given weather station as defined by length\_rng.
-This then gives us some sectors at some distances away from the weather station.
+A sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point.
+In any case the points, that represent these sections, were selected like as shown in code Listing (\ref{code:sectorElevation}).
+That is a range of angles are defined based on the wind direction d at some distance from the given point.
+This means that the resultant points (equal in number to the length of angleRange by k) form sectors at several distances from the given weather station.
 
@@
 We end up with a dataframe that has measured data from AWS, which gives us our target, reanalysis data from Carra, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data.
-An example of what the data looks like can be seen in table \ref{table:trainDataExample}.
+An example of what the data looks like can be seen in Table (\ref{table:trainDataExample}).
 
 \begin{table}[h]
+    \caption{An example of data structure used with model}
+    \label{table:trainDataExample}
     \resizebox{\textwidth}{!}{
@@
     }
-    \caption{An example of data structure used with model}
-    \label{table:trainDataExample}
 \end{table}
 
-Looking at table \ref{table:trainDataExample} we note that the last \nPCA columns represent the principle component analysis of the elevation data.
-We also note that the first four columns represent two variables that describe the stability of the air.
-These are the Richardson number (Ri) and Brunt–Väisälä frequency (N).
-They are calculated using equations \ref{eqn:Ri} and \ref{eqn:N}.
+Looking at Table (\ref{table:trainDataExample}) note that the last \nPCA columns represent the principle component analysis of the elevation data.
+Also note that the first four columns represent two variables that describe the stability of the air.
+These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency\cite{brunt_vaisala_freq_eumtrain} ($N$).
+They are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}.
 They are calculated using reanalysis data about the the weather at two different height levels.
@@
     \label{eqn:Ri}
-    Ri = \frac{g \cdot dT \cdot dz}{T_{ave} \cdot dU^2}
+    Ri = \frac{g \cdot dT \cdot dz}{T_{\textrm ave} \cdot dU^2} \unit{[]}
 \end{equation}
@@
     \label{eqn:N}
-    N = \sqrt{\frac{g \cdot dT }{T_{ave} \cdot dz}}
+    N = \sqrt{\frac{g \cdot dT }{T_{\textrm ave} \cdot dz}} \unit{[Hz]}
 \end{equation}
 
-Here, g is the acceleration due to gravity, dT is the temperature difference between the two height levels, dz is the elevation difference, $T_{ave}$ is the average temperature (that is the average of the two temperatures in the height levels) and dU is the wind speed difference between the two height levels.
-Both of these numbers tell us something about the stability of the air.
-These are derived factors from the reanalysis data and as such there shouldn't be a significant information gain in calculating the as opposed to having the raw data.
-Including these factors instead of the every variable might speed up training as well as making the model more easily explainable with the use of Shapley values or other tools for explainability.
+Here, $g$ is the acceleration due to gravity, $dT$ is the temperature difference between the two height levels, dz is the elevation difference, $T_{\textrm ave}$ is the average temperature (that is the average of the two temperatures in the height levels) and $dU$ is the wind speed difference between the two height levels.
+Both of these numbers provide some insight about the stability of the air.
+A lower value for the Richardson number indicates a higher turbulance.
+A typical range of values could be between 10 and 0.1, with values below 1 indicating significant turbulance\cite{richardson_number_skybrary}.
+When the square of the Brunt-Väisälä frequency is negative, then we have unstable air (air parcel will move away from its original position)\cite{brunt_vaisala_freq_eumtrain}.
+These are derived factors from the reanalysis data and as such there shouldn't be a significant information gain using $Ri$ and $N$ as opposed to having the raw data.
+Including these factors instead of every reanalysis variable requested might speed up training as well as making the model more easily explainable with the use of Shapley values or other tools for explainability.
+Using Shapley a feature importance value is attributed to a given feature by creating all possible permutations of any possible length (up to number of features) and seeing how the predictions are skewed when the given parameter is included or excluded.
+This the needs to be done for all parameters.
+The time complexity of this is very high ($2^n$ coalitions)\cite{shapley_information}.
+Most implementations use some approximations, which still can take a considerable amount of time for models with a high parameter count and many examples.

### 2024-04-03 Chapters/Chapter4.tex
Some updates after review from OPP
Uppfært eftir ábendingar frá ÓPP

@@
 
-\label{Chapter3} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}
+\label{Chapter4} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}
 
@@
 \section{Model structure}
-The structure of the neural network is such that it contains some number n of  fully connected layers and batch normalization for each layer, along with regularization.
+The structure of the neural network is such that it contains some number $n$ of fully connected layers and batch normalization for each layer, along with regularization.
 All of these layers have the same number of units.
@@
 This is simply a dense layer with 1 unit.
-A grid search was performed to figure out the best hyper parameters.
+A grid search was performed to determine the hyperparameters that minmize the loss.
+Hyperparameters are parameters that are set before the training begins\cite{hyperparameter_definition}.
 These hyperparameters include number of units in each layer, number of epochs to train for, number of layers, batch size, optimizer and penalty to enforce in the regularization.
-The possible combinations tried can be seen in table \ref{table:gridSearchHyperparamters}.
+The possible combinations tried can be seen in Table \ref{table:gridSearchHyperparamters}.
 
@@
     \centering
+    \caption{Hyperparamter search with best performing combination colored red}
+    \label{table:gridSearchHyperparamters}
     \begin{tabular}{c|c}
@@
     \end{tabular}
-    \caption{Hyperparamter search with best performing combination colored red}
-    \label{table:gridSearchHyperparamters}
 \end{table}
@@
 \section{Feature selection and importance}
-Using the three datasources, we train a model.
+Using the three datasources, a model is trained.
 The data from Veðurstofa only represents the ground truth and is not used as part of the training data.
-We use the gust ($f_g$) and wind speed ($f$) to calculate the target gust factor.
-From the Carra reanalysis data we get all of our weather information, with variables wind speed, wind direction, temperature, pressure.
+The gust ($f_g$) and wind speed ($f$) are used to calculate the target gust factor.
+All weather parameters used in predictions come from the Carra reanalysis data, with variables wind speed, wind direction, temperature, pressure.
 All of these variables are queried in three height levels (15, 250, 500 meters above ground) at the location of a given weather station.
-In addition we have the elevation above sea level over Iceland.
-We select the elevation in a couple of different shapes.
+The second data source for the training data, is the elevation data.
+Veðurstofa provided a GeoTif file that contains elevation information above sea level, covering Iceland, in a 20 by 20 grid.
+A couple of different landscape distributions were looked at.
 A sector upwind, a sector upwind and downwind as well as a circle surrounding a weather station.
 
-To add to these observed values and reanalysis, we calculate two derived variables, the Richardson number and the Brunt-Väisälä frequency.
-To begin with we simply use all the features we have to train our model along with the derived variables to see.
-We do this to be able to then use tools such as Shapley to see which features are impacting our predictions.
-Using Shapley values to see which features, will then allow us to exclude features that don't affect our model prediction and simplify the training data.
-We thus start with much higher dimensionality of our training data and reduce it as much as we can without impacting our results.
+To add to these observed values and reanalysis, two derived variables, the Richardson number and the Brunt-Väisälä frequency, were calculated.
+To begin with all the available parameters were used to train the model along with the derived variables to see.
+This is done to be able to then use tools such as Shapley to see which features are impacting the predictions.
+Using Shapley values to see which features, will then enable the exclusion of features that don't affect the model prediction and simplify the training data.
+This means starting with much higher dimensionality of our training data and reduce it as much as possible without impacting our results.
 This crystallizes in the landscape points.
-We start with n points that describe the elevation of the landscape around our weather station.
-This n might be in the hundreds.
-If we look at concentric circles outward from the weather station, we might have 10 circles (largest with radius 20 km), each with 72 points (5°spacing) and end up with 720 points for the landscape.
+Starting with $n$ points that describe the elevation of the landscape around our weather station.
+This $n$ might be in the hundreds.
+Looking at concentric circles outward from a given weather station, there might be 10 circles (largest with radius 20 km), each with 72 points (5° spacing) and end up with 720 points for the landscape.
 This might slow down training.
-In addition, we might imagine that we are more interested in certain important landscape features.
-There might be a large mountain up- or downwind of our weather station, that might be the most important feature, in addition to some hills and such.
-To address this we use principal component analysis (PCA) to try to reduce the dimensionality of our landscape data.
-Another thing that might serve the same purpose is to select some number of m points that have the most increase or decrease in elevation from the weather station along with the coordinates of those points in relation to the weather station and the direction of the wind.
+In addition, certain landscape features might be more important than others.
+There might be a large mountain up- or downwind of a given weather station, that might be the most important feature.
+To address this principal component analysis (PCA) is used to try to reduce the dimensionality of our landscape elevation data.
+PCA is a statistical procedure that allows the summarization of multivariate data to lower dimension and thus can ease visualization or increase speed of training a complex model with only minimal adverse effect to performance\cite{pca_information}.
+Another thing that might serve the same purpose is to select some number of $m$ points that have the most increase or decrease in elevation from the weather station along with the coordinates of those points in relation to the weather station and the direction of the wind.
 
-Using all the variables from Carra, along with the two derived factors (Ri, N) and two sectors (upwind and downwind) downscaled to 10 features using PCA we can see the feature importance in figure \ref{fig:ShapleyWaterfallFirstTry}.
-Looking at the waterfall graph we can see that Richardson number for higher levels is most important.
+Using all the variables from Carra, along with the two derived factors ($Ri$, $N$) and two sectors (upwind and downwind) downscaled to 10 features using PCA, the feature importance can be seen in Figure \ref{fig:ShapleyWaterfallFirstTry}.
+The waterfall graph in Figure \ref{fig:ShapleyWaterfallFirstTry} shows that the Richardson number for higher levels is most important.
 
 \begin{figure}[h]
+    \caption{Feature importance of a neural network as described with 5 hidden layers and 256 units in each}
+    \label{fig:ShapleyWaterfallFirstTry}
     \centering
     \includegraphics[scale = 0.6]{Figures/shap_bar_nn_256_example100.png}
-    \caption{Feature importance of a neural network as described with 5 hidden layers and 256 units in each}
-    \label{fig:ShapleyWaterfallFirstTry}
 \end{figure}

### 2024-04-03 MSc.tex
Uppfært eftir ábendingar frá ÓPP

@@
 \usepackage{listings}
+\usepackage{xspace}
 
@@
 Otherwhise: make it empty
-\def\thesiscredits{60 ECTS} 						% Credits awarded for the project
+\def\thesiscredits{60} 						% Credits awarded for the project
 \def\thesissubject{Mechanical Engineering}
@@
 %\def\thesispublisher{XX}					% Publisher (not used)
-\def\thesissupervisors{Kristján Jónasson // Ólafur Pétur Pálsson}					% Names of supervisors (split by \\ if more than one).
-\def\thesisnrofsupervisors{2}					% Number of supervisors (to use "Supervisor" vs. "Supervisors")
+\def\thesissupervisors{Kristján Jónasson}					% Names of supervisors (split by \\ if more than one).
+\def\thesisnrofsupervisors{1}					% Number of supervisors (to use "Supervisor" vs. "Supervisors")
 \def\thesiscommittee{Kristján Jónasson \\ Ólafur Pétur Pálsson}			% Thesis commitee must include the supervisor(s)
@@
   \begingroup\def\\{}#1\endgroup}
+
+% To be able to use units in equations. From https://qerub.se/typesetting-units-in-latex
+\newcommand{\unit}[1]{\ensuremath{\, \mathrm{#1}}}
 

### 2024-04-03 Chapters/Chapter3.tex
Uppfært eftir ábendingar ÓPP

@@
 A limit was set on the average wind speed and it was used to select measurement points.
-Along with the average wind speed having to be above a certain limit, to make sure that we were not essentially getting duplicates, we would select only the top wind speed in any given 48 hour period.
+Along with the average wind speed having to be above a certain limit, to make sure that the same weather for the same location is not duplicated, only the top wind speed in any given 48 hour period is selected.
 That is, for a data point to be included it must have been the highest wind speed in the previous and following 24 hours, so as to not describe the same weather multiple times.
@@
 The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}).
-Even so being able to predict the gust factor is more important for higher average wind speed as there we have the highest wind gusts.
+Even so being able to predict the gust factor is more important for higher average wind speed as there are the highest wind gusts.
 After this stripped dataset over every AWS has been created it is used to query the Carra data by using their API.
@@
 When selecting the Carra data certain height levels are chosen.
-That is at which heights above ground we want the reanalysis parameters that are requested.
+That is at which heights above ground the reanalysis parameters are requested.
 These present as separate lines in the Carra dataframe.
-We need to combine them so that each line indicates a single observation.
+That is information for one observation is represented in as many lines as height levels requested in the reanalysis data.
+These rows need to be combined on the position (the weather station) and time so that each row contains all reanalysis information for a given measurement of a weather station.
 When this is done it is possible to combine the AWS Veðurstofu data and Carra reanalysis data on the location and time columns.
@@
 
-We end up with a dataframe that has measured data from AWS, which gives us our target, reanalysis data from Carra, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data.
+The result is a dataframe that has measured data from AWS, which gives us our target, reanalysis data from Carra, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data.
 An example of what the data looks like can be seen in Table (\ref{table:trainDataExample}).
@@
 A typical range of values could be between 10 and 0.1, with values below 1 indicating significant turbulance\cite{richardson_number_skybrary}.
-When the square of the Brunt-Väisälä frequency is negative, then we have unstable air (air parcel will move away from its original position)\cite{brunt_vaisala_freq_eumtrain}.
+When the square of the Brunt-Väisälä frequency is negative, then the air is unstable (an air parcel will move away from its original position)\cite{brunt_vaisala_freq_eumtrain}.
 These are derived factors from the reanalysis data and as such there shouldn't be a significant information gain using $Ri$ and $N$ as opposed to having the raw data.

### 2024-04-04 Chapters/Chapter2.tex
Uppfært eftir ábendingar ÓPP

@@
 
-Veðurstofan provided recordings from weather stations all around Iceland.
+Veðurstofan provided recordings from 412 weather stations all around Iceland.
+The location of these weather stations can be seen in Figure \ref{fig:aws_map}.
 These automatic weather stations (AWS) are at a height of 10 meters above ground level.
@@
 This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.
+
+\begin{figure}[h]
+    \caption{Locations of Automatic Weather Stations in Iceland}
+    \label{fig:aws_map}
+    \includegraphics[scale = 0.75]{Figures/weather_stations.png}
+\end{figure}
 
@@
 One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly one of the three hour interval times.
-That is if the observation was at 13:00, we would need Carra data for noon and 15:00.
+That is if the observation was at 13:00, Carra data for noon and 15:00 would have to be queried.
 Date and times were generated automatically from the AWS and requested by calling the client.

### 2024-04-04 Chapters/Chapter3.tex
Uppfært eftir ábendingar ÓPP

@@
 
-The procedure of combining these sources was as follows.
+\begin{figure}[h]
+    \caption{A flow chart showing how data sources were combined}
+    \label{fig:data_preprocessing_flow_chart}
+    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart.png}
+\end{figure}
+
+The procedure of combining these sources was as follows and can be seen in Figure (\ref{fig:data_preprocessing_flow_chart}).
 The measured data from the AWS is filtered by using a limit on the average wind speed.
@@
 In light of these restraints, it was decided to query month by month.
-Querying only the days needed (both the days included in the 10 minute measurements and the days needed for bridging if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00 PM) and then take these values and bridge for the 10 minute values.
-After querying and downloading the data for the height levels and variables requested, bridging is done for the points of interest and values stored in a pandas dataframe.
-After this is done the downloaded data is discarded and the next month is queried.
+Querying only the days needed (both the days included in the 10 minute measurements and the days needed for interpolation if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00 PM) and then take these values and interpolate for the 10 minute values.
+After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe.
+After this the downloaded data is discarded and the next month is queried.
 This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (goes from several terabytes to less than a gigabyte).
@@
 For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved.
-For each point retrieved bridging needs to be done.
+For each point retrieved interpolation needs to be done.
 This is done in a similar manner to the interpolation of the Carra data.
 Weights are assigned to each of the four points bounding the point of interest.
-These weights are then used to calculate the weighted average that represents the bridged value.
+These weights are then used to calculate the weighted average that represents the interpolated value.
 This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.
@@
 
-Once data has been retrieved for all three sources and processed, including bridging values, it needs to be made ready to use by the model, for both training, validation and test.
+Once data has been retrieved for all three sources and processed, including interpolating values, it needs to be made ready to use by the model, for both training, validation and test.
 Starting with a dataframe that contains measured information from AWS.

### 2024-04-07 Chapters/Chapter2.tex

@@
 
-Veðurstofan provided recordings from 412 weather stations all around Iceland.
+Veðurstofan provided 10 minute recordings from \nStationsMin weather stations all around Iceland.
+Of these \nStationsMin stations, \nVedurMin were from Veðurstofa Íslands and placed at 10 meters above ground, while the rest (\nVGMin) were from \href{https://www.vegagerdin.is/}{Vegagerðin} and placed at 7.5 meters above ground.
 The location of these weather stations can be seen in Figure \ref{fig:aws_map}.
-These automatic weather stations (AWS) are at a height of 10 meters above ground level.
-The information that is provided by these AWS is presented in two different type of documents, hourly and 10 minute documents.
+The information that is provided by these Automatic Weather Stations (AWS) is presented in two different type of documents, hourly and 10 minute documents.
 The hourly documents are summations of the 10 minute documents, with the exception that any errors (so called nails) still in the 10 minute documents should have been removed from the hourly documents.
+Nails, are sharp increases from the rest of the data and are thought to be measurements errors by experts and are discarded by hand.
 Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another document), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction.
@@
 
-\begin{figure}[h]
+\begin{figure}
     \caption{Locations of Automatic Weather Stations in Iceland}
@@
     \includegraphics[scale = 0.75]{Figures/weather_stations.png}
+\end{figure}
+
+These 10 minute measurements were filtered to only include measurements where the average wind speed was at or above \averageWindSpeedLimit m/s.
+They were also filtered in such a way that only the highest value in any given 48 hour span is considered.
+This can be seen in Code Listing (\ref{code:filter48hr}).
+Filtering by, for each station, the highest wind speed and removing all observations that occurred within a 48 hour period centered on the observation (including the highest wind) and doing this iteratively until reaching an empty dataframe for a given station.
+
+\begin{lstlisting}[language = Python, caption = {Filter points over 48 hour interval}, label = code:filter48hr]
+    for station in tqdm(stations, total = len(stations)):
+        subset_df = vedur_df[station == vedur_df.stod]
+        subset_df = subset_df.reset_index(drop = True)
+
+        while not subset_df.empty:
+            idx = subset_df.f.idxmax()
+            time_of_max = subset_df.iloc[idx].timi
+
+            filtered_data.append(subset_df.iloc[idx])
+
+            subset_df = subset_df[abs(subset_df.timi - time_of_max)
+                >= pd.Timedelta(threshold)]
+
+            subset_df = subset_df.reset_index(drop = True)  
+\end{lstlisting}
+
+When these two filterings have been done, the end result is \nStationsCombined stations that meet these criteria.
+Of these, \nVedurMin belong to Veðurstofa Íslands and \nVGMin belong to Vegagerðin.
+The number of observations after, these filterings, by year can be seen in Figure (\ref{fig:occurrences_by_year}).
+The first observations occurring in \startDateVedur and the latest in 2023.
+
+\begin{figure}
+    \caption{Measurements that fit filtering by year}
+    \label{fig:occurrences_by_year}
+    \begin{center}
+        \includegraphics[scale = 0.75]{Figures/occurrences_by_year.png}
+    \end{center}
+\end{figure}
+
+Looking at the number of measurements that meet these criteria for a given station, Figure (\ref{fig:occurrences_by_station_20}) shows only one station has over 1000 measurements that fit the criterias.
+
+\begin{figure}
+    \caption{Number of stations that have measurements in a range}
+    \label{fig:occurrences_by_station_20}
+    \begin{center}
+        \includegraphics[scale = 0.75]{Figures/station_occurrences_20.png}        
+    \end{center}
 \end{figure}
@@
 This could range from almost immediate to 30 minutes.
-This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during the winter and spring of 2023/2024, which increased the wait time and sometimes resulted in queries not being responded to.
+This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during \href{https://forum.ecmwf.int/t/a-new-cds-soon-to-be-launched-expect-some-disruptions/1607}{the winter and spring of 2023/2024}, which increased the wait time and sometimes resulted in queries not being responded to.
 This meant that the time it would take to retrieve the remaining information went from around a day, when the requests were at their quickest, to many months, something that would not be possible given the time frame of the project.

### 2024-04-07 Chapters/Chapter3.tex

@@
 The three data sources were all in a different format.
-Measurement data from Veðurstofa was in text files, elevation data was in GeoTiff and reanalysis data from Carra was in a GRIB format.
+Measurement data from Veðurstofa was in text files, elevation data was in GeoTiff and reanalysis data from CARRA was in a GRIB format.
 To use the data to train, these three data sources needed to be combined into one file.
@@
 That is, for a data point to be included it must have been the highest wind speed in the previous and following 24 hours, so as to not describe the same weather multiple times.
-The data from Veðurstofan was supplied for 10 minute increments, while Carra data is in 3 hour intervals.
-This means that to use the Carra data to predict the measured values from Veðurstofa, temporal interpolation would need to be done.
-Along with the temporal interpolation, note that the Carra data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the Veðurstofa is given at specific locations.
+The data from Veðurstofan was supplied for 10 minute increments, while CARRA data is in 3 hour intervals.
+This means that to use the CARRA data to predict the measured values from Veðurstofa, temporal interpolation would need to be done.
+Along with the temporal interpolation, note that the CARRA data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the Veðurstofa is given at specific locations.
 The elevation information was given by a 20 by 20 rectangular grid that covers Iceland.
@@
 Even so being able to predict the gust factor is more important for higher average wind speed as there are the highest wind gusts.
-After this stripped dataset over every AWS has been created it is used to query the Carra data by using their API.
-The Carra API needs to be queried for given hours, days, months, years and a given area.
+After this stripped dataset over every AWS has been created it is used to query the CARRA data by using their API.
+The CARRA API needs to be queried for given hours, days, months, years and a given area.
 That is, if queried for a given hour, it returns that hour for every day that is queried.
@@
 
+Once CARRA data has been merged with AWS data, using station and time columns, then this combined file needs to checked for nails.
+This is done by using the hourly data (which is supposedly error free).
+As the data has been filtered in for the highest average wind speed in a 48 hour interval, the hourly data can be used to find nails.
+The hourly data is combined with the merged AWS and 10 min data.
+Then filtering is applied on the average wind.
+If the average wind speed differs the rows are dropped.
+This can be used to look at which stations have high percentage error.
+This can be seen in Figure (\ref{fig:error_count}).
+
+\begin{figure}
+    \caption{The number of stations whose number of nails are in a certain percentage range of the overall}
+    \label{fig:error_count}
+    \includegraphics[scale=0.75]{Figures/error_count.png}
+\end{figure}
+
+Figure (\ref{fig:error_count}) shows that most stations have nails in less that 10\% measurements.
+One has nails in 100\% of measurements.
+This is only because the number of measurements that matched previous criteria was 1.
+
 The elevation data comes in a GeoTif file that covers Iceland.
@@
 For each point retrieved interpolation needs to be done.
-This is done in a similar manner to the interpolation of the Carra data.
+This is done in a similar manner to the interpolation of the CARRA data.
 Weights are assigned to each of the four points bounding the point of interest.
@@
 This includes the average wind, the wind gust, wind direction along with the station number and coordinates.
-When selecting the Carra data certain height levels are chosen.
+When selecting the CARRA data certain height levels are chosen.
 That is at which heights above ground the reanalysis parameters are requested.
-These present as separate lines in the Carra dataframe.
+These present as separate lines in the CARRA dataframe.
 That is information for one observation is represented in as many lines as height levels requested in the reanalysis data.
 These rows need to be combined on the position (the weather station) and time so that each row contains all reanalysis information for a given measurement of a weather station.
-When this is done it is possible to combine the AWS Veðurstofu data and Carra reanalysis data on the location and time columns.
+When this is done it is possible to combine the AWS Veðurstofu data and CARRA reanalysis data on the location and time columns.
 The last data source is the elevation.
@@
 A sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point.
-In any case the points, that represent these sections, were selected like as shown in code Listing (\ref{code:sectorElevation}).
+In any case the points, that represent these sections, were selected like as shown in Code Listing (\ref{code:sectorElevation}).
 That is a range of angles are defined based on the wind direction d at some distance from the given point.
@@
 
-The result is a dataframe that has measured data from AWS, which gives us our target, reanalysis data from Carra, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data.
+The result is a dataframe that has measured data from AWS, which gives us our target, reanalysis data from CARRA, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data.
 An example of what the data looks like can be seen in Table (\ref{table:trainDataExample}).

### 2024-04-09 Chapters/Chapter1.tex
Aðeins uppfært eftir ath frá KJ 9/4/24

@@
 \end{equation}
-Equation \ref{eq:navierstokes} relates the change in wind w.r.t.
+Equation (\ref{eq:navierstokes}) relates the change in wind w.r.t.
 time ($\frac{\delta \boldsymbol{V}}{\delta t}$) and change in wind w.r.t.
@@
 The results were flawed and impractical because of the time it took to calculate was much longer than the forecast period.
-In the 1950's, with the advent of computers we get the first operational forecasts.
+In the 1950's, with the advent of computers the first operational forecasts emerged.
 In September of 1954, Rossby and his Stockholm based team produced the first real-time barotropic forecasts.
@@
 In the last two years there have been even more developments with the emergence of Large AI Weather forecast Models (LWM).
-In 2024 Ling et al.
-tried to standardize the definition of LWM in meteorology and came up with 3 rules that need to be met to count as LWM.
+In 2024, Ling et al.\cite{SecondRevolution} tried to standardize the definition of LWM in meteorology and came up with 3 rules that need to be met to count as LWM.
 
@@
 No model had shown that it could in any way completely replace the traditional systems.
-In early 2022, Pathak et al.
-presented FourCastNet.
+In early 2022, Pathak et al.\cite{FourCastNet} presented FourCastNet.
 FourCastNet uses an Adaptive Fourier Neural Operator model that leverages transfomer architecture rather than the popular convolutional model architechture.
@@
 
-There has been a lot of progress made over the last 6 years (since 2018) and especially in the last 2 years (since 2022).
+There has been a lot of progress made over the last 6 years (since 2018) and especially in the last 2 years (since 2022)\cite{SecondRevolution}.
 The progression from machine learning methods being an interesting idea in the field of numerical weather predictions, to outperforming the standard NWP has been remarkably quick.
@@
 
-In 2004, H.Ágústsson and H.
-Ólafsson looked at the variability of gust factor in complex landscapes.
+In 2004, H.
+Ágústsson and H.
+Ólafsson\cite{mean_gust_HA_HO} looked at the variability of gust factor in complex landscapes.
 They looked at data from automatic weather stations that measure wind at 10 meters above ground.

### 2024-04-09 Chapters/Chapter2.tex
Aðeins uppfært eftir ath frá KJ 9/4/24

--- 2024-04-03 Chapters/Chapter2.tex
@@
 Data was sourced from several streams.
-Veðurstofan provided recordings from weather stations all around the Iceland.
-Data was downloaded from Copernicus Arctic Regional Reanalysis dataset (CARRA).
-A land elevation model was also provided by Veðurstofan.
+The Icelandic Meterological Office (IMO) provided measurements from weather stations all around the Iceland.
+NWP data was downloaded from Copernicus Arctic Regional Reanalysis dataset (CARRA).
+A land elevation model was also provided by IMO.
 
@@
 
-Veðurstofan provided recordings from weather stations all around Iceland.
-These automatic weather stations (AWS) are at a height of 10 meters above ground level.
-The information that is provided by these AWS is presented in two different type of documents, hourly and 10 minute documents.
-The hourly documents are summations of the 10 minute documents, with the exception that any errors (so called nails) still in the 10 minute documents should have been removed from the hourly documents.
-Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another document), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction.
-These features for each of the stations go back at least two decades, but do not have the same start date.
+IMO provided 10 minute measurements from \nStationsMin weather stations all around Iceland.
+The measurements that met the filtering criteria, started in \startDateVedur and ended in 2023 (the year measurements were provided).
+Of these \nStationsMin stations, \nVedurMin were from IMO Íslands and placed at 10 meters above ground, while the rest (\nVGMin) were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)} and placed at 7.5 meters above ground.
+The location of these weather stations can be seen in Figure \ref{fig:aws_map}.
+The information that is provided by these Automatic Weather Stations (AWS) is presented in two different type of data files, hourly and 10 minute files.
+The hourly files are summations of the 10 minute files, with the exception that errors, such as nails, still in the 10 minute files should have been removed from the hourly documents.
+Nails, are sharp increases from the rest of the data and are unrealistic outliers that are considered measurement errors and are discarded.
+Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another data set), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction.
+These measurements were started in the end of the 20th century, when the first AWS stations were installed.
+More have been added in the over two decades since then.
 This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.
 
-\section{Carra Data}
-The CARRA dataset goes back to September 1990 and is currently updated monthly, with a latency of 2-3 months.
-The oldest Vedurstofu data example that fulfills given criteria is from \startDateVedur.
-This is covered by Carra.
-A few of the newer points from AWS were not available from Carra when data was gathered.
-The Carra dataset is available for two regions, west and east.
+\begin{figure}
+    \includegraphics[scale = 0.75]{Figures/weather_stations_2024-04-09.png}
+    \caption{Locations of automatic weather stations in Iceland}
+    \label{fig:aws_map}
+\end{figure}
+
+These 10 minute measurements were filtered to only include measurements where the average wind speed was at or above \averageWindSpeedLimit m/s.
+They were also filtered in such a way that only the highest value in any given 48 hour span is considered.
+This can be seen in Code Listing \ref{code:filter48hr}.
+Any measurement that is within a 48 hour interval centered on of the current highest is removed.
+This is done for each station and iteratively until all measurements have been removed.
+
+\begin{lstlisting}[style = Python, caption = {Filter points over 48 hour interval}, label = code:filter48hr]
+    for station in tqdm(stations, total = len(stations)):
+        subset_df = vedur_df[station == vedur_df.stod]
+        subset_df = subset_df.reset_index(drop = True)
+
+        while not subset_df.empty:
+            idx = subset_df.f.idxmax()
+            time_of_max = subset_df.iloc[idx].timi
+
+            filtered_data.append(subset_df.iloc[idx])
+
+            subset_df = subset_df[abs(subset_df.timi 
+                - time_of_max) >= pd.Timedelta(threshold)]
+
+            subset_df = subset_df.reset_index(drop = True)  
+\end{lstlisting}
+
+When these two filterings have been done, the end result is \nStationsMin stations that meet these criteria.
+Of these, \nVedurMin belong to IMO and \nVGMin belong to IRCA.
+The number of observations by year can be seen in Figure \ref{fig:occurrences_by_year}.
+The first observations occurring in \startDateVedur and the latest in 2023.
+
+\begin{figure}
+    \begin{center}
+        \includegraphics[scale = 0.75]{Figures/occurrences_by_year.png}
+    \end{center}
+    \caption{Measurements that fit filtering by year (from \startDateVedur to 2023)}
+    \label{fig:occurrences_by_year}
+\end{figure}
+
+Looking at the number of measurements that meet these criteria for a given station, Figure \ref{fig:occurrences_by_station_20} shows only one station has over 1000 measurements that fit the criterias.
+
+\begin{figure}
+    \begin{center}
+        \includegraphics[scale = 0.75]{Figures/station_occurrences_20.png}        
+    \end{center}
+    \caption{Number of stations that have measurements in a range}
+    \label{fig:occurrences_by_station_20}
+\end{figure}
+
+\section{CARRA Data}
+The CARRA dataset goes back to September 1991 and is currently updated monthly, with a latency of 2-3 months\cite{carra_information}.
+The oldest IMO data point that fulfills given criteria is from \startDateVedur.
+This is covered by CARRA.
+A few of the newer points from AWS were not available from CARRA when data was gathered.
+The CARRA dataset is available for two regions, west and east.
 Each of these covers a vastly larger area than the area of interest.
-It is possible to request a subsection of the area, a rectangular subsection of the area.
 This leads to having to store a large amount of data, because it is not possible to ask for specific points of interest.
@@
 
-Each request is made by going through all instances of the Veðurstofu data were able to fulfill the requirements (of mean wind speed reaching \averageWindSpeedLimit m/s over a 10 minute interval).
-For each such observation in the Veðurstofa data, two API calls excluding redundancy are needed.
+Each request is made by going through all instances of the IMO data after filtering (by mean wind speed reaching \averageWindSpeedLimit m/s over a 10 minute interval).
+For each such observation in the IMO data, two API calls excluding redundancy are needed.
 One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly one of the three hour interval times.
-That is if the observation was at 13:00, we would need Carra data for noon and 15:00.
+That is if the observation was at 13:00, CARRA data for noon and 15:00 would have to be queried.
 Date and times were generated automatically from the AWS and requested by calling the client.
 Using these datapoints, interpolation was used to get an estimation for the point of the given weather station.
-The Carra data contains several types of layers.
+The CARRA data contains several types of layers.
 These are single levels, model levels, height levels, pressure levels.
@@
 
-After using this method to request terabytes of data, it was discovered that it is apparently possible to query a specific area.
-This decreased the size of each file created after a request from around 50 MB to around 2 MB.
-Downloading and processing this data would take a lot less time.
+After using this method to request terabytes of data, it was discovered that it was possible to query a specific area.
+This decreased the size of each file from around 50 MB to around 2 MB.
 The bottleneck for retrieving the data is the request time.
-That is the time the request is queued and running (server preparing the data) before it begins downloading.
+The time the request is queued and running (server preparing the data) before it began downloading.
 This could range from almost immediate to 30 minutes.
-This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during the winter and spring of 2023/2024, which increased the wait time and sometimes resulted in queries not being responded to.
+This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during \href{https://forum.ecmwf.int/t/a-new-cds-soon-to-be-launched-expect-some-disruptions/1607}{the winter and spring of 2023/2024}, which increased the wait time and sometimes resulted in queries not being responded to.
 This meant that the time it would take to retrieve the remaining information went from around a day, when the requests were at their quickest, to many months, something that would not be possible given the time frame of the project.
@@
 \section{Elevation data}
-Veðurstofa provided a tif file containing the elevation of Iceland on a 20 meter by 20 meter grid.
-This file encompasses Iceland and is around 685 Mb.
+IMO provided a tif file containing the elevation of Iceland on a 20 meter by 20 meter grid.
+This file encompasses Iceland and is around 685 MB.
 A good amount of time was spent finding a data structure that was best for lookup when trying to find points within a certain area.

### 2024-04-09 Chapters/Chapter3.tex
Aðeins uppfært eftir ath frá KJ 9/4/24

@@
 The three data sources were all in a different format.
-Measurement data from Veðurstofa was in text files, elevation data was in GeoTiff and reanalysis data from CARRA was in a GRIB format.
+Measurement data from IMO was in text files, elevation data was in GeoTiff and reanalysis data from CARRA was in a GRIB format.
 To use the data to train, these three data sources needed to be combined into one file.
-This was done based on the measurement data from the Veðurstofa.
+This was done based on the measurement data from the IMO.
 A limit was set on the average wind speed and it was used to select measurement points.
@@
 That is, for a data point to be included it must have been the highest wind speed in the previous and following 24 hours, so as to not describe the same weather multiple times.
-The data from Veðurstofan was supplied for 10 minute increments, while CARRA data is in 3 hour intervals.
-This means that to use the CARRA data to predict the measured values from Veðurstofa, temporal interpolation would need to be done.
-Along with the temporal interpolation, note that the CARRA data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the Veðurstofa is given at specific locations.
+The data from IMO was supplied for 10 minute increments, while CARRA data is in 3 hour intervals.
+This means that to use the CARRA data to predict the measured values from IMO, temporal interpolation would need to be done.
+Along with the temporal interpolation, note that the CARRA data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the IMO is given at specific locations.
 The elevation information was given by a 20 by 20 rectangular grid that covers Iceland.
@@
     \label{fig:data_preprocessing_flow_chart}
-    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart.png}
+    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart-2024-04-08.png}
 \end{figure}
 
-The procedure of combining these sources was as follows and can be seen in Figure (\ref{fig:data_preprocessing_flow_chart}).
+The procedure of combining these sources was as follows and can be seen in Figure \ref{fig:data_preprocessing_flow_chart}.
 The measured data from the AWS is filtered by using a limit on the average wind speed.
@@
 In light of these restraints, it was decided to query month by month.
-Querying only the days needed (both the days included in the 10 minute measurements and the days needed for interpolation if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00 PM) and then take these values and interpolate for the 10 minute values.
+Querying only the days needed (both the days included in the 10 minute measurements and the days needed for interpolation if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00) and then take these values and interpolate for the 10 minute values.
 After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe.
@@
 This can be used to look at which stations have high percentage error.
-This can be seen in Figure (\ref{fig:error_count}).
+This can be seen in Figure \ref{fig:error_count}.
 
 \begin{figure}
+    \includegraphics[scale=0.75]{Figures/error_count.png}
     \caption{The number of stations whose number of nails are in a certain percentage range of the overall}
     \label{fig:error_count}
-    \includegraphics[scale=0.75]{Figures/error_count.png}
 \end{figure}
 
-Figure (\ref{fig:error_count}) shows that most stations have nails in less that 10\% measurements.
+Figure \ref{fig:error_count} shows that most stations have nails in less that 10\% measurements.
 One has nails in 100\% of measurements.
@@
 
-\section{Data structure}
+\section{Data Structure}
 
@@
 These rows need to be combined on the position (the weather station) and time so that each row contains all reanalysis information for a given measurement of a weather station.
-When this is done it is possible to combine the AWS Veðurstofu data and CARRA reanalysis data on the location and time columns.
+When this is done it is possible to combine the AWS IMO data and CARRA reanalysis data on the location and time columns.
 The last data source is the elevation.
@@
 A sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point.
-In any case the points, that represent these sections, were selected like as shown in Code Listing (\ref{code:sectorElevation}).
-That is a range of angles are defined based on the wind direction d at some distance from the given point.
+In any case the points, that represent these sections, were selected like as shown in Code Listing \ref{code:sectorElevation}.
+That is a range of angles are defined based on the wind direction $d$ at some distance from the given point.
 This means that the resultant points (equal in number to the length of angleRange by k) form sectors at several distances from the given weather station.
 
-\begin{lstlisting}[language = Python, caption = {Sector elevation points generated}, label = code:sectorElevation]
+\begin{lstlisting}[style = Python, caption = {Sector elevation points generated}, label = code:sectorElevation]
 angles = [(angle + (90 - d)) * pi/180 for angle in angleRange]
@@
 The result is a dataframe that has measured data from AWS, which gives us our target, reanalysis data from CARRA, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data.
-An example of what the data looks like can be seen in Table (\ref{table:trainDataExample}).
+An example of what the data looks like can be seen in Table \ref{table:trainDataExample}.
 
@@
 
-Looking at Table (\ref{table:trainDataExample}) note that the last \nPCA columns represent the principle component analysis of the elevation data.
+Looking at Table \ref{table:trainDataExample} note that the last \nPCA columns represent the principle component analysis of the elevation data.
 Also note that the first four columns represent two variables that describe the stability of the air.

### 2024-04-09 Chapters/Chapter4.tex
Uppfært eftir ábendingar ÓPP
Aðeins uppfært eftir ath frá KJ 9/4/24

@@
 A grid search was performed to determine the hyperparameters that minmize the loss.
-Hyperparameters are parameters that are set before the training begins\cite{hyperparameter_definition}.
+Hyperparameters are parameters that are set before the training begins\cite{hyperparameters_definition}.
 These hyperparameters include number of units in each layer, number of epochs to train for, number of layers, batch size, optimizer and penalty to enforce in the regularization.
@@
 Using the three datasources, a model is trained.
-The data from Veðurstofa only represents the ground truth and is not used as part of the training data.
+The data from IMO only represents the ground truth and is not used as part of the training data.
 The gust ($f_g$) and wind speed ($f$) are used to calculate the target gust factor.
-All weather parameters used in predictions come from the Carra reanalysis data, with variables wind speed, wind direction, temperature, pressure.
+All weather parameters used in predictions come from the CARRA reanalysis data, with variables wind speed, wind direction, temperature, pressure.
 All of these variables are queried in three height levels (15, 250, 500 meters above ground) at the location of a given weather station.
 The second data source for the training data, is the elevation data.
-Veðurstofa provided a GeoTif file that contains elevation information above sea level, covering Iceland, in a 20 by 20 grid.
+IMO provided a GeoTif file that contains elevation information above sea level, covering Iceland, in a 20 by 20 grid.
 A couple of different landscape distributions were looked at.
@@
 
-Using all the variables from Carra, along with the two derived factors ($Ri$, $N$) and two sectors (upwind and downwind) downscaled to 10 features using PCA, the feature importance can be seen in Figure \ref{fig:ShapleyWaterfallFirstTry}.
+Using all the variables from CARRA, along with the two derived factors ($Ri$, $N$) and two sectors (upwind and downwind) downscaled to 10 features using PCA, the feature importance can be seen in Figure \ref{fig:ShapleyWaterfallFirstTry}.
 The waterfall graph in Figure \ref{fig:ShapleyWaterfallFirstTry} shows that the Richardson number for higher levels is most important.

### 2024-04-09 MSc.tex
Aðeins uppfært eftir ath frá KJ 9/4/24

@@
 \usepackage{xspace}
+\usepackage{float}
 
@@
 \usetikzlibrary{positioning}
+
+% https://tex.stackexchange.com/questions/374661/keyword-highlighting-not-working-in-python-with-listings
+\lstdefinestyle{Python}{
+    language        = Python,
+    basicstyle      = \ttfamily,
+    keywordstyle    = \color{blue},
+    keywordstyle    = [2] \color{teal}, % just to check that it works
+    stringstyle     = \color{green},
+    commentstyle    = \color{red}\ttfamily
+}
 
@@
 \begin{titlepage} % This titlepage environment spans in fact multiple pages
+
+% This must be placed after "\begin{document}":
+\lstset{
+    frame       = single,
+    numbers     = left,
+    showspaces  = false,
+    showstringspaces    = false,
+    captionpos  = t,
+    caption     = \lstname
+}
 

### 2024-04-10 Chapters/Chapter1.tex
.

@@
 
+\subsection{Model architechture}
 To be able to capture the patterns in the data a neural network was constructed.
@@
 If the target would have been the wind gust rather than the gust factor then something like mean absolute error might be more appropriate.
+
+\subsection{Model explainability}
+Neural networks are often considered as mysterious black boxes\cite{nn_black_box}.
+In an attempt to understand the model predictions, methods designed for explainability are used.
+One such method is Shapley values\cite{shapley_information}.
+Shapley values are calculated as the average marginal contribution of a feature value across all possible coalitions.
+For any combination of parameters what is the contribution of a given parameter.
+This means that Shapley values can explain individual predictions.
+Other machine learning tools, like ELI5 (Explain like I am  5), randomly shuffle a feature and look at the effect on model performance\cite{eli5_information}.

### 2024-04-10 Chapters/Chapter2.tex
.

@@
     \end{center}
-    \caption{Measurements that fit filtering by year (from \startDateVedur to 2023)}
+    \caption{Measurements that fit filtering by year (from \startDateVedur to 2023).}
     \label{fig:occurrences_by_year}
@@
     \end{center}
-    \caption{Number of stations that have measurements in a range}
+    \caption{Number of stations that have measurements in a range.}
     \label{fig:occurrences_by_station_20}
@@
 The country was divided into 10 parts (as only around 13\% of the file was able to be read into memory at each time as a part of dictionary object) with boundary boxes.
-A quadtree was constructed for each of the sections, so as to simplify the lookup.
-A quadtree allows for efficient lookup by searching for points that intersect a given boundary.
-A quadtree, like the name indicates, is a tree structure.
-To initialize it a boundary box is given.
-This boundary box is then the parent of exactly four nodes and so on, dividing the area in four at each level and allowing quick lookup.
-This was not necessary as the Python package Rasterio allowed for quick lookup with it's index and the affine transform.
+The Python package Rasterio allowed for quick lookup with it's index and the affine transform.
 Using this package it is possible to quickly look up elevation given coordinates using matrix calculations.

### 2024-04-10 Chapters/Chapter3.tex
.

@@
 A limit was set on the average wind speed and it was used to select measurement points.
-Along with the average wind speed having to be above a certain limit, to make sure that the same weather for the same location is not duplicated, only the top wind speed in any given 48 hour period is selected.
-That is, for a data point to be included it must have been the highest wind speed in the previous and following 24 hours, so as to not describe the same weather multiple times.
+Along with the average wind speed having to be above a certain limit, to ensure that the same weather for the same location is not duplicated.
 The data from IMO was supplied for 10 minute increments, while CARRA data is in 3 hour intervals.
 This means that to use the CARRA data to predict the measured values from IMO, temporal interpolation would need to be done.
-Along with the temporal interpolation, note that the CARRA data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the IMO is given at specific locations.
+Along with the temporal interpolation, note that the CARRA data is given in a rectangular grid where the distance between each point is around 2.5 km while the the information from the IMO is given at specific locations.
 The elevation information was given by a 20 by 20 rectangular grid that covers Iceland.
-When combining these data sources a selection of interpolation method needs to occur.
-The interpolation was linear, both temporally and spatially.
-This might influence the results but was not considered in this study.
+When combining these data sources an interpolation method needs to be decided upon.
+Here linear interpolation was applied, both temporally and spatially.
+The choice of interpolation method, although potentially impactful on the results, was not specifically addressed in this study.
 
@@
 The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}).
-Even so being able to predict the gust factor is more important for higher average wind speed as there are the highest wind gusts.
+Even so being able to predict the gust factor is more important for higher average wind speed due to then higher wind gusts.
 After this stripped dataset over every AWS has been created it is used to query the CARRA data by using their API.
@@
 In light of these restraints, it was decided to query month by month.
-Querying only the days needed (both the days included in the 10 minute measurements and the days needed for interpolation if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00) and then take these values and interpolate for the 10 minute values.
+Querying only the days needed (both the days included in the 10 minute measurements and the days needed for interpolation if close to midnight) but every hour of the day (UTC 00, 03, 06, 09, 12, 15, 18 and 21) and then take these values and interpolate for the 10 minute values.
 After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe.
 After this the downloaded data is discarded and the next month is queried.
-This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (goes from several terabytes to less than a gigabyte).
+This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (a reduction from several terabytes to less than a gigabyte).
 
@@
 Figure \ref{fig:error_count} shows that most stations have nails in less that 10\% measurements.
-One has nails in 100\% of measurements.
-This is only because the number of measurements that matched previous criteria was 1.
+The stations that have higher thatn 10\% error, have only few observations that meet the criteria (< 10) and are ignored.
 
@@
 This is done in a similar manner to the interpolation of the CARRA data.
-Weights are assigned to each of the four points bounding the point of interest.
-These weights are then used to calculate the weighted average that represents the interpolated value.
+The four points bounding the point of interest were used to linearly interpolate the value of the point of interest.
 This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.
@@
 Once data has been retrieved for all three sources and processed, including interpolating values, it needs to be made ready to use by the model, for both training, validation and test.
-Starting with a dataframe that contains measured information from AWS.
+The starting point is a dataframe that contains measured information from AWS.
 This includes the average wind, the wind gust, wind direction along with the station number and coordinates.
-When selecting the CARRA data certain height levels are chosen.
-That is at which heights above ground the reanalysis parameters are requested.
+When selecting the CARRA data certain height levels are chosen, which heights above ground the reanalysis parameters are requested.
 These present as separate lines in the CARRA dataframe.
-That is information for one observation is represented in as many lines as height levels requested in the reanalysis data.
+Information for one observation is represented in as many lines as height levels requested in the reanalysis data.
 These rows need to be combined on the position (the weather station) and time so that each row contains all reanalysis information for a given measurement of a weather station.
@@
 A sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point.
-In any case the points, that represent these sections, were selected like as shown in Code Listing \ref{code:sectorElevation}.
-That is a range of angles are defined based on the wind direction $d$ at some distance from the given point.
+In any case the points, that represent these sections, were selected as shown in Code Listing \ref{code:sectorElevation}.
+A range of angles are defined based on the wind direction $d$ at some distance from the given point.
 This means that the resultant points (equal in number to the length of angleRange by k) form sectors at several distances from the given weather station.
@@
 
-Looking at Table \ref{table:trainDataExample} note that the last \nPCA columns represent the principle component analysis of the elevation data.
+Looking at Table \ref{table:trainDataExample} note that the last column shown before ellipsis contains the first principle component in principles component analysis.
+In total there are \nPCA columns representing the principle component analysis of the elevation data.
 Also note that the first four columns represent two variables that describe the stability of the air.
@@
 They are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}.
-They are calculated using reanalysis data about the the weather at two different height levels.
+They are calculated using reanalysis data at two different height levels.
 Thus $Ri_{01}$ refers to the Richardson number calculated between height levels 0 and 1.
@@
 
-Here, $g$ is the acceleration due to gravity, $dT$ is the temperature difference between the two height levels, dz is the elevation difference, $T_{\textrm ave}$ is the average temperature (that is the average of the two temperatures in the height levels) and $dU$ is the wind speed difference between the two height levels.
+Here, $g$ is the acceleration due to gravity, $dT$ is the temperature difference between the two height levels, $dz$ is the elevation difference, $T_{\textrm ave}$ is the average temperature (that is the average of the two temperatures in the height levels) and $dU$ is the wind speed difference between the two height levels.
 Both of these numbers provide some insight about the stability of the air.
-A lower value for the Richardson number indicates a higher turbulance.
-A typical range of values could be between 10 and 0.1, with values below 1 indicating significant turbulance\cite{richardson_number_skybrary}.
+A lower value for the Richardson number indicates a higher turbulence.
+A typical range of values could be between 10 and 0.1, with values below 1 indicating significant turbulence\cite{richardson_number_skybrary}.
 When the square of the Brunt-Väisälä frequency is negative, then the air is unstable (an air parcel will move away from its original position)\cite{brunt_vaisala_freq_eumtrain}.
 These are derived factors from the reanalysis data and as such there shouldn't be a significant information gain using $Ri$ and $N$ as opposed to having the raw data.
-Including these factors instead of every reanalysis variable requested might speed up training as well as making the model more easily explainable with the use of Shapley values or other tools for explainability.
+However, including these factors instead of every reanalysis variable requested might speed up training as well as making the model more easily explainable with the use of Shapley values or other tools for explainability.
 Using Shapley a feature importance value is attributed to a given feature by creating all possible permutations of any possible length (up to number of features) and seeing how the predictions are skewed when the given parameter is included or excluded.

### 2024-04-10 Chapters/Chapter4.tex
.

@@
 This is simply a dense layer with 1 unit.
-A grid search was performed to determine the hyperparameters that minmize the loss.
+A grid search was performed to determine the hyperparameters that minimize the loss.
 Hyperparameters are parameters that are set before the training begins\cite{hyperparameters_definition}.
@@
 The second data source for the training data, is the elevation data.
-IMO provided a GeoTif file that contains elevation information above sea level, covering Iceland, in a 20 by 20 grid.
-A couple of different landscape distributions were looked at.
+Using the GeoTif file that IMO provided, several different landscape elevation distributions were looked at.
 A sector upwind, a sector upwind and downwind as well as a circle surrounding a weather station.
 
-To add to these observed values and reanalysis, two derived variables, the Richardson number and the Brunt-Väisälä frequency, were calculated.
-To begin with all the available parameters were used to train the model along with the derived variables to see.
+To begin with all the available parameters were used to train the model along with the derived variables.
 This is done to be able to then use tools such as Shapley to see which features are impacting the predictions.

### 2024-04-30 Chapters/Chapter1.tex
.

@@
 The gust factor is defined as the peak gust divided by the mean wind speed over some defined time period.
-The peak wind gust is often defined as the highest 3 second rolling average measured wind speed over a period of 10 minutes, while the mean wind is the average of all measurements in the 10 minute interval.
+The peak wind gust is often defined as the highest 3 second rolling average measured wind speed over a period of 10 minutes, while the mean wind is the average of all measurements in the 10 minute intervalf.
+This thesis uses this definition.
 This varies, with the US using a 1 minute interval, leading to 14\% higher results \cite{why_wind_gusts}.
-As the Navier Stokes Equation (\ref{eq:navierstokes}) shows the variability of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth, and resistance.
-It states that the change in wind is described by the pressure gradient, the oscillation, the acceleration due to gravity and the resistance from the landscape.
+The Navier Stokes Equation (\ref{eq:navierstokes}) shows that the change of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth (the Coriolis force), and frictional force.\cite{uncertainties_in_numerical_weather_predictions}
 \begin{equation}
@@
 \end{equation}
-Equation (\ref{eq:navierstokes}) relates the change in wind w.r.t.
-time ($\frac{\delta \boldsymbol{V}}{\delta t}$) and change in wind w.r.t.
-position ($\nabla\boldsymbol{V}$) to pressure change w.r.t.
-position ($\nabla\boldsymbol{P}$).
-The wind is also influenced by the oscillation, gravity and resistance but the main drive of wind is pressure gradient\cite{uncertainties_in_numerical_weather_predictions}.
 
@@
 These models describe the transition between discretized packages of atmospheric states using partial differential equations based on physical reality.
-These results are usually published for 3-6 hour intervals.They describe the state over the period and so do not necessarily grasp fluctuations well.
+These results are usually published every hour, or at courser time intervals for climate simulations.
+With increasing computer power and efficiency the trend is to output data more often\cite{GNP_vidtal}.They describe the state over the period and so do not necessarily grasp fluctuations well.
 These fluctuations would include fluctuations in the wind speed, wind gusts\cite{canNNBeatNWP}.
@@
 This thesis looks at how best to predict gust factor based on various factors, using several different data sources, including NWP and observations.
-NWP forecasts or outputs analysis, traditionally, at intervals of around 3-6 hours and thus, alone, do not capture the variability that is the essence of wind gusts.
-This is why another model that uses the information produced by the NWP along with observations and landscape information, is needed to try to accurately predict wind gusts.
 Being able to accurately predict the wind gust is important as it is often the peak wind gusts that will cause failures in structures.
@@
 The history of numerical weather predictions goes all the way back to the 1920's when Lewis Fry Richardson pioneered the field and tried to produce forecasts.
-The results were flawed and impractical because of the time it took to calculate was much longer than the forecast period.
+The results were flawed due to noise in the calculations.
+ENIAC was built in 1945, it was a general purpose computer that was used, among other things, to make predictions.
+These predictions took 24 hours to make and were predicting 24 hours into the future.
+It was a proof of concept but not usable\cite{TheENIACForecastsARecreation}.
 In the 1950's, with the advent of computers the first operational forecasts emerged.
@@
 The results were inferior to subjective human-based forecasts but showed that such forecasts were feasible and promoted further development in the area \cite{historyNWP}.
-The field underwent significant transformation over the next 50 years, and predictions got significantly better.
+The field of NWP has taken great strides since then following the development of computer power an efficiency.
 
@@
 It is to be noted that images are grid data with some number of channels.
-Using CNNs could work on any gridded data where there is some spatial features\cite{canNNBeatNWP}.
+Using CNNs could work on any gridded data where there are some spatial features\cite{canNNBeatNWP}.
 Since 2018, there has been significant work done in the weather prediction field using AI.
@@
 Now they are competitive.
+It is worth noting that the training of these large models is based on data from traditional large weather models.
 It will be very interesting to watch what the next few years will have in store for the development of machine learning in weather predictions.
@@
 They looked at data from automatic weather stations that measure wind at 10 meters above ground.
-A common definition of wind gust is the maximum average 1 second measured wind speed over a rolling 3 second interval using all measurements over a ten minute span.
-A common definition of average wind is the average of the measured wind speed over 10 minutes.
-These are the definitions used both by H.
-Ágústsson and H.
-Ólafsson in 2004 as well as this thesis.
 The data that was studied in 2004 comes from the same source as used in this thesis, but limits itself to a smaller section.
@@
 These were $d_m, D, H$, that is direction of wind blowing off a mountain, distance to the mountain and the height of the mountain above the weather station.
-Their main results were these, gust factor is inversely correlated to the distance from a mountain and correlated to the height of the mountain.
+Their main results were that the gust factor is inversely correlated to the distance from a mountain and correlated to the height of the mountain.
 The study in 2004 looked at the effect of a dominant point upwind.

### 2024-04-30 Chapters/Chapter2.tex
.

@@
 The measurements that met the filtering criteria, started in \startDateVedur and ended in 2023 (the year measurements were provided).
-Of these \nStationsMin stations, \nVedurMin were from IMO Íslands and placed at 10 meters above ground, while the rest (\nVGMin) were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)} and placed at 7.5 meters above ground.
+Of these \nStationsMin stations, \nVedurMin were from IMO Íslands and placed at 10 meters above ground, while the rest (\nVGMin) were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)} and placed at 6-7 meters above ground\cite{vegagerdin_postur}.
 The location of these weather stations can be seen in Figure \ref{fig:aws_map}.
@@
 Nails, are sharp increases from the rest of the data and are unrealistic outliers that are considered measurement errors and are discarded.
-Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another data set), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction.
-These measurements were started in the end of the 20th century, when the first AWS stations were installed.
-More have been added in the over two decades since then.
+Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another data set), the average wind speed ($f$), the wind gust ($f_g$), the standard deviation of the wind gust, the direction of the wind ($d$) and the standard deviation for the wind direction.
+These measurement started at the end of the 20th century, when the first AWS stations was installed.
+More have been added in the following decades.
 This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.
@@
 \begin{figure}
+    \centering
     \includegraphics[scale = 0.75]{Figures/weather_stations_2024-04-09.png}
-    \caption{Locations of automatic weather stations in Iceland}
+    \caption[Locations of automatic weather stations in Iceland]{Locations of all 412 stations that were looked at in this study.
+Most of these were from IMO but over a hundred were from IRCA.
+IMO AWS are placed at 10 meters above ground, while IRCA stations are placed at around 6-7 meters above ground.}
     \label{fig:aws_map}
@@
 
-\begin{lstlisting}[style = Python, caption = {Filter points over 48 hour interval}, label = code:filter48hr]
+\begin{lstlisting}[style = Python, caption = {[Filter points over 48 hour interval] Filter all measured points so to only keep the highest value in any 48 hour interval for a given station.
+For a given station, the highest wind speed recorded is selected.
+Then any measurement in a 48 hour interval, centered on the highest wind speed, for that given station, is removed.
+This is then done iteratively until we are left with an empty dataframe.}, label = code:filter48hr]
     for station in tqdm(stations, total = len(stations)):
@@
     \end{center}
-    \caption{Measurements that fit filtering by year (from \startDateVedur to 2023).}
+    \caption[Measurements that fit filtering by year (from \startDateVedur to 2023).]{The total number of measurements by year after filtering by average wind speed limit of \averageWindSpeedLimit m/s and clearing the neighborhood using filtering method described in code listing.
+The measurements provided were sparse for years before 2005, as can be seen for 2004.
+Years before 2005 will thus be ignored.}
     \label{fig:occurrences_by_year}
@@
 \begin{figure}
-    \begin{center}
+    \centering
     \includegraphics[scale = 0.75]{Figures/station_occurrences_20.png}
-    \end{center}
-    \caption{Number of stations that have measurements in a range.}
+    \caption[Number of stations that have measurements in a range.]{The number of stations that have a-b measurements represented in filtered data.
+That is station x may have c measurements represented and $\protect a <= c < b$.
+Station x will then add to the count of column labeled $\protect [a, b[$.}
     \label{fig:occurrences_by_station_20}
@@
 For each such observation in the IMO data, two API calls excluding redundancy are needed.
-One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly one of the three hour interval times.
+One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly on one of the three hour interval times.
 That is if the observation was at 13:00, CARRA data for noon and 15:00 would have to be queried.
@@
 
+\subsection{Statistical Analysis of Measurements}
+The data gathered from AWS showed inconsistencies due to significant gaps in data coverage in the early 2000s and 90s.
+This is shown in Figure \ref{fig:measurements_by_year}.
+It is important to note that Figure \ref{fig:measurements_by_year} and subsequent figures show number of measurements in logarithmic scale.
+Late in the process of making this thesis, it was discovered that data was missing.
+Most data points before 2005 had not been provided.
+Eventually this missing data was provided but the decision was made to exclude that data due to time constraints.
+Using only data from 2005 to 2023, means that all years cover the entire year.
+Thus the data was filtered to only include these years.
+For some reason there were a hundred or so measurements from 1990.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.75]{Figures/measurements_by_year.png}
+    \caption[Number of measurement data points by year.] {Number of measurement data points by year.
+The effect of filtering by average windspeed limit shown in the orange overlay.
+It is important to note that the y-axis scale is logarithmic.
+If this was not the case, then the filtered values would not be clearly displayed.}
+    \label{fig:measurements_by_year}
+\end{figure}
+
+It is a good sanity check to look at figures like Figure \ref{fig:measurements_by_year}.
+If there were massive discrepancies by year, then some data is missing.
+Another way to look at the distribution of measurement over time is to look at it by month, as in Figure \ref{fig:measurements_by_month}.
+The total measurements is evenly distributed over the months.
+As expected higher winds occur during the winter months, with only a few hundred wind measurements reaching \averageWindSpeedLimit.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.75]{Figures/measurements_by_month.png}
+    \caption[Number of measurement data points by month.]{Number of measurement data points by month.
+The effect of filtering by average windspeed limit shown in the orange overlay.
+This filtering shows us what we would expect to see.
+We can see that we have fewer extreme wind speeds during summer as compared to winter.}
+    \label{fig:measurements_by_month}
+\end{figure}
+
+Another thing to look at is the number of average wind speed measurements by interval wind speed interval.
+This is shown in Figure \ref{fig:measurements_by_windspeed}.
+This figure shows the expected, the number of measurements at higher wind speeds decrease.
+The orange overlay shows the section that is considered in this thesis.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.75]{Figures/measurements_by_windspeed.png}
+    \caption[Number of measurement data points by wind speed.]{Number of measurement data points by wind speed.
+The effect of filtering by average windspeed limit shown in the orange overlay.
+The filtering obviously just selects the columns to the right and including column $\protect [20, 25[$.}
+    \label{fig:measurements_by_windspeed}
+\end{figure}
+
+Looking at Figure \ref{fig:measurements_by_windgust} something strange seems to be happening.
+The orange overlay shows the number of gust values where the wind speed is over \averageWindSpeedLimit m/s and some of the gust values is lower than that.
+This shows the relevancy of looking at such figures.
+These data points, where the average wind speed is higher than the gust speed, are filtered out later in the preprocessing pipeline.
+Note that there are seemingly no missing values for wind speed, wind gust and wind direction, but there is missing standard deviations for these parameters.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.75]{Figures/measurements_by_windgust.png}
+    \caption[Number of measurement data points by wind gust.]{Number of measurement data points by wind gust.
+The effect of filtering by average wind speed limit shown in the orange overlay.
+Here we might expect to see some discrepancies in the columns.
+That is, the orange overlays should not completely cover all columns.
+This is because we select by windspeed and not gust.
+There should be no coverage below the average wind speed limit of \averageWindSpeedLimit m/s.}
+    \label{fig:measurements_by_windgust}
+\end{figure}
+
 \section{Elevation data}
-IMO provided a tif file containing the elevation of Iceland on a 20 meter by 20 meter grid.
+IMO provided a TIFF file containing the elevation of Iceland on a 20 meter by 20 meter grid.
 This file encompasses Iceland and is around 685 MB.

### 2024-04-30 Chapters/Chapter3.tex
.

@@
     \label{fig:data_preprocessing_flow_chart}
-    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart-2024-04-08.png}
+    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart.png}
 \end{figure}
@@
 Figure \ref{fig:error_count} shows that most stations have nails in less that 10\% measurements.
-The stations that have higher thatn 10\% error, have only few observations that meet the criteria (< 10) and are ignored.
+The stations that have higher than 10\% error, have only few observations that meet the criteria (< 10 measurements) and are ignored.
 
-The elevation data comes in a GeoTif file that covers Iceland.
+The elevation data comes in a GeoTIFF file that covers Iceland.
 It is a rectangular grid of resolution 20 meters.
@@
 \begin{table}[h]
-    \caption{An example of data structure used with model}
+    \caption[An example of data structure used with model]{An example of structure of data used to train model.
+Data points include the derived variables Ri and N, the elevation of the station, direction of wind and relative direction of the wind (that is the direction of the wind relative to center of Iceland), along with some combination of wind speed, pressure and temperature at the different height levels.
+Finally there are the elevation points around a given station.
+The example below only shows data derived variables elevation and relative corner.}
     \label{table:trainDataExample}
@@
     \begin{tabular}{c|c|c|c|c|c|c|c}
-         Ri\_01 & Ri\_12 & N\_01 & N\_12 & station\_elevation & relative\_corner & PC1 & ...\\\hline
+         Ri\_01 & Ri\_12 & N\_01 & N\_12 & station\_elevation & relative\_corner & elevation\_point\_0 & ...\\\hline
          -1.18e+00 &  2.67e+04 & -8.57e-06 & 6.78e-05 & 3.34e+01 & 2.73e+00 & 1.36e+01 & ... 
@@
 Also note that the first four columns represent two variables that describe the stability of the air.
-These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency\cite{brunt_vaisala_freq_eumtrain} ($N$).
-They are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}.
-They are calculated using reanalysis data at two different height levels.
+These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency\cite{brunt_vaisala_freq_eumtrain} ($N$), and are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}.
+These values are calculated using reanalysis data at three different height levels.
 Thus $Ri_{01}$ refers to the Richardson number calculated between height levels 0 and 1.
 Exactly the same notation is used with the Brunt–Väisälä frequency.
-The height levels used were 15, 250 and 500 meters above the ground.
+The height levels used were 15, 250 and 500 meters above the ground and the three different sets of height levels were (15 m, 250 m), (250 m, 500 m) and (15 m, 500 m).
 
@@
 A lower value for the Richardson number indicates a higher turbulence.
-A typical range of values could be between 10 and 0.1, with values below 1 indicating significant turbulence\cite{richardson_number_skybrary}.
+A typical range of values could be between 0.1 and 10, with values below 1 indicating significant turbulence\cite{richardson_number_skybrary}.
 When the square of the Brunt-Väisälä frequency is negative, then the air is unstable (an air parcel will move away from its original position)\cite{brunt_vaisala_freq_eumtrain}.
@@
 Using Shapley a feature importance value is attributed to a given feature by creating all possible permutations of any possible length (up to number of features) and seeing how the predictions are skewed when the given parameter is included or excluded.
-This the needs to be done for all parameters.
+This   needs to be done for all parameters.
 The time complexity of this is very high ($2^n$ coalitions)\cite{shapley_information}.

### 2024-04-30 Chapters/Chapter4.tex
.

@@
     \centering
-    \caption{Hyperparamter search with best performing combination colored red}
+    \caption[Hyperparamter search with best performing combination.]{Hyperparamter search with best performing combination shown.
+Hyperparameter search was done using hyperband algorithm that initially searches randomly for the best parameters but then hones in on the what is working and as such is neither exhaustive nor completely random.
+This means that an upper limit won't be set on the numbe of combinations to try like with randomsearch.}
     \label{table:gridSearchHyperparamters}
-    \begin{tabular}{c|c}
-        Layers &  4, \textcolor{red}{5}, 6, 7, 8, 9\\\hline
-        Units &  128, 256, 512\\\hline
-        Epochs & 100, 200, 500\\\hline
-        Batch size & 32, 64, 128, 256\\\hline
-        Optimizers & Adam, RMSprop, Adamax\\\hline
-        Penalties & 1, 0.1, 0.01, 0\\\hline
+    \begin{tabular}{c|c|c}
+        Parameter & Range of values & Selected\\\hline
+        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{7}\\\hline
+        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{256}\\\hline
+        Penalties & min\_value = 1e-4, max\_value = 1, sampling = log & \textbf{red}{0.1}\\\hline
+        Epochs & min\_value = 50, max\_value = 1000, step = 50 & \textbf{500}\\\hline
+        Optimizers & Adam, RMSprop, Adamax & \textbf{Adam}\\\hline
+        Activation & ReLU, ELu, Softmax & \textbf{ReLU}\\\hline
     \end{tabular}
@@
 The second data source for the training data, is the elevation data.
-Using the GeoTif file that IMO provided, several different landscape elevation distributions were looked at.
+Using the GeoTIFF file that IMO provided, several different landscape elevation distributions were looked at.
 A sector upwind, a sector upwind and downwind as well as a circle surrounding a weather station.
@@
 PCA is a statistical procedure that allows the summarization of multivariate data to lower dimension and thus can ease visualization or increase speed of training a complex model with only minimal adverse effect to performance\cite{pca_information}.
-Another thing that might serve the same purpose is to select some number of $m$ points that have the most increase or decrease in elevation from the weather station along with the coordinates of those points in relation to the weather station and the direction of the wind.
 
@@
 \begin{figure}[h]
-    \caption{Feature importance of a neural network as described with 5 hidden layers and 256 units in each}
-    \label{fig:ShapleyWaterfallFirstTry}
     \centering
     \includegraphics[scale = 0.6]{Figures/shap_bar_nn_256_example100.png}
+    \caption[Feature importance of a neural network.]{Feature importance of a neural network as described with 5 hidden layers and 256 units in each}
+    \label{fig:ShapleyWaterfallFirstTry}
 \end{figure}

### 2024-04-30 Chapters/Chapter5.tex
.

@@
 
-\section{Model structure}
-This baseline model gave an average error of around 10\%.
+\section{Results}
+A baseline model was constructed.
+This model looked at the gust factor for some training data and took the average of this and predicted this average everytime.
+The baseline model gave an average error of around \baselineerror\%.
+This sets a goal.
+A model that does not significantly improve on this baseline suggests either failure to capture essential patterns in the data or that the data itself may lack the necessary information for substantial improvements upon the baseline.
+Using the previously described neural network architecture a mean absolute percentage error of \modelerror\% was achived.
+This is some improvement upon the baseline error, with a decrease in error of around \errorImprovement\%.
+
+\section{Discussion}
+Looking only at the error, the performance of the model can be gauged.
+However it does not help us explain why the model made these predictions.
+Here feature interpretation can be of great help.
+Using Shapley values and the python SHAP package, the contribution of each attribute can be given a value, as previously mentioned.
+Figure \ref{fig:SHAP_final_model} shows the average contribution of each feature.

### 2024-05-07 Chapters/Chapter4.tex
.

@@
 Hyperparameter search was done using hyperband algorithm that initially searches randomly for the best parameters but then hones in on the what is working and as such is neither exhaustive nor completely random.
-This means that an upper limit won't be set on the numbe of combinations to try like with randomsearch.}
+This means that an upper limit will nont be set on the number of combinations to try like with randomsearch.}
     \label{table:gridSearchHyperparamters}
@@
         Parameter & Range of values & Selected\\\hline
-        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{7}\\\hline
-        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{256}\\\hline
-        Penalties & min\_value = 1e-4, max\_value = 1, sampling = log & \textbf{red}{0.1}\\\hline
-        Epochs & min\_value = 50, max\_value = 1000, step = 50 & \textbf{500}\\\hline
-        Optimizers & Adam, RMSprop, Adamax & \textbf{Adam}\\\hline
-        Activation & ReLU, ELu, Softmax & \textbf{ReLU}\\\hline
+        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{11}\\\hline
+        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{64}\\\hline
+        Penalties & min\_value = 1e-4, max\_value = 1, sampling = log & \textbf{0.000168}\\\hline
+        Epochs & min\_value = 50, max\_value = 1000, step = 50 & \textbf{750}\\\hline
+        Optimizers & Adam, RMSprop, Adamax & \textbf{RMSProp}\\\hline
+        Activation & ReLU, ELu, Softmax & \textbf{eLU}\\\hline
     \end{tabular}

### 2024-05-07 Chapters/Chapter5.tex
.

@@
 This is some improvement upon the baseline error, with a decrease in error of around \errorImprovement\%.
+The power generated by wind mill increases with wind speed cubed \cite{wind_power}.
+The highest wind gusts in Iceland are around 70 m/s.
+Knowing the gust factor to 2 percentage points better than before can allow for the anticipation power generated by wind mills in the kilowatts per square meter of swept area.
 
 \section{Discussion}
+The error improvement is modest in absolute terms but represents 21\% over the baseline.
+This would indicate significant improvement.
+A point to consider, that might heavily impact the results, is that the reanalysis data, sourced from CARRA, represents some kind of 3 hour average and therefore will be shifted towards lower values for wind speed when compared to the 10 minute measurements provided by IMO.
+The baseline model uses measured information to calculate the average gust factor, while the model does not have access to any measured data directly (the reanalysis data is trained on some measured data).
+The expected shape of the measured wind shape distribution, after filtering and nailstripping, is a bell curve with a steep tail to the right and no tail to the left and this proves to be true.
+If the reanalysis data would closely follow the measured wind speed, the same distribution should be observed.
+This is not the case.
+The distributions can be seen in Figure (\ref{fig:difference_measured_reanalysis})
+
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.65]{Figures/difference_measured_reanalysis_cropped.png}
+    \caption[Measured wind speed and corresponding points for reanalysis.]{Measured wind speed and corresponding points for reanalysis.
+The two distributions vary greatly.
+Some variation is expected, as the reanalysis represents some kind of average.}
+    \label{fig:difference_measured_reanalysis}
+\end{figure}
+
+Using any information from the measurement data to improve the data, would induce data leakage.
+Any simple mathematical transform would not be expected to help the model as the model would simply be able to learn the pattern.
+A simple check was done to confirm this.
+The reanalysis wind speed was transformed using code shown code listing \ref{code:custom_distribution_transform}.
+
+\begin{lstlisting}[style = Python, caption = {Custom transform method}, label = code:custom_distribution_transform]
+    def custom_distribution_transform(x, threshold=20):
+        max_decay, max_growth = 1.2, 6
+        decay = lambda z: 1 + (max_decay-1) 
+            * np.exp(threshold - z)
+        growth = lambda z: 1 + (max_growth - 1) 
+            * np.exp(z - threshold)
+        amplification_factors = np.where(x <= threshold,
+             growth(x), decay(x))
+        y = amplification_factors * x + 8
+        return y
+\end{lstlisting}
+
+This shift changes the shape of the reanalysis data distribution and shifts it to the right. The updated distribution can be seen in Figure (\ref{fig:transformed_reanalysis_measured}).
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.5]{Figures/transformed_reanalysis_measured.png}
+    \caption[Transformed reanalysis wind speed and measured wind speed.]{The distributions of the transformed 15 meter height wind speed of reanalysis data, along with the 10 meter measured wind speed.}
+    \label{fig:transformed_reanalysis_measured}
+\end{figure}
+
+Training two identical models, one with the transformed reanalysis data and the other with original reanalysis data, results in two near identical errors, as expected.
+As no new information was input, the data was only manipulated, the model evaluates to the same error.
+The model that used the transformed reanalysis data was slightly quicker to converge.
+
 Looking only at the error, the performance of the model can be gauged.

### 2024-08-27 Chapters/Chapter1.tex
Updated after new CARRA data sourced and run

@@
 The results were inferior to subjective human-based forecasts but showed that such forecasts were feasible and promoted further development in the area \cite{historyNWP}.
-The field of NWP has taken great strides since then following the development of computer power an efficiency.
+The field of NWP has taken great strides since then following the development of computer power and efficiency.
 
-Currently there has been another transformation in the field of weather prediction driven by artificial intelligence.
+In the last decade there has been another transformation in the field of weather prediction driven by artificial intelligence.
 Interest in AI has come in waves.
 Some progress is made, then interest dwindles.
-The interest has been increasing steadily since 2010.
+Interest in AI has been increasing steadily since 2010.
 Notable work that has driven this wave of interest include increase in computational abilities due to parallel processing in graphical processing units (GPU), convolutional neural networks (CNN), which allowed much faster processing of massive (image) datasets and the availability of large datasets online.
@@
 \href{https://www.youtube.com/watch?v=PD1v5PCJs_o&ab_channel=GregBronevetsky}{GraphCast})\cite{SecondRevolution}.
-This is done to show the versability of the models.
+This is done to show the versatility of the models.
 \end{enumerate}
+
 Before 2022, LWM had been shown to be able to compete with traditional NWP for some specifc cases as well as making predictions quicker, after training.
@@
 It uses reanalysis data, along with elevation data to predict the gust factor.
-It looks at the data at the point of interest and does not look at the data as a time series.
+It looks at the data at the point of interest.
+It does not look at the data as a time series.
 This thesis aims to improve on the baseline model of always predicting the mean gust factor and show that some structure can be learned from reanalysis data about gusts.
@@
 It did not look at the effects of the landscape more broadly.
-In this study, landscape upwind is looked at along with landscape downwind.
-Both the upwind and downwind landscape can influence the gustiness at a certain point \cite{GNP_vidtal}.
+In this study, landscape upwind is looked at.
 

### 2024-08-27 Chapters/Chapter2.tex
Updated after new CARRA data sourced and run

--- 2024-04-10 Chapters/Chapter2.tex
@@
 
-\chapter{Data gathering and preprocessing} % Main chapter title
-
-\label{Chapter2} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}
-
-%----------------------------------------------------------------------------------------
-%	SECTION 1
-%----------------------------------------------------------------------------------------
-
+\chapter{Data gathering and preprocessing}
+\label{Chapter2}
 Data was sourced from several streams.
-The Icelandic Meterological Office (IMO) provided measurements from weather stations all around the Iceland.
+The Icelandic Meterological Office (IMO) provided measurements from weather stations all around Iceland.
 NWP data was downloaded from Copernicus Arctic Regional Reanalysis dataset (CARRA).
@@
 IMO provided 10 minute measurements from \nStationsMin weather stations all around Iceland.
-The measurements that met the filtering criteria, started in \startDateVedur and ended in 2023 (the year measurements were provided).
-Of these \nStationsMin stations, \nVedurMin were from IMO Íslands and placed at 10 meters above ground, while the rest (\nVGMin) were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)} and placed at 7.5 meters above ground.
+The measurements that met the filtering criteria, started in \startDateVedur and ended in 2023.
+Of these \nStationsMin stations, \nVedurMin were from IMO and placed at 10 meters above ground, while the rest (\nVGMin) were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)} and placed at 6-7 meters above ground\cite{vegagerdin_postur}.
 The location of these weather stations can be seen in Figure \ref{fig:aws_map}.
-The information that is provided by these Automatic Weather Stations (AWS) is presented in two different type of data files, hourly and 10 minute files.
+The information that is provided by these Automatic Weather Stations (AWS) is presented in two different types of data files, hourly and 10 minute files.
 The hourly files are summations of the 10 minute files, with the exception that errors, such as nails, still in the 10 minute files should have been removed from the hourly documents.
 Nails, are sharp increases from the rest of the data and are unrealistic outliers that are considered measurement errors and are discarded.
-Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another data set), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction.
-These measurements were started in the end of the 20th century, when the first AWS stations were installed.
-More have been added in the over two decades since then.
+Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another data set), the average wind speed ($f$), the wind gust ($f_g$), the standard deviation of the wind gust, the direction of the wind ($d$) and the standard deviation for the wind direction.
+These measurement started at the end of the 20th century, when the first AWS stations was installed.
+More have been added in the following decades.
 This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.
@@
 \begin{figure}
-    \includegraphics[scale = 0.75]{Figures/weather_stations_2024-04-09.png}
-    \caption{Locations of automatic weather stations in Iceland}
+    \centering
+    \includegraphics[scale = 1]{Figures/stationsOverIceland_2024-05-16_stripped_to_frame.png}
+    \caption[Locations of automatic weather stations in Iceland]{Locations of all 412 stations that were looked at in this study.
+Most of these were from IMO but over a hundred were from IRCA.
+IMO AWS are placed at 10 meters above ground, while IRCA stations are placed at around 6-7 meters above ground.}
     \label{fig:aws_map}
-\end{figure}
-
-These 10 minute measurements were filtered to only include measurements where the average wind speed was at or above \averageWindSpeedLimit m/s.
-They were also filtered in such a way that only the highest value in any given 48 hour span is considered.
-This can be seen in Code Listing \ref{code:filter48hr}.
-Any measurement that is within a 48 hour interval centered on of the current highest is removed.
-This is done for each station and iteratively until all measurements have been removed.
-
-\begin{lstlisting}[style = Python, caption = {Filter points over 48 hour interval}, label = code:filter48hr]
-    for station in tqdm(stations, total = len(stations)):
-        subset_df = vedur_df[station == vedur_df.stod]
-        subset_df = subset_df.reset_index(drop = True)
-
-        while not subset_df.empty:
-            idx = subset_df.f.idxmax()
-            time_of_max = subset_df.iloc[idx].timi
-
-            filtered_data.append(subset_df.iloc[idx])
-
-            subset_df = subset_df[abs(subset_df.timi 
-                - time_of_max) >= pd.Timedelta(threshold)]
-
-            subset_df = subset_df.reset_index(drop = True)  
-\end{lstlisting}
-
-When these two filterings have been done, the end result is \nStationsMin stations that meet these criteria.
-Of these, \nVedurMin belong to IMO and \nVGMin belong to IRCA.
-The number of observations by year can be seen in Figure \ref{fig:occurrences_by_year}.
-The first observations occurring in \startDateVedur and the latest in 2023.
-
-\begin{figure}
-    \begin{center}
-        \includegraphics[scale = 0.75]{Figures/occurrences_by_year.png}
-    \end{center}
-    \caption{Measurements that fit filtering by year (from \startDateVedur to 2023).}
-    \label{fig:occurrences_by_year}
-\end{figure}
-
-Looking at the number of measurements that meet these criteria for a given station, Figure \ref{fig:occurrences_by_station_20} shows only one station has over 1000 measurements that fit the criterias.
-
-\begin{figure}
-    \begin{center}
-        \includegraphics[scale = 0.75]{Figures/station_occurrences_20.png}        
-    \end{center}
-    \caption{Number of stations that have measurements in a range.}
-    \label{fig:occurrences_by_station_20}
 \end{figure}
@@
 This is covered by CARRA.
-A few of the newer points from AWS were not available from CARRA when data was gathered.
-The CARRA dataset is available for two regions, west and east.
+The CARRA dataset is available for two regions, West and East.
 Each of these covers a vastly larger area than the area of interest.
-This leads to having to store a large amount of data, because it is not possible to ask for specific points of interest.
+This leads to having to store a large amount of data.
 To get the data one has two options.
@@
 Using the API client is the only realistic option here, as there were thousands of requests made for different times.
+If using the API, it is possible to query a smaller area (such as a rectangular area around Iceland) given a set of coordinates.
 
-Each request is made by going through all instances of the IMO data after filtering (by mean wind speed reaching \averageWindSpeedLimit m/s over a 10 minute interval).
-For each such observation in the IMO data, two API calls excluding redundancy are needed.
-One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly one of the three hour interval times.
-That is if the observation was at 13:00, CARRA data for noon and 15:00 would have to be queried.
-Date and times were generated automatically from the AWS and requested by calling the client.
+The requests to the API are made at each available CARRA hour ([00, 03, 06, 09, 12, 15, 18, 21]) for each available observation.
+Only these hours can be used as the CARRA predicted output is represents the wind speed at those given hours and would thus not encapsulate well the extreme values in between these times.
 Using these datapoints, interpolation was used to get an estimation for the point of the given weather station.
@@
 These are single levels, model levels, height levels, pressure levels.
-The data for this observation was downloaded from height levels.
-That is, data was requested at heights of 15, 150, 250 and 500 meters above ground.
+The data for this thesis was downloaded from height levels.
+That is, data was requested at heights of 15, 250 and 500 meters above ground.
 For each point 4 parameters were requested, wind speed, wind direction, pressure and temperature.
@@
 
-After using this method to request terabytes of data, it was discovered that it was possible to query a specific area.
-This decreased the size of each file from around 50 MB to around 2 MB.
-The bottleneck for retrieving the data is the request time.
-The time the request is queued and running (server preparing the data) before it began downloading.
-This could range from almost immediate to 30 minutes.
-This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during \href{https://forum.ecmwf.int/t/a-new-cds-soon-to-be-launched-expect-some-disruptions/1607}{the winter and spring of 2023/2024}, which increased the wait time and sometimes resulted in queries not being responded to.
-This meant that the time it would take to retrieve the remaining information went from around a day, when the requests were at their quickest, to many months, something that would not be possible given the time frame of the project.
-Fortunately these problems were only particularly time consuming during mid winter.
-
 \section{Elevation data}
-IMO provided a tif file containing the elevation of Iceland on a 20 meter by 20 meter grid.
+IMO provided a TIFF file containing the elevation of Iceland on a 20 meter by 20 meter grid.
 This file encompasses Iceland and is around 685 MB.

### 2024-08-27 Chapters/Chapter3.tex
Updated after new CARRA data sourced and run

@@
-% Chapter Template
-\chapter{Data processing and structure} % Main chapter title
-
-\label{Chapter3} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}
-
-%----------------------------------------------------------------------------------------
-%	SECTION 1
-%----------------------------------------------------------------------------------------
-
+\chapter{Data processing and structure}
+\label{Chapter3} 
 \section{Combining data sources}
@@
 The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}).
-Even so being able to predict the gust factor is more important for higher average wind speed due to then higher wind gusts.
+Even so being able to predict the gust factor is more important for higher average wind speed due to higher wind gusts.
 After this stripped dataset over every AWS has been created it is used to query the CARRA data by using their API.
@@
 In light of these restraints, it was decided to query month by month.
-Querying only the days needed (both the days included in the 10 minute measurements and the days needed for interpolation if close to midnight) but every hour of the day (UTC 00, 03, 06, 09, 12, 15, 18 and 21) and then take these values and interpolate for the 10 minute values.
+Querying only the days needed but every hour of the day (UTC 00, 03, 06, 09, 12, 15, 18 and 21).
 After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe.
@@
 If the average wind speed differs the rows are dropped.
-This can be used to look at which stations have high percentage error.
-This can be seen in Figure \ref{fig:error_count}.
 
-\begin{figure}
-    \includegraphics[scale=0.75]{Figures/error_count.png}
-    \caption{The number of stations whose number of nails are in a certain percentage range of the overall}
-    \label{fig:error_count}
-\end{figure}
-
-Figure \ref{fig:error_count} shows that most stations have nails in less that 10\% measurements.
-The stations that have higher than 10\% error, have only few observations that meet the criteria (< 10 measurements) and are ignored.
+After filtering out nails, most stations have nails in less that 10\% measurements. The stations that have higher than 10\% error are ignored.
 
@@
 
+The error in reanalysis wind speed and measured wind speed can be significant.
+The absolute error increases as the measured wind speed increases, while the percentage wind speed decreases.
+A grouping of these errors by wind speed can be seen in Table \ref{table:measuredVSReanalysis_wind_speed}
+
+\begin{table}[h]
+    \caption[Comparison of measured and reanalysis wind speed]{Comparison of measured and reanalysis wind speed using mean absolute error (MAE) and mean absolute percentage error (MAPE).
+Note that for the computation of MAPE for ranges that otherwise include 0, 0 values have been excluded so as to prevent division by zero and exploding values.
+The comparisons are done using measured wind speed (at 10 meters above ground for IMO and 6-7 meters above ground for IRCA) and reanalysis wind speed at 15 meters above ground.}
+    \label{table:measuredVSReanalysis_wind_speed}
+    \centering
+    \begin{tabular}{c|ccc}
+        f & n & MAE & MAPE \\ \hline
+        {[}0;5{[} & 6.2e6 & 2.1 & 1.6\\
+        {[}5;10{[} & 4.2e6 & 2.2 & 0.3\\
+        {[}10;15{[} & 1.5e6 & 2.5 & 0.2\\
+        {[}15;20{[} & 3.9e5 & 3.0 & 0.2\\
+        {[}20;25{[} & 8.4e4 & 4.0 & 0.2\\
+        {[}25;$\infty${[} & 2.0e4 & 6.6 & 0.2\\
+        {[}0;$\infty${[} & 1.2e7 & 2.2 & 1.0
+    \end{tabular}
+\end{table}
+
+Another thing to look at is the distribution of error by station, both in terms of their coordinates and number of measurements.
+Looking at Figure \ref{fig:station_mae_distribution}, this distribution can be seen.
+
+\begin{figure}
+    \includegraphics[scale=0.6]{Figures/MAEoverIceland.png}
+    \caption[Distribution of mean absolute errors by station]{The distribution of mean absolute errors by station.
+Using mean absolute error instead of mean absolute percentage error allows for all points to be used.
+Mean absolute percentage error can only be used if 0 values are ignored.}
+    \label{fig:station_mae_distribution}
+\end{figure}
+
+Table \ref{table:station_mae_distribution} shows the 5 best and worst stations in terms of MAE.
+
+\begin{table}[h]
+    \caption[]{}
+    \label{table:station_mae_distribution}
+    \resizebox{\textwidth}{!}{
+    \centering
+    \begin{tabular}{c|ccc}
+        Station & Number of measurements & MAE & Location \\\hline
+        1470 & 6.8e3 & 1.17 & Reykjavík Háahlíð \\
+        1350 & 5.2e4 & 1.18 & Keflavíkurflugvöllur \\
+        1482 & 1.4e4 & 1.23 & Reykjavík Víðidalur \\
+        4921 & 1.3e4 & 1.29 & Rif á Melrakkasléttu \\
+        1477 & 5.6e4 & 1.29 & Reykjavíkurflugvöllur\\ \hdashline[0.5pt/1pt]
+        35553 & 4.0e3 & 4.30 & Almannaskarð - göng\\
+        6745 & 1.5e4 & 4.36 & Kerlingarfjöll - Ásgarðsfjall\\
+        35978 & 7.9e3 & 4.40 & Fáskrúðsfjarðargöng suður\\
+        2640 & 1.6e3 & 4.51 & Seljalandsdalur\\
+        32635 & 3.2e4 & 4.95 & Botn í Súgandafirði\\
+    \end{tabular}
+    }
+\end{table}
+
+If the 0 m/s is excluded from the measurement data, then the error distribution using MAPE can be plotted. This plot can be seen in Figure \ref{fig:station_mape_distribution}
+
+\begin{figure}
+    \includegraphics[scale=0.6]{Figures/MAPEoverIceland.png}
+    \caption[Distribution of MAPE by station]{The distribution of MAPE. Using MAPE instead of MAE, necessitates the exclusion of 0 values from the ground truth (measured values).}
+    \label{fig:station_mape_distribution}
+\end{figure}
+
 \section{Data Structure}
@@
 This includes the average wind, the wind gust, wind direction along with the station number and coordinates.
-When selecting the CARRA data certain height levels are chosen, which heights above ground the reanalysis parameters are requested.
+When selecting the CARRA data certain height levels are chosen.
 These present as separate lines in the CARRA dataframe.
 Information for one observation is represented in as many lines as height levels requested in the reanalysis data.
-These rows need to be combined on the position (the weather station) and time so that each row contains all reanalysis information for a given measurement of a weather station.
+These rows need to be combined on the position (the weather station).
 When this is done it is possible to combine the AWS IMO data and CARRA reanalysis data on the location and time columns.
 The last data source is the elevation.
-A couple of different sections of land around the weather stations have been looked at.
-A sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point.
+A sector of a circle looking upwind is looked at.
 In any case the points, that represent these sections, were selected as shown in Code Listing \ref{code:sectorElevation}.
@@
 \begin{table}[h]
-    \caption[An example of data structure used with model]{An example of structure of data used to train model.
-Data points include the derived variables Ri and N, the elevation of the station, direction of wind and relative direction of the wind (that is the direction of the wind relative to center of Iceland), along with some combination of wind speed, pressure and temperature at the different height levels.
-Finally there are the elevation points around a given station.
-The example below only shows data derived variables elevation and relative corner.}
+    \caption[An example of data structure used to train model]{An example of data structure used to train model.
+Data points include the derived variables Ri and N, the elevation of the station, direction of wind and relative direction of the wind (twd, that is the direction of the wind relative to center of Iceland), along with some combination of wind speed, pressure and temperature at the different height levels.
+Finally there are the elevation points around a given station, where the elevation is relative to the station.}
     \label{table:trainDataExample}
@@
     \centering
-    \begin{tabular}{c|c|c|c|c|c|c|c}
-         Ri\_01 & Ri\_12 & N\_01 & N\_12 & station\_elevation & relative\_corner & elevation\_point\_0 & ...\\\hline
-         -1.18e+00 &  2.67e+04 & -8.57e-06 & 6.78e-05 & 3.34e+01 & 2.73e+00 & 1.36e+01 & ... 
+    \begin{tabular}{c|c|c|c|c|c|c|c|c|c}
+         Ri & N\_squared & station\_elevation & twd & ws\_15 & wd\_15 & t\_15 & p\_15 & elevation\_point\_0 & \dots\\\hline
+         -1.18e+00 &  2.67e+04 & 100 & 1.5 & 10 & 5 & 0 & 100 & 2 & \dots 
     \end{tabular}
@@
 
-Looking at Table \ref{table:trainDataExample} note that the last column shown before ellipsis contains the first principle component in principles component analysis.
-In total there are \nPCA columns representing the principle component analysis of the elevation data.
-Also note that the first four columns represent two variables that describe the stability of the air.
+ Looking at Table \ref{table:trainDataExample} note that the first two columns represent two variables that describe the stability of the air.
 These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency\cite{brunt_vaisala_freq_eumtrain} ($N$), and are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}.
-These values are calculated using reanalysis data at three different height levels.
-Thus $Ri_{01}$ refers to the Richardson number calculated between height levels 0 and 1.
-Exactly the same notation is used with the Brunt–Väisälä frequency.
-The height levels used were 15, 250 and 500 meters above the ground and the three different sets of height levels were (15 m, 250 m), (250 m, 500 m) and (15 m, 500 m).
-
+These values are calculated using reanalysis data at two different height levels.
+Thus $Ri$ refers to the Richardson number calculated between height levels 15m and 500m.
+Exactly the same notation is used with the Brunt–Väisälä frequency, except the square is used.
 \begin{equation}

### 2024-08-27 Chapters/Chapter4.tex
Updated after new CARRA data sourced and run

--- 2024-04-10 Chapters/Chapter4.tex
@@
     \centering
-    \caption{Hyperparamter search with best performing combination colored red}
+    \caption[Hyperparamter search with best performing combination.]{Hyperparamter search with best performing combination shown.
+Hyperparameter search was done using hyperband algorithm that initially searches randomly for the best parameters but then hones in on what is working and as such is neither exhaustive nor completely random.
+This means that an upper limit will not be set on the number of combinations to try like with randomsearch.}
     \label{table:gridSearchHyperparamters}
-    \begin{tabular}{c|c}
-        Layers &  4, \textcolor{red}{5}, 6, 7, 8, 9\\\hline
-        Units &  128, 256, 512\\\hline
-        Epochs & 100, 200, 500\\\hline
-        Batch size & 32, 64, 128, 256\\\hline
-        Optimizers & Adam, RMSprop, Adamax\\\hline
-        Penalties & 1, 0.1, 0.01, 0\\\hline
+    \begin{tabular}{c|c|c}
+        Parameter & Range of values & Selected\\\hline
+        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{10}\\\hline
+        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{64}\\\hline
+        Penalties & min\_value = 1e-5, max\_value = 1, sampling = log & \textbf{1e-4}\\\hline
+        Epochs & min\_value = 10, max\_value = 1000, step = 10 & \textbf{30}\\\hline
+        Optimizers & Adam, RMSprop, Adamax & \textbf{Adamax}\\\hline
+        Activation & ReLU, ELu, Softmax & \textbf{ReLU}\\\hline
     \end{tabular}
@@
 The second data source for the training data, is the elevation data.
-Using the GeoTif file that IMO provided, several different landscape elevation distributions were looked at.
-A sector upwind, a sector upwind and downwind as well as a circle surrounding a weather station.
-
-To begin with all the available parameters were used to train the model along with the derived variables.
-This is done to be able to then use tools such as Shapley to see which features are impacting the predictions.
-Using Shapley values to see which features, will then enable the exclusion of features that don't affect the model prediction and simplify the training data.
-This means starting with much higher dimensionality of our training data and reduce it as much as possible without impacting our results.
-This crystallizes in the landscape points.
-Starting with $n$ points that describe the elevation of the landscape around our weather station.
-This $n$ might be in the hundreds.
-Looking at concentric circles outward from a given weather station, there might be 10 circles (largest with radius 20 km), each with 72 points (5° spacing) and end up with 720 points for the landscape.
-This might slow down training.
-In addition, certain landscape features might be more important than others.
-There might be a large mountain up- or downwind of a given weather station, that might be the most important feature.
-To address this principal component analysis (PCA) is used to try to reduce the dimensionality of our landscape elevation data.
-PCA is a statistical procedure that allows the summarization of multivariate data to lower dimension and thus can ease visualization or increase speed of training a complex model with only minimal adverse effect to performance\cite{pca_information}.
-Another thing that might serve the same purpose is to select some number of $m$ points that have the most increase or decrease in elevation from the weather station along with the coordinates of those points in relation to the weather station and the direction of the wind.
-
-Using all the variables from CARRA, along with the two derived factors ($Ri$, $N$) and two sectors (upwind and downwind) downscaled to 10 features using PCA, the feature importance can be seen in Figure \ref{fig:ShapleyWaterfallFirstTry}.
-The waterfall graph in Figure \ref{fig:ShapleyWaterfallFirstTry} shows that the Richardson number for higher levels is most important.
+Using the GeoTIFF file that IMO provided, a sector upwind were looked at.
 
 \begin{figure}[h]
-    \caption{Feature importance of a neural network as described with 5 hidden layers and 256 units in each}
-    \label{fig:ShapleyWaterfallFirstTry}
     \centering
     \includegraphics[scale = 0.6]{Figures/shap_bar_nn_256_example100.png}
+    \caption[Feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in table \ref{table:gridSearchHyperparamters} and data as described in table \ref{table:trainDataExample}.}
+    \label{fig:ShapleyWaterfallFirstTry}
 \end{figure}

### 2024-08-27 Chapters/Chapter5.tex
Updated after new CARRA data sourced and run

--- 2024-04-30 Chapters/Chapter5.tex
@@
-% Chapter Template
-
-\chapter{Results} % Main chapter title
-
-\label{Chapter5} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}
-
-%----------------------------------------------------------------------------------------
-%	SECTION 1
-%----------------------------------------------------------------------------------------
-
+\chapter{Results}
+\label{Chapter5}
 \section{Results}
@@
 This model looked at the gust factor for some training data and took the average of this and predicted this average everytime.
-The baseline model gave an average error of around \baselineerror\%.
+Several baseline guesses were created based on a lower limit assigned to the average wind speed limit (AWSL).
+As the average wind speed increases, then the variability in gust as a percentage decreases\cite{mean_gust_HA_HO}.
+This means looking at a subset of the data where the AWSL is higher, a better result can be expected.
+The results show this.
+Several different AWSL were used for the baseline model, as can be seen in table \ref{table:results}.
 This sets a goal.
 A model that does not significantly improve on this baseline suggests either failure to capture essential patterns in the data or that the data itself may lack the necessary information for substantial improvements upon the baseline.
-Using the previously described neural network architecture a mean absolute percentage error of \modelerror\% was achived.
+Using the previously described neural network architecture setups for each AWSL, with and without landscape elevation information, MAPE was determined.
+The results can be seen in Table \ref{table:results}.
+
+\begin{table}[h]
+    \caption[Model results for different AWSL]{MAPE for each average wind speed limit with and without landscape elevation in a 30°sector around the point of interest into the direction of the reanalysis wind.
+The percentage error is higher for lower wind speeds and thus observing the error for different lower bound of wind speed will produce different results.
+How to determine this lower bound can done by measurements (f) or reanalysis wind speed at 15 m/s ($ws_{15}$).
+The influence of adding elevation data seems to reduce the error.}
+    \label{table:results}
+    \centering
+    \begin{tabular}{c|cccccc}
+        \textbf{AWSL} & \textbf{MAPE} & & & & &\\ 
+        m/s & Baseline & &  Without elevation & & With elevation & \\
+            & $ws_{15}$ & f & $ws_{15}$ & f & $ws_{15}$ & f\\\hline
+        0 & 39.2\% & 39.2\% & 19.7\% & 15.9\% & 18.6\% & 15.3\%\\
+        5 & 28.1\% & 14.8\% & 15.9\% & 9.9\% & 14.1\% & 9.4\%\\
+        7.5 & 25.3\% & 12.6\% & 14.5\% & 8.6\% & 12.5\% & 8.6\%\\
+        10 & 23.9\% & 11.1\%& 13.6\% & 7.6\% & 11.5\% & 7.4\%\\
+        12.5 & 23.3\% & 10.1\% & 13.1\% & 7.1\% & 11.0\% & 6.9\%\\
+        15 & 23.2\% & 9.3\% & 13.0 & 7.0\% & 10.6\% & 6.4\%\\
+        17.5 & 23.7\% & 8.7\% & 12.8\% & 6.4\% & 10.53\% & 6.1\%\\
+        20 & 24.7\% & 8.2\% & 13.6\% & 6.3\% & 11.1\% & 5.8\%\\
+        22.5 & 26.1\% & 7.8\% & 14.5\% & 6.1\% & 12.6\% & 5.7\%\\
+        25 & 27.7\% & 7.3\% & 15.7\% & 5.8\% & 12.9\% & 5.5\% \\
+    \end{tabular}
+\end{table}
+
 This is some improvement upon the baseline error, with a decrease in error of around \errorImprovement\%.
+The power generated by wind mill increases with wind speed cubed\cite{wind_power}.
+The highest wind gusts in Iceland are around 70 m/s.
+Knowing the gust factor to 2 percentage points better than before can allow for the anticipated power generated by wind mills to increase significantly.
 
 \section{Discussion}
+The error improvement is modest in absolute terms but represents 1.8\% over the baseline.
+
+Training two identical models, one with the transformed reanalysis data and the other with original reanalysis data, results in two near identical errors, as expected.
+As no new information was input, the data was only manipulated, the model evaluates to the same error.
+The model that used the transformed reanalysis data was slightly quicker to converge.
+
 Looking only at the error, the performance of the model can be gauged.

### 2024-08-27 Chapters/Chapter4.tex
Take out stuff that was outdated after updating data

@@
 \end{table}
-
-\section{Feature selection and importance}
-Using the three datasources, a model is trained.
-The data from IMO only represents the ground truth and is not used as part of the training data.
-The gust ($f_g$) and wind speed ($f$) are used to calculate the target gust factor.
-All weather parameters used in predictions come from the CARRA reanalysis data, with variables wind speed, wind direction, temperature, pressure.
-All of these variables are queried in three height levels (15, 250, 500 meters above ground) at the location of a given weather station.
-The second data source for the training data, is the elevation data.
-Using the GeoTIFF file that IMO provided, a sector upwind were looked at.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_bar_nn_256_example100.png}
-    \caption[Feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in table \ref{table:gridSearchHyperparamters} and data as described in table \ref{table:trainDataExample}.}
-    \label{fig:ShapleyWaterfallFirstTry}
-\end{figure}

### 2024-08-27 Chapters/Chapter5.tex
Take out stuff that was outdated after updating data

@@
 
-This is some improvement upon the baseline error, with a decrease in error of around \errorImprovement\%.
+This is some improvement upon the baseline error, with a decrease in error from 23.9\% to 1.8\% depending on the setup used.
 The power generated by wind mill increases with wind speed cubed\cite{wind_power}.
@@
 \section{Discussion}
-The error improvement is modest in absolute terms but represents 1.8\% over the baseline.
-
-Training two identical models, one with the transformed reanalysis data and the other with original reanalysis data, results in two near identical errors, as expected.
-As no new information was input, the data was only manipulated, the model evaluates to the same error.
-The model that used the transformed reanalysis data was slightly quicker to converge.
-
+The error improvement is modest in absolute terms but represents at least 1.8\% over the baseline.
 Looking only at the error, the performance of the model can be gauged.
@@
 Using Shapley values and the python SHAP package, the contribution of each attribute can be given a value, as previously mentioned.
-Figure \ref{fig:SHAP_final_model} shows the average contribution of each feature.
+Figure \ref{fig:ShapleyWaterfall} shows the contribution of each feature for a given example.
+
+\begin{figure}[h]
+    \centering
+    \caption[Feature importance for a single observation of a neural network.]{Feature importance of a neural network with model architecture as described in table \ref{table:gridSearchHyperparamters} and data as described in table \ref{table:trainDataExample}.
+In this specific instance the wind direction ($wd_15$) has the highest negative influence and the temperature has the highest positive influence.
+Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them.
+To see their influence on the model output see Table \ref{table:results}.}
+    \includegraphics[scale = 0.6]{Figures/waterfall_plot.png}
+    \label{fig:ShapleyWaterfall}
+\end{figure}
+
+In Figure (\ref{fig:ShapleySummary}) the contribution of each feature, excluding the elevation points, can be seen for the model in general (a significant subset of data is used).
+
+\begin{figure}
+    \centering
+    \caption[Feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in table \ref{table:gridSearchHyperparamters} and data as described in table \ref{table:trainDataExample}.
+We can see that generally multiple factors influence the prediction, with the station elevation being highly influential.
+There is seemingly one outlier for the Richardson number, which usually has very little influence.
+Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them.
+To see their influence on the model output see Table \ref{table:results}.}
+    \includegraphics[scale = 0.6]{Figures/summary_plot.png}
+    \label{fig:ShapleySummary}
+\end{figure}

### 2024-09-26 Chapters/Chapter1.tex
Final preperations

@@
 This varies, with the US using a 1 minute interval, leading to 14\% higher results \cite{why_wind_gusts}.
-The Navier Stokes Equation (\ref{eq:navierstokes}) shows that the change of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth (the Coriolis force), and frictional force.\cite{uncertainties_in_numerical_weather_predictions}
+The Navier Stokes Equation (\ref{eqn:navierstokes}) shows that the change of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth (the Coriolis force), and frictional force.\cite{uncertainties_in_numerical_weather_predictions}
 \begin{equation}
-    \label{eq:navierstokes}
+    \label{eqn:navierstokes}
     \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{pressure} -\overbrace{ f\mathbf{k}x\mathbf{V}}^{oscillation} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{resistance}
@@
 It is also easy to construct different types of neural networks and see how they fit well with parts of the dataset.
-To measure the performance of these models, both to train and test, mean absolute percentage error (MAPE) as defined in Equation (\ref{eq:mape}) was used.
+To measure the performance of these models, both to train and test, mean absolute percentage error (MAPE) as defined in Equation (\ref{eqn:mape}) was used.
 \begin{equation}
-    \label{eq:mape}
-    MAPE = \frac{1}{n}\Sigma_{i=1}^n\frac{|y_{predict} - y_{true}|}{y_{predict}}
+    \label{eqn:mape}
+    \text{MAPE} = \frac{1}{n}\Sigma_{i=1}^n\frac{|y_{predict} - y_{true}|}{y_{predict}}
 \end{equation}

### 2024-09-26 Chapters/Chapter4.tex
Final preperations

@@
     \centering
-    \caption[Hyperparamter search with best performing combination.]{Hyperparamter search with best performing combination shown.
+    \caption[Hyperparameter search with best performing combination.]{Hyperparamter search with best performing combination shown.
 Hyperparameter search was done using hyperband algorithm that initially searches randomly for the best parameters but then hones in on what is working and as such is neither exhaustive nor completely random.
@@
     \label{table:gridSearchHyperparamters}
-    \begin{tabular}{c|c|c}
-        Parameter & Range of values & Selected\\\hline
-        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{10}\\\hline
-        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{64}\\\hline
-        Penalties & min\_value = 1e-5, max\_value = 1, sampling = log & \textbf{1e-4}\\\hline
-        Epochs & min\_value = 10, max\_value = 1000, step = 10 & \textbf{30}\\\hline
-        Optimizers & Adam, RMSprop, Adamax & \textbf{Adamax}\\\hline
-        Activation & ReLU, ELu, Softmax & \textbf{ReLU}\\\hline
+    \begin{tabular}{ccc}
+        Parameter & Range of values & Selected\\
+        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{10}\\
+        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{64}\\
+        Penalties & min\_value = 1e-5, max\_value = 1, sampling = log & \textbf{1e-4}\\
+        Epochs & min\_value = 10, max\_value = 1000, step = 10 & \textbf{50}\\
+        Optimizers & Adam, RMSprop, Adamax & \textbf{Adamax}\\
+        Activation & ReLU, ELu, Softmax & \textbf{ReLU}\\
     \end{tabular}

### 2024-09-26 Chapters/Chapter5.tex
Final preperations

@@
     \caption[Model results for different AWSL]{MAPE for each average wind speed limit with and without landscape elevation in a 30°sector around the point of interest into the direction of the reanalysis wind.
+The influence of adding elevation data seems to reduce the error.
 The percentage error is higher for lower wind speeds and thus observing the error for different lower bound of wind speed will produce different results.
-How to determine this lower bound can done by measurements (f) or reanalysis wind speed at 15 m/s ($ws_{15}$).
-The influence of adding elevation data seems to reduce the error.}
+How to determine this lower bound can done by measurements (f) or reanalysis wind speed at 15 m ($ws_{15}$).
+Using the measured wind will lead to data leakage and thus gives better results.
+This should not be used generally but is included here to show the difference in performance.}
     \label{table:results}
     \centering
-    \begin{tabular}{c|cccccc}
+    \begin{tabular}{lcccccc}
         \textbf{AWSL} & \textbf{MAPE} & & & & &\\ 
-        m/s & Baseline & &  Without elevation & & With elevation & \\
-            & $ws_{15}$ & f & $ws_{15}$ & f & $ws_{15}$ & f\\\hline
-        0 & 39.2\% & 39.2\% & 19.7\% & 15.9\% & 18.6\% & 15.3\%\\
-        5 & 28.1\% & 14.8\% & 15.9\% & 9.9\% & 14.1\% & 9.4\%\\
-        7.5 & 25.3\% & 12.6\% & 14.5\% & 8.6\% & 12.5\% & 8.6\%\\
-        10 & 23.9\% & 11.1\%& 13.6\% & 7.6\% & 11.5\% & 7.4\%\\
-        12.5 & 23.3\% & 10.1\% & 13.1\% & 7.1\% & 11.0\% & 6.9\%\\
-        15 & 23.2\% & 9.3\% & 13.0 & 7.0\% & 10.6\% & 6.4\%\\
-        17.5 & 23.7\% & 8.7\% & 12.8\% & 6.4\% & 10.53\% & 6.1\%\\
-        20 & 24.7\% & 8.2\% & 13.6\% & 6.3\% & 11.1\% & 5.8\%\\
-        22.5 & 26.1\% & 7.8\% & 14.5\% & 6.1\% & 12.6\% & 5.7\%\\
-        25 & 27.7\% & 7.3\% & 15.7\% & 5.8\% & 12.9\% & 5.5\% \\
+         & \multicolumn{2}{c}{\textit{Baseline}} &  \multicolumn{2}{c}{\textit{Without DEM}} & \multicolumn{2}{c}{\textit{With DEM}} \\\hline
+        m/s & $ws_{15}$ & f & $ws_{15}$ & f & $ws_{15}$ & f\\\hline
+        $\geq 0$ & 39.2\% & 39.2\% & 19.7\% & 15.9\% & 18.6\% & 15.3\%\\
+        $\geq 5$ & 28.1\% & 14.8\% & 15.9\% & 9.9\% & 14.1\% & 9.4\%\\
+        $\geq 10$ & 23.9\% & 11.1\%& 13.6\% & 7.6\% & 11.5\% & 7.4\%\\
+        $\geq 15$ & 23.2\% & 9.3\% & 13.0\% & 7.0\% & 10.6\% & 6.4\%\\
+        $\geq 20$ & 24.7\% & 8.2\% & 13.6\% & 6.3\% & 11.1\% & 5.8\%\\
+        $\geq 25$ & 27.7\% & 7.3\% & 15.7\% & 5.8\% & 12.9\% & 5.5\% \\
     \end{tabular}
 \end{table}
+
+%7.5 & 25.3\% & 12.6\% & 14.5\% & 8.6\% & 12.5\% & 8.6\%\\
+%12.5 & 23.3\% & 10.1\% & 13.1\% & 7.1\% & 11.0\% & 6.9\%\\
+%17.5 & 23.7\% & 8.7\% & 12.8\% & 6.4\% & 10.5\% & 6.1\%\\
+%22.5 & 26.1\% & 7.8\% & 14.5\% & 6.1\% & 12.6\% & 5.7\%\\
 
@@
 
-\section{Discussion}
-The error improvement is modest in absolute terms but represents at least 1.8\% over the baseline.
-Looking only at the error, the performance of the model can be gauged.
-However it does not help us explain why the model made these predictions.
-Here feature interpretation can be of great help.
-Using Shapley values and the python SHAP package, the contribution of each attribute can be given a value, as previously mentioned.
-Figure \ref{fig:ShapleyWaterfall} shows the contribution of each feature for a given example.
+Another way to look at the error improvement is by station.
+No location data was directly included in the training data.
+In Table \ref{table:station_mae_distribution} the mean absolute error of predicted average wind speed and measured average wind speed can be seen for the extreme values.
+A question to ask is it possible to achieve better results when only looking at a single station?
+
+\begin{table}[h]
+    \caption[Model result by station]{The MAPE results for selected stations of interest, both when training for the specific site and when the stations are a part of the general data.
+For every station the AWSL is set at 10 m/s.
+In training for a single station at a time, some site specific information can be gauged.
+This does not mean that the a better result can be reached for that site.
+Factors such as the number of datapoints at given location can significantly impact the result.
+This table uses the measured wind speed to determine the cut off for data points.
+This leads to some data leakage and an increased performance compared to using the reanalysis CARRA speed for cut off.}
+    \label{table:specific_sites}
+    \centering
+    \begin{tabular}{l  ccc}
+        \textbf{Station name} & \textbf{Number of measurement points} & \multicolumn{2}{c}{\textbf{MAPE}}\\
+        & & General training & Site training\\\hline
+        Akrafjall & 42,791 & 18.6\% & 93.7\%\\
+        Almannaskarð & 4,014 & 12.2\% & 86.7\%\\
+        Ásgarðsfjall & 15,121 & 9.1\% & 9.4\%\\
+        Jökulheimar & 17,176 & 7.7\% & 7.7\%\\
+        Sandbúðir & 18,718 & 6.8\% & 6.4\%\\
+        Stórholt & 35,126 & 7.1\% & 29.2\% \\
+        Þúfuver & 19,538 & 6.4\% & 6.8\%\\
+    \end{tabular}
+\end{table}
+
+Instead of looking at the exact values of MAPE at select stations, a plot of the error distribution can be created.
+This can be seen in Figure (\ref{fig:errorMap}).
+Looking at Figure (\ref{fig:errorMap}), the worst performing stations can be seen at Vestfirðir and around the coastline.
+Stations further inland seem to have lower error.
+The worst performing station, at Seljalandsdalur, is at Vestfirðir while the best performing station Garðskagaviti is at the South-Western tip of Iceland in Reykjanes.
 
@@
     \centering
-    \caption[Feature importance for a single observation of a neural network.]{Feature importance of a neural network with model architecture as described in table \ref{table:gridSearchHyperparamters} and data as described in table \ref{table:trainDataExample}.
-In this specific instance the wind direction ($wd_15$) has the highest negative influence and the temperature has the highest positive influence.
+    \caption[MAPE error distribution of stations shown on a map of Iceland.]{The MAPE error of each station in data shown as color gradient circles.
+That is each station is represented by one circle, with the error value represented as a color gradient from dark blue to yellow.
+The lowest error at a single station was around 7\% at Garðskagaviti and the highest around 35\% at Seljalandsdalur.
+It is important to note that the model is trained using a cutoff of 10 m/s and this cutoff point is determined using the reanalysis wind at 15 meters above ground.}
+    \includegraphics[scale = 0.5]{Figures/errorMap.png}
+    \label{fig:errorMap}
+\end{figure}
+
+\begin{table}[h]
+    \caption[Model result looking at closed wind speed intervals]{The MAPE results for different AWSL intervals.
+Here instead of training for all data above a certain threshold put of the wind, training is done only on data between two wind speeds.
+The percentage variance in gust factor as a function of wind speed increases with decreasing wind speed.
+Measured wind speed is used for the cutoff and thus have data leakage.
+This results should thus be somewhat comparable to the last column in Table (\ref{table:results})}
+    \label{table:closed_intervals}
+    \centering
+    \begin{tabular}{l | cc}
+        \textbf{Interval} &  \multicolumn{2}{c}{\textbf{MAPE}}\\
+        $[m/s]$ & Without Elevation & With Elevation\\\hline
+        $[5, 10[$ & 17.2\% & 15.4\%\\ %10.6\%\\
+        $[10, 15[$ & 14.1\% & 11.9\%\\ %7.8\%\\
+        $[15, 20[$ & 13.2\% & 10.9\%\\ %6.4\%\\
+        $[20, 25[$ & 14.9 \% & 11.6\%\\ %6.3\%\\
+        $[25, 30[$ & 17.0\% & 16.0\%\\ %7.1\%\\
+    \end{tabular}
+\end{table}
+
+Table (\ref{table:closed_intervals}) shows no improvement over the values in the last column in Table (\ref{table:results}).
+As previously mentioned, the gust factor decreases with increasing wind speed and thus, training on intervals and lessening this effect might be expected to give better results.
+This does not seem to be the case.
+
+Some interesting sites to look closer at for drivers might include places like Kjalarnes, Ingólfsfjall, Þrengsli and others. These can be seen in Table (\ref{table:specific_sites}).
+
+\begin{table}[h]
+    \caption[Model result by stations of interest]{The MAPE results of different stations for several stations of interest, both when training for the specific site and when the stations are a part of the general data.
+For every station the AWSL is set at 10 m/s.
+In training for a single station at a time, some site specific information can be gauged.
+This does not mean that the a better result can be reached for that site.
+Factors such as the number of datapoints at given location can significantly impact the result.
+This table uses the measured wind speed to determine the cut off for data points.
+This leads to some data leakage and an increased performance compared to using the reanalysis CARRA speed for cut off.}
+    \label{table:specific_sites}
+    \centering
+    \begin{tabular}{l | cc}
+        \textbf{Station name} & \multicolumn{2}{c}{\textbf{MAPE}}\\
+         & Baseline & Model\\\hline
+        Fáskrúðsfjörður & 28.2\% & 21.8\%\\
+        Ingólfsfjall & 30.0\% & 19.6\%\\
+        Kjalarnes & 20.7\% & 13.5\% \\
+        Sandskeið & 13.0\% & 10.2\%\\
+        Seyðisfjörður & 32.1\% & 23.0\%\\
+        Þjórsárdalur & 12.2\% & 11.4\%\\
+        Þrengsli & 13.6\% & 11.3\%\\
+    \end{tabular}
+\end{table}
+
+\begin{figure}[h]
+    \centering
+    \caption[Feature importance for a single observation of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+In this specific instance the wind direction ($wd_{15}$) has the highest negative influence and the temperature has the highest positive influence.
 Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them.
 To see their influence on the model output see Table \ref{table:results}.}
-    \includegraphics[scale = 0.6]{Figures/waterfall_plot.png}
+    \includegraphics[scale = 0.6]{Figures/shap_plots/waterfall_plot.png}
     \label{fig:ShapleyWaterfall}
@@
     \centering
-    \caption[Feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in table \ref{table:gridSearchHyperparamters} and data as described in table \ref{table:trainDataExample}.
+    \caption[Summary feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
 We can see that generally multiple factors influence the prediction, with the station elevation being highly influential.
@@
 To see their influence on the model output see Table \ref{table:results}.}
-    \includegraphics[scale = 0.6]{Figures/summary_plot.png}
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot.png}
     \label{fig:ShapleySummary}
 \end{figure}
+
+Looking at Figure (\ref{fig:ShapleySummary}), there is an outlier.
+Exactly calculating the Shapley values is time intensive, in the subset shown there is an outlier that skews the figure and makes it so that viewing the importance distribution excluding the outlier is difficult.
+For this reason, another shapley summary vas created that looked at different, and larger distribution.
+This can be seen in Figure (\ref{fig:ShapleySummary2}).
+
+\begin{figure}
+    \centering
+    \caption[Summary feature importance of a neural network using a larger distribution of data.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+Generally multiple factors influence the prediction, with the station elevation being highly influential.
+Elevation data is excluded when working with Shapley values, as the contribution of each elevation.
+In contrast to Figure (\ref{fig:ShapleySummary}), the distribution doesn't have as extreme outliers.
+This means that more details can be seen in the figure.
+The X-axis shows the influence of feature values on the model.
+The color gradient shows the value of each feature.
+As an example, there is a very red value for station elevation (top line, all the way to the left).
+This means that in this instance, the station elevation contributed around -0.25 to the final output and that the station had an elevation significantly above average.}
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_.png}
+    \label{fig:ShapleySummary2}
+\end{figure}
+
+The Figures (\ref{fig:ShapleySummary}) and (\ref{fig:ShapleySummary2}) show the summary for a model trained only on the features shown and not on landscape elevation of the surrounding area.
+This is done as the number of points there is too high to display in one figure (70 total points).
+Looking specifically at Figure (\ref{fig:ShapleySummary2}), the station elevation is most influential and the Richardson number's influence is very low.
+Most of the feature values bunch up in the middle, while the station elevation is elongated compared to other features.
+Station elevation seems to have two bunches on either side of 0.
+Overall, the values seem to be skewed to the right of 0.
+This would be expected as the predicted values are expected to be in the range of 1.2-2, or at least always above 1 by definition.
+
+Finally, for the Shapley values, looking that all the data there are again outliers that skew the data so that spotting general distribution is difficult.
+This can be seen in Figure (\ref{fig:ShapleySummary3}).
+
+\begin{figure}
+    \centering
+    \caption[Summary feature importance of a neural network using entire dataset.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+The distribution seems to be the same as before, discounting the outliers.}
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_full_10ms.png}
+    \label{fig:ShapleySummary3}
+\end{figure}
+
+These plots can also be done by looking at individual stations.
+The model is then trained on all examples and afterwards the summary plot created using only points for each station.
+Stations of interest include stations listed in Table \ref{table:specific_sites}.
+These plots can all be seen in Figures (\ref{fig:ShapleySummaryAkrafjall} - \ref{fig:ShapleySummaryKeflavikurflugvollur}).
+
+\begin{figure}
+    \centering
+    \caption[Summary feature importance of a neural network only looking at AWS at Akrafjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at datapoints from Akrafjall.
+This seems to show the same distribution as previous summary plots.
+Station elevation is influential and Richardson number has no impact.}
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_31572.png}
+    \label{fig:ShapleySummaryAkrafjall}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \caption[Summary feature importance of a neural network only looking at AWS at Almannaskarð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at datapoints from Almannaskarð.
+This seems to show the same distribution as previous summary plots.
+Station elevation is influential and Richardson number has no impact.}
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_35553.png}
+    \label{fig:ShapleySummaryAlmannaskarð}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \caption[Summary feature importance of a neural network only looking at AWS at Ásgarðsfjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at datapoints from Ásgarðsfjall.
+This seems to show the same distribution as previous summary plots.
+Station elevation is influential and Richardson number has no impact.}
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_6745.png}
+    \label{fig:ShapleySummaryAsgarðsfjall}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \caption[Summary feature importance of a neural network only looking at AWS at Háahlíð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at datapoints from Háahlíð.
+This seems to show the same distribution as previous summary plots.
+Station elevation is influential and Richardson number has no impact.}
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1470.png}
+    \label{fig:ShapleySummaryHaahlid}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \caption[Summary feature importance of a neural network only looking at AWS at Keflavíkurflugvöllur.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at datapoints from Keflavíkurflugvöllur.
+This seems to show the same distribution as previous summary plots.
+Station elevation is influential and Richardson number has no impact.}
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1350.png}
+    \label{fig:ShapleySummaryKeflavikurflugvollur}
+\end{figure}
+
+In each of these plots, even without the feature labels, the station elevation is easily noticed as the value is constant for a station.
+This is not noteworthy.
+What is noteworthy is the range of impact from this single value.
+For simpler models, this would not happen.
+It is important to note that SHAP assumes feature independece\cite{Salih_2024}.
+This might explain why the impact of the Richardson number is so low.
+Both the squared Brunt–Väisälä frequency and the Richardson number are derived features from reanalysis data.
+They carry with them some extra information over the other features in the dataset.
+This is because both are variables over elevation ranges.
+That is, as seen in Equations (\ref{eqn:Ri}, \ref{eqn:N}), both are dependent on values at lower and upper elevations and try to describe the stability of that range.
+Shapley tries to assign contribution values for each feature for each observation.
+SHAP assumes that the features are independent, but this is not the case.
+It is clearly not the case for the derived variables, but how the contribution should be distributed between the features is not clear.
+Seemingly the SHAP python package is giving all the impact to the Brunt–Väisälä squared frequency and none to Richardson number.
+If the Brunt–Väisälä would be excluded from the data, the impact of the Richardson number would likely increase.
+Another point to note is that the features are ordered by their impact.
+This means that the station elevation is the most impactful for each plot, but the ordering of other variables changes.
+Looking at Figure (\ref{fig:ShapleySummary3}), which shows the Shapley summary plot for all data, the wind speed is the second most important feature.
+This is reversed in Figure (\ref{fig:ShapleySummaryAkrafjall}).
+This is not unexpected as Akrafjall station was specifically selected as the MAE for reanalysis wind speed was very high as can be seen in Table (\ref{fig:ShapleySummaryAkrafjall}), where you will also find the stations whose summary plot is shown in Figures (\ref{fig:ShapleySummaryAlmannaskarð}, \ref{fig:ShapleySummaryAsgarðsfjall}) and these also fall into the category of very high MAE for wind speed.
+What is interesting is that the reanalysis wind speed is also of low impact at stations like Háahlíð and Keflavíkurflugvöllur, as shown in Figures (\ref{fig:ShapleySummaryHaahlid}, \ref{fig:ShapleySummaryKeflavikurflugvollur}).
+These stations had the lowest of MAE for difference between measured wind speed and reanalysis wind speed.
+As the summary plot over all stations (Figure (\ref{fig:ShapleySummary3})) shows that reanalysis wind speed is impactful, might lead to the conclusion that the reanalysis wind speed is not a good predictor at these locations or something else is skewing the data.

### 2024-09-26 MSc.tex
.
Updated after new CARRA data sourced and run
Take out stuff that was outdated after updating data
Final preperations

@@
 % Packages
+% BETRI TÖFLUR OG LISTAR
+\usepackage{booktabs}         % Fyrir betri töflur
+\usepackage{caption}          % M.a. aukabil á undan töflufyrirsögnum
+\captionsetup{font=small,   % Smærra letur á mynda- og töflutexta,
+              labelfont=bf, % "Mynd 1" og "Tafla 2" feitletrað 
+              margin=2cm,
+              skip = 0.5\baselineskip} 
 \usepackage[utf8]{inputenc}
@@
 \usepackage{placeins}
-\usepackage{tabularx}
+%\usepackage{tabularx}
 \usepackage{enumitem}
@@
 \usepackage{float}
-
+\usepackage{caption}
 % Packages used for title page layout
@@
 \usepackage{tikz}
+\usepackage{arydshln}
 \usetikzlibrary{positioning}
+\pgfkeys{/pgf/number format/.cd,fixed,precision=2}
 
@@
 \def\thesisyear{2024}       					% Year thesis submitted
-\def\thesismonth{June}					% Month thesis submitted, e.g. "December"
+\def\thesismonth{September}					% Month thesis submitted, e.g. "December"
 \def\thesisauthor{Brynjar Geir Sigurðsson}				% Thesis authoreiningaraðferðinni
-\def\thesistitle{A NN approach to predicting gust factors in complex landscape} % Title of thesis
+\def\thesistitle{A Neural Network Approach to Predicting Gust Factors in Complex Landscape} % Title of thesis
 %\def\thesissubtitle{XXSubtitle is rarely usedXX}		% Subtitle of thesis (optional)
@@
 \def\thesisnrofsupervisors{1}					% Number of supervisors (to use "Supervisor" vs. "Supervisors")
-\def\thesiscommittee{Kristján Jónasson \\ Ólafur Pétur Pálsson}			% Thesis commitee must include the supervisor(s)
+\def\thesiscommittee{Kristján Jónasson \\ Ólafur Pétur Pálsson \\ Guðrún Nína Petersen}			% Thesis commitee must include the supervisor(s)
 \def\thesisexaminer{XXNN3XX}				        % Examiner (prófdómari)
@@
 \listoftables
+\lstlistoflistings
 
@@
 \include{Chapters/Chapter4}
+\include{Chapters/Chapter5}
 

### 2024-10-03 Chapters/Chapter4.tex
Updated tables in 5 and added 6

@@
 
-\chapter{Model architecture} % Main chapter title
+\chapter{Model Architecture and Training}
 
-\label{Chapter4} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}
-
-%----------------------------------------------------------------------------------------
-%	SECTION 1
-%----------------------------------------------------------------------------------------
+\label{Chapter4}
 
@@
 Hyperparameter search was done using hyperband algorithm that initially searches randomly for the best parameters but then hones in on what is working and as such is neither exhaustive nor completely random.
-This means that an upper limit will not be set on the number of combinations to try like with randomsearch.}
+This means that a hard upper limit will not be set on the number of combinations to try like with randomsearch.}
     \label{table:gridSearchHyperparamters}
     \begin{tabular}{ccc}
+        \toprule
         Parameter & Range of values & Selected\\
+        \midrule
         Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{10}\\
@@
         Penalties & min\_value = 1e-5, max\_value = 1, sampling = log & \textbf{1e-4}\\
-        Epochs & min\_value = 10, max\_value = 1000, step = 10 & \textbf{50}\\
+        Epochs & min\_value = 10, max\_value = 1000, step = 10 & \textbf{250}\\
         Optimizers & Adam, RMSprop, Adamax & \textbf{Adamax}\\
         Activation & ReLU, ELu, Softmax & \textbf{ReLU}\\
+        \bottomrule
     \end{tabular}
 \end{table}
+
+As mentioned, hyperband doesn't set a hard upper limit to the number of epochs it will train in total.
+When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{hyperband class} several factors can be set.
+One of interest here is the $hyperband_iterations$ argument.
+This determines how often the hyperband algorithm is run and defaults to 1.
+For each iteration the epochs are distributed between tries (that is each set of hyperparameters) with the total amount of epochs approximately $n_{epochs} = max_{epochs} * log^2(max_{epochs})\approx 10^4$, where $max_{epochs} = 1000$ gives the maximum number of epochs that one set of hyperparameters can be trained for.
+Searching a space generally takes a lot of time but this drastically improves on gridsearch.
+If each epoch takes around 10 seconds to run then the total search would take around 28 hours on a shared resource.
+This is resource intensive and cannot be repeated often.
+Another question that remains is whether the ranges given are optimal.
+Are a 1000 epochs even enough?
+
+\section{Model Training}
+To determine the required number of epochs a high number of epochs can be selected and the loss plotted.
+These plots can be seen in Figures (\ref{fig:loss_plot_wo_elevation}, \ref{fig:loss_plot_w_elevation}).
+Looking at these two plots, training loss decreases for over 1900 epochs in both plots, while the validation loss stops decreasing at around 200 epochs for DEM model and 1800 epochs for model without DEM.
+This is unexpected.
+The DEM model has vastly more parameters and as such should contain more useful information to be learned, which one might expect to take longer to learn.
+This does not seem to be the case.
+The DEM model only gets trivially better after 1000 epochs while the non DEM model keeps meaningfully improving until around 1800 epochs.
+With respect to time constraints of shared resources models were trained for 250 epochs.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.6]{Figures/loss_plot_wo_elevation.png}
+    \caption[Loss plot of model trained for 2000 without DEM.]{Loss plot of model trained for 2000 without DEM.}
+    \label{fig:loss_plot_wo_elevation}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.6]{Figures/loss_plot_w_elevation.png}
+    \caption[Loss plot of model trained for 2000 with DEM.]{Loss plot of model trained for 2000 with DEM.}
+    \label{fig:loss_plot_w_elevation}
+\end{figure}

### 2024-10-03 Chapters/Chapter5.tex
Updated tables in 5 and added 6

@@
 \begin{table}[h]
-    \caption[Model results for different AWSL]{MAPE for each average wind speed limit with and without landscape elevation in a 30°sector around the point of interest into the direction of the reanalysis wind.
+    \caption[Model results for different AWSL]{MAPE for each average wind speed limit with and without landscape elevation in a 30° sector around the point of interest into the direction of the reanalysis wind.
 The influence of adding elevation data seems to reduce the error.
 The percentage error is higher for lower wind speeds and thus observing the error for different lower bound of wind speed will produce different results.
-How to determine this lower bound can done by measurements (f) or reanalysis wind speed at 15 m ($ws_{15}$).
-Using the measured wind will lead to data leakage and thus gives better results.
-This should not be used generally but is included here to show the difference in performance.}
+This lower bound is determined using the reanalysis wind speed at 15 meters ($ws_{15}$).}
     \label{table:results}
     \centering
-    \begin{tabular}{lcccccc}
-        \textbf{AWSL} & \textbf{MAPE} & & & & &\\ 
-         & \multicolumn{2}{c}{\textit{Baseline}} &  \multicolumn{2}{c}{\textit{Without DEM}} & \multicolumn{2}{c}{\textit{With DEM}} \\\hline
-        m/s & $ws_{15}$ & f & $ws_{15}$ & f & $ws_{15}$ & f\\\hline
-        $\geq 0$ & 39.2\% & 39.2\% & 19.7\% & 15.9\% & 18.6\% & 15.3\%\\
-        $\geq 5$ & 28.1\% & 14.8\% & 15.9\% & 9.9\% & 14.1\% & 9.4\%\\
-        $\geq 10$ & 23.9\% & 11.1\%& 13.6\% & 7.6\% & 11.5\% & 7.4\%\\
-        $\geq 15$ & 23.2\% & 9.3\% & 13.0\% & 7.0\% & 10.6\% & 6.4\%\\
-        $\geq 20$ & 24.7\% & 8.2\% & 13.6\% & 6.3\% & 11.1\% & 5.8\%\\
-        $\geq 25$ & 27.7\% & 7.3\% & 15.7\% & 5.8\% & 12.9\% & 5.5\% \\
+    \begin{tabular}{lccc}
+        \toprule
+        \textbf{AWSL} & \textbf{MAPE} & &\\ 
+        $[m/s]$ & \textit{Baseline} &  \textit{Without DEM} & \textit{With DEM} \\\hline
+        $\geq 0$ & 39.2\% & 19.7\% & 18.6\% \\
+        $\geq 5$ & 28.1\% & 15.9\% & 14.1\%\\
+        $\geq 10$ & 23.9\% & 13.6\% & 11.5\%\\
+        $\geq 15$ & 23.2\% & 13.0\% & 10.6\%\\
+        $\geq 20$ & 24.7\% & 13.6\% & 11.1\%\\
+        $\geq 25$ & 27.7\% & 15.7\% & 12.9\%\\
+        \bottomrule
     \end{tabular}
@@
 
+%m/s & $ws_{15}$ & f & $ws_{15}$ & f & $ws_{15}$ & f\\\hline
+%$\geq 0$ & 39.2\% & 39.2\% & 19.7\% & 15.9\% & 18.6\% & 15.3\%\\
+%$\geq 5$ & 28.1\% & 14.8\% & 15.9\% & 9.9\% & 14.1\% & 9.4\%\\
+%$\geq 10$ & 23.9\% & 11.1\%& 13.6\% & 7.6\% & 11.5\% & 7.4\%\\
+%$\geq 15$ & 23.2\% & 9.3\% & 13.0\% & 7.0\% & 10.6\% & 6.4\%\\
+%$\geq 20$ & 24.7\% & 8.2\% & 13.6\% & 6.3\% & 11.1\% & 5.8\%\\
+%$\geq 25$ & 27.7\% & 7.3\% & 15.7\% & 5.8\% & 12.9\% & 5.5\% \\
 %7.5 & 25.3\% & 12.6\% & 14.5\% & 8.6\% & 12.5\% & 8.6\%\\
@@
 
-This is some improvement upon the baseline error, with a decrease in error from 23.9\% to 1.8\% depending on the setup used.
+This is some improvement upon the baseline error, with a decrease in error from 23.9\% to 13.6\% and 10.6\% for the baseline, model without DEM and with model with DEM for 10 m/s cutoff.
 The power generated by wind mill increases with wind speed cubed\cite{wind_power}.
 The highest wind gusts in Iceland are around 70 m/s.
-Knowing the gust factor to 2 percentage points better than before can allow for the anticipated power generated by wind mills to increase significantly.
-
+Knowing the gust factor with half as much error as before can allow better anticipation and thus spare turbines for high wind gusts.
 Another way to look at the error improvement is by station.
@@
     \centering
-    \begin{tabular}{l  ccc}
+    \begin{tabular}{lccc}
         \textbf{Station name} & \textbf{Number of measurement points} & \multicolumn{2}{c}{\textbf{MAPE}}\\
@@
     \centering
+    \includegraphics[scale = 0.5]{Figures/errorMap.png}
     \caption[MAPE error distribution of stations shown on a map of Iceland.]{The MAPE error of each station in data shown as color gradient circles.
@@
 It is important to note that the model is trained using a cutoff of 10 m/s and this cutoff point is determined using the reanalysis wind at 15 meters above ground.}
-    \includegraphics[scale = 0.5]{Figures/errorMap.png}
     \label{fig:errorMap}
@@
 This does not seem to be the case.
-
-Some interesting sites to look closer at for drivers might include places like Kjalarnes, Ingólfsfjall, Þrengsli and others. These can be seen in Table (\ref{table:specific_sites}).
+Some interesting sites to look closer at for drivers might include places like Kjalarnes, Ingólfsfjall, Þrengsli and others.
+These can be seen in Table (\ref{table:specific_sites}).
 
@@
     \begin{tabular}{l | cc}
+        \toprule
         \textbf{Station name} & \multicolumn{2}{c}{\textbf{MAPE}}\\
-         & Baseline & Model\\\hline
+         & Baseline & Model\\
+         \midrule
         Fáskrúðsfjörður & 28.2\% & 21.8\%\\
@@
         Þrengsli & 13.6\% & 11.3\%\\
+        \bottomrule
     \end{tabular}
@@
     \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/waterfall_plot.png}
     \caption[Feature importance for a single observation of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
@@
 To see their influence on the model output see Table \ref{table:results}.}
-    \includegraphics[scale = 0.6]{Figures/shap_plots/waterfall_plot.png}
     \label{fig:ShapleyWaterfall}
@@
 In Figure (\ref{fig:ShapleySummary}) the contribution of each feature, excluding the elevation points, can be seen for the model in general (a significant subset of data is used).
-
-\begin{figure}
-    \centering
-    \caption[Summary feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
-We can see that generally multiple factors influence the prediction, with the station elevation being highly influential.
-There is seemingly one outlier for the Richardson number, which usually has very little influence.
-Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them.
-To see their influence on the model output see Table \ref{table:results}.}
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot.png}
-    \label{fig:ShapleySummary}
-\end{figure}
-
 Looking at Figure (\ref{fig:ShapleySummary}), there is an outlier.
@@
     \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot.png}
+    \caption[Summary feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+We can see that generally multiple factors influence the prediction, with the station elevation being highly influential.
+There is seemingly one outlier for the Richardson number, which usually has very little influence.
+Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them.
+To see their influence on the model output see Table \ref{table:results}.}
+    \label{fig:ShapleySummary}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_.png}
     \caption[Summary feature importance of a neural network using a larger distribution of data.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
@@
 This means that in this instance, the station elevation contributed around -0.25 to the final output and that the station had an elevation significantly above average.}
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_.png}
     \label{fig:ShapleySummary2}
@@
 This would be expected as the predicted values are expected to be in the range of 1.2-2, or at least always above 1 by definition.
-
 Finally, for the Shapley values, looking that all the data there are again outliers that skew the data so that spotting general distribution is difficult.
@@
     \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_full_10ms.png}
     \caption[Summary feature importance of a neural network using entire dataset.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
 The distribution seems to be the same as before, discounting the outliers.}
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_full_10ms.png}
     \label{fig:ShapleySummary3}
@@
     \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_31572.png}
     \caption[Summary feature importance of a neural network only looking at AWS at Akrafjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
@@
 Station elevation is influential and Richardson number has no impact.}
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_31572.png}
     \label{fig:ShapleySummaryAkrafjall}
@@
     \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_35553.png}
     \caption[Summary feature importance of a neural network only looking at AWS at Almannaskarð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
@@
 Station elevation is influential and Richardson number has no impact.}
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_35553.png}
     \label{fig:ShapleySummaryAlmannaskarð}
@@
     \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_6745.png}
     \caption[Summary feature importance of a neural network only looking at AWS at Ásgarðsfjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
@@
 Station elevation is influential and Richardson number has no impact.}
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_6745.png}
     \label{fig:ShapleySummaryAsgarðsfjall}
@@
     \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1470.png}
     \caption[Summary feature importance of a neural network only looking at AWS at Háahlíð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
@@
 Station elevation is influential and Richardson number has no impact.}
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1470.png}
     \label{fig:ShapleySummaryHaahlid}
@@
     \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1350.png}
     \caption[Summary feature importance of a neural network only looking at AWS at Keflavíkurflugvöllur.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
@@
 Station elevation is influential and Richardson number has no impact.}
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1350.png}
     \label{fig:ShapleySummaryKeflavikurflugvollur}
@@
 As the summary plot over all stations (Figure (\ref{fig:ShapleySummary3})) shows that reanalysis wind speed is impactful, might lead to the conclusion that the reanalysis wind speed is not a good predictor at these locations or something else is skewing the data.
+
+A simpler way to look at feature importance is to create models that are trained on and use to predict different sets of parameters.
+The results of such a comparison can be seen in Table \ref{table:setsOfParams}.
+
+\begin{table}[h]
+    \caption[Model results for different sets of parameters.]{...}
+    \label{table:setsOfParams}
+    \centering
+    \begin{tabular}{lc}
+        \toprule
+        \textbf{Model parameters} & \textbf{MAPE}\\ 
+        \midrule
+        Baseline constant & 23.9\%\\
+        $ws_{15}$ & 17.2\%\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ 16.7\% & ---\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd]$ & ---\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ & ---\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ + DEM & ---\\
+        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, ASL, twd, N, Ri]$
+        \bottomrule
+    \end{tabular}
+\end{table}

### 2024-10-04 MSc.tex
Updated tables in 5 and added 6
Updated abstract

@@
 \section*{Abstract}
-English abstract (ca. 250 words).
+In many industries, whether it be windmill farming or transportation, being able to precisely predict the load caused by weather is crucial to prevent harm coming to people and machinery as well as increasing productivity.
+The efficiency of wind turbines in low wind speed conditions is a critical challenge in optimizing renewable energy systems.
+Gust predictions can be instrumental in loss prevention.
+If looking only at wind, the most destructive moments are gusts, by definition.
+To predict these values there are two ways, traditional numerical weather prediction systems and machine learning or other pattern recognition systems.
+Computer aided numerical weather prediction systems originated in the 1950's, while the use of machine learning in numerical weather predictions is much more recent, having truly gained traction in the last decade.
 
+The goal of this thesis is try to combine these two methods to try and predict wind gust factor, where the gust factor is defined as the ratio between wind gust and average wind speed.
+This is done by having the input data to the machine learning model be reanalysis variables that were generated by traditional numerical weather prediction systems.
+In addition to these variables and derived variables, a digital elevation model (DEM) data was used.
+A deep neural network was constructed using some set of variables and the results are compared to a baseline as well as each other so as to show the impact each feature has.
 \vfill \vspace*{1cm}
@@
 \section*{Útdráttur}
-Hér kemur útdráttur á íslensku sem er að hámarki 250 orð.
+Í mörgum iðnuðum, hvort sem það er vindmyllubúgarður eða flutningar, að geta 
 \vfill
@@
 \include{Chapters/Chapter5}
+\include{Chapters/Chapter6}
 

### 2024-10-07 Chapters/Chapter3.tex
Final preperations
Updated tables in 5 and added 6
Added abstract. For some reason the citations are not being loaded so just get [descriptive_name].

@@
 \begin{figure}[h]
+    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart.png}
     \caption{A flow chart showing how data sources were combined}
     \label{fig:data_preprocessing_flow_chart}
-    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart.png}
 \end{figure}
@@
     \centering
-    \begin{tabular}{c|ccc}
-        f & n & MAE & MAPE \\ \hline
+    \begin{tabular}{cccc}
+
+        f & n & MAE & MAPE \\
+
         {[}0;5{[} & 6.2e6 & 2.1 & 1.6\\
@@
         {[}25;$\infty${[} & 2.0e4 & 6.6 & 0.2\\
-        {[}0;$\infty${[} & 1.2e7 & 2.2 & 1.0
+        {[}0;$\infty${[} & 1.2e7 & 2.2 & 1.0\\
+
     \end{tabular}
@@
 \begin{table}[h]
-    \caption[]{}
+    \caption[Mean absolute difference of measured wind speed and reanalysis wind speed at select stations]{Mean absolute error for reanalysis wind speed as compared to measured wind speed, for the five stations with the highest difference and the five stations with the lowest difference.}
     \label{table:station_mae_distribution}
@@
     \centering
-    \begin{tabular}{c|ccc}
-        Station & Number of measurements & MAE & Location \\\hline
+    \begin{tabular}{cccc}
+
+        Station & Number of measurements & MAE & Location \\
+
         1470 & 6.8e3 & 1.17 & Reykjavík Háahlíð \\
@@
         32635 & 3.2e4 & 4.95 & Botn í Súgandafirði\\
+
     \end{tabular}

### 2024-10-07 Chapters/Chapter5.tex
Updated tables in 5 and added 6
Added abstract. For some reason the citations are not being loaded so just get [descriptive_name].

@@
         \textbf{AWSL} & \textbf{MAPE} & &\\ 
-        $[m/s]$ & \textit{Baseline} &  \textit{Without DEM} & \textit{With DEM} \\\hline
+        $[m/s]$ & \textit{Baseline} &  \textit{Without DEM} & \textit{With DEM} \\
+        \midrule
         $\geq 0$ & 39.2\% & 19.7\% & 18.6\% \\
@@
 
-%m/s & $ws_{15}$ & f & $ws_{15}$ & f & $ws_{15}$ & f\\\hline
+%m/s & $ws_{15}$ & f & $ws_{15}$ & f & $ws_{15}$ & f\\
 %$\geq 0$ & 39.2\% & 39.2\% & 19.7\% & 15.9\% & 18.6\% & 15.3\%\\
@@
     \begin{tabular}{lccc}
+        \toprule
         \textbf{Station name} & \textbf{Number of measurement points} & \multicolumn{2}{c}{\textbf{MAPE}}\\
-        & & General training & Site training\\\hline
+        & & General training & Site training\\
+        \midrule
         Akrafjall & 42,791 & 18.6\% & 93.7\%\\
@@
         Þúfuver & 19,538 & 6.4\% & 6.8\%\\
+        \bottomrule
     \end{tabular}
@@
     \centering
-    \begin{tabular}{l | cc}
+    \begin{tabular}{lcc}
+        \toprule
         \textbf{Interval} &  \multicolumn{2}{c}{\textbf{MAPE}}\\
-        $[m/s]$ & Without Elevation & With Elevation\\\hline
+        $[m/s]$ & Without Elevation & With Elevation\\
+        \midrule
         $[5, 10[$ & 17.2\% & 15.4\%\\ %10.6\%\\
@@
         $[25, 30[$ & 17.0\% & 16.0\%\\ %7.1\%\\
+        \bottomrule
     \end{tabular}
@@
 This leads to some data leakage and an increased performance compared to using the reanalysis CARRA speed for cut off.}
-    \label{table:specific_sites}
+    \label{table:more_specific_sites}
     \centering
-    \begin{tabular}{l | cc}
+    \begin{tabular}{lcc}
         \toprule
@@
 As the summary plot over all stations (Figure (\ref{fig:ShapleySummary3})) shows that reanalysis wind speed is impactful, might lead to the conclusion that the reanalysis wind speed is not a good predictor at these locations or something else is skewing the data.
-
 A simpler way to look at feature importance is to create models that are trained on and use to predict different sets of parameters.
@@
         $ws_{15}$ & 17.2\%\\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ 16.7\% & ---\\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd]$ & ---\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd]$ & 13.8\% \\
         $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ & ---\\
         $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ + DEM & ---\\
-        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, ASL, twd, N, Ri]$
+        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, ASL, twd, N, Ri]$ & --- \\
         \bottomrule

### 2024-10-07 MSc.tex
Added abstract. For some reason the citations are not being loaded so just get [descriptive_name].

@@
 
-\setkomafont{section}{\huge} % The title "Abstract" and "Útfdráttur" should look like the chapters, i.e. use \huge (\chapter cannot be used as this would create a new page)
+\setkomafont{section}{\huge} % The title "Abstract" and "Útdráttur" should look like the chapters, i.e. use \huge (\chapter cannot be used as this would create a new page)
+
 \section*{Abstract}
 In many industries, whether it be windmill farming or transportation, being able to precisely predict the load caused by weather is crucial to prevent harm coming to people and machinery as well as increasing productivity.
-The efficiency of wind turbines in low wind speed conditions is a critical challenge in optimizing renewable energy systems.
+The efficiency of wind turbines in high wind speed conditions is a vital challenge in optimizing renewable energy systems.
 Gust predictions can be instrumental in loss prevention.
@@
 Computer aided numerical weather prediction systems originated in the 1950's, while the use of machine learning in numerical weather predictions is much more recent, having truly gained traction in the last decade.
-
 The goal of this thesis is try to combine these two methods to try and predict wind gust factor, where the gust factor is defined as the ratio between wind gust and average wind speed.
@@
 \section*{Útdráttur}
-Í mörgum iðnuðum, hvort sem það er vindmyllubúgarður eða flutningar, að geta 
+Í mörgum iðnuðum, hvort sem það er vindmyllubúgarður eða flutningar er getan til þess að spá fyrir um álag vegna veðurs mikilvæg til að koma í veg fyrir tjóns á vélum og byggingum ásamt því að koma í veg fyrir skaða á fólki.
+Einnig er hægt að auka framleiðni.
+Nýtni vindtúrbínna í hávinda aðstæðum er mikilvæg áskorun í bestun endurnýjanlega orkukerfisins.
+Hviðuspár geta verið stór þáttur í tjónaminnkun.
+Ef einungis er horft á vind, þá eru hviður hæsti álagspunktur út frá skilgreiningu.
+Sögulega hefur verið notast við hefðbundin spálíkön sem reiða sig á töluleg eðlisfræðileg líkön til að spá fyrir um niðurstöður.
+Á síðustu árum hefur gervigreind þróast mikið og líkön verið þjálfuð til þess að giska á veður.
+Svo miklar framfarir hafa orðið að sum líkön geta keppt við hefðbundin veðurlíkön.
+Markmið þessa verkefnis er reyna að nýta bæði hefðbundin veðurlíkön og gervigreind til þess að giska á hviðustuðla, þar sem hviðustuðullinn er skilgreindur sem hlutfall hæstu hviðu og meðalvinds.
+Þetta er gert með því að nota endurgreiningargögn úr hefðbundnu veðurlíkani sem grunngögn í gervigreindarlíkan.
+Einnig eru notað hæðarlíkan til að lýsa umhverfi.
+Djúpt tauganet var búið til með því að velja breytur úr endurgreiningargögnunum ásamt hæðarpunktum og afleiddum breytum.
+Þetta líkan er svo þjálfað og niðurstöður bornar saman við grunnlíkan sem og önnur líkön með öðrum breytum til að skoða áhrif breyta.
 \vfill

### 2024-10-07 Chapters/Chapter3.tex
Able to build with [#] citations. Also added line oto table setsOfParams

@@
     \begin{tabular}{cccc}
-
+        \toprule
         f & n & MAE & MAPE \\
-
-        {[}0;5{[} & 6.2e6 & 2.1 & 1.6\\
-        {[}5;10{[} & 4.2e6 & 2.2 & 0.3\\
-        {[}10;15{[} & 1.5e6 & 2.5 & 0.2\\
-        {[}15;20{[} & 3.9e5 & 3.0 & 0.2\\
-        {[}20;25{[} & 8.4e4 & 4.0 & 0.2\\
-        {[}25;$\infty${[} & 2.0e4 & 6.6 & 0.2\\
-        {[}0;$\infty${[} & 1.2e7 & 2.2 & 1.0\\
-
+        \midrule
+        $[0;5[$ & 6.2e6 & 2.1 & 1.6\\
+        $[5;10[$ & 4.2e6 & 2.2 & 0.3\\
+        $[10;15[$ & 1.5e6 & 2.5 & 0.2\\
+        $[15;20[$ & 3.9e5 & 3.0 & 0.2\\
+        $[20;25[$ & 8.4e4 & 4.0 & 0.2\\
+        $[25;\infty[$ & 2.0e4 & 6.6 & 0.2\\
+        $[0;\infty[$ & 1.2e7 & 2.2 & 1.0\\
+        \bottomrule
     \end{tabular}
@@
     \begin{tabular}{cccc}
-
+        \toprule
         Station & Number of measurements & MAE & Location \\
-
+        \midrule
         1470 & 6.8e3 & 1.17 & Reykjavík Háahlíð \\
@@
         32635 & 3.2e4 & 4.95 & Botn í Súgandafirði\\
-
+        \bottomrule
     \end{tabular}

### 2024-10-07 Chapters/Chapter5.tex
Able to build with [#] citations. Also added line oto table setsOfParams
Updated table with different sets of parameters model fully.

@@
 \begin{table}[h]
-    \caption[Model results for different sets of parameters.]{...}
+    \caption[Model results for different sets of parameters.]{Showing the results of different models.
+Models defined by different sets of parameters, with increasing complexity.
+This table shows the feature importance as the loss decreases as more variables are added.
+Adding all height levels decreases the loss but very little.}
     \label{table:setsOfParams}
@@
         $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd]$ & 13.8\% \\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ & ---\\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ + DEM & ---\\
-        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, ASL, twd, N, Ri]$ & --- \\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ & 13.7\%\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ + DEM & 12.0\%\\
+        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, ASL, twd, N, Ri]$ & 12.0\% \\
         \bottomrule

### 2024-10-13 Chapters/Chapter2.tex
Updated abbreviations, some text and added to appendix

@@
 Data was sourced from several streams.
-The Icelandic Meterological Office (IMO) provided measurements from weather stations all around Iceland.
+The Icelandic Meteorological Office (IMO) provided measurements from weather stations all around Iceland.
 NWP data was downloaded from Copernicus Arctic Regional Reanalysis dataset (CARRA).
@@
 This file encompasses Iceland and is around 685 MB.
-A good amount of time was spent finding a data structure that was best for lookup when trying to find points within a certain area.
-The country was divided into 10 parts (as only around 13\% of the file was able to be read into memory at each time as a part of dictionary object) with boundary boxes.
 The Python package Rasterio allowed for quick lookup with it's index and the affine transform.
 Using this package it is possible to quickly look up elevation given coordinates using matrix calculations.
+This package allowed for lookup by coordinates and by positioning inside the grid encompassing Iceland.
+It also allows for index lookup using coordinates.
+Using index lookup, neighboring points can be looked up and used to bridge the elevation for exact posisiton of given point.

### 2024-10-13 Chapters/Chapter3.tex
Updated abbreviations, some text and added to appendix

@@
 If the average wind speed differs the rows are dropped.
-
-After filtering out nails, most stations have nails in less that 10\% measurements. The stations that have higher than 10\% error are ignored.
+After filtering out nails, most stations have nails in less that 10\% measurements.
+The stations that have higher than 10\% error are ignored.
 
@@
 The last data source is the elevation.
-A sector of a circle looking upwind is looked at.
+A circle sector upwind is looked at.
 In any case the points, that represent these sections, were selected as shown in Code Listing \ref{code:sectorElevation}.
 A range of angles are defined based on the wind direction $d$ at some distance from the given point.
-This means that the resultant points (equal in number to the length of angleRange by k) form sectors at several distances from the given weather station.
+This means that the resultant points (equal in number to the length of angleRange by k) from arcs at several distances from the given weather station.
 
@@
     \centering
-    \begin{tabular}{c|c|c|c|c|c|c|c|c|c}
-         Ri & N\_squared & station\_elevation & twd & ws\_15 & wd\_15 & t\_15 & p\_15 & elevation\_point\_0 & \dots\\\hline
-         -1.18e+00 &  2.67e+04 & 100 & 1.5 & 10 & 5 & 0 & 100 & 2 & \dots 
+    \begin{tabular}{cccccccccc}
+        \toprule
+        Ri & $N^2$ & station elevation & twd & $ws_{15}$ & $wd_{15}$ & $t_{15}$ & $p_{15}$ & $elevation_0$ & \dots\\
+        \midrule 
+        -1.18e+00 &  2.67e+04 & 100 & 1.5 & 10 & 5 & 0 & 100 & 2 & \dots\\
+        \bottomrule
     \end{tabular}
@@
 Exactly the same notation is used with the Brunt–Väisälä frequency, except the square is used.
+
 \begin{equation}
@@
 Most implementations use some approximations, which still can take a considerable amount of time for models with a high parameter count and many examples.
+The Richardson number includes the difference in wind speeds in the denominator.
+In certain cases, where the difference in wind speed between two levels is very, can blow up to infinity.
+This can cause problems and distort the predicitons.

### 2024-10-13 Chapters/Chapter5.tex
Updated abbreviations, some text and added to appendix

@@
         \toprule
-        \textbf{Station name} & \textbf{Number of measurement points} & \multicolumn{2}{c}{\textbf{MAPE}}\\
+        \textbf{Station name} & \textbf{\# points} & \multicolumn{2}{c}{\textbf{MAPE}}\\
         & & General training & Site training\\
@@
 This does not seem to be the case.
-Some interesting sites to look closer at for drivers might include places like Kjalarnes, Ingólfsfjall, Þrengsli and others.
-These can be seen in Table (\ref{table:specific_sites}).
+Some interesting sites to look closer at for drivers might include places such as Ingólfsfjall, Kjalarnes and others.
+These can be seen in Table (\ref{table:more_specific_sites}).
 
@@
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/waterfall_plot.png}
-    \caption[Feature importance for a single observation of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
-In this specific instance the wind direction ($wd_{15}$) has the highest negative influence and the temperature has the highest positive influence.
-Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them.
-To see their influence on the model output see Table \ref{table:results}.}
-    \label{fig:ShapleyWaterfall}
-\end{figure}
-
 In Figure (\ref{fig:ShapleySummary}) the contribution of each feature, excluding the elevation points, can be seen for the model in general (a significant subset of data is used).
@@
 This can be seen in Figure (\ref{fig:ShapleySummary2}).
+The Figures (\ref{fig:ShapleySummary}) and (\ref{fig:ShapleySummary2}) show the summary for a model trained only on the features shown and not on landscape elevation of the surrounding area.
+This is done as the number of points there is too high to display in one figure (70 total points).
+Looking specifically at Figure (\ref{fig:ShapleySummary2}), the station elevation is most influential and the Richardson number's influence is very low.
+Most of the feature values bunch up in the middle, while the station elevation is elongated compared to other features.
+Station elevation seems to have two bunches on either side of 0.
+Overall, the values seem to be skewed to the right of 0.
+This would be expected as the predicted values are expected to be in the range of 1.2-2, or at least always above 1 by definition.
+Finally, for the Shapley values, looking that all the data there are again outliers that skew the data so that spotting general distribution is difficult.
+This can be seen in Figure (\ref{fig:ShapleySummary3}).
 
@@
 
-The Figures (\ref{fig:ShapleySummary}) and (\ref{fig:ShapleySummary2}) show the summary for a model trained only on the features shown and not on landscape elevation of the surrounding area.
-This is done as the number of points there is too high to display in one figure (70 total points).
-Looking specifically at Figure (\ref{fig:ShapleySummary2}), the station elevation is most influential and the Richardson number's influence is very low.
-Most of the feature values bunch up in the middle, while the station elevation is elongated compared to other features.
-Station elevation seems to have two bunches on either side of 0.
-Overall, the values seem to be skewed to the right of 0.
-This would be expected as the predicted values are expected to be in the range of 1.2-2, or at least always above 1 by definition.
-Finally, for the Shapley values, looking that all the data there are again outliers that skew the data so that spotting general distribution is difficult.
-This can be seen in Figure (\ref{fig:ShapleySummary3}).
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_full_10ms.png}
-    \caption[Summary feature importance of a neural network using entire dataset.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
-The distribution seems to be the same as before, discounting the outliers.}
-    \label{fig:ShapleySummary3}
-\end{figure}
-
-These plots can also be done by looking at individual stations.
-The model is then trained on all examples and afterwards the summary plot created using only points for each station.
-Stations of interest include stations listed in Table \ref{table:specific_sites}.
-These plots can all be seen in Figures (\ref{fig:ShapleySummaryAkrafjall} - \ref{fig:ShapleySummaryKeflavikurflugvollur}).
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_31572.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Akrafjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Akrafjall.
-This seems to show the same distribution as previous summary plots.
-Station elevation is influential and Richardson number has no impact.}
-    \label{fig:ShapleySummaryAkrafjall}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_35553.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Almannaskarð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Almannaskarð.
-This seems to show the same distribution as previous summary plots.
-Station elevation is influential and Richardson number has no impact.}
-    \label{fig:ShapleySummaryAlmannaskarð}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_6745.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Ásgarðsfjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Ásgarðsfjall.
-This seems to show the same distribution as previous summary plots.
-Station elevation is influential and Richardson number has no impact.}
-    \label{fig:ShapleySummaryAsgarðsfjall}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1470.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Háahlíð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Háahlíð.
-This seems to show the same distribution as previous summary plots.
-Station elevation is influential and Richardson number has no impact.}
-    \label{fig:ShapleySummaryHaahlid}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1350.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Keflavíkurflugvöllur.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Keflavíkurflugvöllur.
-This seems to show the same distribution as previous summary plots.
-Station elevation is influential and Richardson number has no impact.}
-    \label{fig:ShapleySummaryKeflavikurflugvollur}
-\end{figure}
-
+Further Shapley figures can be seen in Appendix~\ref{appendix:A}.
 In each of these plots, even without the feature labels, the station elevation is easily noticed as the value is constant for a station.

### 2024-10-13 MSc.tex
Updated abbreviations, some text and added to appendix

@@
 \usepackage{arydshln}
+\usepackage{acronym}
 \usetikzlibrary{positioning}
@@
 \markboth{Abbreviations}{Abbreviations}
-
-Í þessum kafla mega koma fram listar yfir skammstafanir og/eða breytuheiti. Gefið kaflanum nafn við hæfi, t.d.\ Skammstafanir eða Breytuheiti. Þessum kafla má sleppa ef hans er ekki þörf. \\
-
-The section could be titled: Glossary, Variable Names, etc.
-
-If you use acronyms, it is strongly recommended to use a LaTeX
-acronyms package. For example, to use the package \emph{acronym}, add
-in the preamble:
-\begin{verbatim}
-\usepackage{acronym}
-\end{verbatim}
-and then here in this chapter (to create the list of all acronyms), use:
-\begin{verbatim}
-\begin{acronym}[SENS]
+\begin{acronym}[GeoTIFF]
+  \acro{API}{Application Programming Interface}
+  \acro{ASL}{Above Sea Level}
+  \acro{AWS}{Automatic Weather Stations}
+  \acro{AWSL}{Average Wind Speed Limit}
+  \acro{CARRA}{Copernicus Artic Regional ReAnalysis dataset}
+  \acro{CNN}{Convolutional Neural Networks}
+  \acro{DEM}{Digital Elevaiton Model}
+  \acro{ELI5}{Explain Like I am 5}
+  \acro{ECMWF}{European Centre for Medium-Range Weather Forecast}
+  \acro{GCM}{General Circulation Model}
+  \acro{GeoTIFF}{Georeferenced TIFF}
+  \acro{GPU}{Graphical Processing Unit}
+  \acro{HRES}{High Resolution forecas}
+  \acro{IMO}{Icelandic Meteoroligcal Office}
+  \acro{IRCA}{Icelandic Road and Coastal Administration}
+  \acro{JNWPU}{Joint Numerical Weather Prediction Unit}
+  \acro{LWM}{Large AI Weather forecast Model}
+  \acro{MAE}{Mean Asbolute Error}
+  \acro{MAPE}{Mean Absolute Percentage Error}
+  \acro{NN}{Neural Network}
+  \acro{NWP}{Numerical Weather Prediction}
   \acro{SENS}{School of Engineering and Natural Sciences}
+  \acro{TIFF}{Tag Image File Format}
   \acro{UoI}{University of Iceland}
 \end{acronym}
-\end{verbatim}
-In the square bracket above, you need to put the longest acronym, so
-that the list gets proper indentation based on the longest acronym.
-Also note that you need to manually sort here that list.
-
-Then, whereever you use the acronym in your text: \verb|\ac{UoI}|:
-that will automatically expand the acronym at the first use, but use
-only the short version after the first use.  (Use \verb|\acp| if need
-the plural form. \verb|\acl| if you explicitly want to have the long
-form only, \verb|\acf| if you explicitly want to have the full long
-and short form, i.e.\ like first use of \verb|\ac|.)
-
-Use \verb|\acresetall| to forget about earlier usage (=expansion) of
-acronyms, e.g.\ if despite already used in Abstract or Introduction,
-but you want to expand them later once again (starting from, e.g., in
-Foundations chapter): add \verb|\acresetall| at start of Introduction
-chapter (and maybe also again at start of Foundations chapter).
 
@@
 \addcontentsline{toc}{chapter}{Acknowledgments}
-Í þessum kafla koma fram þakkir til þeirra sem hafa styrkt rannsóknina með fjárframlögum, aðstöðu eða vinnu.
-t.d.\ styrktarsjóðir, fyrirtæki, leiðbeinendur, og aðrir aðilar sem hafa á einhvern hátt aðstoðað við gerð verkefnisins, þ.m.t.\ vinir og fjölskylda ef við á.
-Þakkir byrja á oddatölusíðu (hægri síðu).
+Special thanks go to my advisor, Kristján Jónasson.
+He would help me with the actual work of the thesis and sit with go over the code with me.
+Thanks also go to Ólafur Pétur Pálsson, who would read over the thesis with notes and give helpful suggestions.
+Guðrún Nína Petersen, also provided valuable insight whenever questioned about meteoroligcal matters as well as giving notes on the thesis.
+
+Finally, I would like to thank my parents without whom I would probably never had any want to finish.
 
@@
 \include{Chapters/Chapter6}
+\include{Appendicies/AppendixA}
 

### 2025-01-29 Appendicies/AppendixA.tex
Updated tables and appendices

\chapter*{Appendix A: Feature importance on Shapley plots}
\label{appendix:A}

\renewcommand{\thefigure}{A.\arabic{figure}} 
\setcounter{figure}{0} % Reset figure counter

\begin{figure}[h]
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_plots/waterfall_plot.png}
    \caption[Feature importance for a single observation of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}. In this specific instance the wind direction ($wd_{15}$) has the highest negative influence and the temperature has the highest positive influence. Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them. To see their influence on the model output see Table \ref{table:results}.}
    \label{fig:ShapleyWaterfall}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot.png}
    \caption[Summary feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}. We can see that generally multiple factors influence the prediction, with the station elevation being highly influential. There is seemingly one outlier for the Richardson number, which usually has very little influence. Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them. To see their influence on the model output see Table \ref{table:results}.}
    %\label{fig:ShapleySummary}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_.png}
    \caption[Summary feature importance of a neural network using a larger distribution of data.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}. Generally multiple factors influence the prediction, with the station elevation being highly influential. Elevation data is excluded when working with Shapley values, as the contribution of each elevation. In contrast to Figure (\ref{fig:ShapleySummary}), the distribution doesn't have as extreme outliers. This means that more details can be seen in the figure. The X-axis shows the influence of feature values on the model. The color gradient shows the value of each feature. As an example, there is a very red value for station elevation (top line, all the way to the left). This means that in this instance, the station elevation contributed around -0.25 to the final output and that the station had an elevation significantly above average.}
    %\label{fig:ShapleySummary2}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_full_10ms.png}
    \caption[Summary feature importance of a neural network using entire dataset.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}. The distribution seems to be the same as before, discounting the outliers.}
    \label{fig:ShapleySummary3}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_31572.png}
    \caption[Summary feature importance of a neural network only looking at AWS at Akrafjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}. This plot only looks at datapoints from Akrafjall. This seems to show the same distribution as previous summary plots. Station elevation is influential and Richardson number has no impact.}
    \label{fig:ShapleySummaryAkrafjall}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_35553.png}
    \caption[Summary feature importance of a neural network only looking at AWS at Almannaskarð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}. This plot only looks at datapoints from Almannaskarð. This seems to show the same distribution as previous summary plots. Station elevation is influential and Richardson number has no impact.}
    \label{fig:ShapleySummaryAlmannaskarð}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_6745.png}
    \caption[Summary feature importance of a neural network only looking at AWS at Ásgarðsfjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}. This plot only looks at datapoints from Ásgarðsfjall. This seems to show the same distribution as previous summary plots. Station elevation is influential and Richardson number has no impact.}
    \label{fig:ShapleySummaryAsgarðsfjall}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1470.png}
    \caption[Summary feature importance of a neural network only looking at AWS at Háahlíð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}. This plot only looks at datapoints from Háahlíð. This seems to show the same distribution as previous summary plots. Station elevation is influential and Richardson number has no impact.}
    \label{fig:ShapleySummaryHaahlid}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1350.png}
    \caption[Summary feature importance of a neural network only looking at AWS at Keflavíkurflugvöllur.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}. This plot only looks at datapoints from Keflavíkurflugvöllur. This seems to show the same distribution as previous summary plots. Station elevation is influential and Richardson number has no impact.}
    \label{fig:ShapleySummaryKeflavikurflugvollur}
\end{figure}

### 2025-01-29 Chapters/Chapter3.tex
Updated tables and appendices

@@
 Here linear interpolation was applied, both temporally and spatially.
+As the measurement points chosen were not outliers, that is the strongest average wind speed in a given range, interpolating in time does not work well.
+This necessitates using only measurements that are exactly at three hour intervals (00, 03, 06, 09, 12, 15, 18, 21 hours) at interpolating only in space.
 The choice of interpolation method, although potentially impactful on the results, was not specifically addressed in this study.
@@
 Then filtering is applied on the average wind.
-If the average wind speed differs the rows are dropped.
-After filtering out nails, most stations have nails in less that 10\% measurements.
+If the average wind speed for the 10 minute data is higher than the hourly the rows are dropped.
+Most stations have nails in less that 10\% measurements.
 The stations that have higher than 10\% error are ignored.

### 2025-01-29 Chapters/Chapter6.tex
Updated tables and appendices

\chapter{Discussion}
\label{Chapter6}
The goal of this tesis was to research whether it was possible to use reanalysis data to increase the predictability of wind gusts and see what influence including the landscape elevation would have. A quick statistical analysis will give a p-value of 0, that is these results are statistically significant. The improvement shown in Table (\ref{table:results}) is not chance. There is some pattern to be learned in the interval that leads to a higher predictability of wind gusts over baseline.

### 2025-03-26 Chapters/Chapter2.tex
Combined C2/3 and added data figures.

@@
 
-\chapter{Data gathering and preprocessing}
+\chapter{Data gathering and processing}
 \label{Chapter2}
@@
 Using index lookup, neighboring points can be looked up and used to bridge the elevation for exact posisiton of given point.
+
+\label{Chapter3} 
+\section{Combining data sources}
+
+This project used three main data sources, which need to be queried, filtered and combined to prepare the data for use in the models.
+When working with hundred of thousands of rows, the efficiency of the code is very important.
+Iterating through those rows might be necessary at times but will increase the time exponentially as compared to using vectorizing methods were possible.
+The three data sources were all in a different format.
+Measurement data from IMO was in text files, elevation data was in GeoTiff and reanalysis data from CARRA was in a GRIB format.
+To use the data to train, these three data sources needed to be combined into one file.
+This was done based on the measurement data from the IMO.
+A limit was set on the average wind speed and it was used to select measurement points.
+Along with the average wind speed having to be above a certain limit, to ensure that the same weather for the same location is not duplicated.
+The data from IMO was supplied for 10 minute increments, while CARRA data is in 3 hour intervals.
+This means that to use the CARRA data to predict the measured values from IMO, temporal interpolation would need to be done.
+Along with the temporal interpolation, note that the CARRA data is given in a rectangular grid where the distance between each point is around 2.5 km while the the information from the IMO is given at specific locations.
+The elevation information was given by a 20 by 20 rectangular grid that covers Iceland.
+When combining these data sources an interpolation method needs to be decided upon.
+Here linear interpolation was applied, both temporally and spatially.
+As the measurement points chosen were not outliers, that is the strongest average wind speed in a given range, interpolating in time does not work well.
+This necessitates using only measurements that are exactly at three hour intervals (00, 03, 06, 09, 12, 15, 18, 21 hours) at interpolating only in space.
+The choice of interpolation method, although potentially impactful on the results, was not specifically addressed in this study.
+
+\begin{figure}[h]
+    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart.png}
+    \caption{A flow chart showing how data sources were combined}
+    \label{fig:data_preprocessing_flow_chart}
+\end{figure}
+
+The procedure of combining these sources was as follows and can be seen in Figure \ref{fig:data_preprocessing_flow_chart}.
+The measured data from the AWS is filtered by using a limit on the average wind speed.
+The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}).
+Even so being able to predict the gust factor is more important for higher average wind speed due to higher wind gusts.
+After this stripped dataset over every AWS has been created it is used to query the CARRA data by using their API.
+The CARRA API needs to be queried for given hours, days, months, years and a given area.
+That is, if queried for a given hour, it returns that hour for every day that is queried.
+Similarly if queried for a given day, it returns that day for every month.
+In light of these restraints, it was decided to query month by month.
+Querying only the days needed but every hour of the day (UTC 00, 03, 06, 09, 12, 15, 18 and 21).
+After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe.
+After this the downloaded data is discarded and the next month is queried.
+This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (a reduction from several terabytes to less than a gigabyte).
+
+Once CARRA data has been merged with AWS data, using station and time columns, then this combined file needs to checked for nails.
+This is done by using the hourly data (which is supposedly error free).
+As the data has been filtered in for the highest average wind speed in a 48 hour interval, the hourly data can be used to find nails.
+The hourly data is combined with the merged AWS and 10 min data.
+Then filtering is applied on the average wind.
+If the average wind speed for the 10 minute data is higher than the hourly the rows are dropped.
+Most stations have nails in less that 10\% measurements.
+The stations that have higher than 10\% error are ignored.
+
+The elevation data comes in a GeoTIFF file that covers Iceland.
+It is a rectangular grid of resolution 20 meters.
+For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved.
+For each point retrieved interpolation needs to be done.
+This is done in a similar manner to the interpolation of the CARRA data.
+The four points bounding the point of interest were used to linearly interpolate the value of the point of interest.
+This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.
+
+The error in reanalysis wind speed and measured wind speed can be significant.
+The absolute error increases as the measured wind speed increases, while the percentage wind speed decreases.
+A grouping of these errors by wind speed can be seen in Table \ref{table:measuredVSReanalysis_wind_speed}
+
+\begin{table}[h]
+    \caption[Comparison of measured and reanalysis wind speed]{Comparison of measured and reanalysis wind speed using mean absolute error (MAE) and mean absolute percentage error (MAPE).
+Note that for the computation of MAPE for ranges that otherwise include 0, 0 values have been excluded so as to prevent division by zero and exploding values.
+The comparisons are done using measured wind speed (at 10 meters above ground for IMO and 6-7 meters above ground for IRCA) and reanalysis wind speed at 15 meters above ground.}
+    \label{table:measuredVSReanalysis_wind_speed}
+    \centering
+    \begin{tabular}{ccc}%c}
+        \toprule
+        f & n & MAE \\%& MAPE \\
+        \midrule
+        $[0;5[$ & 6.2e6 & 2.1 \\%& 1.6\\
+        $[5;10[$ & 4.2e6 & 2.2 \\%& 0.3\\
+        $[10;15[$ & 1.5e6 & 2.5 \\%& 0.2\\
+        $[15;20[$ & 3.9e5 & 3.0 \\%& 0.2\\
+        $[20;25[$ & 8.4e4 & 4.0 \\%& 0.2\\
+        $[25;\infty[$ & 2.0e4 & 6.6 \\%& 0.2\\
+        $[0;\infty[$ & 1.2e7 & 2.2 \\%& 1.0\\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+Another thing to look at is the distribution of error by station, both in terms of their coordinates and number of measurements.
+Looking at Figure \ref{fig:station_mae_distribution}, this distribution can be seen.
+
+\begin{figure}
+    \includegraphics[scale=0.6]{Figures/MAEoverIceland.png}
+    \caption[Distribution of mean absolute errors by station]{The distribution of mean absolute errors by station.
+Using mean absolute error instead of mean absolute percentage error allows for all points to be used.
+Mean absolute percentage error can only be used if 0 values are ignored.}
+    \label{fig:station_mae_distribution}
+\end{figure}
+
+Table \ref{table:station_mae_distribution} shows the 5 best and worst stations in terms of MAE.
+
+\begin{table}[h]
+    \caption[Mean absolute difference of measured wind speed and reanalysis wind speed at select stations]{Mean absolute error for reanalysis wind speed as compared to measured wind speed, for the five stations with the highest difference and the five stations with the lowest difference.}
+    \label{table:station_mae_distribution}
+    \resizebox{\textwidth}{!}{
+    \centering
+    \begin{tabular}{cccc}
+        \toprule
+        Station & Number of measurements & MAE & Location \\
+        \midrule
+        1470 & 6.8e3 & 1.17 & Reykjavík Háahlíð \\
+        1350 & 5.2e4 & 1.18 & Keflavíkurflugvöllur \\
+        1482 & 1.4e4 & 1.23 & Reykjavík Víðidalur \\
+        4921 & 1.3e4 & 1.29 & Rif á Melrakkasléttu \\
+        1477 & 5.6e4 & 1.29 & Reykjavíkurflugvöllur\\ \hdashline[0.5pt/1pt]
+        35553 & 4.0e3 & 4.30 & Almannaskarð - göng\\
+        6745 & 1.5e4 & 4.36 & Kerlingarfjöll - Ásgarðsfjall\\
+        35978 & 7.9e3 & 4.40 & Fáskrúðsfjarðargöng suður\\
+        2640 & 1.6e3 & 4.51 & Seljalandsdalur\\
+        32635 & 3.2e4 & 4.95 & Botn í Súgandafirði\\
+        \bottomrule
+    \end{tabular}
+    }
+\end{table}
+
+If the 0 m/s is excluded from the measurement data, then the error distribution using MAPE can be plotted. This plot can be seen in Figure \ref{fig:station_mape_distribution}
+
+\section{Data Structure}
+
+Once data has been retrieved for all three sources and processed, including interpolating values, it needs to be made ready to use by the model, for both training, validation and test.
+The starting point is a dataframe that contains measured information from AWS.
+This includes the average wind, the wind gust, wind direction along with the station number and coordinates.
+When selecting the CARRA data certain height levels are chosen.
+These present as separate lines in the CARRA dataframe.
+Information for one observation is represented in as many lines as height levels requested in the reanalysis data.
+These rows need to be combined on the position (the weather station).
+When this is done it is possible to combine the AWS IMO data and CARRA reanalysis data on the location and time columns.
+The last data source is the elevation.
+A circle sector upwind is looked at.
+In any case the points, that represent these sections, were selected as shown in Code Listing \ref{code:sectorElevation}.
+A range of angles are defined based on the wind direction $d$ at some distance from the given point.
+This means that the resultant points (equal in number to the length of angleRange by k) from arcs at several distances from the given weather station.
+
+\begin{lstlisting}[style = Python, caption = {Sector elevation points generated}, label = code:sectorElevation]
+angles = [(angle + (90 - d)) * pi/180 for angle in angleRange]
+length_rng = [(exp(i * log(n + 1)/ k) - 1) * 1000 
+                for i in range(1, k + 1)]
+points = np.array([[(X + l * cos(angle), Y + l * sin(angle))
+                    for angle in angles] for l in length_rng])   
+\end{lstlisting}
+
+The result is a dataframe that has measured data from AWS, which gives us our target, reanalysis data from CARRA, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data.
+An example of what the data looks like can be seen in Table \ref{table:trainDataExample}.
+
+\begin{table}[h]
+    \caption[An example of data structure used to train model]{An example of data structure used to train model.
+Data points include the derived variables Ri and N, the elevation of the station, direction of wind and relative direction of the wind (twd, that is the direction of the wind relative to center of Iceland), along with some combination of wind speed, pressure and temperature at the different height levels.
+Finally there are the elevation points around a given station, where the elevation is relative to the station.}
+    \label{table:trainDataExample}
+    \resizebox{\textwidth}{!}{
+    \centering
+    \begin{tabular}{cccccccccc}
+        \toprule
+        Ri & $N^2$ & station elevation & twd & $ws_{15}$ & $wd_{15}$ & $t_{15}$ & $p_{15}$ & $elevation_0$ & \dots\\
+        \midrule 
+        -1.18e+00 &  2.67e+04 & 100 & 1.5 & 10 & 5 & 0 & 100 & 2 & \dots\\
+        \bottomrule
+    \end{tabular}
+    }
+\end{table}
+
+Looking at Table \ref{table:trainDataExample} note that the first two columns represent two variables that describe the stability of the air.
+These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency\cite{brunt_vaisala_freq_eumtrain} ($N$), and are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}.
+These values are calculated using reanalysis data at two different height levels.
+Thus $Ri$ refers to the Richardson number calculated between height levels 15m and 500m.
+Exactly the same notation is used with the Brunt–Väisälä frequency, except the square is used.
+
+\begin{equation}
+    \label{eqn:Ri}
+    Ri = \frac{g \cdot dT \cdot dz}{T_{\textrm ave} \cdot dU^2} \unit{[]}
+\end{equation}
+
+\begin{equation}
+    \label{eqn:N}
+    N = \sqrt{\frac{g \cdot dT }{T_{\textrm ave} \cdot dz}} \unit{[Hz]}
+\end{equation}
+
+Here, $g$ is the acceleration due to gravity, $dT$ is the temperature difference between the two height levels, $dz$ is the elevation difference, $T_{\textrm ave}$ is the average temperature (that is the average of the two temperatures in the height levels) and $dU$ is the wind speed difference between the two height levels.
+Both of these numbers provide some insight about the stability of the air.
+A lower value for the Richardson number indicates a higher turbulence.
+A typical range of values could be between 0.1 and 10, with values below 1 indicating significant turbulence\cite{richardson_number_skybrary}.
+When the square of the Brunt-Väisälä frequency is negative, then the air is unstable (an air parcel will move away from its original position)\cite{brunt_vaisala_freq_eumtrain}.
+These are derived factors from the reanalysis data and as such there shouldn't be a significant information gain using $Ri$ and $N$ as opposed to having the raw data.
+However, including these factors instead of every reanalysis variable requested might speed up training as well as making the model more easily explainable with the use of Shapley values or other tools for explainability.
+Using Shapley a feature importance value is attributed to a given feature by creating all possible permutations of any possible length (up to number of features) and seeing how the predictions are skewed when the given parameter is included or excluded.
+This needs to be done for all parameters.
+The time complexity of this is very high ($2^n$ coalitions)\cite{shapley_information}.
+Most implementations use some approximations, which still can take a considerable amount of time for models with a high parameter count and many examples.
+The Richardson number includes the difference in wind speeds in the denominator.
+In certain cases, where the difference in wind speed between two levels is very, can blow up to infinity.
+This can cause problems and distort the predicitons.
+
+\section{Data distribution}
+
+The CARRA data is reanalysis and as such might have a bias or some systemic distortion when compared to the measured data.
+The distribution of the observed and reanalysis wind speed can be seen in Figure \ref{fig:obs_carra_wind_speeds}.
+Looking at the figure, the reanalysis wind tends to be higher even though the distribution is similar.
+
+\begin{figure}[ht]
+    \centering
+    % First subfigure (top)
+    \begin{subfigure}[b]{0.8\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{Figures/obs_wind_speeds.png}
+        \caption[Histogram of observed wind speeds]{A histogram distribution of observed wind speeds provided by IMO and IRCA for all used weather stations.}
+        \label{fig:obs_wind_speeds}
+    \end{subfigure}
+
+    \vspace{0.5cm} % Space between the two images
+
+    % Second subfigure (bottom)
+    \begin{subfigure}[b]{0.8\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{Figures/carra_wind_speeds.png}
+        \caption[Distribution of interpolated CARRA wind speeds at weather stations]{Distribution of interpolated CARRA reanalysis values at weather stations.
+The reanalysis data is only available at 3 hour intervals and 2.5km grid.
+These values are interpolated from data given by CARRA at the weather stations.}
+        \label{fig:carra_wind_speeds}
+    \end{subfigure}
+
+    \caption[Distribution of CARRA and observed wind speeds]{Distribution of CARRA reanalysis wind speed and observed wind speed at weather stations provided by}
+    \label{fig:obs_carra_wind_speeds}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.8\textwidth]{Figures/station_heights.png}
+    \caption[Distribution of weather station heights above see level]{Distribution of ...}
+    \label{fig:station_heights}
+\end{figure}

### 2025-03-26 Chapters/Chapter2.tex
Added a couple of data figures in c2 and added text.

@@
     \includegraphics[width=0.8\textwidth]{Figures/station_heights.png}
-    \caption[Distribution of weather station heights above see level]{Distribution of ...}
+    \caption[Distribution of weather station heights above sea level]{Distribution of heights of weather stations above sea level.
+One station has been excluded as an outlier of well above 1000 meters, having few and inconsistent datapoints.}
     \label{fig:station_heights}
 \end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.8\textwidth]{Figures/gust_factor_2025.png}
+    \caption[Distribution of gust factors]{Histogram of gust factors.
+By definition the lower bound of gust factor is 1.
+The majority of observed gust factors fall in the range 1.2 to 2.
+Gust factor decreases with increasing wind speed.}
+    \label{fig:gust_factors}
+\end{figure}

### 2025-04-07 Chapters/Chapter1.tex
To be read by KJ

@@
 
-\subsection{Model architechture}
+\subsection{Neural networks}
 To be able to capture the patterns in the data a neural network was constructed.
@@
 It is also easy to construct different types of neural networks and see how they fit well with parts of the dataset.
+A NN uses a lot of matrix calculations to weight input parameters and predict an output.
+A NN has some number of layers, a deep neural network (DNN) has an input layer, output layer and some number of hidden layers.
+Hidden layers are middle layers that take data from input layer and eventually reach the output layer.
+The input layer has a width equal to the number of parameters and the ouput has width equal to the number of predictants.
+Models were created for several number of parameters as a way to gauge the influence of each parameter.
+Each model has one numerical output variable.
+Each layer also has an activation function.
+There are several popular activation functions such as Rectified Linear Unit (ReLU), Exponential Linear Unit (ELU) and Hyperbolic tangetn (tanh).
+The definitions of ReLU, ELU and tanh can be seen in Equations (\ref{eqn:relu}), (\ref{eqn:elu}) and (\ref{eqn:tanh}).
+
+\begin{equation}
+    \label{eqn:relu}
+    r(x):=max(0, x)
+\end{equation}
+
+\begin{align}
+    \label{eqn:elu}
+    f(x) := x, x > 0\\
+    f(x) := \alpha (e^x-1), x\leq 0, \alpha > 0
+\end{align}
+
+\begin{equation}
+    \label{eqn:tanh}
+    tanh(x):=\frac{e^x-e^{-x}}{e^x+e^{-x}}
+\end{equation}
+
+These activation control neural activation and can help stabilize the network.
+
+\subsection{Model evaluation}
  To measure the performance of these models, both to train and test, mean absolute percentage error (MAPE) as defined in Equation (\ref{eqn:mape}) was used.

### 2025-04-07 Chapters/Chapter2.tex
To be read by KJ

@@
 
-If the 0 m/s is excluded from the measurement data, then the error distribution using MAPE can be plotted. This plot can be seen in Figure \ref{fig:station_mape_distribution}
-
 \section{Data Structure}
@@
         \includegraphics[width=\textwidth]{Figures/obs_wind_speeds.png}
-        \caption[Histogram of observed wind speeds]{A histogram distribution of observed wind speeds provided by IMO and IRCA for all used weather stations.}
         \label{fig:obs_wind_speeds}
@@
         \includegraphics[width=\textwidth]{Figures/carra_wind_speeds.png}
-        \caption[Distribution of interpolated CARRA wind speeds at weather stations]{Distribution of interpolated CARRA reanalysis values at weather stations.
-The reanalysis data is only available at 3 hour intervals and 2.5km grid.
-These values are interpolated from data given by CARRA at the weather stations.}
         \label{fig:carra_wind_speeds}
@@
 
-    \caption[Distribution of CARRA and observed wind speeds]{Distribution of CARRA reanalysis wind speed and observed wind speed at weather stations provided by}
+    \caption[Distribution of CARRA and observed wind speeds]{Upper figure shows a histogram of observed wind speeds provided by IMO and IRCA for all used weather stations.
+Lower figure shows the interpolated CARRA reanalysis values at weather stations.
+The reanalysis data is only available at 3 hour intervals and 2.5km grid.
+As such the observed values are only chosen at these specific times (00, 03, 06, ..., 21).
+Spatial interpolation is needed and is linearly weighted based on distance of CARRA grid points from stations.}
     \label{fig:obs_carra_wind_speeds}

### 2025-04-07 Chapters/Chapter3.tex
To be read by KJ

--- 2024-10-07 Chapters/Chapter3.tex
@@
-\chapter{Data processing and structure}
+% Chapter Template
+
+\chapter{Model Architecture and Training}
+
 \label{Chapter3}
-\section{Combining data sources}
 
-This project used three main data sources, which need to be queried, filtered and combined to prepare the data for use in the models.
-When working with hundred of thousands of rows, the efficiency of the code is very important.
-Iterating through those rows might be necessary at times but will increase the time exponentially as compared to using vectorizing methods were possible.
-The three data sources were all in a different format.
-Measurement data from IMO was in text files, elevation data was in GeoTiff and reanalysis data from CARRA was in a GRIB format.
-To use the data to train, these three data sources needed to be combined into one file.
-This was done based on the measurement data from the IMO.
-A limit was set on the average wind speed and it was used to select measurement points.
-Along with the average wind speed having to be above a certain limit, to ensure that the same weather for the same location is not duplicated.
-The data from IMO was supplied for 10 minute increments, while CARRA data is in 3 hour intervals.
-This means that to use the CARRA data to predict the measured values from IMO, temporal interpolation would need to be done.
-Along with the temporal interpolation, note that the CARRA data is given in a rectangular grid where the distance between each point is around 2.5 km while the the information from the IMO is given at specific locations.
-The elevation information was given by a 20 by 20 rectangular grid that covers Iceland.
-When combining these data sources an interpolation method needs to be decided upon.
-Here linear interpolation was applied, both temporally and spatially.
-The choice of interpolation method, although potentially impactful on the results, was not specifically addressed in this study.
-
-\begin{figure}[h]
-    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart.png}
-    \caption{A flow chart showing how data sources were combined}
-    \label{fig:data_preprocessing_flow_chart}
-\end{figure}
-
-The procedure of combining these sources was as follows and can be seen in Figure \ref{fig:data_preprocessing_flow_chart}.
-The measured data from the AWS is filtered by using a limit on the average wind speed.
-The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}).
-Even so being able to predict the gust factor is more important for higher average wind speed due to higher wind gusts.
-After this stripped dataset over every AWS has been created it is used to query the CARRA data by using their API.
-The CARRA API needs to be queried for given hours, days, months, years and a given area.
-That is, if queried for a given hour, it returns that hour for every day that is queried.
-Similarly if queried for a given day, it returns that day for every month.
-In light of these restraints, it was decided to query month by month.
-Querying only the days needed but every hour of the day (UTC 00, 03, 06, 09, 12, 15, 18 and 21).
-After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe.
-After this the downloaded data is discarded and the next month is queried.
-This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (a reduction from several terabytes to less than a gigabyte).
-
-Once CARRA data has been merged with AWS data, using station and time columns, then this combined file needs to checked for nails.
-This is done by using the hourly data (which is supposedly error free).
-As the data has been filtered in for the highest average wind speed in a 48 hour interval, the hourly data can be used to find nails.
-The hourly data is combined with the merged AWS and 10 min data.
-Then filtering is applied on the average wind.
-If the average wind speed differs the rows are dropped.
-
-After filtering out nails, most stations have nails in less that 10\% measurements. The stations that have higher than 10\% error are ignored.
-
-The elevation data comes in a GeoTIFF file that covers Iceland.
-It is a rectangular grid of resolution 20 meters.
-For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved.
-For each point retrieved interpolation needs to be done.
-This is done in a similar manner to the interpolation of the CARRA data.
-The four points bounding the point of interest were used to linearly interpolate the value of the point of interest.
-This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.
-
-The error in reanalysis wind speed and measured wind speed can be significant.
-The absolute error increases as the measured wind speed increases, while the percentage wind speed decreases.
-A grouping of these errors by wind speed can be seen in Table \ref{table:measuredVSReanalysis_wind_speed}
+\section{Model structure}
+The structure of the neural network is such that it contains some number $n$ of fully connected layers and batch normalization for each layer, along with regularization.
+All of these layers have the same number of units.
+The last layer has a dropout of 50\%.
+In addition to these layers there is one more output layer.
+This is simply a dense layer with 1 unit.
+A grid search was performed to determine the hyperparameters that minimize the loss.
+Hyperparameters are parameters that are set before the training begins\cite{hyperparameters_definition}.
+These hyperparameters include number of units in each layer, number of epochs to train for, number of layers, batch size, optimizer and penalty to enforce in the regularization.
+The possible combinations tried can be seen in Table \ref{table:gridSearchHyperparamters}.
 
 \begin{table}[h]
-    \caption[Comparison of measured and reanalysis wind speed]{Comparison of measured and reanalysis wind speed using mean absolute error (MAE) and mean absolute percentage error (MAPE).
-Note that for the computation of MAPE for ranges that otherwise include 0, 0 values have been excluded so as to prevent division by zero and exploding values.
-The comparisons are done using measured wind speed (at 10 meters above ground for IMO and 6-7 meters above ground for IRCA) and reanalysis wind speed at 15 meters above ground.}
-    \label{table:measuredVSReanalysis_wind_speed}
     \centering
-    \begin{tabular}{cccc}
+    \caption[Hyperparameter search with best performing combination.]{Hyperparamter search with best performing combination shown.
+Hyperparameter search was done using hyperband algorithm that initially searches randomly for the best parameters but then hones in on what is working and as such is neither exhaustive nor completely random.
+This means that a hard upper limit will not be set on the number of combinations to try like with randomsearch.}
+    \label{table:gridSearchHyperparamters}
+    \begin{tabular}{ccc}
         \toprule
-        f & n & MAE & MAPE \\
+        Parameter & Range of values & Selected\\
         \midrule
-        $[0;5[$ & 6.2e6 & 2.1 & 1.6\\
-        $[5;10[$ & 4.2e6 & 2.2 & 0.3\\
-        $[10;15[$ & 1.5e6 & 2.5 & 0.2\\
-        $[15;20[$ & 3.9e5 & 3.0 & 0.2\\
-        $[20;25[$ & 8.4e4 & 4.0 & 0.2\\
-        $[25;\infty[$ & 2.0e4 & 6.6 & 0.2\\
-        $[0;\infty[$ & 1.2e7 & 2.2 & 1.0\\
+        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{10}\\
+        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{64}\\
+        Penalties & min\_value = 1e-5, max\_value = 1, sampling = log & \textbf{1e-4}\\
+        Epochs & min\_value = 10, max\_value = 1000, step = 10 & \textbf{250}\\
+        Optimizers & Adam, RMSprop, Adamax & \textbf{Adamax}\\
+        Activation & ReLU, ELu, tanh & \textbf{ReLU}\\
         \bottomrule
@@
 
-Another thing to look at is the distribution of error by station, both in terms of their coordinates and number of measurements.
-Looking at Figure \ref{fig:station_mae_distribution}, this distribution can be seen.
+As mentioned, hyperband doesn't set a hard upper limit to the number of epochs it will train in total.
+When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{hyperband class} several factors can be set.
+One of interest here is the $hyperband_iterations$ argument.
+This determines how often the hyperband algorithm is run and defaults to 1.
+For each iteration the epochs are distributed between tries (that is each set of hyperparameters) with the total amount of epochs approximately $n_{epochs} = max_{epochs} * log^2(max_{epochs})\approx 10^4$, where $max_{epochs} = 1000$ gives the maximum number of epochs that one set of hyperparameters can be trained for.
+Searching a space generally takes a lot of time but this drastically improves on gridsearch.
+If each epoch takes around 10 seconds to run then the total search would take around 28 hours on a shared resource.
+This is resource intensive and cannot be repeated often.
+Another question that remains is whether the ranges given are optimal.
+
+\section{Model Training}
+To determine the required number of epochs a high number of epochs can be selected and the loss plotted.
+These plots can be seen in Figures (\ref{fig:loss_plot_wo_elevation}, \ref{fig:loss_plot_w_elevation}).
+Looking at these two plots, training loss decreases for over 1900 epochs in both plots, while the validation loss stops decreasing at around 200 epochs for DEM model and 1800 epochs for model without DEM.
+This is unexpected.
+The DEM model has vastly more parameters and as such should contain more useful information to be learned, which one might expect to take longer to learn.
+This does not seem to be the case.
+The DEM model only gets trivially better after 1000 epochs while the non DEM model keeps meaningfully improving until around 1800 epochs.
+With respect to time constraints of shared resources models were trained for 250 epochs.
 
 \begin{figure}
-    \includegraphics[scale=0.6]{Figures/MAEoverIceland.png}
-    \caption[Distribution of mean absolute errors by station]{The distribution of mean absolute errors by station.
-Using mean absolute error instead of mean absolute percentage error allows for all points to be used.
-Mean absolute percentage error can only be used if 0 values are ignored.}
-    \label{fig:station_mae_distribution}
+    \centering
+    \includegraphics[scale = 0.6]{Figures/loss_plot_wo_elevation.png}
+    \caption[Loss plot of model trained for 2000 without DEM.]{Loss plot of model trained for 2000 without DEM.}
+    \label{fig:loss_plot_wo_elevation}
 \end{figure}
 
-Table \ref{table:station_mae_distribution} shows the 5 best and worst stations in terms of MAE.
-
-\begin{table}[h]
-    \caption[Mean absolute difference of measured wind speed and reanalysis wind speed at select stations]{Mean absolute error for reanalysis wind speed as compared to measured wind speed, for the five stations with the highest difference and the five stations with the lowest difference.}
-    \label{table:station_mae_distribution}
-    \resizebox{\textwidth}{!}{
+\begin{figure}
     \centering
-    \begin{tabular}{cccc}
-        \toprule
-        Station & Number of measurements & MAE & Location \\
-        \midrule
-        1470 & 6.8e3 & 1.17 & Reykjavík Háahlíð \\
-        1350 & 5.2e4 & 1.18 & Keflavíkurflugvöllur \\
-        1482 & 1.4e4 & 1.23 & Reykjavík Víðidalur \\
-        4921 & 1.3e4 & 1.29 & Rif á Melrakkasléttu \\
-        1477 & 5.6e4 & 1.29 & Reykjavíkurflugvöllur\\ \hdashline[0.5pt/1pt]
-        35553 & 4.0e3 & 4.30 & Almannaskarð - göng\\
-        6745 & 1.5e4 & 4.36 & Kerlingarfjöll - Ásgarðsfjall\\
-        35978 & 7.9e3 & 4.40 & Fáskrúðsfjarðargöng suður\\
-        2640 & 1.6e3 & 4.51 & Seljalandsdalur\\
-        32635 & 3.2e4 & 4.95 & Botn í Súgandafirði\\
-        \bottomrule
-    \end{tabular}
-    }
-\end{table}
-
-If the 0 m/s is excluded from the measurement data, then the error distribution using MAPE can be plotted. This plot can be seen in Figure \ref{fig:station_mape_distribution}
-
-\begin{figure}
-    \includegraphics[scale=0.6]{Figures/MAPEoverIceland.png}
-    \caption[Distribution of MAPE by station]{The distribution of MAPE. Using MAPE instead of MAE, necessitates the exclusion of 0 values from the ground truth (measured values).}
-    \label{fig:station_mape_distribution}
+    \includegraphics[scale = 0.6]{Figures/loss_plot_w_elevation.png}
+    \caption[Loss plot of model trained for 2000 with DEM.]{Loss plot of model trained for 2000 with DEM.}
+    \label{fig:loss_plot_w_elevation}
 \end{figure}
-
-\section{Data Structure}
-
-Once data has been retrieved for all three sources and processed, including interpolating values, it needs to be made ready to use by the model, for both training, validation and test.
-The starting point is a dataframe that contains measured information from AWS.
-This includes the average wind, the wind gust, wind direction along with the station number and coordinates.
-When selecting the CARRA data certain height levels are chosen.
-These present as separate lines in the CARRA dataframe.
-Information for one observation is represented in as many lines as height levels requested in the reanalysis data.
-These rows need to be combined on the position (the weather station).
-When this is done it is possible to combine the AWS IMO data and CARRA reanalysis data on the location and time columns.
-The last data source is the elevation.
-A sector of a circle looking upwind is looked at.
-In any case the points, that represent these sections, were selected as shown in Code Listing \ref{code:sectorElevation}.
-A range of angles are defined based on the wind direction $d$ at some distance from the given point.
-This means that the resultant points (equal in number to the length of angleRange by k) form sectors at several distances from the given weather station.
-
-\begin{lstlisting}[style = Python, caption = {Sector elevation points generated}, label = code:sectorElevation]
-angles = [(angle + (90 - d)) * pi/180 for angle in angleRange]
-length_rng = [(exp(i * log(n + 1)/ k) - 1) * 1000 
-                for i in range(1, k + 1)]
-points = np.array([[(X + l * cos(angle), Y + l * sin(angle))
-                    for angle in angles] for l in length_rng])   
-\end{lstlisting}
-
-The result is a dataframe that has measured data from AWS, which gives us our target, reanalysis data from CARRA, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data.
-An example of what the data looks like can be seen in Table \ref{table:trainDataExample}.
-
-\begin{table}[h]
-    \caption[An example of data structure used to train model]{An example of data structure used to train model.
-Data points include the derived variables Ri and N, the elevation of the station, direction of wind and relative direction of the wind (twd, that is the direction of the wind relative to center of Iceland), along with some combination of wind speed, pressure and temperature at the different height levels.
-Finally there are the elevation points around a given station, where the elevation is relative to the station.}
-    \label{table:trainDataExample}
-    \resizebox{\textwidth}{!}{
-    \centering
-    \begin{tabular}{c|c|c|c|c|c|c|c|c|c}
-         Ri & N\_squared & station\_elevation & twd & ws\_15 & wd\_15 & t\_15 & p\_15 & elevation\_point\_0 & \dots\\\hline
-         -1.18e+00 &  2.67e+04 & 100 & 1.5 & 10 & 5 & 0 & 100 & 2 & \dots 
-    \end{tabular}
-    }
-\end{table}
-
- Looking at Table \ref{table:trainDataExample} note that the first two columns represent two variables that describe the stability of the air.
-These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency\cite{brunt_vaisala_freq_eumtrain} ($N$), and are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}.
-These values are calculated using reanalysis data at two different height levels.
-Thus $Ri$ refers to the Richardson number calculated between height levels 15m and 500m.
-Exactly the same notation is used with the Brunt–Väisälä frequency, except the square is used.
-\begin{equation}
-    \label{eqn:Ri}
-    Ri = \frac{g \cdot dT \cdot dz}{T_{\textrm ave} \cdot dU^2} \unit{[]}
-\end{equation}
-
-\begin{equation}
-    \label{eqn:N}
-    N = \sqrt{\frac{g \cdot dT }{T_{\textrm ave} \cdot dz}} \unit{[Hz]}
-\end{equation}
-
-Here, $g$ is the acceleration due to gravity, $dT$ is the temperature difference between the two height levels, $dz$ is the elevation difference, $T_{\textrm ave}$ is the average temperature (that is the average of the two temperatures in the height levels) and $dU$ is the wind speed difference between the two height levels.
-Both of these numbers provide some insight about the stability of the air.
-A lower value for the Richardson number indicates a higher turbulence.
-A typical range of values could be between 0.1 and 10, with values below 1 indicating significant turbulence\cite{richardson_number_skybrary}.
-When the square of the Brunt-Väisälä frequency is negative, then the air is unstable (an air parcel will move away from its original position)\cite{brunt_vaisala_freq_eumtrain}.
-These are derived factors from the reanalysis data and as such there shouldn't be a significant information gain using $Ri$ and $N$ as opposed to having the raw data.
-However, including these factors instead of every reanalysis variable requested might speed up training as well as making the model more easily explainable with the use of Shapley values or other tools for explainability.
-Using Shapley a feature importance value is attributed to a given feature by creating all possible permutations of any possible length (up to number of features) and seeing how the predictions are skewed when the given parameter is included or excluded.
-This needs to be done for all parameters.
-The time complexity of this is very high ($2^n$ coalitions)\cite{shapley_information}.
-Most implementations use some approximations, which still can take a considerable amount of time for models with a high parameter count and many examples.

### 2025-04-07 Chapters/Chapter4.tex
To be read by KJ

--- 2024-09-26 Chapters/Chapter4.tex
@@
-% Chapter Template
-
-\chapter{Model architecture} % Main chapter title
-
-\label{Chapter4} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}
-
-%----------------------------------------------------------------------------------------
-%	SECTION 1
-%----------------------------------------------------------------------------------------
-
-\section{Model structure}
-The structure of the neural network is such that it contains some number $n$ of fully connected layers and batch normalization for each layer, along with regularization.
-All of these layers have the same number of units.
-The last layer has a dropout of 50\%.
-In addition to these layers there is one more output layer.
-This is simply a dense layer with 1 unit.
-A grid search was performed to determine the hyperparameters that minimize the loss.
-Hyperparameters are parameters that are set before the training begins\cite{hyperparameters_definition}.
-These hyperparameters include number of units in each layer, number of epochs to train for, number of layers, batch size, optimizer and penalty to enforce in the regularization.
-The possible combinations tried can be seen in Table \ref{table:gridSearchHyperparamters}.
+\chapter{Results}
+\label{Chapter4}
+\section{Results}
+A baseline model was constructed.
+This model looked at the gust factor for some training data and took the average of this and predicted this average everytime.
+Several baseline guesses were created based on a lower limit assigned to the average wind speed limit (AWSL).
+As the average wind speed increases, then the variability in gust as a percentage decreases\cite{mean_gust_HA_HO}.
+This means looking at a subset of the data where the AWSL is higher, a better result can be expected.
+The results show this.
+Several different AWSL were used for the baseline model, as can be seen in table \ref{table:results}.
+This sets a goal.
+A model that does not significantly improve on this baseline suggests either failure to capture essential patterns in the data or that the data itself may lack the necessary information for substantial improvements upon the baseline.
+Using the previously described neural network architecture setups for each AWSL, with and without landscape elevation information, MAPE was determined.
+The results can be seen in Table \ref{table:results}.
 
 \begin{table}[h]
+    \caption[Model results for different AWSL]{MAPE for each average wind speed limit with and without landscape elevation in a 30° sector around the point of interest into the direction of the reanalysis wind.
+The influence of adding elevation data seems to reduce the error.
+The percentage error is higher for lower wind speeds and thus observing the error for different lower bound of wind speed will produce different results.
+This lower bound is determined using the reanalysis wind speed at 15 meters ($ws_{15}$).}
+    \label{table:results}
     \centering
-    \caption[Hyperparameter search with best performing combination.]{Hyperparamter search with best performing combination shown.
-Hyperparameter search was done using hyperband algorithm that initially searches randomly for the best parameters but then hones in on what is working and as such is neither exhaustive nor completely random.
-This means that an upper limit will not be set on the number of combinations to try like with randomsearch.}
-    \label{table:gridSearchHyperparamters}
-    \begin{tabular}{ccc}
-        Parameter & Range of values & Selected\\
-        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{10}\\
-        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{64}\\
-        Penalties & min\_value = 1e-5, max\_value = 1, sampling = log & \textbf{1e-4}\\
-        Epochs & min\_value = 10, max\_value = 1000, step = 10 & \textbf{50}\\
-        Optimizers & Adam, RMSprop, Adamax & \textbf{Adamax}\\
-        Activation & ReLU, ELu, Softmax & \textbf{ReLU}\\
+    \begin{tabular}{lccc}
+        \toprule
+        \textbf{AWSL} & \textbf{MAPE} & &\\ 
+        $[m/s]$ & \textit{Baseline} &  \textit{Without DEM} & \textit{With DEM} \\
+        \midrule
+        $\geq 0$ & 39.2\% & 19.2\% & 18.9\% \\
+        $\geq 5$ & 28.1\% & 15.3\% & 14.9\%\\
+        $\geq 10$ & 23.9\% & 13.3\% & 12.5\%\\
+        $\geq 15$ & 23.2\% & 14.4\% & 11.9\%\\
+        $\geq 20$ & 24.7\% & 15.7\% & 13.3\%\\
+        $\geq 25$ & 27.7\% & 19.4\% & 17.3\%\\
+        \bottomrule
     \end{tabular}
 \end{table}
+
+%m/s & $ws_{15}$ & f & $ws_{15}$ & f & $ws_{15}$ & f\\
+%$\geq 0$ & 39.2\% & 39.2\% & 19.7\% & 15.9\% & 18.6\% & 15.3\%\\
+%$\geq 5$ & 28.1\% & 14.8\% & 15.9\% & 9.9\% & 14.1\% & 9.4\%\\
+%$\geq 10$ & 23.9\% & 11.1\%& 13.6\% & 7.6\% & 11.5\% & 7.4\%\\
+%$\geq 15$ & 23.2\% & 9.3\% & 13.0\% & 7.0\% & 10.6\% & 6.4\%\\
+%$\geq 20$ & 24.7\% & 8.2\% & 13.6\% & 6.3\% & 11.1\% & 5.8\%\\
+%$\geq 25$ & 27.7\% & 7.3\% & 15.7\% & 5.8\% & 12.9\% & 5.5\% \\
+%7.5 & 25.3\% & 12.6\% & 14.5\% & 8.6\% & 12.5\% & 8.6\%\\
+%12.5 & 23.3\% & 10.1\% & 13.1\% & 7.1\% & 11.0\% & 6.9\%\\
+%17.5 & 23.7\% & 8.7\% & 12.8\% & 6.4\% & 10.5\% & 6.1\%\\
+%22.5 & 26.1\% & 7.8\% & 14.5\% & 6.1\% & 12.6\% & 5.7\%\\
+
+This is some improvement upon the baseline error, with a decrease in error from 23.9\% to 13.6\% and 10.6\% for the baseline, model without DEM and with model with DEM for 10 m/s cutoff.
+The power generated by wind mill increases with wind speed cubed\cite{wind_power}.
+The highest wind gusts in Iceland are around 70 m/s.
+Knowing the gust factor with half as much error as before can allow better anticipation and thus spare turbines for high wind gusts.
+Another way to look at the error improvement is by station.
+No location data was directly included in the training data.
+In Table \ref{table:station_mae_distribution} the mean absolute error of predicted average wind speed and measured average wind speed can be seen for the extreme values.
+A question to ask is it possible to achieve better results when only looking at a single station?
+
+\begin{table}[h]
+    \caption[Model result by station]{The MAPE results for selected stations of interest, both when training for the specific site and when the stations are a part of the general data.
+For every station the AWSL is set at 10 m/s.
+In training for a single station at a time, some site specific information can be gauged.
+This does not mean that the a better result can be reached for that site.
+Factors such as the number of datapoints at given location can significantly impact the result.
+This table uses the measured wind speed to determine the cut off for data points.
+This leads to some data leakage and an increased performance compared to using the reanalysis CARRA speed for cut off.}
+    \label{table:specific_sites}
+    \centering
+    \begin{tabular}{lccc}
+        \toprule
+        \textbf{Station name} & \textbf{\# points} & \multicolumn{2}{c}{\textbf{MAPE}}\\
+        & & General training & Site training\\
+        \midrule
+        Akrafjall & 42,791 & 18.6\% & 93.7\%\\
+        Almannaskarð & 4,014 & 12.2\% & 86.7\%\\
+        Ásgarðsfjall & 15,121 & 9.1\% & 9.4\%\\
+        Jökulheimar & 17,176 & 7.7\% & 7.7\%\\
+        Sandbúðir & 18,718 & 6.8\% & 6.4\%\\
+        Stórholt & 35,126 & 7.1\% & 29.2\% \\
+        Þúfuver & 19,538 & 6.4\% & 6.8\%\\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+Instead of looking at the exact values of MAPE at select stations, a plot of the error distribution can be created.
+This can be seen in Figure (\ref{fig:errorMap}).
+Looking at Figure (\ref{fig:errorMap}), the worst performing stations can be seen at Vestfirðir and around the coastline.
+Stations further inland seem to have lower error.
+The worst performing station, at Seljalandsdalur, is at Vestfirðir while the best performing station Garðskagaviti is at the South-Western tip of Iceland in Reykjanes.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale = 0.5]{Figures/errorMap.png}
+    \caption[MAPE error distribution of stations shown on a map of Iceland.]{The MAPE error of each station in data shown as color gradient circles.
+That is each station is represented by one circle, with the error value represented as a color gradient from dark blue to yellow.
+The lowest error at a single station was around 7\% at Garðskagaviti and the highest around 35\% at Seljalandsdalur.
+It is important to note that the model is trained using a cutoff of 10 m/s and this cutoff point is determined using the reanalysis wind at 15 meters above ground.}
+    \label{fig:errorMap}
+\end{figure}
+
+\begin{table}[h]
+    \caption[Model result looking at closed wind speed intervals]{The MAPE results for different AWSL intervals.
+Here instead of training for all data above a certain threshold put of the wind, training is done only on data between two wind speeds.
+The percentage variance in gust factor as a function of wind speed increases with decreasing wind speed.
+Measured wind speed is used for the cutoff and thus have data leakage.
+This results should thus be somewhat comparable to the last column in Table (\ref{table:results})}
+    \label{table:closed_intervals}
+    \centering
+    \begin{tabular}{lcc}
+        \toprule
+        \textbf{Interval} &  \multicolumn{2}{c}{\textbf{MAPE}}\\
+        $[m/s]$ & Without Elevation & With Elevation\\
+        \midrule
+        $[5, 10[$ & 17.1\% & 16.4\%\\
+        $[10, 15[$ & 14.5\% & 13.0\%\\
+        $[15, 20[$ & 15.0\% & 12.0\%\\
+        $[20, 25[$ & 15.6 \% & 13.1\%\\
+        $[25, 30[$ & 18.4\% & 19.0\%\\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+Table (\ref{table:closed_intervals}) shows no improvement over the values in the last column in Table (\ref{table:results}).
+As previously mentioned, the gust factor decreases with increasing wind speed and thus, training on intervals and lessening this effect might be expected to give better results.
+This does not seem to be the case.
+Some interesting sites to look closer at for drivers might include places such as Ingólfsfjall, Kjalarnes and others.
+These can be seen in Table (\ref{table:more_specific_sites}).
+
+\begin{table}[h]
+    \caption[Model result by stations of interest]{The MAPE results of different stations for several stations of interest, both when training for the specific site and when the stations are a part of the general data.
+For every station the AWSL is set at 10 m/s.
+In training for a single station at a time, some site specific information can be gauged.
+This does not mean that the a better result can be reached for that site.
+Factors such as the number of datapoints at given location can significantly impact the result.
+This table uses the measured wind speed to determine the cut off for data points.
+This leads to some data leakage and an increased performance compared to using the reanalysis CARRA speed for cut off.}
+    \label{table:more_specific_sites}
+    \centering
+    \begin{tabular}{lcc}
+        \toprule
+        \textbf{Station name} & \multicolumn{2}{c}{\textbf{MAPE}}\\
+         & Baseline & Model\\
+        \midrule
+        Fáskrúðsfjörður & 28.2\% & 21.8\%\\
+        Ingólfsfjall & 30.0\% & 19.6\%\\
+        Kjalarnes & 20.7\% & 13.5\% \\
+        Sandskeið & 13.0\% & 10.2\%\\
+        Seyðisfjörður & 32.1\% & 23.0\%\\
+        Þjórsárdalur & 12.2\% & 11.4\%\\
+        Þrengsli & 13.6\% & 11.3\%\\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+In Figure (\ref{fig:ShapleySummary}) the contribution of each feature, excluding the elevation points, can be seen for the model in general (a significant subset of data is used).
+Looking at Figure (\ref{fig:ShapleySummary}), there is an outlier.
+Exactly calculating the Shapley values is time intensive, in the subset shown there is an outlier that skews the figure and makes it so that viewing the importance distribution excluding the outlier is difficult.
+For this reason, another shapley summary vas created that looked at different, and larger distribution.
+This can be seen in Figure (\ref{fig:ShapleySummary2}).
+The Figures (\ref{fig:ShapleySummary}) and (\ref{fig:ShapleySummary2}) show the summary for a model trained only on the features shown and not on landscape elevation of the surrounding area.
+This is done as the number of points there is too high to display in one figure (70 total points).
+Looking specifically at Figure (\ref{fig:ShapleySummary2}), the station elevation is most influential and the Richardson number's influence is very low.
+Most of the feature values bunch up in the middle, while the station elevation is elongated compared to other features.
+Station elevation seems to have two bunches on either side of 0.
+Overall, the values seem to be skewed to the right of 0.
+This would be expected as the predicted values are expected to be in the range of 1.2-2, or at least always above 1 by definition.
+Finally, for the Shapley values, looking that all the data there are again outliers that skew the data so that spotting general distribution is difficult.
+This can be seen in Figure (\ref{fig:ShapleySummary3}).
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot.png}
+    \caption[Summary feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+We can see that generally multiple factors influence the prediction, with the station elevation being highly influential.
+There is seemingly one outlier for the Richardson number, which usually has very little influence.
+Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them.
+To see their influence on the model output see Table \ref{table:results}.}
+    \label{fig:ShapleySummary}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_.png}
+    \caption[Summary feature importance of a neural network using a larger distribution of data.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+Generally multiple factors influence the prediction, with the station elevation being highly influential.
+Elevation data is excluded when working with Shapley values, as the contribution of each elevation.
+In contrast to Figure (\ref{fig:ShapleySummary}), the distribution doesn't have as extreme outliers.
+This means that more details can be seen in the figure.
+The X-axis shows the influence of feature values on the model.
+The color gradient shows the value of each feature.
+As an example, there is a very red value for station elevation (top line, all the way to the left).
+This means that in this instance, the station elevation contributed around -0.25 to the final output and that the station had an elevation significantly above average.}
+    \label{fig:ShapleySummary2}
+\end{figure}
+
+Further Shapley figures can be seen in Appendix~\ref{appendix:A}.
+In each of these plots, even without the feature labels, the station elevation is easily noticed as the value is constant for a station.
+This is not noteworthy.
+What is noteworthy is the range of impact from this single value.
+For simpler models, this would not happen.
+It is important to note that SHAP assumes feature independece\cite{Salih_2024}.
+This might explain why the impact of the Richardson number is so low.
+Both the squared Brunt–Väisälä frequency and the Richardson number are derived features from reanalysis data.
+They carry with them some extra information over the other features in the dataset.
+This is because both are variables over elevation ranges.
+That is, as seen in Equations (\ref{eqn:Ri}, \ref{eqn:N}), both are dependent on values at lower and upper elevations and try to describe the stability of that range.
+Shapley tries to assign contribution values for each feature for each observation.
+SHAP assumes that the features are independent, but this is not the case.
+It is clearly not the case for the derived variables, but how the contribution should be distributed between the features is not clear.
+Seemingly the SHAP python package is giving all the impact to the Brunt–Väisälä squared frequency and none to Richardson number.
+If the Brunt–Väisälä would be excluded from the data, the impact of the Richardson number would likely increase.
+Another point to note is that the features are ordered by their impact.
+This means that the station elevation is the most impactful for each plot, but the ordering of other variables changes.
+Looking at Figure (\ref{fig:ShapleySummary3}), which shows the Shapley summary plot for all data, the wind speed is the second most important feature.
+This is reversed in Figure (\ref{fig:ShapleySummaryAkrafjall}).
+This is not unexpected as Akrafjall station was specifically selected as the MAE for reanalysis wind speed was very high as can be seen in Table (\ref{fig:ShapleySummaryAkrafjall}), where you will also find the stations whose summary plot is shown in Figures (\ref{fig:ShapleySummaryAlmannaskarð}, \ref{fig:ShapleySummaryAsgarðsfjall}) and these also fall into the category of very high MAE for wind speed.
+What is interesting is that the reanalysis wind speed is also of low impact at stations like Háahlíð and Keflavíkurflugvöllur, as shown in Figures (\ref{fig:ShapleySummaryHaahlid}, \ref{fig:ShapleySummaryKeflavikurflugvollur}).
+These stations had the lowest of MAE for difference between measured wind speed and reanalysis wind speed.
+As the summary plot over all stations (Figure (\ref{fig:ShapleySummary3})) shows that reanalysis wind speed is impactful, might lead to the conclusion that the reanalysis wind speed is not a good predictor at these locations or something else is skewing the data.
+A simpler way to look at feature importance is to create models that are trained on and use to predict different sets of parameters.
+The results of such a comparison can be seen in Table \ref{table:setsOfParams}.
+
+\begin{table}[h]
+    \caption[Model results for different sets of parameters.]{Showing the results of different models.
+Models defined by different sets of parameters, with increasing complexity.
+This table shows the feature importance as the loss decreases as more variables are added.
+Adding all height levels decreases the loss but very little.}
+    \label{table:setsOfParams}
+    \centering
+    \begin{tabular}{lc}
+        \toprule
+        \textbf{Model parameters} & \textbf{MAPE}\\
+        \midrule
+        Baseline constant & 23.9\%\\
+        $ws_{15}$ & 17.2\%\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd]$ & 13.8\% \\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ & 13.7\%\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ + DEM & 12.0\%\\
+        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, ASL, twd, N, Ri]$ & 12.0\% \\
+        \bottomrule
+    \end{tabular}
+\end{table}

### 2025-04-07 Chapters/Chapter5.tex
Combined C2/3 and added data figures.
To be read by KJ

@@
-\chapter{Results}
+\chapter{Discussion}
 \label{Chapter5}
-\section{Results}
-A baseline model was constructed.
-This model looked at the gust factor for some training data and took the average of this and predicted this average everytime.
-Several baseline guesses were created based on a lower limit assigned to the average wind speed limit (AWSL).
-As the average wind speed increases, then the variability in gust as a percentage decreases\cite{mean_gust_HA_HO}.
-This means looking at a subset of the data where the AWSL is higher, a better result can be expected.
-The results show this.
-Several different AWSL were used for the baseline model, as can be seen in table \ref{table:results}.
-This sets a goal.
-A model that does not significantly improve on this baseline suggests either failure to capture essential patterns in the data or that the data itself may lack the necessary information for substantial improvements upon the baseline.
-Using the previously described neural network architecture setups for each AWSL, with and without landscape elevation information, MAPE was determined.
-The results can be seen in Table \ref{table:results}.
-
-\begin{table}[h]
-    \caption[Model results for different AWSL]{MAPE for each average wind speed limit with and without landscape elevation in a 30° sector around the point of interest into the direction of the reanalysis wind.
-The influence of adding elevation data seems to reduce the error.
-The percentage error is higher for lower wind speeds and thus observing the error for different lower bound of wind speed will produce different results.
-This lower bound is determined using the reanalysis wind speed at 15 meters ($ws_{15}$).}
-    \label{table:results}
-    \centering
-    \begin{tabular}{lccc}
-        \toprule
-        \textbf{AWSL} & \textbf{MAPE} & &\\ 
-        $[m/s]$ & \textit{Baseline} &  \textit{Without DEM} & \textit{With DEM} \\
-        \midrule
-        $\geq 0$ & 39.2\% & 19.7\% & 18.6\% \\
-        $\geq 5$ & 28.1\% & 15.9\% & 14.1\%\\
-        $\geq 10$ & 23.9\% & 13.6\% & 11.5\%\\
-        $\geq 15$ & 23.2\% & 13.0\% & 10.6\%\\
-        $\geq 20$ & 24.7\% & 13.6\% & 11.1\%\\
-        $\geq 25$ & 27.7\% & 15.7\% & 12.9\%\\
-        \bottomrule
-    \end{tabular}
-\end{table}
-
-%m/s & $ws_{15}$ & f & $ws_{15}$ & f & $ws_{15}$ & f\\
-%$\geq 0$ & 39.2\% & 39.2\% & 19.7\% & 15.9\% & 18.6\% & 15.3\%\\
-%$\geq 5$ & 28.1\% & 14.8\% & 15.9\% & 9.9\% & 14.1\% & 9.4\%\\
-%$\geq 10$ & 23.9\% & 11.1\%& 13.6\% & 7.6\% & 11.5\% & 7.4\%\\
-%$\geq 15$ & 23.2\% & 9.3\% & 13.0\% & 7.0\% & 10.6\% & 6.4\%\\
-%$\geq 20$ & 24.7\% & 8.2\% & 13.6\% & 6.3\% & 11.1\% & 5.8\%\\
-%$\geq 25$ & 27.7\% & 7.3\% & 15.7\% & 5.8\% & 12.9\% & 5.5\% \\
-%7.5 & 25.3\% & 12.6\% & 14.5\% & 8.6\% & 12.5\% & 8.6\%\\
-%12.5 & 23.3\% & 10.1\% & 13.1\% & 7.1\% & 11.0\% & 6.9\%\\
-%17.5 & 23.7\% & 8.7\% & 12.8\% & 6.4\% & 10.5\% & 6.1\%\\
-%22.5 & 26.1\% & 7.8\% & 14.5\% & 6.1\% & 12.6\% & 5.7\%\\
-
-This is some improvement upon the baseline error, with a decrease in error from 23.9\% to 13.6\% and 10.6\% for the baseline, model without DEM and with model with DEM for 10 m/s cutoff.
-The power generated by wind mill increases with wind speed cubed\cite{wind_power}.
-The highest wind gusts in Iceland are around 70 m/s.
-Knowing the gust factor with half as much error as before can allow better anticipation and thus spare turbines for high wind gusts.
-Another way to look at the error improvement is by station.
-No location data was directly included in the training data.
-In Table \ref{table:station_mae_distribution} the mean absolute error of predicted average wind speed and measured average wind speed can be seen for the extreme values.
-A question to ask is it possible to achieve better results when only looking at a single station?
-
-\begin{table}[h]
-    \caption[Model result by station]{The MAPE results for selected stations of interest, both when training for the specific site and when the stations are a part of the general data.
-For every station the AWSL is set at 10 m/s.
-In training for a single station at a time, some site specific information can be gauged.
-This does not mean that the a better result can be reached for that site.
-Factors such as the number of datapoints at given location can significantly impact the result.
-This table uses the measured wind speed to determine the cut off for data points.
-This leads to some data leakage and an increased performance compared to using the reanalysis CARRA speed for cut off.}
-    \label{table:specific_sites}
-    \centering
-    \begin{tabular}{lccc}
-        \toprule
-        \textbf{Station name} & \textbf{\# points} & \multicolumn{2}{c}{\textbf{MAPE}}\\
-        & & General training & Site training\\
-        \midrule
-        Akrafjall & 42,791 & 18.6\% & 93.7\%\\
-        Almannaskarð & 4,014 & 12.2\% & 86.7\%\\
-        Ásgarðsfjall & 15,121 & 9.1\% & 9.4\%\\
-        Jökulheimar & 17,176 & 7.7\% & 7.7\%\\
-        Sandbúðir & 18,718 & 6.8\% & 6.4\%\\
-        Stórholt & 35,126 & 7.1\% & 29.2\% \\
-        Þúfuver & 19,538 & 6.4\% & 6.8\%\\
-        \bottomrule
-    \end{tabular}
-\end{table}
-
-Instead of looking at the exact values of MAPE at select stations, a plot of the error distribution can be created.
-This can be seen in Figure (\ref{fig:errorMap}).
-Looking at Figure (\ref{fig:errorMap}), the worst performing stations can be seen at Vestfirðir and around the coastline.
-Stations further inland seem to have lower error.
-The worst performing station, at Seljalandsdalur, is at Vestfirðir while the best performing station Garðskagaviti is at the South-Western tip of Iceland in Reykjanes.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale = 0.5]{Figures/errorMap.png}
-    \caption[MAPE error distribution of stations shown on a map of Iceland.]{The MAPE error of each station in data shown as color gradient circles.
-That is each station is represented by one circle, with the error value represented as a color gradient from dark blue to yellow.
-The lowest error at a single station was around 7\% at Garðskagaviti and the highest around 35\% at Seljalandsdalur.
-It is important to note that the model is trained using a cutoff of 10 m/s and this cutoff point is determined using the reanalysis wind at 15 meters above ground.}
-    \label{fig:errorMap}
-\end{figure}
-
-\begin{table}[h]
-    \caption[Model result looking at closed wind speed intervals]{The MAPE results for different AWSL intervals.
-Here instead of training for all data above a certain threshold put of the wind, training is done only on data between two wind speeds.
-The percentage variance in gust factor as a function of wind speed increases with decreasing wind speed.
-Measured wind speed is used for the cutoff and thus have data leakage.
-This results should thus be somewhat comparable to the last column in Table (\ref{table:results})}
-    \label{table:closed_intervals}
-    \centering
-    \begin{tabular}{lcc}
-        \toprule
-        \textbf{Interval} &  \multicolumn{2}{c}{\textbf{MAPE}}\\
-        $[m/s]$ & Without Elevation & With Elevation\\
-        \midrule
-        $[5, 10[$ & 17.2\% & 15.4\%\\ %10.6\%\\
-        $[10, 15[$ & 14.1\% & 11.9\%\\ %7.8\%\\
-        $[15, 20[$ & 13.2\% & 10.9\%\\ %6.4\%\\
-        $[20, 25[$ & 14.9 \% & 11.6\%\\ %6.3\%\\
-        $[25, 30[$ & 17.0\% & 16.0\%\\ %7.1\%\\
-        \bottomrule
-    \end{tabular}
-\end{table}
-
-Table (\ref{table:closed_intervals}) shows no improvement over the values in the last column in Table (\ref{table:results}).
-As previously mentioned, the gust factor decreases with increasing wind speed and thus, training on intervals and lessening this effect might be expected to give better results.
-This does not seem to be the case.
-Some interesting sites to look closer at for drivers might include places such as Ingólfsfjall, Kjalarnes and others.
-These can be seen in Table (\ref{table:more_specific_sites}).
-
-\begin{table}[h]
-    \caption[Model result by stations of interest]{The MAPE results of different stations for several stations of interest, both when training for the specific site and when the stations are a part of the general data.
-For every station the AWSL is set at 10 m/s.
-In training for a single station at a time, some site specific information can be gauged.
-This does not mean that the a better result can be reached for that site.
-Factors such as the number of datapoints at given location can significantly impact the result.
-This table uses the measured wind speed to determine the cut off for data points.
-This leads to some data leakage and an increased performance compared to using the reanalysis CARRA speed for cut off.}
-    \label{table:more_specific_sites}
-    \centering
-    \begin{tabular}{lcc}
-        \toprule
-        \textbf{Station name} & \multicolumn{2}{c}{\textbf{MAPE}}\\
-         & Baseline & Model\\
-        \midrule
-        Fáskrúðsfjörður & 28.2\% & 21.8\%\\
-        Ingólfsfjall & 30.0\% & 19.6\%\\
-        Kjalarnes & 20.7\% & 13.5\% \\
-        Sandskeið & 13.0\% & 10.2\%\\
-        Seyðisfjörður & 32.1\% & 23.0\%\\
-        Þjórsárdalur & 12.2\% & 11.4\%\\
-        Þrengsli & 13.6\% & 11.3\%\\
-        \bottomrule
-    \end{tabular}
-\end{table}
-
-In Figure (\ref{fig:ShapleySummary}) the contribution of each feature, excluding the elevation points, can be seen for the model in general (a significant subset of data is used).
-Looking at Figure (\ref{fig:ShapleySummary}), there is an outlier.
-Exactly calculating the Shapley values is time intensive, in the subset shown there is an outlier that skews the figure and makes it so that viewing the importance distribution excluding the outlier is difficult.
-For this reason, another shapley summary vas created that looked at different, and larger distribution.
-This can be seen in Figure (\ref{fig:ShapleySummary2}).
-The Figures (\ref{fig:ShapleySummary}) and (\ref{fig:ShapleySummary2}) show the summary for a model trained only on the features shown and not on landscape elevation of the surrounding area.
-This is done as the number of points there is too high to display in one figure (70 total points).
-Looking specifically at Figure (\ref{fig:ShapleySummary2}), the station elevation is most influential and the Richardson number's influence is very low.
-Most of the feature values bunch up in the middle, while the station elevation is elongated compared to other features.
-Station elevation seems to have two bunches on either side of 0.
-Overall, the values seem to be skewed to the right of 0.
-This would be expected as the predicted values are expected to be in the range of 1.2-2, or at least always above 1 by definition.
-Finally, for the Shapley values, looking that all the data there are again outliers that skew the data so that spotting general distribution is difficult.
-This can be seen in Figure (\ref{fig:ShapleySummary3}).
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot.png}
-    \caption[Summary feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
-We can see that generally multiple factors influence the prediction, with the station elevation being highly influential.
-There is seemingly one outlier for the Richardson number, which usually has very little influence.
-Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them.
-To see their influence on the model output see Table \ref{table:results}.}
-    \label{fig:ShapleySummary}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_.png}
-    \caption[Summary feature importance of a neural network using a larger distribution of data.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
-Generally multiple factors influence the prediction, with the station elevation being highly influential.
-Elevation data is excluded when working with Shapley values, as the contribution of each elevation.
-In contrast to Figure (\ref{fig:ShapleySummary}), the distribution doesn't have as extreme outliers.
-This means that more details can be seen in the figure.
-The X-axis shows the influence of feature values on the model.
-The color gradient shows the value of each feature.
-As an example, there is a very red value for station elevation (top line, all the way to the left).
-This means that in this instance, the station elevation contributed around -0.25 to the final output and that the station had an elevation significantly above average.}
-    \label{fig:ShapleySummary2}
-\end{figure}
-
-Further Shapley figures can be seen in Appendix~\ref{appendix:A}.
-In each of these plots, even without the feature labels, the station elevation is easily noticed as the value is constant for a station.
-This is not noteworthy.
-What is noteworthy is the range of impact from this single value.
-For simpler models, this would not happen.
-It is important to note that SHAP assumes feature independece\cite{Salih_2024}.
-This might explain why the impact of the Richardson number is so low.
-Both the squared Brunt–Väisälä frequency and the Richardson number are derived features from reanalysis data.
-They carry with them some extra information over the other features in the dataset.
-This is because both are variables over elevation ranges.
-That is, as seen in Equations (\ref{eqn:Ri}, \ref{eqn:N}), both are dependent on values at lower and upper elevations and try to describe the stability of that range.
-Shapley tries to assign contribution values for each feature for each observation.
-SHAP assumes that the features are independent, but this is not the case.
-It is clearly not the case for the derived variables, but how the contribution should be distributed between the features is not clear.
-Seemingly the SHAP python package is giving all the impact to the Brunt–Väisälä squared frequency and none to Richardson number.
-If the Brunt–Väisälä would be excluded from the data, the impact of the Richardson number would likely increase.
-Another point to note is that the features are ordered by their impact.
-This means that the station elevation is the most impactful for each plot, but the ordering of other variables changes.
-Looking at Figure (\ref{fig:ShapleySummary3}), which shows the Shapley summary plot for all data, the wind speed is the second most important feature.
-This is reversed in Figure (\ref{fig:ShapleySummaryAkrafjall}).
-This is not unexpected as Akrafjall station was specifically selected as the MAE for reanalysis wind speed was very high as can be seen in Table (\ref{fig:ShapleySummaryAkrafjall}), where you will also find the stations whose summary plot is shown in Figures (\ref{fig:ShapleySummaryAlmannaskarð}, \ref{fig:ShapleySummaryAsgarðsfjall}) and these also fall into the category of very high MAE for wind speed.
-What is interesting is that the reanalysis wind speed is also of low impact at stations like Háahlíð and Keflavíkurflugvöllur, as shown in Figures (\ref{fig:ShapleySummaryHaahlid}, \ref{fig:ShapleySummaryKeflavikurflugvollur}).
-These stations had the lowest of MAE for difference between measured wind speed and reanalysis wind speed.
-As the summary plot over all stations (Figure (\ref{fig:ShapleySummary3})) shows that reanalysis wind speed is impactful, might lead to the conclusion that the reanalysis wind speed is not a good predictor at these locations or something else is skewing the data.
-A simpler way to look at feature importance is to create models that are trained on and use to predict different sets of parameters.
-The results of such a comparison can be seen in Table \ref{table:setsOfParams}.
-
-\begin{table}[h]
-    \caption[Model results for different sets of parameters.]{Showing the results of different models.
-Models defined by different sets of parameters, with increasing complexity.
-This table shows the feature importance as the loss decreases as more variables are added.
-Adding all height levels decreases the loss but very little.}
-    \label{table:setsOfParams}
-    \centering
-    \begin{tabular}{lc}
-        \toprule
-        \textbf{Model parameters} & \textbf{MAPE}\\
-        \midrule
-        Baseline constant & 23.9\%\\
-        $ws_{15}$ & 17.2\%\\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd]$ & 13.8\% \\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ & 13.7\%\\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ + DEM & 12.0\%\\
-        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, ASL, twd, N, Ri]$ & 12.0\% \\
-        \bottomrule
-    \end{tabular}
-\end{table}
+The goal of this tesis was to research whether it was possible to use reanalysis data to increase the predictability of wind gusts and see what influence including the landscape elevation would have.
+A quick statistical analysis will give a p-value of 0, that is these results are statistically significant.
+The improvement shown in Table (\ref{table:results}) is not chance.
+There is some pattern to be learned in the interval that leads to a higher predictability of wind gusts over baseline.
+There is also a large variability in the predictability based on weather stations.
+A large part of that can be attributed to the differing amount of data points for each station.
+It would be interesting to know how much of the predictability difference can be explained away with more data for these stations that have few data points or if there is some inherent difference.
+If there is inherent less predictability at those stations, what could be contributing factors in that?

### 2025-04-25 Chapters/Chapter1.tex
Updated after comments from OPP

@@
 \begin{enumerate}[label = Rule \arabic*:]
-    \item Large Parameter Count. The number of parameters can vary wildly but a general range might be from tens of millions to billions of parameters
-    \item Large Number of Predictands: predicting on different levels (such as pressure levels or height levels) and offering detailed information on the atmospheric vertical structure and surface conditions
-    \item Scalability and downstream applicability.
-This might crystallize in predicting cyclones.
+    \item \textbf{Large Parameter Count:} The number of parameters can vary wildly but a general range might be from tens of millions to billions of parameters
+    \item \textbf{Large Number of Predictands:} predicting on different levels (such as pressure levels or height levels) and offering detailed information on the atmospheric vertical structure and surface conditions
+    \item \textbf{Scalability and downstream applicability:} This might crystallize in predicting cyclones.
 Often, the teams responsible for creating these models try to show their applicability to predict cyclones when not trained specifically on cyclone data (e.g.
@@
 In 2023, Remi Lam and the GraphCast team at Google introduced GraphCast.
-This model was able to outperform the industry standard High Resolution Forecast (HRES) produced by the European Centre for Medium-Range Weather Forecasts (ECMWF).
+This model was able to outperform the industry standard High RESolution Forecast (HRES) produced by the European Centre for Medium-Range Weather Forecasts (ECMWF).
 This model as the name suggests leverages graphing connections rather than traditional grid like data structure.
@@
 \section{Methodology and related work}
-This study looks at data from three sources and an attempt is made to predict the gust factor in a given place in Iceland.
-It uses reanalysis data, along with elevation data to predict the gust factor.
-It looks at the data at the point of interest.
-It does not look at the data as a time series.
+In this study data looks at data from three sources and an attempt is made to predict the gust factor in a given place in Iceland.
+Reanalysis data, along with elevation data is used to predict the gust factor.
+This study looks at data at the point of interest but does not look at the data as a time series.
 This thesis aims to improve on the baseline model of always predicting the mean gust factor and show that some structure can be learned from reanalysis data about gusts.
@@
 Their main results were that the gust factor is inversely correlated to the distance from a mountain and correlated to the height of the mountain.
-The study in 2004 looked at the effect of a dominant point upwind.
+Ágústsson and Ólafsson (2004) looked at the effect of a dominant point upwind.
 It did not look at the effects of the landscape more broadly.
@@
     \label{eqn:relu}
-    r(x):=max(0, x)
+    r(x):=\mathrm{max(0, x)}
 \end{equation}
@@
     f(x) := x, x > 0\\
-    f(x) := \alpha (e^x-1), x\leq 0, \alpha > 0
+    f(x) := \mathrm{\alpha} (e^x-1), x\leq 0, \alpha > 0
 \end{align}
@@
     \label{eqn:tanh}
-    tanh(x):=\frac{e^x-e^{-x}}{e^x+e^{-x}}
+    \mathrm{tanh(x)}:=\frac{e^x-e^{-x}}{e^x+e^{-x}}
 \end{equation}
@@
     \label{eqn:mape}
-    \text{MAPE} = \frac{1}{n}\Sigma_{i=1}^n\frac{|y_{predict} - y_{true}|}{y_{predict}}
+    \text{MAPE} = \frac{1}{n}\Sigma_{i=1}^n\frac{|y_{\mathrm{predict}} - y_{\mathrm{true}}|}{y_{\mathrm{predict}}}
 \end{equation}
-This was chosen because the target is the gust factor (the wind gust over the average wind).
+This measure was chosen because the target is the gust factor (the wind gust over the average wind).
 If the target would have been the wind gust rather than the gust factor then something like mean absolute error might be more appropriate.
@@
 Shapley values are calculated as the average marginal contribution of a feature value across all possible coalitions.
+The contribution for a single feature $j$ to a prediciton $\hat{f}(X)$ for Shapley values is given by Equation (\ref{eqn:shapley}).
+In Equation (\ref{eqn:shapley}), $x_j$ is the feature value, with $\beta_j$ as the weight of that feature and $\beta_j E\left[X_j\right]$ is the mean effect estimate for feature j.
+
+\begin{equation}
+    \label{eqn:shapley}
+    \phi_j(\hat{f}) = \beta_jx_j-\beta_jE \left[ X_j \right]
+\end{equation}
+
 For any combination of parameters what is the contribution of a given parameter.

### 2025-04-25 Chapters/Chapter2.tex
Updated after comments from OPP

@@
 More have been added in the following decades.
-This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.
+In this thesis the data is not considered as a time series, instead the predictions are made using only data at a given time.
 
@@
 The CARRA dataset goes back to September 1991 and is currently updated monthly, with a latency of 2-3 months\cite{carra_information}.
-The oldest IMO data point that fulfills given criteria is from \startDateVedur.
+The oldest IMO data point that fulfills given criteria of consistent available datapoints is from \startDateVedur.
 This is covered by CARRA.
 The CARRA dataset is available for two regions, West and East.
+These regions cover a large part of the North Atlantic, including Greenland and Scandinavia.
 Each of these covers a vastly larger area than the area of interest.
@@
 To get the data one has two options.
-Their web interface or using their API client.
+CARRA web interface or using the API client provided by CARRA.
 Using the API client is the only realistic option here, as there were thousands of requests made for different times.
@@
 For each point 4 parameters were requested, wind speed, wind direction, pressure and temperature.
-Each of these features needed to be interpolated to create data for model to be trained on.
+Each of these features needed to be interpolated to create data for model to be trained on, as the CARRA grid points do not coincide with weather stations.
 
@@
     \caption[An example of data structure used to train model]{An example of data structure used to train model.
-Data points include the derived variables Ri and N, the elevation of the station, direction of wind and relative direction of the wind (twd, that is the direction of the wind relative to center of Iceland), along with some combination of wind speed, pressure and temperature at the different height levels.
+Data points include the derived variables Richardson number ($Ri$) and Brunt-Väisäla frequency ($N$) (defined below), the elevation of the station, direction of wind and relative direction of the wind (twd, that is the direction of the wind relative to center of Iceland), along with some combination of wind speed, pressure and temperature at the different height levels.
 Finally there are the elevation points around a given station, where the elevation is relative to the station.}
@@
 Looking at Table \ref{table:trainDataExample} note that the first two columns represent two variables that describe the stability of the air.
-These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency\cite{brunt_vaisala_freq_eumtrain} ($N$), and are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}.
+These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency ($N$)\cite{brunt_vaisala_freq_eumtrain}, and are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}.
 These values are calculated using reanalysis data at two different height levels.

### 2025-04-25 Chapters/Chapter3.tex
Updated after comments from OPP

@@
         Optimizers & Adam, RMSprop, Adamax & \textbf{Adamax}\\
-        Activation & ReLU, ELu, tanh & \textbf{ReLU}\\
+        Activation & ReLU, ELU, tanh & \textbf{ReLU}\\
         \bottomrule
@@
 To determine the required number of epochs a high number of epochs can be selected and the loss plotted.
-These plots can be seen in Figures (\ref{fig:loss_plot_wo_elevation}, \ref{fig:loss_plot_w_elevation}).
+These plots can be seen in Figures \ref{fig:loss_plot_wo_elevation} and \ref{fig:loss_plot_w_elevation}.
 Looking at these two plots, training loss decreases for over 1900 epochs in both plots, while the validation loss stops decreasing at around 200 epochs for DEM model and 1800 epochs for model without DEM.

### 2025-04-25 MSc.tex
Combined C2/3 and added data figures.
To be read by KJ
Updated after comments from OPP

@@
 \usepackage{amssymb}
-\usepackage[sf,normalsize]{subfigure}
+\usepackage{subcaption}
+%\usepackage[sf,normalsize]{subfigure}
 \usepackage[format=plain,labelformat=simple,labelsep=colon]{caption}
@@
 
-\def\thesisyear{2024}       					% Year thesis submitted
-\def\thesismonth{September}					% Month thesis submitted, e.g. "December"
+\def\thesisyear{2025}       					% Year thesis submitted
+\def\thesismonth{June}					% Month thesis submitted, e.g. "December"
 \def\thesisauthor{Brynjar Geir Sigurðsson}				% Thesis authoreiningaraðferðinni
@@
 \def\thesiskind{M.Sc.}					% E.g. M.Sc. or Ph.D. thesis
-\def\thesiskindformal{CCMagister Scientiarum}			% E.g. Magister Scientiarum
+\def\thesiskindformal{Magister Scientiarum}			% E.g. Magister Scientiarum
 \def\thesisschool{School of Engineering and {Natural Sciences}}		% School
@@
 A deep neural network was constructed using some set of variables and the results are compared to a baseline as well as each other so as to show the impact each feature has.
+Neural networks can help make wind gust predictions in complex ladscape.
 \vfill \vspace*{1cm}
@@
 \section*{Útdráttur}
-Í mörgum iðnuðum, hvort sem það er vindmyllubúgarður eða flutningar er getan til þess að spá fyrir um álag vegna veðurs mikilvæg til að koma í veg fyrir tjóns á vélum og byggingum ásamt því að koma í veg fyrir skaða á fólki.
-Einnig er hægt að auka framleiðni.
-Nýtni vindtúrbínna í hávinda aðstæðum er mikilvæg áskorun í bestun endurnýjanlega orkukerfisins.
+Hvort sem um er að ræða vindmyllubúgarða eða flutningar er spágetan um álag vegna vinds mikilvæg til að koma í veg fyrir tjón á iðnviðum auk þess er hægt að auka framleiðni.
 Hviðuspár geta verið stór þáttur í tjónaminnkun.
@@
 Þetta líkan er svo þjálfað og niðurstöður bornar saman við grunnlíkan sem og önnur líkön með öðrum breytum til að skoða áhrif breyta.
+Djúp tauganet geta búið til hjálplegar spár fyrir vindhviður í flóknu landslagi.
 \vfill
@@
 \include{Chapters/Chapter5}
-\include{Chapters/Chapter6}
+
+\printbibliography
+\appendix
 \include{Appendicies/AppendixA}

### 2025-04-30 Chapters/Chapter1.tex
komið að grein 2.5

--- 2024-09-26 Chapters/Chapter1.tex
@@
 %----------------------------------------------------------------------------------------
-Wind gusts are brief increase in wind speed (lasting seconds) as compared to mean wind speed.
-The gust factor is defined as the peak gust divided by the mean wind speed over some defined time period.
-The peak wind gust is often defined as the highest 3 second rolling average measured wind speed over a period of 10 minutes, while the mean wind is the average of all measurements in the 10 minute intervalf.
-This thesis uses this definition.
-This varies, with the US using a 1 minute interval, leading to 14\% higher results \cite{why_wind_gusts}.
-The Navier Stokes Equation (\ref{eqn:navierstokes}) shows that the change of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth (the Coriolis force), and frictional force.\cite{uncertainties_in_numerical_weather_predictions}
+Wind gusts are brief increases in wind speed (lasting seconds) compared to the mean wind speed over intervals of one or several minutes.
+The gust factor is defined as the peak gust divided by the mean wind speed over a specified time period.
+The peak wind gust is often defined as the highest 3-second rolling-average wind speed measured over a 10-minute period, while the mean wind speed is the average of all measurements in the same interval.
+This thesis uses that definition.
+However, definitions vary: for example, the US uses a 1-minute interval, leading to approximately 14\% higher values \cite{why_wind_gusts}.
+
+The Navier–Stokes equation (\ref{eqn:navierstokes}) shows that changes in wind, both in time and space, depend on the pressure gradient, the oscillating force of the Earth (the Coriolis force), and frictional forces \cite{uncertainties_in_numerical_weather_predictions}.
 \begin{equation}
     \label{eqn:navierstokes}
-    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{pressure} -\overbrace{ f\mathbf{k}x\mathbf{V}}^{oscillation} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{resistance}
+    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{\text{pressure}} -\overbrace{ f\mathbf{k}\times\mathbf{V}}^{\text{Coriolis}} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{\text{resistance}}
 \end{equation}
 
-Traditionally, numerical weather prediction (NWP) systems are used to forecast and analyze weather patterns\cite{medium_range_3d_weather_forecasting_NN}.
-These models describe the transition between discretized packages of atmospheric states using partial differential equations based on physical reality.
-These results are usually published every hour, or at courser time intervals for climate simulations.
-With increasing computer power and efficiency the trend is to output data more often\cite{GNP_vidtal}.They describe the state over the period and so do not necessarily grasp fluctuations well.
-These fluctuations would include fluctuations in the wind speed, wind gusts\cite{canNNBeatNWP}.
+Traditionally, numerical weather prediction (NWP) systems are used to forecast and analyze weather patterns \cite{medium_range_3d_weather_forecasting_NN}.
+These models describe the evolution of discretized atmospheric states using partial differential equations grounded in physics.
+Forecasts are typically produced every hour, or at coarser intervals for climate simulations.
+With increasing computational power and efficiency, the trend is to output data more frequently \cite{GNP_vidtal}.
+However, such outputs summarize conditions over each period and may not capture short-term fluctuations well, including variations in wind speed and gusts \cite{canNNBeatNWP}.
 
-This thesis looks at how best to predict gust factor based on various factors, using several different data sources, including NWP and observations.
-Being able to accurately predict the wind gust is important as it is often the peak wind gusts that will cause failures in structures.
-A problem that will become increasingly prevalent in the near future\cite{nasa_extreme_weather}.
+This thesis examines methods to predict the gust factor using various predictors and multiple data sources, including NWP outputs and observational data.
+Accurate gust predictions are crucial, as it is ofteh the peak wind gust that will cause damaging incidents, such as structural failures, traffic accidents and downed power lines.
+Moreover, the prevalence of extreme wind events is expected to increase in the future \cite{nasa_extreme_weather}.
 
 \section{Background}
-The history of numerical weather predictions goes all the way back to the 1920's when Lewis Fry Richardson pioneered the field and tried to produce forecasts.
-The results were flawed due to noise in the calculations.
-ENIAC was built in 1945, it was a general purpose computer that was used, among other things, to make predictions.
-These predictions took 24 hours to make and were predicting 24 hours into the future.
-It was a proof of concept but not usable\cite{TheENIACForecastsARecreation}.
-In the 1950's, with the advent of computers the first operational forecasts emerged.
-In September of 1954, Rossby and his Stockholm based team produced the first real-time barotropic forecasts.
-The next year the Joint Numerical Weather Prediction Unit (JNWPU), based in Princeton New Jersey, released their first forecasts.
-These forecasts were for 36 hours at 400, 700 and 900 mb.
-The results were inferior to subjective human-based forecasts but showed that such forecasts were feasible and promoted further development in the area \cite{historyNWP}.
-The field of NWP has taken great strides since then following the development of computer power and efficiency.
 
-In the last decade there has been another transformation in the field of weather prediction driven by artificial intelligence.
-Interest in AI has come in waves.
-Some progress is made, then interest dwindles.
-Interest in AI has been increasing steadily since 2010.
-Notable work that has driven this wave of interest include increase in computational abilities due to parallel processing in graphical processing units (GPU), convolutional neural networks (CNN), which allowed much faster processing of massive (image) datasets and the availability of large datasets online.
-It is to be noted that images are grid data with some number of channels.
-Using CNNs could work on any gridded data where there are some spatial features\cite{canNNBeatNWP}.
-Since 2018, there has been significant work done in the weather prediction field using AI.
-In 2018, Dueben and Bauer showed that you can build a NN that can outperform a simple persistence forecast and is competitive with very coarse-resolution atmosphere models of similar complexity for short lead times\cite{dueben2018}.
-Also in 2018, Scher created a deep convolutional neural network (CNN) to emulate a general circulation model (GCM, a numerical model representing the physical processes), training on the GCM which allows it to emulate the dynamics of the model and maintain stability for much longer than Dueben\cite{scher2018}.
-These two papers were more proof of concept rather than production ready models to replace NWP.
-They showed that models based on deep learning might, with further development, compete with standard models in the field.
+\subsection*{Numerical weather prediction}
+The history of numerical weather prediction dates back to the 1920s, when Lewis Fry Richardson pioneered the field and attempted to produce forecasts.
+His results were flawed due to numerical noise.
+The ENIAC, built in 1945, was a general-purpose computer used—among other tasks—for weather prediction.
+These forecasts took 24 hours to compute and predicted 24 hours into the future.
+While a proof of concept, they were not operationally useful \cite{TheENIACForecastsARecreation}.
 
-In the last two years there have been even more developments with the emergence of Large AI Weather forecast Models (LWM).
-In 2024, Ling et al.\cite{SecondRevolution} tried to standardize the definition of LWM in meteorology and came up with 3 rules that need to be met to count as LWM.
+With the advent of electronic computers in the 1950s, the first operational forecasts emerged.
+In September 1954, Carl-Gustaf Rossby and his Stockholm-based team produced the first real-time barotropic forecasts.
+The following year, the Joint Numerical Weather Prediction Unit (JNWPU), based in Princeton, New Jersey, released its first 36-hour forecasts at 400, 700, and 900 mb.
+Although these forecasts were inferior to subjective human analyses, they demonstrated feasibility and spurred further development in the field \cite{historyNWP}.
+Since then, NWP has made tremendous strides in parallel with increases in computational power and efficiency.
 
-\begin{enumerate}[label = Rule \arabic*:]
-    \item Large Parameter Count. The number of parameters can vary wildly but a general range might be from tens of millions to billions of parameters
-    \item Large Number of Predictands: predicting on different levels (such as pressure levels or height levels) and offering detailed information on the atmospheric vertical structure and surface conditions
-    \item Scalability and downstream applicability.
-This might crystallize in predicting cyclones.
-Often, the teams responsible for creating these models try to show their applicability to predict cyclones when not trained specifically on cyclone data (e.g.
-\href{https://www.youtube.com/watch?v=PD1v5PCJs_o&ab_channel=GregBronevetsky}{GraphCast})\cite{SecondRevolution}.
-This is done to show the versatility of the models.
+\subsection*{AI in weather forecasts}
+In the last decade, there has been another transformation in weather prediction driven by artificial intelligence (AI).
+Interest in AI has come in waves: progress is made, then interest diminishes, but over the past 15 years growth has been steady.
+Notable drivers of this wave include advances in computational power (notably parallel processing on graphics processing units, GPUs), the developement of neural networks (NNs) for processing massive datasets, and the availability of large online datasets.
+One type of NN is the convolutional neural network (CNN), originally designed for image processing but applicable to any gridded data with spatial structure \cite{canNNBeatNWP}.
+Since 2018, significant work has applied AI to weather prediction.
+
+There are two common approaches to AI-based forecasting: combining NWP outputs with NNs, or using NNs alone.
+In the former, NWP forecasts can be used as inputs to NN training or integrated in other ways; in the latter, NNs are trained directly on meteorological observations, bypassing NWP entirely.
+
+In 2018, Düben and Bauer showed that an NN could outperform a simple persistence forecast and compete with very coarse-resolution atmospheric models of similar complexity for short lead times \cite{dueben2018}.
+Also in 2018, Scher developed a deep CNN to emulate a general circulation model (GCM)—a numerical model representing physical processes—by training on GCM output.
+This approach allowed stable emulation of model dynamics for much longer horizons \cite{scher2018}.
+These papers were proofs of concept rather than production-ready replacements for NWP, but they demonstrated that deep-learning–based models could, with further development, compete with standard models in the field.
+
+\subsection*{Large weather models}
+In the last two years, there have been even more developments with the emergence of Large AI Weather forecast Models (LWMs).
+In 2024, Ling et al.\ \cite{SecondRevolution} proposed a standardized definition of LWMs in meteorology, outlining three criteria, referred to as the "Three Large Rules":
+
+\begin{enumerate}[label=\textbf{\arabic*}.,rightmargin=1.5em]
+  \item \textbf{Large parameter count:} Typically ranging from tens of millions to billions of parameters.
+  \item \textbf{Multiple predictands:} Forecasting at various levels (e.g., pressure or height levels) to provide detailed atmospheric vertical structure and surface conditions.
+  \item \textbf{Scalability and downstream applicability:} Demonstrated, for example, by predicting cyclones even when not explicitly trained on cyclone data (e.g., \href{https://www.youtube.com/watch?v=PD1v5PCJs_o&ab_channel=GregBronevetsky}{GraphCast}) to showcase model versatility \cite{SecondRevolution}.
 \end{enumerate}
 
-Before 2022, LWM had been shown to be able to compete with traditional NWP for some specifc cases as well as making predictions quicker, after training.
-No model had shown that it could in any way completely replace the traditional systems.
-In early 2022, Pathak et al.\cite{FourCastNet} presented FourCastNet.
-FourCastNet uses an Adaptive Fourier Neural Operator model that leverages transfomer architecture rather than the popular convolutional model architechture.
-FourCastNet matches the performance of standard forecasting techniques at short lead times for large-scale variables and outperforms for smaller variables.
-It generates a week-long forecast in less than 2 seconds, orders of magnitude faster than standard physical methods\cite{FourCastNet}.
-In 2022, machine learning methods were presented that made predictions much faster than traditional NWP, after a one time training (or at least training that wouldn't have to be redone often).
-These were in some cases performing better than NWP.
-In 2023, Remi Lam and the GraphCast team at Google introduced GraphCast.
-This model was able to outperform the industry standard High Resolution Forecast (HRES) produced by the European Centre for Medium-Range Weather Forecasts (ECMWF).
-This model as the name suggests leverages graphing connections rather than traditional grid like data structure.
-The base data is given in latitude and longitude degrees at a resolution of 0.25 degrees.
-This means points are closer to each other at the poles.
-Using the graphing structure is supposed to help with bias incurred as a result of this\cite{GraphCast}.
+Before 2022, LWMs had been shown to compete with traditional NWP in specific cases and to generate forecasts much faster after training, No model had been shown to be able to completely replace tradional NWP systems.
+In early 2022, Pathak et al.\ \cite{FourCastNet} introduced FourCastNet, which employs an Adaptive Fourier Neural Operator model leveraging transformer architecture instead of convolution.
+FourCastNet matches the performance of standard forecasting techniques at short lead times for large-scale variables and outperforms them for smaller-scale features.
+It generates a one-week forecast in under two seconds—orders of magnitude faster than conventional physical methods \cite{FourCastNet}.
+In 2022, several ML-based models demonstrated faster-than-NWP predictions after one-time training, with performance rivaling or exceeding NWP in some cases.
 
-There has been a lot of progress made over the last 6 years (since 2018) and especially in the last 2 years (since 2022)\cite{SecondRevolution}.
-The progression from machine learning methods being an interesting idea in the field of numerical weather predictions, to outperforming the standard NWP has been remarkably quick.
-Two years ago, machine learning methods were able to predict quickly and in some niche cases outperform traditional models.
-They were not generally competitive with standard weather models.
-Now they are competitive.
-It is worth noting that the training of these large models is based on data from traditional large weather models.
-It will be very interesting to watch what the next few years will have in store for the development of machine learning in weather predictions.
+In 2023, Remi Lam and the GraphCast team at Google released GraphCast, which outperformed the European Centre for Medium-Range Weather Forecasts’ (ECMWF) industry-standard High-Resolution Forecast (HRES).
+GraphCast uses a graph-based representation instead of a regular grid, operating on a global latitude–longitude grid at 0.25° resolution, which introduces nonuniform point spacing near the poles; the graph structure helps mitigate this bias \cite{GraphCast}.
+
+\subsection*{Conclusion}
+Substantial progress has been made since 2018 and especially since 2022 \cite{SecondRevolution}.
+The evolution from proof-of-concept ML methods to models competitive with—or surpassing—standard NWP has been remarkably rapid.
+It is important to note that training data for these large models are drawn from traditional NWP outputs, underscoring the synergy between ML and physics-based approaches.
+The coming years promise further exciting developments in machine-learning–based weather prediction.
 
 \section{Methodology and related work}
-This study looks at data from three sources and an attempt is made to predict the gust factor in a given place in Iceland.
-It uses reanalysis data, along with elevation data to predict the gust factor.
-It looks at the data at the point of interest.
-It does not look at the data as a time series.
-This thesis aims to improve on the baseline model of always predicting the mean gust factor and show that some structure can be learned from reanalysis data about gusts.
-To do this a neural network was created.
-Any significant improvement on a base model, that always guesses the mean gust factor, would indicate that the final model has something to contribute.
+In this study, data from three sources is used:
 
-In 2004, H.
-Ágústsson and H.
-Ólafsson\cite{mean_gust_HA_HO} looked at the variability of gust factor in complex landscapes.
-They looked at data from automatic weather stations that measure wind at 10 meters above ground.
-The data that was studied in 2004 comes from the same source as used in this thesis, but limits itself to a smaller section.
-They only looked at the years 1999-2001.
-They looked at three factors and how these three parameters effected the gust factor.
-These were $d_m, D, H$, that is direction of wind blowing off a mountain, distance to the mountain and the height of the mountain above the weather station.
-Their main results were that the gust factor is inversely correlated to the distance from a mountain and correlated to the height of the mountain.
-The study in 2004 looked at the effect of a dominant point upwind.
-It did not look at the effects of the landscape more broadly.
-In this study, landscape upwind is looked at.
+\begin{enumerate}[label=\textbf{\arabic*}.,rightmargin=1.5em]
+  \item Three-hourly reanalysis data for Iceland (2004–2023)
+  \item Dense (20 by 20~m) gridded elevation data for Iceland
+  \item Three-hourly observations from Icelandic meteorological stations
+\end{enumerate}
 
-\subsection{Model architechture}
-To be able to capture the patterns in the data a neural network was constructed.
-A NN architechture was chosen as they are known to be able to capture patterns well in complex data and handle high parameter counts.
-This comes in handy when training on different types of data.
-It is also easy to construct different types of neural networks and see how they fit well with parts of the dataset.
-To measure the performance of these models, both to train and test, mean absolute percentage error (MAPE) as defined in Equation (\ref{eqn:mape}) was used.
+The aim of the study is to improve prediction of gust factors that can be applied at any given place in Iceland.
+Currently, gust factor predictions can be obtained directly from NWP output, but one may also create a simple regression model (involving, e.g., wind speed and location) to predict them.
+To accomplish this, an NN is used as a backend to the NWP.
+The network is trained with gust factors from observations as ground truth, and model quality is measured by comparing predictions with either the NWP or the regression values.
+Note that only data at the prediction time is used; the time-series aspect of the data is not utilized.
+
+In 2004, Hálfdán Ágústsson and Haraldur Ólafsson \cite{mean_gust_HA_HO} investigated gust factor variability in complex landscapes.
+They used data from automatic weather stations measuring wind at 10~m above ground during 1999–2001.
+They examined how three parameters affected the gust factor: \(d_m\), \(D\), and \(H\)—the wind direction relative to a mountain, the distance to the mountain, and the height of the mountain above the station.
+Their main results showed that the gust factor is inversely correlated with distance and directly correlated with mountain height.
+Ágústsson and Ólafsson considered the effect of a dominant point upwind but did not consider the broader landscape.
+
+\subsection{Neural networks}
+To capture patterns in the data, a neural network was constructed.
+An NN architecture was chosen because NNs handle complex data well and can manage large numbers of parameters.
+This flexibility is useful when training on different types of data and allows testing various network designs.
+An NN uses many matrix calculations to weight input parameters and predict an output.
+A deep neural network (DNN) has an input layer, an output layer, and one or more hidden layers.
+Hidden layers transform data from the input layer until it reaches the output layer.
+The input layer width equals the number of input parameters, and the output layer width equals the number of predictands.
+Models were created with different parameter sets to gauge each parameter’s influence.
+Each model produces a single numerical output.
+Every layer also has an activation function; common choices include Rectified Linear Unit (ReLU), Exponential Linear Unit (ELU), and hyperbolic tangent (tanh), defined in Equations (\ref{eqn:relu}), (\ref{eqn:elu}), and (\ref{eqn:tanh}).
+
+\begin{equation}
+    \label{eqn:relu}
+    r(x):=\max(0, x)
+\end{equation}
+
+\begin{align}
+    \label{eqn:elu}
+    f(x) := x,\ x > 0\\
+    f(x) := \alpha (e^x-1),\ x\leq 0,\ \alpha > 0
+\end{align}
+
+\begin{equation}
+    \label{eqn:tanh}
+    \operatorname{tanh}(x):=\frac{e^x-e^{-x}}{e^x+e^{-x}}
+\end{equation}
+
+These activation control neural activation and can help stabilize the network.
+\subsection{Model evaluation}
+To measure the performance of these models, both in training and testing, mean absolute percentage error (MAPE) as defined in Equation (\ref{eqn:mape}) was used.
 \begin{equation}
     \label{eqn:mape}
-    \text{MAPE} = \frac{1}{n}\Sigma_{i=1}^n\frac{|y_{predict} - y_{true}|}{y_{predict}}
+    \text{MAPE} = \frac{1}{n}\sum_{i=1}^n\frac{|y_{\mathrm{predict}} - y_{\mathrm{true}}|}{y_{\mathrm{predict}}}
 \end{equation}
-This was chosen because the target is the gust factor (the wind gust over the average wind).
-If the target would have been the wind gust rather than the gust factor then something like mean absolute error might be more appropriate.
+This measure was chosen because the target is the gust factor (the wind gust over the average wind).
+If the target had been the wind gust rather than the gust factor, mean absolute error might be more appropriate.
 
 \subsection{Model explainability}
-Neural networks are often considered as mysterious black boxes\cite{nn_black_box}.
-In an attempt to understand the model predictions, methods designed for explainability are used.
-One such method is Shapley values\cite{shapley_information}.
+Neural networks are often considered mysterious black boxes \cite{nn_black_box}.
+To understand model predictions, explainability methods are used.
+One such method is Shapley values \cite{shapley_information}.
 Shapley values are calculated as the average marginal contribution of a feature value across all possible coalitions.
-For any combination of parameters what is the contribution of a given parameter.
-This means that Shapley values can explain individual predictions.
-Other machine learning tools, like ELI5 (Explain like I am 5), randomly shuffle a feature and look at the effect on model performance\cite{eli5_information}.
+The contribution for a single feature \(j\) to a prediction \(\hat{f}(X)\) is given by Equation (\ref{eqn:shapley}).
+In Equation (\ref{eqn:shapley}), \(x_j\) is the feature value, \(\beta_j\) its weight, and \(\beta_j E[X_j]\) the mean effect estimate for feature \(j\).
+\begin{equation}
+    \label{eqn:shapley}
+    \phi_j(\hat{f}) = \beta_j x_j - \beta_j E[X_j]
+\end{equation}
+For any combination of parameters, Shapley values explain the individual prediction by attributing a contribution to each feature.
+Other methods, like ELI5 (Explain Like I’m 5), randomly shuffle a feature and measure the effect on model performance \cite{eli5_information}.

### 2025-04-30 Chapters/Chapter2.tex
komið að grein 2.5

@@
 \label{Chapter2}
-Data was sourced from several streams.
-The Icelandic Meteorological Office (IMO) provided measurements from weather stations all around Iceland.
-NWP data was downloaded from Copernicus Arctic Regional Reanalysis dataset (CARRA).
-A land elevation model was also provided by IMO.
+Data were sourced from several streams.
+The Icelandic Meteorological Office (IMO) provided measurements from weather stations across Iceland, NWP data were downloaded from the Copernicus Arctic Regional Reanalysis dataset (CARRA), and finally a land-elevation model was also provided by the IMO.
 
@@
 
-IMO provided 10 minute measurements from \nStationsMin weather stations all around Iceland.
-The measurements that met the filtering criteria, started in \startDateVedur and ended in 2023.
-Of these \nStationsMin stations, \nVedurMin were from IMO and placed at 10 meters above ground, while the rest (\nVGMin) were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)} and placed at 6-7 meters above ground\cite{vegagerdin_postur}.
-The location of these weather stations can be seen in Figure \ref{fig:aws_map}.
-The information that is provided by these Automatic Weather Stations (AWS) is presented in two different types of data files, hourly and 10 minute files.
-The hourly files are summations of the 10 minute files, with the exception that errors, such as nails, still in the 10 minute files should have been removed from the hourly documents.
-Nails, are sharp increases from the rest of the data and are unrealistic outliers that are considered measurement errors and are discarded.
-Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another data set), the average wind speed ($f$), the wind gust ($f_g$), the standard deviation of the wind gust, the direction of the wind ($d$) and the standard deviation for the wind direction.
-These measurement started at the end of the 20th century, when the first AWS stations was installed.
-More have been added in the following decades.
-In this thesis the data is not considered as a time series, instead the predictions are made using only data at a given time.
+IMO provided measurements from \nStationsMin weather stations across Iceland.
+The measurements that met the filtering criteria began in \startDateVedur and ended in 2023.
+Of these \nStationsMin stations, \nVedurMin were from the IMO, with the anemometer at 10~m above ground, while the remaining \nVGMin stations were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)}, with the anemometer at 6–7~m above ground \cite{vegagerdin_postur}.
+The locations of these weather stations are shown in Figure \ref{fig:aws_map}.
+
+Data from these automatic weather stations (AWS) are stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases.
+Each record contains the following information: date and time; station number (convertible to coordinates using another dataset of Icelandic meteorological stations); average wind speed (\(f\)); wind gust (\(f_g\)); standard deviation of the wind gust; wind direction (\(d\)); and standard deviation of the wind direction.
+
+These measurements began at the end of the 20th century with the installation of the first AWSs, and more stations have been added in subsequent decades.
 
@@
 Most of these were from IMO but over a hundred were from IRCA.
-IMO AWS are placed at 10 meters above ground, while IRCA stations are placed at around 6-7 meters above ground.}
+IMO anemometers are placed at 10 meters above ground, while IRCA ones are placed at 6--7 meters above ground.}
     \label{fig:aws_map}
@@
 \section{CARRA Data}
-The CARRA dataset goes back to September 1991 and is currently updated monthly, with a latency of 2-3 months\cite{carra_information}.
-The oldest IMO data point that fulfills given criteria of consistent available datapoints is from \startDateVedur.
-This is covered by CARRA.
-The CARRA dataset is available for two regions, West and East.
-These regions cover a large part of the North Atlantic, including Greenland and Scandinavia.
-Each of these covers a vastly larger area than the area of interest.
+CARRA is a high-resolution atmospheric reanalysis produced by the Copernicus Climate Change Service and run by ECMWF.
+It covers two regions, a west region covering Greenland and Iceland and an east region covering the European Arctic.
+It has a 2.5 km horizontal resolution and dates from 1991 to the present, with monthly updates.
+CARRA provides three-hourly analysis fields and short-term forecasts (hourly for lead times under 6 h and three-hourly beyond) of surface and near-surface variables—wind, temperature, pressure, precipitation, etc.
+It is based on the HARMONIE-AROME limited-area NWP model, forced at its boundaries by ERA5 (ECMWF Re-Analysis v.
+5) and enhanced by local observations to better represent complex terrain, land–sea contrasts, and sea-ice processes.
+It is updated monthly, with a latency of 2-3 months\cite{carra_information}.
+
+The CARRA dataset covers all the IMO observations that fulfill criteria of consistent availability – the oldest observation is from \startDateVedur.
+The CARRA-West region covers a vastly larger area than the area of interest.
 This leads to having to store a large amount of data.
-To get the data one has two options.
-CARRA web interface or using the API client provided by CARRA.
-Using the API client is the only realistic option here, as there were thousands of requests made for different times.
-If using the API, it is possible to query a smaller area (such as a rectangular area around Iceland) given a set of coordinates.
+To download CARRA data one has two options, a web interface or using an API client provided by CARRA.
+Using the API client is the only realistic option here, as there are thousands of requests made for different times.
+If using the API, it is possible to query a smaller area (such as a rectangular area around Iceland) given a set of coordinates, but this is not possible with the web interface.
 
-The requests to the API are made at each available CARRA hour ([00, 03, 06, 09, 12, 15, 18, 21]) for each available observation.
-Only these hours can be used as the CARRA predicted output is represents the wind speed at those given hours and would thus not encapsulate well the extreme values in between these times.
-Using these datapoints, interpolation was used to get an estimation for the point of the given weather station.
-The CARRA data contains several types of layers.
-These are single levels, model levels, height levels, pressure levels.
+The requests to the API were made at each available CARRA hour ([00, 03, 06, 09, 12, 15, 18, 21]) on a grid covering Iceland, for each available observation time.
+The downloaded data were interpolated to get values at the weather stations.
+CARRA contains several types of layers: single levels, model levels, height levels, and pressure levels.
 The data for this thesis was downloaded from height levels.
-That is, data was requested at heights of 15, 250 and 500 meters above ground.
-For each point 4 parameters were requested, wind speed, wind direction, pressure and temperature.
-Each of these features needed to be interpolated to create data for model to be trained on, as the CARRA grid points do not coincide with weather stations.
+They were requested at heights of 15, 250 and 500 meters above ground.
+For each point 4 parameters were requested, wind speed, wind direction, pressure, and temperature.
 
 \section{Elevation data}
-IMO provided a TIFF file containing the elevation of Iceland on a 20 meter by 20 meter grid.
-This file encompasses Iceland and is around 685 MB.
-The Python package Rasterio allowed for quick lookup with it's index and the affine transform.
-Using this package it is possible to quickly look up elevation given coordinates using matrix calculations.
-This package allowed for lookup by coordinates and by positioning inside the grid encompassing Iceland.
-It also allows for index lookup using coordinates.
-Using index lookup, neighboring points can be looked up and used to bridge the elevation for exact posisiton of given point.
+A GeoTIFF file containing Iceland’s elevation on a 20~m by 20~m grid was provided by the Icelandic Meteorological Office (IMO).
+The entire country is covered by this file, and its size is approximately 685\,MB.
 
-\label{Chapter3} 
+The Python package \texttt{rasterio} is used, enabling rapid elevation lookup via its spatial indexing and affine‐transform capabilities.
+Elevation at specified geographic coordinates can be retrieved directly, and grid indices may be used for efficient access.
+Elevation for any exact location can be interpolated by fetching neighboring grid points.
+
 \section{Combining data sources}
 
-This project used three main data sources, which need to be queried, filtered and combined to prepare the data for use in the models.
-When working with hundred of thousands of rows, the efficiency of the code is very important.
-Iterating through those rows might be necessary at times but will increase the time exponentially as compared to using vectorizing methods were possible.
-The three data sources were all in a different format.
-Measurement data from IMO was in text files, elevation data was in GeoTiff and reanalysis data from CARRA was in a GRIB format.
-To use the data to train, these three data sources needed to be combined into one file.
-This was done based on the measurement data from the IMO.
-A limit was set on the average wind speed and it was used to select measurement points.
-Along with the average wind speed having to be above a certain limit, to ensure that the same weather for the same location is not duplicated.
-The data from IMO was supplied for 10 minute increments, while CARRA data is in 3 hour intervals.
-This means that to use the CARRA data to predict the measured values from IMO, temporal interpolation would need to be done.
-Along with the temporal interpolation, note that the CARRA data is given in a rectangular grid where the distance between each point is around 2.5 km while the the information from the IMO is given at specific locations.
-The elevation information was given by a 20 by 20 rectangular grid that covers Iceland.
-When combining these data sources an interpolation method needs to be decided upon.
-Here linear interpolation was applied, both temporally and spatially.
-As the measurement points chosen were not outliers, that is the strongest average wind speed in a given range, interpolating in time does not work well.
-This necessitates using only measurements that are exactly at three hour intervals (00, 03, 06, 09, 12, 15, 18, 21 hours) at interpolating only in space.
-The choice of interpolation method, although potentially impactful on the results, was not specifically addressed in this study.
+Three main data sources were used, each requiring querying, filtering, and merging to prepare the combined dataset.
+When handling hundreds of thousands of rows, code efficiency is essential: row-by-row iteration can increase execution time dramatically compared to vectorized operations.
+
+The sources were provided in different formats: IMO measurement data in text files, elevation data in GeoTIFF, and CARRA reanalysis data in GRIB.
+To train the models, these datasets were combined into a single file using the IMO measurements as a reference.
+CARRA data are supplied on a rectangular grid with approximately 2.5~km spacing, while IMO observations are tied to specific station locations.
+Elevation data are on a 20~m by 20~m grid covering Iceland.
+Linear interpolation was applied to merge the sources.
 
 \begin{figure}[h]
-    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart.png}
-    \caption{A flow chart showing how data sources were combined}
+  \centering
+  \includegraphics[scale=0.5]{Figures/data-preprocessing-flow-chart.png}
+  \caption{Flow chart illustrating the data-combination workflow}
   \label{fig:data_preprocessing_flow_chart}
@@
 
-The procedure of combining these sources was as follows and can be seen in Figure \ref{fig:data_preprocessing_flow_chart}.
-The measured data from the AWS is filtered by using a limit on the average wind speed.
-The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}).
-Even so being able to predict the gust factor is more important for higher average wind speed due to higher wind gusts.
-After this stripped dataset over every AWS has been created it is used to query the CARRA data by using their API.
-The CARRA API needs to be queried for given hours, days, months, years and a given area.
-That is, if queried for a given hour, it returns that hour for every day that is queried.
-Similarly if queried for a given day, it returns that day for every month.
-In light of these restraints, it was decided to query month by month.
-Querying only the days needed but every hour of the day (UTC 00, 03, 06, 09, 12, 15, 18 and 21).
-After querying and downloading the data for the height levels and variables requested, points of interest are interpolated and values stored in a pandas dataframe.
-After this the downloaded data is discarded and the next month is queried.
-This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (a reduction from several terabytes to less than a gigabyte).
+The merging procedure, shown in Figure \ref{fig:data_preprocessing_flow_chart}, was as follows: for each AWS observation, a query was constructed for the CARRA API by specifying year, month, day, hour, and spatial extent.
+Because the API returns all specified days when queried by hour, and all specified months when queried by day, monthly queries were issued for only the required days, retrieving all eight three-hourly time points (00, 03, 06, 09, 12, 15, 18, and 21~UTC).
+After downloading the requested variables at the desired pressure levels, point values were interpolated and appended to a pandas dataframe.
+The monthly GRIB files were then discarded before proceeding to the next month.
+This strategy reduced storage needs from several terabytes to under one gigabyte.
 
-Once CARRA data has been merged with AWS data, using station and time columns, then this combined file needs to checked for nails.
-This is done by using the hourly data (which is supposedly error free).
-As the data has been filtered in for the highest average wind speed in a 48 hour interval, the hourly data can be used to find nails.
-The hourly data is combined with the merged AWS and 10 min data.
-Then filtering is applied on the average wind.
-If the average wind speed for the 10 minute data is higher than the hourly the rows are dropped.
-Most stations have nails in less that 10\% measurements.
-The stations that have higher than 10\% error are ignored.
+Elevation values from the GeoTIFF were interpolated in the same manner: the four surrounding grid points were used in a linear interpolation to estimate the elevation at each station location.
+These interpolated values were included in the dataframe, since topography influences both average wind speed and gustiness \cite{GNP_vidtal}.
 
-The elevation data comes in a GeoTIFF file that covers Iceland.
-It is a rectangular grid of resolution 20 meters.
-For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved.
-For each point retrieved interpolation needs to be done.
-This is done in a similar manner to the interpolation of the CARRA data.
-The four points bounding the point of interest were used to linearly interpolate the value of the point of interest.
-This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.
+\section{Comparison of observed and reanalysis wind speed}
 

### 2025-05-02 Chapters/Chapter2.tex
komið að kafla 4

@@
 
-The error in reanalysis wind speed and measured wind speed can be significant.
-The absolute error increases as the measured wind speed increases, while the percentage wind speed decreases.
-A grouping of these errors by wind speed can be seen in Table \ref{table:measuredVSReanalysis_wind_speed}
+Differences between reanalysis and measured wind speeds can be substantial.
+Absolute error increases with wind speed, while percentage error decreases.
+An overview of these errors by wind-speed range is shown in Table \ref{table:measuredVSReanalysis_wind_speed}.
 
 \begin{table}[h]
-    \caption[Comparison of measured and reanalysis wind speed]{Comparison of measured and reanalysis wind speed using mean absolute error (MAE) and mean absolute percentage error (MAPE).
-Note that for the computation of MAPE for ranges that otherwise include 0, 0 values have been excluded so as to prevent division by zero and exploding values.
-The comparisons are done using measured wind speed (at 10 meters above ground for IMO and 6-7 meters above ground for IRCA) and reanalysis wind speed at 15 meters above ground.}
+  \centering
+  \caption[Measured vs.\ reanalysis wind-speed errors]{Comparison of measured and reanalysis wind speeds using mean absolute error (MAE) and mean absolute percentage error (MAPE).
+Values for observed wind speeds below 1~m/s are excluded to avoid inflated MAPE values.
+Measured speeds are at 10~m above ground (IMO) or 6–7~m (IRCA); reanalysis speeds are at 15~m.}
   \label{table:measuredVSReanalysis_wind_speed}
-    \centering
-    \begin{tabular}{ccc}%c}
+  \begin{tabular}{cccc}
     \toprule
-        f & n & MAE \\%& MAPE \\
+    $f$ & $n$ & MAE & MAPE \\
     \midrule
-        $[0;5[$ & 6.2e6 & 2.1 \\%& 1.6\\
-        $[5;10[$ & 4.2e6 & 2.2 \\%& 0.3\\
-        $[10;15[$ & 1.5e6 & 2.5 \\%& 0.2\\
-        $[15;20[$ & 3.9e5 & 3.0 \\%& 0.2\\
-        $[20;25[$ & 8.4e4 & 4.0 \\%& 0.2\\
-        $[25;\infty[$ & 2.0e4 & 6.6 \\%& 0.2\\
-        $[0;\infty[$ & 1.2e7 & 2.2 \\%& 1.0\\
+    $[1; 5[$     & 5,260,814  & 2.0 & 83.9\% \\
+    $[5; 10[$    & 4,150,923  & 2.2 & 31.3\% \\
+    $[10; 15[$   & 1,480,487  & 2.5 & 21.0\% \\
+    $[15; 20[$   &   388,905  & 3.0 & 17.8\% \\
+    $[20; 25[$   &    84,099  & 4.0 & 18.4\% \\
+    $[25; \infty[$ &   20,288 & 6.6 & 23.0\% \\
+    $[1; \infty[$  &11,385,516 & 2.2 & 53.7\% \\
     \bottomrule
@@
 
-Another thing to look at is the distribution of error by station, both in terms of their coordinates and number of measurements.
-Looking at Figure \ref{fig:station_mae_distribution}, this distribution can be seen.
+Next, the distribution of MAE by station is examined, considering station location and number of observations.
+Figure \ref{fig:station_mae_distribution} shows this distribution, and Table \ref{table:station_mae_distribution} lists the five stations with the lowest MAE and the five with the highest MAE.
 
-\begin{figure}
+\begin{figure}[h]
+  \centering
   \includegraphics[scale=0.6]{Figures/MAEoverIceland.png}
-    \caption[Distribution of mean absolute errors by station]{The distribution of mean absolute errors by station.
-Using mean absolute error instead of mean absolute percentage error allows for all points to be used.
-Mean absolute percentage error can only be used if 0 values are ignored.}
+  \caption[MAE distribution by station]{Distribution of mean absolute error (MAE) by station for all observed wind speeds.}
   \label{fig:station_mae_distribution}
@@
 
-Table \ref{table:station_mae_distribution} shows the 5 best and worst stations in terms of MAE.
-
 \begin{table}[h]
-    \caption[Mean absolute difference of measured wind speed and reanalysis wind speed at select stations]{Mean absolute error for reanalysis wind speed as compared to measured wind speed, for the five stations with the highest difference and the five stations with the lowest difference.}
+  \centering
+  \caption[Station MAE extremes]{Mean absolute error (MAE) between reanalysis and observed wind speeds at the five stations with lowest and highest errors.}
   \label{table:station_mae_distribution}
-    \resizebox{\textwidth}{!}{
-    \centering
-    \begin{tabular}{cccc}
+  \begin{tabular}{lcc}
     \toprule
-        Station & Number of measurements & MAE & Location \\
+    Station               & $n$     & MAE  \\
     \midrule
-        1470 & 6.8e3 & 1.17 & Reykjavík Háahlíð \\
-        1350 & 5.2e4 & 1.18 & Keflavíkurflugvöllur \\
-        1482 & 1.4e4 & 1.23 & Reykjavík Víðidalur \\
-        4921 & 1.3e4 & 1.29 & Rif á Melrakkasléttu \\
-        1477 & 5.6e4 & 1.29 & Reykjavíkurflugvöllur\\ \hdashline[0.5pt/1pt]
-        35553 & 4.0e3 & 4.30 & Almannaskarð - göng\\
-        6745 & 1.5e4 & 4.36 & Kerlingarfjöll - Ásgarðsfjall\\
-        35978 & 7.9e3 & 4.40 & Fáskrúðsfjarðargöng suður\\
-        2640 & 1.6e3 & 4.51 & Seljalandsdalur\\
-        32635 & 3.2e4 & 4.95 & Botn í Súgandafirði\\
+    Reykjavík Háahlíð     & 6,800   & 1.17 \\
+    Keflavíkurflugvöllur  & 52,000  & 1.18 \\
+    Reykjavík Víðidalur   & 14,000  & 1.23 \\
+    Rif á Melrakkasléttu  & 13,000  & 1.29 \\
+    Reykjavíkurflugvöllur & 56,000  & 1.29 \\
+    \midrule
+    Almannaskarð          &  4,000  & 4.30 \\
+    Kerlingarfjöll        & 15,000  & 4.36 \\
+    Fáskrúðsfjarðargöng   &  7,900  & 4.40 \\
+    Seljalandsdalur       &  1,600  & 4.51 \\
+    Botn í Súgandafirði   & 32,000  & 4.95 \\
     \bottomrule
   \end{tabular}
-    }
 \end{table}
 
-\section{Data Structure}
+\section{Layout of combined data}
 
-Once data has been retrieved for all three sources and processed, including interpolating values, it needs to be made ready to use by the model, for both training, validation and test.
-The starting point is a dataframe that contains measured information from AWS.
-This includes the average wind, the wind gust, wind direction along with the station number and coordinates.
-When selecting the CARRA data certain height levels are chosen.
-These present as separate lines in the CARRA dataframe.
-Information for one observation is represented in as many lines as height levels requested in the reanalysis data.
-These rows need to be combined on the position (the weather station).
-When this is done it is possible to combine the AWS IMO data and CARRA reanalysis data on the location and time columns.
-The last data source is the elevation.
-A circle sector upwind is looked at.
-In any case the points, that represent these sections, were selected as shown in Code Listing \ref{code:sectorElevation}.
-A range of angles are defined based on the wind direction $d$ at some distance from the given point.
-This means that the resultant points (equal in number to the length of angleRange by k) from arcs at several distances from the given weather station.
+Once data from all three sources have been retrieved and processed—including interpolation—they must be combined and formatted for modeling (training, validation, and testing).
+The starting point is a DataFrame containing AWS observations: average wind speed, wind gust, wind direction, station number, and coordinates.
+CARRA data are provided at selected height levels, each as a separate row in the CARRA DataFrame.
+Thus a single observation will span multiple rows, one per level.
+These rows are merged by time and station location, enabling AWS and CARRA data to be joined on station and time fields.
 
-\begin{lstlisting}[style = Python, caption = {Sector elevation points generated}, label = code:sectorElevation]
+Elevation data are handled by defining an upwind sector: a range of angles relative to the wind direction \(d\) and radial distances from each station.
+Points within this sector are generated as shown in Code Listing \ref{code:sectorElevation}.
+
+\begin{lstlisting}[
+    style=Python,
+    basicstyle=\small\ttfamily,
+    caption={Generation of elevation points in the upwind sector},
+    label=code:sectorElevation
+]
 angles = [(angle + (90 - d)) * pi/180 for angle in angleRange]
@@
 
-The result is a dataframe that has measured data from AWS, which gives us our target, reanalysis data from CARRA, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data.
-An example of what the data looks like can be seen in Table \ref{table:trainDataExample}.
+The resulting DataFrame includes AWS measurements (our target), CARRA weather variables, and elevation points. An example of the combined data structure is shown in Table \ref{table:trainDataExample}.
 
 \begin{table}[h]
-    \caption[An example of data structure used to train model]{An example of data structure used to train model.
-Data points include the derived variables Richardson number ($Ri$) and Brunt-Väisäla frequency ($N$) (defined below), the elevation of the station, direction of wind and relative direction of the wind (twd, that is the direction of the wind relative to center of Iceland), along with some combination of wind speed, pressure and temperature at the different height levels.
-Finally there are the elevation points around a given station, where the elevation is relative to the station.}
+    \centering
+    \caption[Example of combined data structure]{Example of the combined data structure used for modeling.
+Data include derived variables Richardson number (\(Ri\)) and squared Brunt–Väisälä frequency (\(N^2\)), station elevation, wind direction and relative wind direction (twd), along with wind speed, pressure, and temperature at various height levels, and elevation points around each station relative to station elevation.}
     \label{table:trainDataExample}
     \resizebox{\textwidth}{!}{
-    \centering
-    \begin{tabular}{cccccccccc}
+    \begin{tabular}{ccccccccccc}
         \toprule
-        Ri & $N^2$ & station elevation & twd & $ws_{15}$ & $wd_{15}$ & $t_{15}$ & $p_{15}$ & $elevation_0$ & \dots\\
+        Ri & \(N^2\) & \shortstack{station\\elevation} & twd & ws\(_{15}\) & wd\(_{15}\)
+        & \(t_{15}\) & \(p_{15}\) & \(\text{elevation}_0\) & elevation\(_1\) & \dots\\
         \midrule 
-        -1.18e+00 &  2.67e+04 & 100 & 1.5 & 10 & 5 & 0 & 100 & 2 & \dots\\
+        -1.18 &  26700 & 100 & 1.5 & 10 & 5 & 0 & 100 & 2 & 4 & \dots\\
+        \dots\\
         \bottomrule
@@
 
-Looking at Table \ref{table:trainDataExample} note that the first two columns represent two variables that describe the stability of the air.
-These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency ($N$)\cite{brunt_vaisala_freq_eumtrain}, and are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}.
-These values are calculated using reanalysis data at two different height levels.
-Thus $Ri$ refers to the Richardson number calculated between height levels 15m and 500m.
-Exactly the same notation is used with the Brunt–Väisälä frequency, except the square is used.
+In Table \ref{table:trainDataExample}, the first two columns describe atmospheric stability: the Richardson number (\(Ri\)) and squared Brunt–Väisälä frequency (\(N^2\)), defined in Equations (\ref{eqn:Ri}) and (\ref{eqn:N}) \cite{richardson_number_skybrary,brunt_vaisala_freq_eumtrain,mean_gust_HA_HO}.
+Here, \(g\) is gravitational acceleration; \(dT\) the temperature difference between levels; \(dz\) the height difference; \(T_{\mathrm{ave}}\) the average temperature; and \(dU\) the wind-speed difference.
+Lower \(Ri\) indicates greater turbulence, with typical values between 0.1 and 10, and values below 1 signifying significant turbulence \cite{richardson_number_skybrary}.
+Negative \(N^2\) denotes instability, as an air parcel will accelerate away from its original position \cite{brunt_vaisala_freq_eumtrain}.
 
-\begin{equation}
-    \label{eqn:Ri}
-    Ri = \frac{g \cdot dT \cdot dz}{T_{\textrm ave} \cdot dU^2} \unit{[]}
-\end{equation}
+Since \(Ri\) and \(N^2\) are derived from reanalysis variables, they add limited new information compared to the raw data, but they can reduce input dimensionality and improve model explainability via Shapley-value analysis \cite{shapley_information}.
+Shapley values assess feature importance by evaluating all possible feature subsets—an operation of computational complexity \(2^n\).
+In practice, approximations are used but can still be costly for large models.
+Because \(dU\) appears squared in the denominator of \(Ri\), if the wind-speed difference between levels is very small, \(Ri\) can become arbitrarily large, potentially distorting predictions.
 
-\begin{equation}
-    \label{eqn:N}
-    N = \sqrt{\frac{g \cdot dT }{T_{\textrm ave} \cdot dz}} \unit{[Hz]}
-\end{equation}
+\section{Distributions of observed and reanalysis data}
 
-Here, $g$ is the acceleration due to gravity, $dT$ is the temperature difference between the two height levels, $dz$ is the elevation difference, $T_{\textrm ave}$ is the average temperature (that is the average of the two temperatures in the height levels) and $dU$ is the wind speed difference between the two height levels.
-Both of these numbers provide some insight about the stability of the air.
-A lower value for the Richardson number indicates a higher turbulence.
-A typical range of values could be between 0.1 and 10, with values below 1 indicating significant turbulence\cite{richardson_number_skybrary}.
-When the square of the Brunt-Väisälä frequency is negative, then the air is unstable (an air parcel will move away from its original position)\cite{brunt_vaisala_freq_eumtrain}.
-These are derived factors from the reanalysis data and as such there shouldn't be a significant information gain using $Ri$ and $N$ as opposed to having the raw data.
-However, including these factors instead of every reanalysis variable requested might speed up training as well as making the model more easily explainable with the use of Shapley values or other tools for explainability.
-Using Shapley a feature importance value is attributed to a given feature by creating all possible permutations of any possible length (up to number of features) and seeing how the predictions are skewed when the given parameter is included or excluded.
-This needs to be done for all parameters.
-The time complexity of this is very high ($2^n$ coalitions)\cite{shapley_information}.
-Most implementations use some approximations, which still can take a considerable amount of time for models with a high parameter count and many examples.
-The Richardson number includes the difference in wind speeds in the denominator.
-In certain cases, where the difference in wind speed between two levels is very, can blow up to infinity.
-This can cause problems and distort the predicitons.
-
-\section{Data distribution}
-
-The CARRA data is reanalysis and as such might have a bias or some systemic distortion when compared to the measured data.
-The distribution of the observed and reanalysis wind speed can be seen in Figure \ref{fig:obs_carra_wind_speeds}.
-Looking at the figure, the reanalysis wind tends to be higher even though the distribution is similar.
+CARRA reanalysis data may exhibit bias or systematic differences when compared to measurements.
+Figure \ref{fig:obs_carra_wind_speeds} shows the distributions of observed and reanalysis wind speeds.
+Although the shapes are similar, reanalysis values tend to be higher.
 
@@
   \centering
-    % First subfigure (top)
   \begin{subfigure}[b]{0.8\textwidth}
@@
     \includegraphics[width=\textwidth]{Figures/obs_wind_speeds.png}
+    \caption{Histogram of observed wind speeds from IMO and IRCA at all stations, sampled at 3-hour intervals (00,~03,~06,~…~,21~UTC).}
     \label{fig:obs_wind_speeds}
@@
 
-    \vspace{0.5cm} % Space between the two images
+  \vspace{0.5cm}
 
-    % Second subfigure (bottom)
   \begin{subfigure}[b]{0.8\textwidth}
@@
     \includegraphics[width=\textwidth]{Figures/carra_wind_speeds.png}
+    \caption{Histogram of interpolated CARRA wind speeds at station locations on a 2.5~km grid and 3-hour intervals.}
     \label{fig:carra_wind_speeds}
@@
 
-    \caption[Distribution of CARRA and observed wind speeds]{Upper figure shows a histogram of observed wind speeds provided by IMO and IRCA for all used weather stations.
-Lower figure shows the interpolated CARRA reanalysis values at weather stations.
-The reanalysis data is only available at 3 hour intervals and 2.5km grid.
-As such the observed values are only chosen at these specific times (00, 03, 06, ..., 21).
-Spatial interpolation is needed and is linearly weighted based on distance of CARRA grid points from stations.}
+  \caption{Comparison of observed and reanalysis wind-speed distributions. CARRA values are spatially interpolated by linear weighting of surrounding grid points.}
   \label{fig:obs_carra_wind_speeds}
@@
   \includegraphics[width=0.8\textwidth]{Figures/station_heights.png}
-    \caption[Distribution of weather station heights above sea level]{Distribution of heights of weather stations above sea level.
-One station has been excluded as an outlier of well above 1000 meters, having few and inconsistent datapoints.}
+  \caption{Distribution of weather station elevations above sea level. One outlier near 2000~m was excluded due to limited and inconsistent data.}
   \label{fig:station_heights}
@@
   \includegraphics[width=0.8\textwidth]{Figures/gust_factor_2025.png}
-    \caption[Distribution of gust factors]{Histogram of gust factors.
-By definition the lower bound of gust factor is 1.
-The majority of observed gust factors fall in the range 1.2 to 2.
-Gust factor decreases with increasing wind speed.}
+  \caption{Histogram of gust factors. By definition, gust factor \(\ge1\); most values lie between 1.2 and 2. The gust factor declines as wind speed increases.}
   \label{fig:gust_factors}

### 2025-05-02 Chapters/Chapter3.tex
komið að kafla 4

@@
 \section{Model structure}
-The structure of the neural network is such that it contains some number $n$ of fully connected layers and batch normalization for each layer, along with regularization.
-All of these layers have the same number of units.
-The last layer has a dropout of 50\%.
-In addition to these layers there is one more output layer.
-This is simply a dense layer with 1 unit.
-A grid search was performed to determine the hyperparameters that minimize the loss.
-Hyperparameters are parameters that are set before the training begins\cite{hyperparameters_definition}.
-These hyperparameters include number of units in each layer, number of epochs to train for, number of layers, batch size, optimizer and penalty to enforce in the regularization.
-The possible combinations tried can be seen in Table \ref{table:gridSearchHyperparamters}.
+The network consists of \(n\) fully connected layers, with \(n\) around 10.
+Each hidden layer has the same number of units and is followed by batch normalization.
+A dropout layer with rate 50\% is applied after the final hidden layer.
+A dense output layer with one unit produces the predicted gust factor.
+A grid search was used to find the hyperparameters that minimize the loss.
+Hyperparameters are settings defined before training begins \cite{hyperparameters_definition}.
+They include the number of units per layer, the number of layers, the number of training epochs, batch size, optimizer choice, and regularization penalty.
+The ranges tested are listed in Table \ref{table:gridSearchHyperparamters}.
 
@@
     \centering
-    \caption[Hyperparameter search with best performing combination.]{Hyperparamter search with best performing combination shown.
-Hyperparameter search was done using hyperband algorithm that initially searches randomly for the best parameters but then hones in on what is working and as such is neither exhaustive nor completely random.
-This means that a hard upper limit will not be set on the number of combinations to try like with randomsearch.}
+    \caption[Hyperparameter search with Hyperband]{Hyperparameter search was performed using the Hyperband algorithm, which begins by sampling configurations at random and then focuses on the most promising ones.
+The table shows the ranges tested and the best-performing combination.
+Unlike random search, Hyperband does not impose a fixed limit on the number of trials.}
     \label{table:gridSearchHyperparamters}
-    \begin{tabular}{ccc}
+    \begin{tabular}{lll}
         \toprule
@@
 
-\section{Model Training}
-To determine the required number of epochs a high number of epochs can be selected and the loss plotted.
-These plots can be seen in Figures \ref{fig:loss_plot_wo_elevation} and \ref{fig:loss_plot_w_elevation}.
-Looking at these two plots, training loss decreases for over 1900 epochs in both plots, while the validation loss stops decreasing at around 200 epochs for DEM model and 1800 epochs for model without DEM.
-This is unexpected.
-The DEM model has vastly more parameters and as such should contain more useful information to be learned, which one might expect to take longer to learn.
-This does not seem to be the case.
-The DEM model only gets trivially better after 1000 epochs while the non DEM model keeps meaningfully improving until around 1800 epochs.
-With respect to time constraints of shared resources models were trained for 250 epochs.
+As noted, Hyperband does not impose a fixed limit on the total number of training epochs.
+When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband tuner}, the parameter \texttt{hyperband\_iterations} controls how many times Hyperband is run and defaults to 1.
+On each run, the total number of epochs is distributed across the hyperparameter trials.
+The total number of epochs is approximately
 
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/loss_plot_wo_elevation.png}
-    \caption[Loss plot of model trained for 2000 without DEM.]{Loss plot of model trained for 2000 without DEM.}
-    \label{fig:loss_plot_wo_elevation}
-\end{figure}
+\begin{equation*}
+n_{\mathrm{epochs}} = \text{max\_epochs} \times \bigl(\log(\text{max\_epochs})\bigr)^2 \approx 10^4
+\end{equation*}
 
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/loss_plot_w_elevation.png}
-    \caption[Loss plot of model trained for 2000 with DEM.]{Loss plot of model trained for 2000 with DEM.}
-    \label{fig:loss_plot_w_elevation}
-\end{figure}
+for $ \text{max\_epochs} = 1000$.
+Although this method is more efficient than grid search, if each epoch takes about 10 seconds, a complete Hyperband search can still require about 28 hours on shared resources, making repeated searches impractical.
+It remains unclear whether the current hyperparameter ranges are optimal.

### 2025-05-03 Chapters/Chapter2.tex
komið lengra

@@
 Data from these automatic weather stations (AWS) are stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases.
-Each record contains the following information: date and time; station number (convertible to coordinates using another dataset of Icelandic meteorological stations); average wind speed (\(f\)); wind gust (\(f_g\)); standard deviation of the wind gust; wind direction (\(d\)); and standard deviation of the wind direction.
+Each record contains the following information: date and time; station number (convertible to coordinates using another dataset of Icelandic meteorological stations); average wind speed ($f$); wind gust ($f_g$); standard deviation of the wind gust; wind direction ($d$); and standard deviation of the wind direction.
 
@@
 \section{Elevation data}
-A GeoTIFF file containing Iceland’s elevation on a 20~m by 20~m grid was provided by the Icelandic Meteorological Office (IMO).
+A GeoTIFF file containing a digital elevation model (DEM) for Iceland on a 20~m by 20~m grid was provided by the Icelandic Meteorological Office (IMO).
 The entire country is covered by this file, and its size is approximately 685\,MB.
@@
 
-Elevation data are handled by defining an upwind sector: a range of angles relative to the wind direction \(d\) and radial distances from each station.
+The DEM elevation data are handled by defining an upwind sector: a range of angles relative to the wind direction $d$ and radial distances from each station.
 Points within this sector are generated as shown in Code Listing \ref{code:sectorElevation}.
@@
     \centering
-    \caption[Example of combined data structure]{Example of the combined data structure used for modeling.
-Data include derived variables Richardson number (\(Ri\)) and squared Brunt–Väisälä frequency (\(N^2\)), station elevation, wind direction and relative wind direction (twd), along with wind speed, pressure, and temperature at various height levels, and elevation points around each station relative to station elevation.}
+    \caption[Example of combined data structure]{Example of the combined data structure of features used for modeling.
+Data include derived variables Richardson number (Ri) and squared Brunt–Väisälä frequency ($N^2$), station altitude (meters above sea level), transformed wind direction (twd), wind speed (ws$_{15}$), wind direction (wd$_{15}$), temperature ($t_{15}$), pressure ($p_{15}$) (CARRA values at 15 m height), and elevation points (from DEM) in a sector pointing upwind.}
     \label{table:trainDataExample}
@@
         \toprule
-        Ri & \(N^2\) & \shortstack{station\\elevation} & twd & ws\(_{15}\) & wd\(_{15}\)
-        & \(t_{15}\) & \(p_{15}\) & \(\text{elevation}_0\) & elevation\(_1\) & \dots\\
+        Ri & $N^2$ & \shortstack{station\\altitude} & twd & ws$_{15}$ & wd$_{15}$
+        & $t_{15}$ & $p_{15}$ & $\text{elevation}_0$ & elevation$_1$ & \dots\\
         \midrule 
@@
 
-In Table \ref{table:trainDataExample}, the first two columns describe atmospheric stability: the Richardson number (\(Ri\)) and squared Brunt–Väisälä frequency (\(N^2\)), defined in Equations (\ref{eqn:Ri}) and (\ref{eqn:N}) \cite{richardson_number_skybrary,brunt_vaisala_freq_eumtrain,mean_gust_HA_HO}.
-Here, \(g\) is gravitational acceleration; \(dT\) the temperature difference between levels; \(dz\) the height difference; \(T_{\mathrm{ave}}\) the average temperature; and \(dU\) the wind-speed difference.
-Lower \(Ri\) indicates greater turbulence, with typical values between 0.1 and 10, and values below 1 signifying significant turbulence \cite{richardson_number_skybrary}.
-Negative \(N^2\) denotes instability, as an air parcel will accelerate away from its original position \cite{brunt_vaisala_freq_eumtrain}.
+The transformed wind direction is a proxy showing whether the wind is coming from land or sea.
+It is computed as the angle between the wind direction and a vector from the center of Iceland to the station.
 
-Since \(Ri\) and \(N^2\) are derived from reanalysis variables, they add limited new information compared to the raw data, but they can reduce input dimensionality and improve model explainability via Shapley-value analysis \cite{shapley_information}.
-Shapley values assess feature importance by evaluating all possible feature subsets—an operation of computational complexity \(2^n\).
+The Richardson number and the Brunt–Väisälä frequency describe atmospheric stability, see equations (\ref{eqn:Ri}) and (\ref{eqn:N})  \cite{richardson_number_skybrary,brunt_vaisala_freq_eumtrain,mean_gust_HA_HO}.
+These values are calculated using reanalysis data at two different height levels.
+Thus Ri refers to the Richardson number calculated between height levels 15m and 500m.
+Exactly the same notation is used with the Brunt–Väisälä frequency, except the square is used.
+%
+\begin{equation}
+  \label{eqn:Ri}
+  Ri = \frac{g \cdot dT \cdot dz}{T_{\textrm ave} \cdot dU^2} \unit{[]}
+\end{equation}
+%
+\begin{equation}
+  \label{eqn:N}
+  N = \sqrt{\frac{g \cdot dT }{T_{\textrm ave} \cdot dz}} \unit{[Hz]}
+\end{equation}
+%
+Here, $g$ is gravitational acceleration; $dT$ the temperature difference between levels; $dz$ the height difference; $T_{\mathrm{ave}}$ the average temperature; and $dU$ the wind-speed difference.
+Lower Ri indicates greater turbulence, with typical values between 0.1 and 10, and values below 1 signifying significant turbulence \cite{richardson_number_skybrary}.
+Negative $N^2$ denotes instability, as an air parcel will accelerate away from its original position \cite{brunt_vaisala_freq_eumtrain}.
+
+Since Ri and $N^2$ are derived from reanalysis variables, they add limited new information compared to the raw data, but they can reduce input dimensionality and improve model explainability via Shapley-value analysis \cite{shapley_information}.
+Shapley values assess feature importance by evaluating all possible feature subsets—an operation of computational complexity $2^n$.
 In practice, approximations are used but can still be costly for large models.
-Because \(dU\) appears squared in the denominator of \(Ri\), if the wind-speed difference between levels is very small, \(Ri\) can become arbitrarily large, potentially distorting predictions.
+Because $dU$ appears squared in the denominator of Ri, if the wind-speed difference between levels is very small, Ri can become arbitrarily large, potentially distorting predictions.
 
@@
   \includegraphics[width=0.8\textwidth]{Figures/station_heights.png}
-  \caption{Distribution of weather station elevations above sea level. One outlier near 2000~m was excluded due to limited and inconsistent data.}
+  \caption{Distribution of weather station altitudes above sea level. One outlier near 2000~m was excluded due to limited and inconsistent data.}
   \label{fig:station_heights}
@@
   \includegraphics[width=0.8\textwidth]{Figures/gust_factor_2025.png}
-  \caption{Histogram of gust factors. By definition, gust factor \(\ge1\); most values lie between 1.2 and 2. The gust factor declines as wind speed increases.}
+  \caption{Histogram of gust factors. By definition, gust factor $\ge1$; most values lie between 1.2 and 2. The gust factor declines as wind speed increases.}
   \label{fig:gust_factors}

### 2025-05-03 Chapters/Chapter4.tex
Update Chapter4.tex
Skipti kafla 4 í þrjá undirkafla.
komið lengra

@@
 \label{Chapter4}
-\section{Results}
-A baseline model was constructed.
-This model looked at the gust factor for some training data and took the average of this and predicted this average everytime.
-Several baseline guesses were created based on a lower limit assigned to the average wind speed limit (AWSL).
+\section{Baseline models}
+The simplest model for the gust factor is to guess the average gust factor for all available observations, i.e.
+use no variables from the training data.
+A slightly better model is obtained by using the wind speed as a training feature, either through regression or NN training.
+The MAPEs for the constructed base models are shown in Table \ref{table:baseline_models}.
+The ws$_{15}$ model was obtained with NN training but regression produces an almost identical result.
+
+\begin{table}[h]
+  \caption[MAPE for baseline models]{MAPE for several lower limits on the reanalysis 15 m wind speed ($ws_{15}$). Note that the ws$_{15}$ model was only evaluated for the 10 m/s lower bound.}
+    \label{table:baseline_models}
+    \centering
+    \begin{tabular}{lcc}
+        \toprule
+        Wind speed & \multicolumn{2}{c}{MAPE}\\ 
+        limit $[m/s]$ & mean model & ws$_{15}$ model\\
+        \midrule
+        $\geq 0$ & 39.2\%  & \\
+        $\geq 5$ & 28.1\%  & \\
+        $\geq 10$ & 23.9\% & 17.2\%\\
+        $\geq 15$ & 23.2\% & \\
+        $\geq 20$ & 24.7\% & \\
+        $\geq 25$ & 27.7\% & \\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
 As the average wind speed increases, then the variability in gust as a percentage decreases\cite{mean_gust_HA_HO}.
-This means looking at a subset of the data where the AWSL is higher, a better result can be expected.
+Therefore lower MAPE woult be expected if the training data is restricted to higher wind speeds.
+This fact explains why several models were created for different lower bounds.
+
+\section{Modeling for CARRA wind speed above 10 m/s}
+As noted in Chapter 1, it is more important to predict accurate gust factor when it is windy than when the weather is calm.
+Therefore, at first, the data is restricted to ws$_{15} \geq 10$ m/s.
+The first constructed model incorporates the features temperature, pressure and wind direction at 15 m height.
+This gives a moderately reduced MAPE of 16.7\% compared with 17.2\% for the baseline ws$_{15}$ model.
+When the next pair of features, station altitude and transformed wind direction, is included, the MAPE is reduced considerably, down to 13.8\%.
+The atmospheric stability indicators, Ri and $N^2$, add little to the model performance, contrary to what might have been expected.
+When the atmosphere is unstable, wind gusts from higher up might be expected.
+However, another considerable perfomance increase is observed when digital elevation model (DEM) information is included, down to a MAPE of 12.0\%.
+This represents over 30\% reduction in the MAPE from the ws$_{15}$ baseline model.
+
+\begin{table}[h]
+    \caption[Model results for different sets of parameters.]{Showing the results of different models.
+Models defined by different sets of parameters, with increasing complexity.
+This table shows the feature importance as the loss decreases as more variables are added.
+Adding all height levels decreases the loss but very little.}
+    \label{table:setsOfParams}
+    \centering
+    \begin{tabular}{lc}
+        \toprule
+        \textbf{Model variables} & \textbf{MAPE}\\
+        \midrule
+        Baseline constant & 23.9\%\\
+        $ws_{15}$ & 17.2\%\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd]$ & 13.8\% \\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ & 13.7\%\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ + DEM & 12.0\%\\
+        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, \text{altitude}, twd, N^2, Ri]$ & 12.0\% \\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
 The results show this.
@@
 A question to ask is it possible to achieve better results when only looking at a single station?
+
+\section{Results for specific locations}
 
@@
 
+\section{Comparison for models with variable parameters}
+
 In Figure (\ref{fig:ShapleySummary}) the contribution of each feature, excluding the elevation points, can be seen for the model in general (a significant subset of data is used).
@@
 The results of such a comparison can be seen in Table \ref{table:setsOfParams}.
-
-\begin{table}[h]
-    \caption[Model results for different sets of parameters.]{Showing the results of different models.
-Models defined by different sets of parameters, with increasing complexity.
-This table shows the feature importance as the loss decreases as more variables are added.
-Adding all height levels decreases the loss but very little.}
-    \label{table:setsOfParams}
-    \centering
-    \begin{tabular}{lc}
-        \toprule
-        \textbf{Model parameters} & \textbf{MAPE}\\
-        \midrule
-        Baseline constant & 23.9\%\\
-        $ws_{15}$ & 17.2\%\\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd]$ & 13.8\% \\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ & 13.7\%\\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, ASL, twd, N, Ri]$ + DEM & 12.0\%\\
-        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, ASL, twd, N, Ri]$ & 12.0\% \\
-        \bottomrule
-    \end{tabular}
-\end{table}

### 2025-05-05 Chapters/Chapter1.tex
Bætti Stórhöfða við, uppfærði síðasta kaflann. Tók út * fyrir undirkafla og eitthvað. Notaði svo GPT til þess að benda mér á stafsetningarvillur og uppfærði. Held að þetta sé nokkuð langt komið.

@@
 This thesis examines methods to predict the gust factor using various predictors and multiple data sources, including NWP outputs and observational data.
-Accurate gust predictions are crucial, as it is ofteh the peak wind gust that will cause damaging incidents, such as structural failures, traffic accidents and downed power lines.
+Accurate gust predictions are crucial, as it is often the peak wind gust that will cause damaging incidents, such as structural failures, traffic accidents and downed power lines.
 Moreover, the prevalence of extreme wind events is expected to increase in the future \cite{nasa_extreme_weather}.
@@
 
-\subsection*{Numerical weather prediction}
+\subsection{Numerical weather prediction}
 The history of numerical weather prediction dates back to the 1920s, when Lewis Fry Richardson pioneered the field and attempted to produce forecasts.
@@
 Interest in AI has come in waves: progress is made, then interest diminishes, but over the past 15 years growth has been steady.
-Notable drivers of this wave include advances in computational power (notably parallel processing on graphics processing units, GPUs), the developement of neural networks (NNs) for processing massive datasets, and the availability of large online datasets.
+Notable drivers of this wave include advances in computational power (notably parallel processing on graphics processing units, GPUs), the development of neural networks (NNs) for processing massive datasets, and the availability of large online datasets.
 One type of NN is the convolutional neural network (CNN), originally designed for image processing but applicable to any gridded data with spatial structure \cite{canNNBeatNWP}.
@@
 
-\subsection*{Large weather models}
+\subsection{Large weather models}
 In the last two years, there have been even more developments with the emergence of Large AI Weather forecast Models (LWMs).
@@
 
-Before 2022, LWMs had been shown to compete with traditional NWP in specific cases and to generate forecasts much faster after training, No model had been shown to be able to completely replace tradional NWP systems.
+Before 2022, LWMs had been shown to compete with traditional NWP in specific cases and to generate forecasts much faster after training, No model had been shown to be able to completely replace traditional NWP systems.
 In early 2022, Pathak et al.\ \cite{FourCastNet} introduced FourCastNet, which employs an Adaptive Fourier Neural Operator model leveraging transformer architecture instead of convolution.
@@
 
-\subsection*{Conclusion}
+\subsection{Conclusion}
 Substantial progress has been made since 2018 and especially since 2022 \cite{SecondRevolution}.
@@
 \section{Methodology and related work}
-In this study, data from three sources is used:
+In this study, data from three sources are used:
 
@@
   \item Dense (20 by 20~m) gridded elevation data for Iceland
-  \item Three-hourly observations from Icelandic meteorological stations
+  \item Hourly observations from Icelandic meteorological stations
 \end{enumerate}
@@
 
-These activation control neural activation and can help stabilize the network.
+These activation functions control neural activation and can help stabilize the network.
 \subsection{Model evaluation}

### 2025-05-05 Chapters/Chapter2.tex
Update Chapter2.tex
Updated images of flow chart and station distribution over iceland
Bætti Stórhöfða við, uppfærði síðasta kaflann. Tók út * fyrir undirkafla og eitthvað. Notaði svo GPT til þess að benda mér á stafsetningarvillur og uppfærði. Held að þetta sé nokkuð langt komið.

@@
 
-IMO provided measurements from \nStationsMin weather stations across Iceland.
-The measurements that met the filtering criteria began in \startDateVedur and ended in 2023.
-Of these \nStationsMin stations, \nVedurMin were from the IMO, with the anemometer at 10~m above ground, while the remaining \nVGMin stations were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)}, with the anemometer at 6–7~m above ground \cite{vegagerdin_postur}.
+Automatic weather stations (AWS) have been set up all around the world, including Iceland.
+AWSs might include a variety of sensors, such as barometers, thermometers, and anemometer along with solar panels (or other power source) and some kind of data logger.
+IMO provided measurements from AWSs across Iceland.
+One of the stations used in this study can be seen in Figure (\ref{fig:storhofdi}).
+Not every station had consistently available data.
+Data from \nStationsMin AWSs were included looking at a period from \startDateVedur and ended in 2023.
+Of these \nStationsMin stations provided by IMO, \nVedurMin were IMO stations, with the anemometer at 10~m above ground, while the remaining \nVGMin stations were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)}, with the anemometer at 6–7~m above ground \cite{vegagerdin_postur}.
 The locations of these weather stations are shown in Figure \ref{fig:aws_map}.
 
-Data from these automatic weather stations (AWS) are stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases.
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.3]{Figures/storhofdi.jpeg}
+  \caption[AWS at Stórhöfði in Vestmannaeyjar, south of Iceland.]{AWS at Stórhöfði in Vestmannaeyjar, south of Iceland.
+The anemometer can be seen as three pins in the upper-central of the figure just slightly to the left.}
+\end{figure}
+
+Data from these AWS were stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases.
 Each record contains the following information: date and time; station number (convertible to coordinates using another dataset of Icelandic meteorological stations); average wind speed ($f$); wind gust ($f_g$); standard deviation of the wind gust; wind direction ($d$); and standard deviation of the wind direction.
@@
     \centering
-    \includegraphics[scale = 1]{Figures/stationsOverIceland_2024-05-16_stripped_to_frame.png}
+    \includegraphics[scale = 0.6]{Figures/station_elevations.png}
     \caption[Locations of automatic weather stations in Iceland]{Locations of all 412 stations that were looked at in this study.
@@
   \centering
-  \includegraphics[scale=0.5]{Figures/data-preprocessing-flow-chart.png}
+  \includegraphics[scale=0.6]{Figures/data_combination.drawio.png}
   \caption{Flow chart illustrating the data-combination workflow}

### 2025-05-05 Chapters/Chapter3.tex
komið lengra
Bætti Stórhöfða við, uppfærði síðasta kaflann. Tók út * fyrir undirkafla og eitthvað. Notaði svo GPT til þess að benda mér á stafsetningarvillur og uppfærði. Held að þetta sé nokkuð langt komið.

@@
 
-\chapter{Model Architecture and Training}
+\chapter{Model Architecture}
 
@@
 
-\section{Model structure}
-The network consists of \(n\) fully connected layers, with \(n\) around 10.
+The network consists of $n$ fully connected layers, with $n$ around 10.
 Each hidden layer has the same number of units and is followed by batch normalization.
@@
 They include the number of units per layer, the number of layers, the number of training epochs, batch size, optimizer choice, and regularization penalty.
-The ranges tested are listed in Table \ref{table:gridSearchHyperparamters}.
+The ranges tested are listed in Table \ref{table:gridSearchHyperparameters}.
 
@@
 Unlike random search, Hyperband does not impose a fixed limit on the number of trials.}
-    \label{table:gridSearchHyperparamters}
+    \label{table:gridSearchHyperparameters}
     \begin{tabular}{lll}
@@
 
-As mentioned, hyperband doesn't set a hard upper limit to the number of epochs it will train in total.
-When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{hyperband class} several factors can be set.
-One of interest here is the $hyperband_iterations$ argument.
-This determines how often the hyperband algorithm is run and defaults to 1.
+As mentioned, Hyperband doesn't set a hard upper limit to the number of epochs it will train in total.
+When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband class} several factors can be set.
+One of interest here is the Hyperband iterations argument.
+This determines how often the Hyperband algorithm is run and defaults to 1.
 For each iteration the epochs are distributed between tries (that is each set of hyperparameters) with the total amount of epochs approximately $n_{epochs} = max_{epochs} * log^2(max_{epochs})\approx 10^4$, where $max_{epochs} = 1000$ gives the maximum number of epochs that one set of hyperparameters can be trained for.
-Searching a space generally takes a lot of time but this drastically improves on gridsearch.
+Searching a space generally takes a lot of time but this drastically improves on grid search.
 If each epoch takes around 10 seconds to run then the total search would take around 28 hours on a shared resource.
-This is resource intensive and cannot be repeated often.
+This is resource-intensive and cannot be repeated often.
 Another question that remains is whether the ranges given are optimal.
@@
 As noted, Hyperband does not impose a fixed limit on the total number of training epochs.
-When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband tuner}, the parameter \texttt{hyperband\_iterations} controls how many times Hyperband is run and defaults to 1.
+When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband tuner}, the parameter \texttt{Hyperband\_iterations} controls how many times Hyperband is run and defaults to 1.
 On each run, the total number of epochs is distributed across the hyperparameter trials.
 The total number of epochs is approximately
-
+%
 \begin{equation*}
@@
 \end{equation*}
-
+%
 for $ \text{max\_epochs} = 1000$.

### 2025-05-05 Chapters/Chapter4.tex
Bætti Stórhöfða við, uppfærði síðasta kaflann. Tók út * fyrir undirkafla og eitthvað. Notaði svo GPT til þess að benda mér á stafsetningarvillur og uppfærði. Held að þetta sé nokkuð langt komið.

@@
 As the average wind speed increases, then the variability in gust as a percentage decreases\cite{mean_gust_HA_HO}.
-Therefore lower MAPE woult be expected if the training data is restricted to higher wind speeds.
+Therefore lower MAPE would be expected if the training data is restricted to higher wind speeds.
 This fact explains why several models were created for different lower bounds.
@@
 \section{Modeling for CARRA wind speed above 10 m/s}
-As noted in Chapter 1, it is more important to predict accurate gust factor when it is windy than when the weather is calm.
+As noted in Chapter \ref{Chapter1}, it is more important to predict accurate gust factor when it is windy than when the weather is calm.
 Therefore, at first, the data is restricted to ws$_{15} \geq 10$ m/s.
@@
 When the atmosphere is unstable, wind gusts from higher up might be expected.
-However, another considerable perfomance increase is observed when digital elevation model (DEM) information is included, down to a MAPE of 12.0\%.
+However, another considerable performance increase is observed when digital elevation model (DEM) information is included, down to a MAPE of 12.0\%.
 This represents over 30\% reduction in the MAPE from the ws$_{15}$ baseline model.
@@
         \toprule
-        \textbf{Model variables} & \textbf{MAPE}\\
+        Model variables & MAPE\\
         \midrule
@@
 The results show this.
-Several different AWSL were used for the baseline model, as can be seen in table \ref{table:results}.
+Several different wind speed intervals were used for the baseline model, as can be seen in table \ref{table:results}.
 This sets a goal.
 A model that does not significantly improve on this baseline suggests either failure to capture essential patterns in the data or that the data itself may lack the necessary information for substantial improvements upon the baseline.
-Using the previously described neural network architecture setups for each AWSL, with and without landscape elevation information, MAPE was determined.
+Using the previously described neural network architecture setups for each wind speed interval, with and without landscape elevation information, MAPE was determined.
 The results can be seen in Table \ref{table:results}.
@@
 \begin{table}[h]
-    \caption[Model results for different AWSL]{MAPE for each average wind speed limit with and without landscape elevation in a 30° sector around the point of interest into the direction of the reanalysis wind.
+    \caption[Model results for different wind speed limits]{MAPE for each average wind speed limit with and without landscape elevation in a 30° sector around the point of interest into the direction of the reanalysis wind.
 The influence of adding elevation data seems to reduce the error.
@@
         \toprule
-        \textbf{AWSL} & \textbf{MAPE} & &\\ 
-        $[m/s]$ & \textit{Baseline} &  \textit{Without DEM} & \textit{With DEM} \\
+        \shortstack{Wind Speed\\limit} & MAPE & &\\
+        $[m/s]$ & \shortstack{Baseline\\constant} &  Without DEM & With DEM \\
         \midrule
@@
     \caption[Model result by station]{The MAPE results for selected stations of interest, both when training for the specific site and when the stations are a part of the general data.
-For every station the AWSL is set at 10 m/s.
+For every station the wind speed limit is set at 10 m/s.
 In training for a single station at a time, some site specific information can be gauged.
@@
         \toprule
-        \textbf{Station name} & \textbf{\# points} & \multicolumn{2}{c}{\textbf{MAPE}}\\
+        Station name & n & \multicolumn{2}{c}{MAPE}\\
         & & General training & Site training\\
@@
 \begin{table}[h]
-    \caption[Model result looking at closed wind speed intervals]{The MAPE results for different AWSL intervals.
-Here instead of training for all data above a certain threshold put of the wind, training is done only on data between two wind speeds.
+    \caption[Model result looking at closed wind speed intervals]{The MAPE results for different wind speed limit intervals.
+Here instead of training for all data above a certain threshold of wind speed, training is done only on data between two wind speeds.
 The percentage variance in gust factor as a function of wind speed increases with decreasing wind speed.
 Measured wind speed is used for the cutoff and thus have data leakage.
-This results should thus be somewhat comparable to the last column in Table (\ref{table:results})}
+These results should thus be somewhat comparable to the last column in Table (\ref{table:results})}
     \label{table:closed_intervals}
@@
         \toprule
-        \textbf{Interval} &  \multicolumn{2}{c}{\textbf{MAPE}}\\
+        Interval &  \multicolumn{2}{c}{MAPE}\\
         $[m/s]$ & Without Elevation & With Elevation\\
@@
     \caption[Model result by stations of interest]{The MAPE results of different stations for several stations of interest, both when training for the specific site and when the stations are a part of the general data.
-For every station the AWSL is set at 10 m/s.
+For every station the wind speed limit is set at 10 m/s.
 In training for a single station at a time, some site specific information can be gauged.
@@
         \toprule
-        \textbf{Station name} & \multicolumn{2}{c}{\textbf{MAPE}}\\
+        Station name & \multicolumn{2}{c}{MAPE}\\
          & Baseline & Model\\
@@
 Exactly calculating the Shapley values is time intensive, in the subset shown there is an outlier that skews the figure and makes it so that viewing the importance distribution excluding the outlier is difficult.
-For this reason, another shapley summary vas created that looked at different, and larger distribution.
+For this reason, another shapley summary was created that looked at different, and larger distribution.
 This can be seen in Figure (\ref{fig:ShapleySummary2}).
@@
 For simpler models, this would not happen.
-It is important to note that SHAP assumes feature independece\cite{Salih_2024}.
+It is important to note that SHAP assumes feature independence\cite{Salih_2024}.
 This might explain why the impact of the Richardson number is so low.

### 2025-05-05 Chapters/Chapter5.tex
Update Chapter5.tex
Breytti nafni á fimmta kafla.
Bætti Stórhöfða við, uppfærði síðasta kaflann. Tók út * fyrir undirkafla og eitthvað. Notaði svo GPT til þess að benda mér á stafsetningarvillur og uppfærði. Held að þetta sé nokkuð langt komið.

@@
-\chapter{Discussion}
+\chapter{Discussion and conclusions}
 \label{Chapter5}
-The goal of this tesis was to research whether it was possible to use reanalysis data to increase the predictability of wind gusts and see what influence including the landscape elevation would have.
-A quick statistical analysis will give a p-value of 0, that is these results are statistically significant.
-The improvement shown in Table (\ref{table:results}) is not chance.
-There is some pattern to be learned in the interval that leads to a higher predictability of wind gusts over baseline.
-There is also a large variability in the predictability based on weather stations.
-A large part of that can be attributed to the differing amount of data points for each station.
-It would be interesting to know how much of the predictability difference can be explained away with more data for these stations that have few data points or if there is some inherent difference.
-If there is inherent less predictability at those stations, what could be contributing factors in that?
+As shown in Table \ref{table:results}, the neural-network model achieves a significant reduction in mean absolute percentage error (MAPE) compared to a constant baseline.
+Table \ref{table:setsOfParams} further demonstrates a meaningful improvement over a simple regression model.
+A statistical test yields $p<0.01$, confirming that these improvements are unlikely to be due to chance.
+The influence of individual features on model performance is illustrated by Shapley plots (Figure \ref{fig:ShapleySummaryKeflavikurflugvollur} and Appendix \ref{appendix:A}).
+Although one might expect stability indicators—such as the Richardson number and the Brunt--Väisälä frequency—to be among the strongest predictors of gust variability, they contributed less than other variables.
+For reanalysis wind speeds above 10~m/s, the network reduces error by about 30\% relative to the regression baseline.
+The largest gains appear in the 10--25~m/s range (Table \ref{table:closed_intervals}).
+At lower speeds, percentage errors inflate even when absolute gust--wind differences remain small; at higher speeds, sample sizes become sparse.
+
+Predictive skill varies notably from station to station.
+This variability can be partly attributed to discrepancies between reanalysis and observations (Table \ref{table:station_mae_distribution}), differences in local terrain, and the number of measurements per site.
+Training separate models for individual stations did not improve performance (Tables \ref{table:specific_sites} and \ref{table:more_specific_sites}); in fact, some well-sampled stations performed worse when modeled in isolation.
+Future work should investigate whether these anomalies arise from systematic measurement errors, inherent atmospheric variability at those sites, or other factors.
+
+This study did not include station coordinates (XY) as direct inputs.
+Adding location information could improve site-specific accuracy without losing the benefits of training on the full dataset.
+Likewise, instead of applying a fixed linear interpolation to CARRA and elevation grids, the network could receive raw neighboring grid values and their relative positions, allowing it to learn an optimal interpolation scheme.
+
+Finally, treating the data as a time series may offer further gains.
+Here, only instantaneous inputs at forecast time were used, partly because CARRA fields are available every three hours and may not capture rapid stability changes.
+Access to higher-frequency data or the inclusion of temporal history could enable models to exploit evolving atmospheric conditions and further enhance gust-prediction skill.

### 2025-05-05 MSc.tex
Bætti Stórhöfða við, uppfærði síðasta kaflann. Tók út * fyrir undirkafla og eitthvað. Notaði svo GPT til þess að benda mér á stafsetningarvillur og uppfærði. Held að þetta sé nokkuð langt komið.

@@
 \def\thesiscommittee{Kristján Jónasson \\ Ólafur Pétur Pálsson \\ Guðrún Nína Petersen}			% Thesis commitee must include the supervisor(s)
-\def\thesisexaminer{XXNN3XX}				        % Examiner (prófdómari)
+\def\thesisexaminer{Halldór Björnsson}				        % Examiner (prófdómari)
 \def\thesisISBN{}           					% Thesis ISBN number (keep empty: not used anymore)
@@
 
-%%%%%%%%%%% START MODIFYING HERE AGAIN %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-  \newpage % Dedication page: remove completely if you have no dedication
-
-  \thispagestyle{empty} \mbox{}
-
-  \vfill
-
-  \begin{center}
-    \textit{
-      To all the students who made the wise decision to use \LaTeX. % Replace this by your dedication
-    }
-  \end{center} \vspace*{5cm}
-
-  \vfill
-
-%%%%%%%%%%%%%%%%%%%%%%%%% If you have no dedication: remove until here %%%%%%%%%%%%%%%%%%%%
-
 \end{titlepage}
@@
 If looking only at wind, the most destructive moments are gusts, by definition.
-To predict these values there are two ways, traditional numerical weather prediction systems and machine learning or other pattern recognition systems.
-Computer aided numerical weather prediction systems originated in the 1950's, while the use of machine learning in numerical weather predictions is much more recent, having truly gained traction in the last decade.
-The goal of this thesis is try to combine these two methods to try and predict wind gust factor, where the gust factor is defined as the ratio between wind gust and average wind speed.
+To predict these values there are two ways, traditional numerical weather prediction systems and machine learning (or other pattern recognition) methods.
+Computer-aided numerical weather prediction systems originated in the 1950s, while the use of machine learning in numerical weather predictions is much more recent, having gained traction in the last decade.
+The goal of this thesis is to combine these two methods to predict the wind gust factor, defined as the ratio between wind gust and average wind speed.
 This is done by having the input data to the machine learning model be reanalysis variables that were generated by traditional numerical weather prediction systems.
-In addition to these variables and derived variables, a digital elevation model (DEM) data was used.
+In addition to these variables (and derived variables), a digital elevation model (DEM) was used.
 A deep neural network was constructed using some set of variables and the results are compared to a baseline as well as each other so as to show the impact each feature has.
@@
 \section*{Útdráttur}
-Hvort sem um er að ræða vindmyllubúgarða eða flutningar er spágetan um álag vegna vinds mikilvæg til að koma í veg fyrir tjón á iðnviðum auk þess er hægt að auka framleiðni.
+Hvort sem um er að ræða vindmyllubúgarða eða stöðuleiki flutningabíla er spágetan um álag vegna vinds mikilvæg til að koma í veg fyrir tjón á iðnviðum, auk þess er hægt að auka framleiðni.
 Hviðuspár geta verið stór þáttur í tjónaminnkun.
-Ef einungis er horft á vind, þá eru hviður hæsti álagspunktur út frá skilgreiningu.
+Ef einungis er horft á vind, þá eru hviður hæsti álagspunktur samkvæmt skilgreiningu.
 Sögulega hefur verið notast við hefðbundin spálíkön sem reiða sig á töluleg eðlisfræðileg líkön til að spá fyrir um niðurstöður.
-Á síðustu árum hefur gervigreind þróast mikið og líkön verið þjálfuð til þess að giska á veður.
+Á síðustu árum hefur gervigreind þróast mikið og líkön verið þjálfuð til þess að spá fyrir um veður.
 Svo miklar framfarir hafa orðið að sum líkön geta keppt við hefðbundin veðurlíkön.
@@
   \acro{AWS}{Automatic Weather Stations}
-  \acro{AWSL}{Average Wind Speed Limit}
   \acro{CARRA}{Copernicus Artic Regional ReAnalysis dataset}

### 2025-05-06 Appendicies/AppendixA.tex
Forgot to push?

@@
     \includegraphics[scale = 0.6]{Figures/shap_plots/waterfall_plot.png}
-    \caption[Feature importance for a single observation of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+    \caption[Feature importance for a single observation of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
 In this specific instance the wind direction ($wd_{15}$) has the highest negative influence and the temperature has the highest positive influence.
@@
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot.png}
-    \caption[Summary feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+    \caption[Summary feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
 We can see that generally multiple factors influence the prediction, with the station elevation being highly influential.
@@
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_.png}
-    \caption[Summary feature importance of a neural network using a larger distribution of data.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+    \caption[Summary feature importance of a neural network using a larger distribution of data.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
 Generally multiple factors influence the prediction, with the station elevation being highly influential.
@@
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_full_10ms.png}
-    \caption[Summary feature importance of a neural network using entire dataset.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+    \caption[Summary feature importance of a neural network using entire dataset.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
 The distribution seems to be the same as before, discounting the outliers.}
@@
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_31572.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Akrafjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+    \caption[Summary feature importance of a neural network only looking at AWS at Akrafjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
 This plot only looks at datapoints from Akrafjall.
@@
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_35553.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Almannaskarð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+    \caption[Summary feature importance of a neural network only looking at AWS at Almannaskarð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
 This plot only looks at datapoints from Almannaskarð.
@@
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_6745.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Ásgarðsfjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+    \caption[Summary feature importance of a neural network only looking at AWS at Ásgarðsfjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
 This plot only looks at datapoints from Ásgarðsfjall.
@@
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1470.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Háahlíð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+    \caption[Summary feature importance of a neural network only looking at AWS at Háahlíð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
 This plot only looks at datapoints from Háahlíð.
@@
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1350.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Keflavíkurflugvöllur.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparamters} and data as described in Table \ref{table:trainDataExample}.
+    \caption[Summary feature importance of a neural network only looking at AWS at Keflavíkurflugvöllur.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
 This plot only looks at datapoints from Keflavíkurflugvöllur.

### 2025-05-06 Chapters/Chapter1.tex
Forgot to push?

@@
 
-\subsection*{AI in weather forecasts}
+\subsection{AI in weather forecasts}
 In the last decade, there has been another transformation in weather prediction driven by artificial intelligence (AI).
@@
 The evolution from proof-of-concept ML methods to models competitive with—or surpassing—standard NWP has been remarkably rapid.
-It is important to note that training data for these large models are drawn from traditional NWP outputs, underscoring the synergy between ML and physics-based approaches.
-The coming years promise further exciting developments in machine-learning–based weather prediction.
+It is important to note that training data for these large models are drawn from traditional NWP outputs, underscoring how ML and physics-based approaches can complement each other.
+The coming years promise further exciting developments in machine learning based weather prediction.
 

### 2025-05-06 Chapters/Chapter2.tex
Forgot to push?

@@
   \caption[AWS at Stórhöfði in Vestmannaeyjar, south of Iceland.]{AWS at Stórhöfði in Vestmannaeyjar, south of Iceland.
-The anemometer can be seen as three pins in the upper-central of the figure just slightly to the left.}
+The anemometer can be seen as three pins in the upper-central area of the figure just slightly to the left.}
+  \label{fig:storhofdi}
 \end{figure}
 
-Data from these AWS were stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases.
+Data from these AWSs were stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases.
 Each record contains the following information: date and time; station number (convertible to coordinates using another dataset of Icelandic meteorological stations); average wind speed ($f$); wind gust ($f_g$); standard deviation of the wind gust; wind direction ($d$); and standard deviation of the wind direction.
@@
 5) and enhanced by local observations to better represent complex terrain, land–sea contrasts, and sea-ice processes.
-It is updated monthly, with a latency of 2-3 months\cite{carra_information}.
+It is updated monthly, with a latency of 2-3 months \cite{carra_information}.
 

### 2025-05-06 Chapters/Chapter3.tex
Forgot to push?

@@
 As noted, Hyperband does not impose a fixed limit on the total number of training epochs.
-When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband tuner}, the parameter \texttt{Hyperband\_iterations} controls how many times Hyperband is run and defaults to 1.
+When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband tuner}, the Hyperband iterations parameter controls how many times Hyperband is run and defaults to 1.
 On each run, the total number of epochs is distributed across the hyperparameter trials.

### 2025-05-06 Chapters/Chapter4.tex
Forgot to push?

@@
 
-As the average wind speed increases, then the variability in gust as a percentage decreases\cite{mean_gust_HA_HO}.
+As the average wind speed increases, then the variability in gust as a percentage decreases \cite{mean_gust_HA_HO}.
 Therefore lower MAPE would be expected if the training data is restricted to higher wind speeds.
@@
         Baseline constant & 23.9\%\\
-        $ws_{15}$ & 17.2\%\\
+        Simple regression $[ws_{15}]$ & 17.2\%\\
         $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
@@
 This is some improvement upon the baseline error, with a decrease in error from 23.9\% to 13.6\% and 10.6\% for the baseline, model without DEM and with model with DEM for 10 m/s cutoff.
-The power generated by wind mill increases with wind speed cubed\cite{wind_power}.
+The power generated by wind mill increases with wind speed cubed \cite{wind_power}.
 The highest wind gusts in Iceland are around 70 m/s.
@@
 For simpler models, this would not happen.
-It is important to note that SHAP assumes feature independence\cite{Salih_2024}.
+It is important to note that SHAP assumes feature independence \cite{Salih_2024}.
 This might explain why the impact of the Richardson number is so low.

### 2025-05-06 Chapters/Chapter5.tex
Forgot to push?

@@
 As shown in Table \ref{table:results}, the neural-network model achieves a significant reduction in mean absolute percentage error (MAPE) compared to a constant baseline.
-Table \ref{table:setsOfParams} further demonstrates a meaningful improvement over a simple regression model.
+Table \ref{table:setsOfParams} further demonstrates an improvement over a simple regression model.
 A statistical test yields $p<0.01$, confirming that these improvements are unlikely to be due to chance.
@@
 
-Predictive skill varies notably from station to station.
+Predictive skill varies significantly from station to station.
 This variability can be partly attributed to discrepancies between reanalysis and observations (Table \ref{table:station_mae_distribution}), differences in local terrain, and the number of measurements per site.
@@
 Adding location information could improve site-specific accuracy without losing the benefits of training on the full dataset.
-Likewise, instead of applying a fixed linear interpolation to CARRA and elevation grids, the network could receive raw neighboring grid values and their relative positions, allowing it to learn an optimal interpolation scheme.
+Similarly, instead of applying a fixed linear interpolation to CARRA and elevation grids, the network could receive raw neighboring grid values and their relative positions, allowing it to learn an optimal interpolation scheme.
 

### 2025-05-06 MSc.tex
Forgot to push?

@@
-\documentclass[a4paper,12pt,twoside,BCOR=10mm]{scrbook}
+\documentclass[a4paper,12pt,twoside,BCOR=10mm,openany]{scrbook}
 
@@
 \usepackage{acronym}
+% Packages used for header vertical space along with the methods that set the spacing
+% Adjust spacing before and after chapters
+\RedeclareSectionCommand[
+  beforeskip=-0pt, % Space before the chapter title
+  afterskip=0pt    % Space after the chapter title
+]{chapter}
+
+% Adjust spacing before and after sections
+\RedeclareSectionCommand[
+  beforeskip=0pt,   % Space before the section title
+  afterskip=1pt     % Space after the section title
+]{section}
+
+% Adjust spacing before and after subsections
+\RedeclareSectionCommand[
+  beforeskip=0pt,   % Space before the subsection title
+  afterskip=1pt     % Space after the subsection title
+]{subsection}
+
 \usetikzlibrary{positioning}
@@
 \include{Appendicies/AppendixA}
+\include{Appendicies/AppendixB}
 

### 2025-05-06 Chapters/Chapter2.tex
inn í grein 4.3

@@
   \centering
-  \caption[Station MAE extremes]{Mean absolute error (MAE) between reanalysis and observed wind speeds at the five stations with lowest and highest errors.}
+  \caption[Station MAE extremes]{Mean absolute error (MAE) between reanalysis and observed wind speeds at the five stations with lowest and highest errors. The observation counts have been rounded.}
   \label{table:station_mae_distribution}

### 2025-05-06 Chapters/Chapter4.tex
inn í grein 4.3

@@
 
-\section{Modeling for CARRA wind speed above 10 m/s}
+\section{Modeling for ws$_{15}$ above 10 m/s}
 As noted in Chapter \ref{Chapter1}, it is more important to predict accurate gust factor when it is windy than when the weather is calm.
@@
 \begin{table}[h]
-    \caption[Model results for different sets of parameters.]{Showing the results of different models.
-Models defined by different sets of parameters, with increasing complexity.
-This table shows the feature importance as the loss decreases as more variables are added.
-Adding all height levels decreases the loss but very little.}
+  \caption[Model results for different sets of parameters.]{Models defined by a selection of parameter sets, with increasing complexity. The table shows decreasing MAPE as more variables are added.}
     \label{table:setsOfParams}
@@
         \midrule
-        Baseline constant & 23.9\%\\
-        Simple regression $[ws_{15}]$ & 17.2\%\\
+        $[\,]$ Baseline & 23.9\%\\
+        $[ws_{15}]$ Baseline & 17.2\%\\
         $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
@@
 
-The results show this.
-Several different wind speed intervals were used for the baseline model, as can be seen in table \ref{table:results}.
-This sets a goal.
-A model that does not significantly improve on this baseline suggests either failure to capture essential patterns in the data or that the data itself may lack the necessary information for substantial improvements upon the baseline.
-Using the previously described neural network architecture setups for each wind speed interval, with and without landscape elevation information, MAPE was determined.
+\section{Modeling for various ws$_{15}$ lower bounds}
+To assess how model performance depends on ws$_{15}$ lower bound several models were trained using the variable sets in rows 5 and 6 in Table \ref{table:setsOfParams}, i.e.
+$[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$, as well as the same set together with DEM.
 The results can be seen in Table \ref{table:results}.
+For this investigation comparison was only made with the constant baseline model.
+In all cases adding the DEM information is seen to increase the model performance.
+The reason for the difference between the 10 m/s lower bound results seen in Tables \ref{table:setsOfParams} and \ref{table:results} is they were obtained with different training setups.
 
 \begin{table}[h]
-    \caption[Model results for different wind speed limits]{MAPE for each average wind speed limit with and without landscape elevation in a 30° sector around the point of interest into the direction of the reanalysis wind.
-The influence of adding elevation data seems to reduce the error.
-The percentage error is higher for lower wind speeds and thus observing the error for different lower bound of wind speed will produce different results.
-This lower bound is determined using the reanalysis wind speed at 15 meters ($ws_{15}$).}
+  \caption[Model results for different wind speed limits]{MAPE for each ws$_{15}$ lower bound without and with landscape elevation data (DEM) in a 30° sector into the direction of the reanalysis wind.
+Adding DEM data reduces the error noticably.
+Note that the MAPE is higher for both the lower lower bounds and for the higher ones.}
     \label{table:results}
@@
         \toprule
-        \shortstack{Wind Speed\\limit} & MAPE & &\\
-        $[m/s]$ & \shortstack{Baseline\\constant} &  Without DEM & With DEM \\
+        Wind Speed & \multicolumn{3}{c}{MAPE} \\
+        limit $[m/s]$ & Baseline & Without DEM & With DEM \\
         \midrule
@@
 
-%m/s & $ws_{15}$ & f & $ws_{15}$ & f & $ws_{15}$ & f\\
-%$\geq 0$ & 39.2\% & 39.2\% & 19.7\% & 15.9\% & 18.6\% & 15.3\%\\
-%$\geq 5$ & 28.1\% & 14.8\% & 15.9\% & 9.9\% & 14.1\% & 9.4\%\\
-%$\geq 10$ & 23.9\% & 11.1\%& 13.6\% & 7.6\% & 11.5\% & 7.4\%\\
-%$\geq 15$ & 23.2\% & 9.3\% & 13.0\% & 7.0\% & 10.6\% & 6.4\%\\
-%$\geq 20$ & 24.7\% & 8.2\% & 13.6\% & 6.3\% & 11.1\% & 5.8\%\\
-%$\geq 25$ & 27.7\% & 7.3\% & 15.7\% & 5.8\% & 12.9\% & 5.5\% \\
-%7.5 & 25.3\% & 12.6\% & 14.5\% & 8.6\% & 12.5\% & 8.6\%\\
-%12.5 & 23.3\% & 10.1\% & 13.1\% & 7.1\% & 11.0\% & 6.9\%\\
-%17.5 & 23.7\% & 8.7\% & 12.8\% & 6.4\% & 10.5\% & 6.1\%\\
-%22.5 & 26.1\% & 7.8\% & 14.5\% & 6.1\% & 12.6\% & 5.7\%\\
-
-This is some improvement upon the baseline error, with a decrease in error from 23.9\% to 13.6\% and 10.6\% for the baseline, model without DEM and with model with DEM for 10 m/s cutoff.
-The power generated by wind mill increases with wind speed cubed \cite{wind_power}.
-The highest wind gusts in Iceland are around 70 m/s.
-Knowing the gust factor with half as much error as before can allow better anticipation and thus spare turbines for high wind gusts.
-Another way to look at the error improvement is by station.
-No location data was directly included in the training data.
-In Table \ref{table:station_mae_distribution} the mean absolute error of predicted average wind speed and measured average wind speed can be seen for the extreme values.
-A question to ask is it possible to achieve better results when only looking at a single station?
-
 \section{Results for specific locations}
+There are two ways to incorporate location in the model building: To include the station location in the training data, and to train on subsets of the data, e.g.
+by station or groups of stations.
+In this section the latter approach is considered.
+Results of an investigation for a few stations can be seen in Table \ref{table:specific_sites}.
+For many of the stations, the results are worse than expected, calling for further investigation.
 
 \begin{table}[h]
-    \caption[Model result by station]{The MAPE results for selected stations of interest, both when training for the specific site and when the stations are a part of the general data.
-For every station the wind speed limit is set at 10 m/s.
-In training for a single station at a time, some site specific information can be gauged.
-This does not mean that the a better result can be reached for that site.
-Factors such as the number of datapoints at given location can significantly impact the result.
-This table uses the measured wind speed to determine the cut off for data points.
-This leads to some data leakage and an increased performance compared to using the reanalysis CARRA speed for cut off.}
+  \caption[Model result by station]{MAPE for a few selected stations, for a model trained with the whole data set (general training), and for a model trained only with the data of the station itself (site training), with a lower bound of 10 m/s.
+This investigation was done in an earlier phase of the study, when $f$ instead of ws$_{15}$ was being used as the lower bound cut off, and that leads to some data leakage and a somewhat lower MAPE.}
     \label{table:specific_sites}
@@
         \midrule
-        Akrafjall & 42,791 & 18.6\% & 93.7\%\\
-        Almannaskarð & 4,014 & 12.2\% & 86.7\%\\
-        Ásgarðsfjall & 15,121 & 9.1\% & 9.4\%\\
-        Jökulheimar & 17,176 & 7.7\% & 7.7\%\\
-        Sandbúðir & 18,718 & 6.8\% & 6.4\%\\
-        Stórholt & 35,126 & 7.1\% & 29.2\% \\
-        Þúfuver & 19,538 & 6.4\% & 6.8\%\\
+        Akrafjall & 43,000 & 18.6\% & 93.7\%\\
+        Almannaskarð & 4,000 & 12.2\% & 86.7\%\\
+        Ásgarðsfjall & 15,000 & 9.1\% & 9.4\%\\
+        Jökulheimar & 17,000 & 7.7\% & 7.7\%\\
+        Sandbúðir & 19,000 & 6.8\% & 6.4\%\\
+        Stórholt & 35,000 & 7.1\% & 29.2\% \\
+        Þúfuver & 20,000 & 6.4\% & 6.8\%\\
         \bottomrule
@@
 \end{table}
+
+% Færum Mynd 4.1 fram fyrir töflu 4.4.   
+% KOMNIR HINGAÐ
 

### 2025-05-07 Chapters/Chapter4.tex
temp

@@
 
+The resulting model will be referred to as the \textit{primary model} and the corresponding parameters as the \textit{primary parameter set}. The .
+
 \begin{table}[h]
@@
         $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ & 13.7\%\\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ + DEM & 12.0\%\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ + DEM* & 12.0\%\\
         $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, \text{altitude}, twd, N^2, Ri]$ & 12.0\% \\
         \bottomrule
+        *\small{Primary model}
     \end{tabular}
@@
 
-\section{Modeling for various ws$_{15}$ lower bounds}
+\section{Modeling for various ws$_\textbf{15}$ lower bounds}
 To assess how model performance depends on ws$_{15}$ lower bound several models were trained using the variable sets in rows 5 and 6 in Table \ref{table:setsOfParams}, i.e.
-$[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$, as well as the same set together with DEM.
+ws$_{15}$, $t_{15}$, $p_{15}$, wd$_{15}$, altitude, twd, $N^2$, Ri, as well as the same set together with DEM (the primary model).
 The results can be seen in Table \ref{table:results}.
@@
 \section{Results for specific locations}
-There are two ways to incorporate location in the model building: To include the station location in the training data, and to train on subsets of the data, e.g.
+In this section, the distribution of MAPE by station will be considered.
+For each station the primary model MAPE was computed using just the ws$_{15} \geq 10$ data for that station was calculated and the results are shown in Figure (\ref{fig:errorMap}).
+The stations with low performance are all near mountains and often in fjords.
+The worst cases are in Vestfirðir and Austfirðir, and several bad stations are e.g.
+in Tröllaskagi and Snæfellsnes, and near Öræfajökull, Esja, Eyjafjallajökull.
+Inland stations seem to have lower error.
+The worst performing station is Seljalandsdalur in Vestfirðir while the best performing station is Garðskagaviti at the tip of the Reykjanes peninsula.
+Evidently rugged landscape makes it difficult to estimate the gust factor accurately.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale = 0.5]{Figures/errorMap.png}
+\caption[MAPE distribution by station]{Mean absolute percentage error (MAPE) at each station, shown as colored circles (dark blue = low error; yellow = high error).
+The lowest station error is about 7\% at Garðskagaviti and the highest about 35\% at Seljalandsdalur.
+The model uses a 10 m/s lower bound based on ws$_{15}$.}
+\label{fig:errorMap}
+\end{figure}
+
+After examining the MAPE at each station individually, it is logical to try accounting for the station (or its location) in the model’s design.
+There are two ways to do this: To include the station location in the training data, and to train on subsets of the data, e.g.
 by station or groups of stations.
-In this section the latter approach is considered.
+Here the latter approach is considered.
 Results of an investigation for a few stations can be seen in Table \ref{table:specific_sites}.
-For many of the stations, the results are worse than expected, calling for further investigation.
+For many of the stations, the results are worse than expected, sometimes much worse, calling for further investigation.
 
@@
 
-% Færum Mynd 4.1 fram fyrir töflu 4.4.   
-% KOMNIR HINGAÐ
-
-Instead of looking at the exact values of MAPE at select stations, a plot of the error distribution can be created.
-This can be seen in Figure (\ref{fig:errorMap}).
-Looking at Figure (\ref{fig:errorMap}), the worst performing stations can be seen at Vestfirðir and around the coastline.
-Stations further inland seem to have lower error.
-The worst performing station, at Seljalandsdalur, is at Vestfirðir while the best performing station Garðskagaviti is at the South-Western tip of Iceland in Reykjanes.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale = 0.5]{Figures/errorMap.png}
-    \caption[MAPE error distribution of stations shown on a map of Iceland.]{The MAPE error of each station in data shown as color gradient circles.
-That is each station is represented by one circle, with the error value represented as a color gradient from dark blue to yellow.
-The lowest error at a single station was around 7\% at Garðskagaviti and the highest around 35\% at Seljalandsdalur.
-It is important to note that the model is trained using a cutoff of 10 m/s and this cutoff point is determined using the reanalysis wind at 15 meters above ground.}
-    \label{fig:errorMap}
-\end{figure}
+\section{Results for wind speed intervals}
+In addition to training for individual stations one can also try to train for individual wind speed intervals. Table \ref{table:closed_intervals} shows the results of training the primary model
 
@@
 Some interesting sites to look closer at for drivers might include places such as Ingólfsfjall, Kjalarnes and others.
-These can be seen in Table (\ref{table:more_specific_sites}).
+These can be seen in Table \ref{table:more_specific_sites}.
 

### 2025-05-07 Appendicies/AppendixA.tex
búnir með kafla 4

@@
 \setcounter{figure}{0} % Reset figure counter
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/waterfall_plot.png}
-    \caption[Feature importance for a single observation of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-In this specific instance the wind direction ($wd_{15}$) has the highest negative influence and the temperature has the highest positive influence.
-Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them.
-To see their influence on the model output see Table \ref{table:results}.}
-    \label{fig:ShapleyWaterfall}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot.png}
-    \caption[Summary feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-We can see that generally multiple factors influence the prediction, with the station elevation being highly influential.
-There is seemingly one outlier for the Richardson number, which usually has very little influence.
-Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them.
-To see their influence on the model output see Table \ref{table:results}.}
-    %\label{fig:ShapleySummary}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_.png}
-    \caption[Summary feature importance of a neural network using a larger distribution of data.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-Generally multiple factors influence the prediction, with the station elevation being highly influential.
-Elevation data is excluded when working with Shapley values, as the contribution of each elevation.
-In contrast to Figure (\ref{fig:ShapleySummary}), the distribution doesn't have as extreme outliers.
-This means that more details can be seen in the figure.
-The X-axis shows the influence of feature values on the model.
-The color gradient shows the value of each feature.
-As an example, there is a very red value for station elevation (top line, all the way to the left).
-This means that in this instance, the station elevation contributed around -0.25 to the final output and that the station had an elevation significantly above average.}
-    %\label{fig:ShapleySummary2}
-\end{figure}
 
@@
     \label{fig:ShapleySummary3}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_31572.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Akrafjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Akrafjall.
-This seems to show the same distribution as previous summary plots.
-Station elevation is influential and Richardson number has no impact.}
-    \label{fig:ShapleySummaryAkrafjall}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_35553.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Almannaskarð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Almannaskarð.
-This seems to show the same distribution as previous summary plots.
-Station elevation is influential and Richardson number has no impact.}
-    \label{fig:ShapleySummaryAlmannaskarð}
 \end{figure}
@@
 \end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1350.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Keflavíkurflugvöllur.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Keflavíkurflugvöllur.
-This seems to show the same distribution as previous summary plots.
-Station elevation is influential and Richardson number has no impact.}
-    \label{fig:ShapleySummaryKeflavikurflugvollur}
-\end{figure}

### 2025-05-07 Chapters/Chapter1.tex
búnir með kafla 4

@@
 
-\section{Methodology and related work}
+\section{Methodology and related work}\label{sec:methodology}
+
 In this study, data from three sources are used:
@@
 Other methods, like ELI5 (Explain Like I’m 5), randomly shuffle a feature and measure the effect on model performance \cite{eli5_information}.
+
+Apart from aiding in understanding final constructed models and the role of individual features, such explainability measures can also be used to select features to use in the modelling.
+For example, if the computed Shapley distribution of a feature is clustered at 0, then the corresponding feature has little or no influence, and can be safely removed from the model.

### 2025-05-07 Chapters/Chapter2.tex
búnir með kafla 4

@@
 IMO provided measurements from AWSs across Iceland.
-One of the stations used in this study can be seen in Figure (\ref{fig:storhofdi}).
+One of the stations used in this study can be seen in Figure \ref{fig:storhofdi}.
 Not every station had consistently available data.
@@
 
-\section{Layout of combined data}
+\section{Layout of combined data}\label{sec:layout}
 
@@
 
-The Richardson number and the Brunt–Väisälä frequency describe atmospheric stability, see equations (\ref{eqn:Ri}) and (\ref{eqn:N})  \cite{richardson_number_skybrary,brunt_vaisala_freq_eumtrain,mean_gust_HA_HO}.
+The Richardson number and the Brunt–Väisälä frequency describe atmospheric stability, see Equations (\ref{eqn:Ri}) and (\ref{eqn:N}) \cite{richardson_number_skybrary,brunt_vaisala_freq_eumtrain,mean_gust_HA_HO}.
 These values are calculated using reanalysis data at two different height levels.
@@
   \label{eqn:Ri}
-  Ri = \frac{g \cdot dT \cdot dz}{T_{\textrm ave} \cdot dU^2} \unit{[]}
+  Ri = \frac{g \cdot d\theta \cdot dz}{\theta_{\textrm ave} \cdot dU^2} \unit{[]}
 \end{equation}
@@
   \label{eqn:N}
-  N = \sqrt{\frac{g \cdot dT }{T_{\textrm ave} \cdot dz}} \unit{[Hz]}
+  N = \sqrt{\frac{g \cdot d\theta }{\theta_{\textrm ave} \cdot dz}} \unit{[Hz]}
 \end{equation}
 %
-Here, $g$ is gravitational acceleration; $dT$ the temperature difference between levels; $dz$ the height difference; $T_{\mathrm{ave}}$ the average temperature; and $dU$ the wind-speed difference.
+Here, $g$ is gravitational acceleration; $d\theta$ the potential temperature difference between levels; $dz$ the height difference; $\theta_{\mathrm{ave}}$ the average potential temperature; and $dU$ the wind-speed difference.
 Lower Ri indicates greater turbulence, with typical values between 0.1 and 10, and values below 1 signifying significant turbulence \cite{richardson_number_skybrary}.
 Negative $N^2$ denotes instability, as an air parcel will accelerate away from its original position \cite{brunt_vaisala_freq_eumtrain}.
+The potential temperature is computed as:
+%
+\begin{equation}
+  \label{eqn:potential_temp}
+  \theta = T\left(\dfrac{P0}{P}\right)^{R/c_p}
+\end{equation}
+%
+where $P$ is the pressure and $R = 287$ and $c_p = 1004$ are the specifc gas constant and the specific heat capacity of air.
 

### 2025-05-07 Chapters/Chapter4.tex
búnir með kafla 4

@@
 
-\section{Modeling for ws$_{15}$ above 10 m/s}
+\section{Modeling for ws$_{\textbf{15}}$ above 10 m/s}
 As noted in Chapter \ref{Chapter1}, it is more important to predict accurate gust factor when it is windy than when the weather is calm.
@@
 This represents over 30\% reduction in the MAPE from the ws$_{15}$ baseline model.
+Adding all height levels decreases the loss but very little, as seen in the last line of the table.
 
-The resulting model will be referred to as the \textit{primary model} and the corresponding parameters as the \textit{primary parameter set}. The .
+The second last model in the Table \ref{table:setsOfParams} will be referred to as the \textit{primary model} and the corresponding parameters as the \textit{primary parameter set}.
+The model immediately before that is the \textit{secondary model}.
+In some cases the terminology ``primary parameter set with and without DEM'' will be used to refer to the parameter sets of these models.
 
@@
         $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd]$ & 13.8\% \\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ & 13.7\%\\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ + DEM* & 12.0\%\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$* & 13.7\%\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ + DEM** & 12.0\%\\
         $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, \text{altitude}, twd, N^2, Ri]$ & 12.0\% \\
         \bottomrule
-        *\small{Primary model}
+        *\small{Secondary model}
+        **\small{Primary model}
     \end{tabular}
@@
 \section{Modeling for various ws$_\textbf{15}$ lower bounds}
-To assess how model performance depends on ws$_{15}$ lower bound several models were trained using the variable sets in rows 5 and 6 in Table \ref{table:setsOfParams}, i.e.
-ws$_{15}$, $t_{15}$, $p_{15}$, wd$_{15}$, altitude, twd, $N^2$, Ri, as well as the same set together with DEM (the primary model).
+To assess how model performance depends on ws$_{15}$ lower bound several models were trained using the primary variable set with and without DEM.
 The results can be seen in Table \ref{table:results}.
@@
 In this section, the distribution of MAPE by station will be considered.
-For each station the primary model MAPE was computed using just the ws$_{15} \geq 10$ data for that station was calculated and the results are shown in Figure (\ref{fig:errorMap}).
+For each station the primary model MAPE was computed using just the ws$_{15} \geq 10$ data for that station was calculated and the results are shown in Figure \ref{fig:errorMap}.
 The stations with low performance are all near mountains and often in fjords.
@@
 in Tröllaskagi and Snæfellsnes, and near Öræfajökull, Esja, Eyjafjallajökull.
+Table \ref{table:more_specific_sites} shows the primary model MAPE together with the constant baseline reference for a few selected stations.
+
 Inland stations seem to have lower error.
@@
 \end{figure}
+
+\begin{table}[h]
+    \caption[Primary model MAPE by selected stations]{Constant baseline and primary model MAPE for a few selected stations.
+High MAPE values is seen in fjords near tall mountains in Austfirðir (Fáskrúðsfjörður and Seyðisfjörður), intermediate values at the foothills of mountains (Ingólfsfjall and Kjalarnes), and low values in flatter places (Sandskeið, Þrengsli).}
+    \label{table:more_specific_sites}
+    \centering
+    \begin{tabular}{lcc}
+        \toprule
+        Station name & \multicolumn{2}{c}{MAPE}\\
+         & Baseline & Model\\
+        \midrule
+        Fáskrúðsfjörður & 28.2\% & 21.8\%\\
+        Ingólfsfjall & 30.0\% & 19.6\%\\
+        Kjalarnes & 20.7\% & 13.5\% \\
+        Sandskeið & 13.0\% & 10.2\%\\
+        Seyðisfjörður & 32.1\% & 23.0\%\\
+        Þjórsárdalur & 12.2\% & 11.4\%\\
+        Þrengsli & 13.6\% & 11.3\%\\
+        \bottomrule
+    \end{tabular}
+\end{table}
 
@@
 \section{Results for wind speed intervals}
-In addition to training for individual stations one can also try to train for individual wind speed intervals. Table \ref{table:closed_intervals} shows the results of training the primary model
+In addition to training for individual stations one can also try to train for individual wind speed intervals.
+Table \ref{table:closed_intervals} shows the results of training with the primary parameter set, with and without DEM.
+The table does not show noticeable improvement over the values in the last column in Table \ref{table:results}, in fact the two tables show quite similar results.
 
@@
 Measured wind speed is used for the cutoff and thus have data leakage.
-These results should thus be somewhat comparable to the last column in Table (\ref{table:results})}
+These results should thus be somewhat comparable to the last column in Table \ref{table:results}}
     \label{table:closed_intervals}
@@
         Interval &  \multicolumn{2}{c}{MAPE}\\
-        $[m/s]$ & Without Elevation & With Elevation\\
+        $[m/s]$ & Without DEM & With DEM \\
         \midrule
@@
 
-Table (\ref{table:closed_intervals}) shows no improvement over the values in the last column in Table (\ref{table:results}).
-As previously mentioned, the gust factor decreases with increasing wind speed and thus, training on intervals and lessening this effect might be expected to give better results.
-This does not seem to be the case.
-Some interesting sites to look closer at for drivers might include places such as Ingólfsfjall, Kjalarnes and others.
-These can be seen in Table \ref{table:more_specific_sites}.
+\section{Feature importance with Shapley values}
+In this section Shapley values (see section \ref{sec:methodology} and \ref{sec:layout}) for the features of the secondary parameter set will be considered.
+Each feature value in each data row will contribute some share in the final prediction of an NN model, that can be computed by shifting the respective value and observe the change in the model output.
+This procedure is illustrated in Figure \ref{fig:ShapleyWaterfall}.
 
-\begin{table}[h]
-    \caption[Model result by stations of interest]{The MAPE results of different stations for several stations of interest, both when training for the specific site and when the stations are a part of the general data.
-For every station the wind speed limit is set at 10 m/s.
-In training for a single station at a time, some site specific information can be gauged.
-This does not mean that the a better result can be reached for that site.
-Factors such as the number of datapoints at given location can significantly impact the result.
-This table uses the measured wind speed to determine the cut off for data points.
-This leads to some data leakage and an increased performance compared to using the reanalysis CARRA speed for cut off.}
-    \label{table:more_specific_sites}
+\begin{figure}[h]
   \centering
-    \begin{tabular}{lcc}
-        \toprule
-        Station name & \multicolumn{2}{c}{MAPE}\\
-         & Baseline & Model\\
-        \midrule
-        Fáskrúðsfjörður & 28.2\% & 21.8\%\\
-        Ingólfsfjall & 30.0\% & 19.6\%\\
-        Kjalarnes & 20.7\% & 13.5\% \\
-        Sandskeið & 13.0\% & 10.2\%\\
-        Seyðisfjörður & 32.1\% & 23.0\%\\
-        Þjórsárdalur & 12.2\% & 11.4\%\\
-        Þrengsli & 13.6\% & 11.3\%\\
-        \bottomrule
-    \end{tabular}
-\end{table}
+  \includegraphics[scale = 0.6]{Figures/shap_plots/waterfall_plot.png}
+  \caption[Shapley contributions for a single observation]{Shapley contributions for a single data row.
+The figure shows an example of the individual values contributing to a Shapley summary plot.
+Each point in a summary plot corresponds to one such instance.
+Several values with the same SHAP value will result in a thicker bulge in the summary.
+In this specific instance the wind direction ($wd_{15}$) has the highest negative contribution and the temperature has the highest positive contribution.
+The last three features contribute nothing.}
+  \label{fig:ShapleyWaterfall}
+\end{figure}
 
-\section{Comparison for models with variable parameters}
+\subsection{Shapley summary plots for all stations}
 
-In Figure (\ref{fig:ShapleySummary}) the contribution of each feature, excluding the elevation points, can be seen for the model in general (a significant subset of data is used).
-Looking at Figure (\ref{fig:ShapleySummary}), there is an outlier.
-Exactly calculating the Shapley values is time intensive, in the subset shown there is an outlier that skews the figure and makes it so that viewing the importance distribution excluding the outlier is difficult.
-For this reason, another shapley summary was created that looked at different, and larger distribution.
-This can be seen in Figure (\ref{fig:ShapleySummary2}).
-The Figures (\ref{fig:ShapleySummary}) and (\ref{fig:ShapleySummary2}) show the summary for a model trained only on the features shown and not on landscape elevation of the surrounding area.
-This is done as the number of points there is too high to display in one figure (70 total points).
-Looking specifically at Figure (\ref{fig:ShapleySummary2}), the station elevation is most influential and the Richardson number's influence is very low.
-Most of the feature values bunch up in the middle, while the station elevation is elongated compared to other features.
-Station elevation seems to have two bunches on either side of 0.
-Overall, the values seem to be skewed to the right of 0.
-This would be expected as the predicted values are expected to be in the range of 1.2-2, or at least always above 1 by definition.
-Finally, for the Shapley values, looking that all the data there are again outliers that skew the data so that spotting general distribution is difficult.
-This can be seen in Figure (\ref{fig:ShapleySummary3}).
+Figure \ref{fig:ShapleySummary} provides a Shapley summary plot showing the contribution of each feature in the set.
+Computation of these values is expensive so subset of data is used ($\sim$1000 randomly selected data rows).
+There is an obvious outlier in the figure, making it difficult to assess the distribution.
+A second Shapley summary was computed with more data rows ($\sim$10000) and outliers excluded, see the lower panel of the figure.
+Note that it is not feasible to include the DEM data in this analysis, as they contain far too many variables.
+
+Looking specifically at the lower panel in Figure \ref{fig:ShapleySummary}, the station elevation (altitude) is the most influential and the Richardson number's influence is very low, with the other features falling in between.
+For most features the Shapley values are bunch up around 0, while the station elevation, and to some extent ws$_{15}$ have a wider distribution.
+Station altitude/elevation seems to have a wide flat part between –0.1 and 0.1.
+Overall, the values seem to have a longer right tail.
+This is expected as gust factors are positive ($\sim$1.2 to 2).
+Notice also that Ri has virtually no contribution and could safely be removed from the model.
+
+The coloring of the points in the displayed distributions show the corresponding feature values.
+For example, in the lower panel of Figure \ref{figure:ShapleySummary}, high station altitude/elevation corresponds to negative contribution and vice versa, and same for $p_{15}$.
+For ws$_{15}$ and $t_{15}$ the relationship is opposite, with high feature values corresponding to positive contribution.
+When the coloring is more mixed the contribution of the feature is more mixed.
+For example, high twd might have a positve contribution at some locations but negative elsewhere.
 
@@
   \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot.png}
-    \caption[Summary feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-We can see that generally multiple factors influence the prediction, with the station elevation being highly influential.
-There is seemingly one outlier for the Richardson number, which usually has very little influence.
-Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them.
-To see their influence on the model output see Table \ref{table:results}.}
   \label{fig:ShapleySummary}
+  \begin{subfigure}{0.9\textwidth}
+    \includegraphics[width=\textwidth]{Figures/shap_plots/summary_plot.png}
+  \end{subfigure}
+  \vspace{0.5cm}
+  \begin{subfigure}{0.9\textwidth}
+    \includegraphics[width=\textwidth]{Figures/shap_plots/summary_plot_190924_.png}
+  \end{subfigure}
+  \caption[Shapley summary plots]{Shapley summary plots showing feature importance for the secondary model described in Table \ref{table:setsOfParams}, with randomly chosen subsets of the data.
+SHAP values with high absolute value indicate high contribution of the corresponding feature.
+Coloring shows the respective feature value, blue for low and red for high.
+For example high station altitude/elevation corresponds to negative contribution and vice versa, and for ws$_{15}$ the relationship is opposite, with high feature values corresponding to positive contribution.
+It can be seen that generally multiple factors influence the prediction, with the station elevation being highly influential.
+There is one outlier in the upper panel, for the Richardson number, which otherwise has very little influence.
+The upper panel is based on $\sim$1000 data rows.
+The lower panel is based on more data rows ($\sim$10000), and furthermore, outliers have been removed there to make the plot more legible.}
+\end{figure}
+
+Figure \ref{fig:ShapleySummary3} in Appendix A shows a Shapley summary plot for all the data rows. However this plot is detrimentally affected outliers.
+
+\subsection{Shapley summary plots for individual stations}
+It can also be informative to look at summary plots for individual stations.
+In this subsection two such examples are considered, for Akrafjall and Keflavíkurflugvöllur.
+It is noteworthy to see that the constant value of a stations altitude produces quite different contributions to the predictions, presumable because of diffent interactions with other features.
+For (very) simple models, such as linear regression without cross terms, this would not happen.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1350.png}
+    \caption[Summary feature importance of a neural network only looking at AWS at Keflavíkurflugvöllur.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at datapoints from Keflavíkurflugvöllur.
+This seems to show the same distribution as previous summary plots.
+Station elevation is influential and Richardson number has no impact.}
+    \label{fig:ShapleySummaryKeflavikurflugvollur}
 \end{figure}
@@
     \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_.png}
-    \caption[Summary feature importance of a neural network using a larger distribution of data.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-Generally multiple factors influence the prediction, with the station elevation being highly influential.
-Elevation data is excluded when working with Shapley values, as the contribution of each elevation.
-In contrast to Figure (\ref{fig:ShapleySummary}), the distribution doesn't have as extreme outliers.
-This means that more details can be seen in the figure.
-The X-axis shows the influence of feature values on the model.
-The color gradient shows the value of each feature.
-As an example, there is a very red value for station elevation (top line, all the way to the left).
-This means that in this instance, the station elevation contributed around -0.25 to the final output and that the station had an elevation significantly above average.}
-    \label{fig:ShapleySummary2}
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_35553.png}
+    \caption[Summary feature importance of a neural network only looking at AWS at Almannaskarð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at datapoints from Almannaskarð.
+This seems to show the same distribution as previous summary plots.
+Station elevation is influential and Richardson number has no impact.}
+    \label{fig:ShapleySummaryAlmannaskarð}
 \end{figure}
 
-Further Shapley figures can be seen in Appendix~\ref{appendix:A}.
-In each of these plots, even without the feature labels, the station elevation is easily noticed as the value is constant for a station.
-This is not noteworthy.
-What is noteworthy is the range of impact from this single value.
-For simpler models, this would not happen.
-It is important to note that SHAP assumes feature independence \cite{Salih_2024}.
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_31572.png}
+    \caption[Summary feature importance of a neural network only looking at AWS at Akrafjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at datapoints from Akrafjall.
+This seems to show the same distribution as previous summary plots.
+Station elevation is influential and Richardson number has no impact.}
+    \label{fig:ShapleySummaryAkrafjall}
+\end{figure}
+
+Further examples of Shapley summary plots for individual stations may be found in Appendix A.
+
+It is important to note that Shapley summary plot analysis assumes feature independence, because only one feature at a time is changed \cite{Salih_2024}.
 This might explain why the impact of the Richardson number is so low.
-Both the squared Brunt–Väisälä frequency and the Richardson number are derived features from reanalysis data.
-They carry with them some extra information over the other features in the dataset.
-This is because both are variables over elevation ranges.
-That is, as seen in Equations (\ref{eqn:Ri}, \ref{eqn:N}), both are dependent on values at lower and upper elevations and try to describe the stability of that range.
-Shapley tries to assign contribution values for each feature for each observation.
-SHAP assumes that the features are independent, but this is not the case.
-It is clearly not the case for the derived variables, but how the contribution should be distributed between the features is not clear.
-Seemingly the SHAP python package is giving all the impact to the Brunt–Väisälä squared frequency and none to Richardson number.
+Both the squared Brunt–Väisälä frequency and the Richardson number are measures of atmospheric stability and partly depend on the same variables, i.e.
+pressure and temperature at different height levels, as seen in Equations (\ref{eqn:Ri}) and (\ref{eqn:N}).
+Shapley tries to assign contribution values for each feature for each observation, but how the contribution should be distributed between the two features is not clear.
 If the Brunt–Väisälä would be excluded from the data, the impact of the Richardson number would likely increase.
+
 Another point to note is that the features are ordered by their impact.
-This means that the station elevation is the most impactful for each plot, but the ordering of other variables changes.
-Looking at Figure (\ref{fig:ShapleySummary3}), which shows the Shapley summary plot for all data, the wind speed is the second most important feature.
-This is reversed in Figure (\ref{fig:ShapleySummaryAkrafjall}).
-This is not unexpected as Akrafjall station was specifically selected as the MAE for reanalysis wind speed was very high as can be seen in Table (\ref{fig:ShapleySummaryAkrafjall}), where you will also find the stations whose summary plot is shown in Figures (\ref{fig:ShapleySummaryAlmannaskarð}, \ref{fig:ShapleySummaryAsgarðsfjall}) and these also fall into the category of very high MAE for wind speed.
-What is interesting is that the reanalysis wind speed is also of low impact at stations like Háahlíð and Keflavíkurflugvöllur, as shown in Figures (\ref{fig:ShapleySummaryHaahlid}, \ref{fig:ShapleySummaryKeflavikurflugvollur}).
-These stations had the lowest of MAE for difference between measured wind speed and reanalysis wind speed.
-As the summary plot over all stations (Figure (\ref{fig:ShapleySummary3})) shows that reanalysis wind speed is impactful, might lead to the conclusion that the reanalysis wind speed is not a good predictor at these locations or something else is skewing the data.
-A simpler way to look at feature importance is to create models that are trained on and use to predict different sets of parameters.
-The results of such a comparison can be seen in Table \ref{table:setsOfParams}.
+The station altitude is the most impactful for all the plots except Keflavíkurflugvöllur (Figure \ref{fig:ShapleySummaryKeflavikurflugvollur}), but the ordering of other variables changes.
+Looking at the lower panel of Figure \ref{fig:ShapleySummary}, the wind speed is the second most important feature.
+This is reversed in Figure \ref{fig:ShapleySummaryAlmannaskarð}.
+This is not unexpected as the Almannaskarð station was specifically selected as a station with high MAE between observed and reanalysis wind speed (Table \ref{table:station_mae_distribution}.
+Another station with high MAE is Akrafjall, with a Shapley summary plot in Figure \ref{fig:ShapleySummaryAkrafjall}).
+
+What is interesting is that the reanalysis wind speed is also of low impact at stations like Keflavíkurflugvöllur, as shown in Figure \ref{fig:ShapleySummaryKeflavikurflugvollur} which had the second lowest MAE difference between measured wind speed and reanalysis wind speed (see Table \ref{table:station_mae_distribution}).
+Another such station, in fact the one with the lowest MAE, is Háahlíð in Reykjavík.
+The Shapley summary plot for this station is shown in Appendix A (\ref{fig:ShapleySummaryHaahlid}).

### 2025-05-08 Chapters/Chapter1.tex
Updated and added neural network figure.

@@
 The input layer width equals the number of input parameters, and the output layer width equals the number of predictands.
+
+\subsubsection{Structure of neural networks}
+The general structure of a neural network is shown in Figure \ref{fig:nn_structure}.
+The input layer consists of $n$ neurons, each representing an input parameter.
+The output layer has $m$ neurons, each representing a predictand.
+Between these layers are one or more hidden layers, each with $h_i$ neurons, where $i$ is the layer index.
+Each neuron in a layer is connected to every neuron in the previous and next layers.
+The connections between neurons have weights that are adjusted during training to minimize prediction error.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\textwidth]{figures/neural_network.drawio.png}
+    \caption[General structure of a fully connected neural network]{General structure of a fully connected neural network.
+The figure shows a network with three input variables and one output variable.
+It also shows one hidden layer (and others are implied).
+This hidden layer has four neurons that each connect to all input neurons and all connect to all neurons in the next hidden layer (not shown).
+This continues for all hidden layers until the output layer, when there is only one neuron.}
+    \label{fig:nn_structure}
+\end{figure}
+
 Models were created with different parameter sets to gauge each parameter’s influence.

### 2025-05-08 Chapters/Chapter2.tex
Updated and added neural network figure.

@@
 \end{figure}
+
+\section{Model architecture}
+\label{sec:model_architecture}
+
+The network consists of $n$ fully connected layers, with $n$ around 10.
+Each hidden layer has the same number of units and is followed by batch normalization.
+A dropout layer with rate 50\% is applied after the final hidden layer.
+A dense output layer with one unit produces the predicted gust factor.
+A grid search was used to find the hyperparameters that minimize the loss.
+Hyperparameters are settings defined before training begins \cite{hyperparameters_definition}.
+They include the number of units per layer, the number of layers, the number of training epochs, batch size, optimizer choice, and regularization penalty.
+The ranges tested are listed in Table \ref{table:gridSearchHyperparameters}.
+
+\begin{table}[h]
+    \centering
+    \caption[Hyperparameter search with Hyperband]{Hyperparameter search was performed using the Hyperband algorithm, which begins by sampling configurations at random and then focuses on the most promising ones.
+The table shows the ranges tested and the best-performing combination.
+Unlike random search, Hyperband does not impose a fixed limit on the number of trials.}
+    \label{table:gridSearchHyperparameters}
+    \begin{tabular}{lll}
+        \toprule
+        Parameter & Range of values & Selected\\
+        \midrule
+        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{10}\\
+        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{64}\\
+        Penalties & min\_value = 1e-5, max\_value = 1, sampling = log & \textbf{1e-4}\\
+        Epochs & min\_value = 10, max\_value = 1000, step = 10 & \textbf{250}\\
+        Optimizers & Adam, RMSprop, Adamax & \textbf{Adamax}\\
+        Activation & ReLU, ELU, tanh & \textbf{ReLU}\\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+As mentioned, Hyperband doesn't set a hard upper limit to the number of epochs it will train in total.
+When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband class} several factors can be set.
+One of interest here is the Hyperband iterations argument.
+This determines how often the Hyperband algorithm is run and defaults to 1.
+For each iteration the epochs are distributed between tries (that is each set of hyperparameters) with the total amount of epochs approximately $n_{epochs} = max_{epochs} * log^2(max_{epochs})\approx 10^4$, where $max_{epochs} = 1000$ gives the maximum number of epochs that one set of hyperparameters can be trained for.
+Searching a space generally takes a lot of time but this drastically improves on grid search.
+If each epoch takes around 10 seconds to run then the total search would take around 28 hours on a shared resource.
+This is resource-intensive and cannot be repeated often.
+Another question that remains is whether the ranges given are optimal.
+
+As noted, Hyperband does not impose a fixed limit on the total number of training epochs.
+When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband tuner}, the Hyperband iterations parameter controls how many times Hyperband is run and defaults to 1.
+On each run, the total number of epochs is distributed across the hyperparameter trials.
+The total number of epochs is approximately
+%
+\begin{equation*}
+n_{\mathrm{epochs}} = \text{max\_epochs} \times \bigl(\log(\text{max\_epochs})\bigr)^2 \approx 10^4
+\end{equation*}
+%
+for $ \text{max\_epochs} = 1000$.
+Although this method is more efficient than grid search, if each epoch takes about 10 seconds, a complete Hyperband search can still require about 28 hours on shared resources, making repeated searches impractical.
+It remains unclear whether the current hyperparameter ranges are optimal.

### 2025-05-08 Chapters/Chapter3.tex
Updated and added neural network figure.

@@
-% Chapter Template
-
-\chapter{Model Architecture}
-
-\label{Chapter3}
-
-The network consists of $n$ fully connected layers, with $n$ around 10.
-Each hidden layer has the same number of units and is followed by batch normalization.
-A dropout layer with rate 50\% is applied after the final hidden layer.
-A dense output layer with one unit produces the predicted gust factor.
-A grid search was used to find the hyperparameters that minimize the loss.
-Hyperparameters are settings defined before training begins \cite{hyperparameters_definition}.
-They include the number of units per layer, the number of layers, the number of training epochs, batch size, optimizer choice, and regularization penalty.
-The ranges tested are listed in Table \ref{table:gridSearchHyperparameters}.
+\chapter{Results}
+\label{Chapter4}
+\section{Baseline models}
+The simplest model for the gust factor is to guess the average gust factor for all available observations, i.e.
+use no variables from the training data.
+A slightly better model is obtained by using the wind speed as a training feature, either through regression or NN training.
+The MAPEs for the constructed base models are shown in Table \ref{table:baseline_models}.
+The ws$_{15}$ model was obtained with NN training but regression produces an almost identical result.
 
 \begin{table}[h]
+  \caption[MAPE for baseline models]{MAPE for several lower limits on the reanalysis 15 m wind speed ($ws_{15}$). Note that the ws$_{15}$ model was only evaluated for the 10 m/s lower bound.}
+    \label{table:baseline_models}
     \centering
-    \caption[Hyperparameter search with Hyperband]{Hyperparameter search was performed using the Hyperband algorithm, which begins by sampling configurations at random and then focuses on the most promising ones.
-The table shows the ranges tested and the best-performing combination.
-Unlike random search, Hyperband does not impose a fixed limit on the number of trials.}
-    \label{table:gridSearchHyperparameters}
-    \begin{tabular}{lll}
+    \begin{tabular}{lcc}
         \toprule
-        Parameter & Range of values & Selected\\
+        Wind speed & \multicolumn{2}{c}{MAPE}\\ 
+        limit $[m/s]$ & mean model & ws$_{15}$ model\\
         \midrule
-        Layers &  min\_value = 4, max\_value = 15, step = 1 & \textbf{10}\\
-        Units &  min\_value = 32, max\_value = 512, step = 32 & \textbf{64}\\
-        Penalties & min\_value = 1e-5, max\_value = 1, sampling = log & \textbf{1e-4}\\
-        Epochs & min\_value = 10, max\_value = 1000, step = 10 & \textbf{250}\\
-        Optimizers & Adam, RMSprop, Adamax & \textbf{Adamax}\\
-        Activation & ReLU, ELU, tanh & \textbf{ReLU}\\
+        $\geq 0$ & 39.2\%  & \\
+        $\geq 5$ & 28.1\%  & \\
+        $\geq 10$ & 23.9\% & 17.2\%\\
+        $\geq 15$ & 23.2\% & \\
+        $\geq 20$ & 24.7\% & \\
+        $\geq 25$ & 27.7\% & \\
         \bottomrule
@@
 
-As mentioned, Hyperband doesn't set a hard upper limit to the number of epochs it will train in total.
-When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband class} several factors can be set.
-One of interest here is the Hyperband iterations argument.
-This determines how often the Hyperband algorithm is run and defaults to 1.
-For each iteration the epochs are distributed between tries (that is each set of hyperparameters) with the total amount of epochs approximately $n_{epochs} = max_{epochs} * log^2(max_{epochs})\approx 10^4$, where $max_{epochs} = 1000$ gives the maximum number of epochs that one set of hyperparameters can be trained for.
-Searching a space generally takes a lot of time but this drastically improves on grid search.
-If each epoch takes around 10 seconds to run then the total search would take around 28 hours on a shared resource.
-This is resource-intensive and cannot be repeated often.
-Another question that remains is whether the ranges given are optimal.
+As the average wind speed increases, then the variability in gust as a percentage decreases \cite{mean_gust_HA_HO}.
+Therefore lower MAPE would be expected if the training data is restricted to higher wind speeds.
+This fact explains why several models were created for different lower bounds.
 
-As noted, Hyperband does not impose a fixed limit on the total number of training epochs.
-When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband tuner}, the Hyperband iterations parameter controls how many times Hyperband is run and defaults to 1.
-On each run, the total number of epochs is distributed across the hyperparameter trials.
-The total number of epochs is approximately
-%
-\begin{equation*}
-n_{\mathrm{epochs}} = \text{max\_epochs} \times \bigl(\log(\text{max\_epochs})\bigr)^2 \approx 10^4
-\end{equation*}
-%
-for $ \text{max\_epochs} = 1000$.
-Although this method is more efficient than grid search, if each epoch takes about 10 seconds, a complete Hyperband search can still require about 28 hours on shared resources, making repeated searches impractical.
-It remains unclear whether the current hyperparameter ranges are optimal.
+\section{Modeling for ws$_{\textbf{15}}$ above 10 m/s}
+As noted in Chapter \ref{Chapter1}, it is more important to predict accurate gust factor when it is windy than when the weather is calm.
+Therefore, at first, the data is restricted to ws$_{15} \geq 10$ m/s.
+The first constructed model incorporates the features temperature, pressure and wind direction at 15 m height.
+This gives a moderately reduced MAPE of 16.7\% compared with 17.2\% for the baseline ws$_{15}$ model.
+When the next pair of features, station altitude and transformed wind direction, is included, the MAPE is reduced considerably, down to 13.8\%.
+The atmospheric stability indicators, Ri and $N^2$, add little to the model performance, contrary to what might have been expected.
+When the atmosphere is unstable, wind gusts from higher up might be expected.
+However, another considerable performance increase is observed when digital elevation model (DEM) information is included, down to a MAPE of 12.0\%.
+This represents over 30\% reduction in the MAPE from the ws$_{15}$ baseline model.
+Adding all height levels decreases the loss but very little, as seen in the last line of the table.
+
+The second last model in the Table \ref{table:setsOfParams} will be referred to as the \textit{primary model} and the corresponding parameters as the \textit{primary parameter set}.
+The model immediately before that is the \textit{secondary model}.
+In some cases the terminology ``primary parameter set with and without DEM'' will be used to refer to the parameter sets of these models.
+
+\begin{table}[h]
+  \caption[Model results for different sets of parameters.]{Models defined by a selection of parameter sets, with increasing complexity. The table shows decreasing MAPE as more variables are added.}
+    \label{table:setsOfParams}
+    \centering
+    \begin{tabular}{lc}
+        \toprule
+        Model variables & MAPE\\
+        \midrule
+        $[\,]$ Baseline & 23.9\%\\
+        $[ws_{15}]$ Baseline & 17.2\%\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd]$ & 13.8\% \\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$* & 13.7\%\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ + DEM** & 12.0\%\\
+        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, \text{altitude}, twd, N^2, Ri]$ & 12.0\% \\
+        \bottomrule
+        *\small{Secondary model}
+        **\small{Primary model}
+    \end{tabular}
+\end{table}
+
+\section{Modeling for various ws$_\textbf{15}$ lower bounds}
+To assess how model performance depends on ws$_{15}$ lower bound several models were trained using the primary variable set with and without DEM.
+The results can be seen in Table \ref{table:results}.
+For this investigation comparison was only made with the constant baseline model.
+In all cases adding the DEM information is seen to increase the model performance.
+The reason for the difference between the 10 m/s lower bound results seen in Tables \ref{table:setsOfParams} and \ref{table:results} is they were obtained with different training setups.
+
+\begin{table}[h]
+  \caption[Model results for different wind speed limits]{MAPE for each ws$_{15}$ lower bound without and with landscape elevation data (DEM) in a 30° sector into the direction of the reanalysis wind.
+Adding DEM data reduces the error noticably.
+Note that the MAPE is higher for both the lower lower bounds and for the higher ones.}
+    \label{table:results}
+    \centering
+    \begin{tabular}{lccc}
+        \toprule
+        Wind Speed & \multicolumn{3}{c}{MAPE} \\
+        limit $[m/s]$ & Baseline & Without DEM & With DEM \\
+        \midrule
+        $\geq 0$ & 39.2\% & 19.2\% & 18.9\% \\
+        $\geq 5$ & 28.1\% & 15.3\% & 14.9\%\\
+        $\geq 10$ & 23.9\% & 13.3\% & 12.5\%\\
+        $\geq 15$ & 23.2\% & 14.4\% & 11.9\%\\
+        $\geq 20$ & 24.7\% & 15.7\% & 13.3\%\\
+        $\geq 25$ & 27.7\% & 19.4\% & 17.3\%\\
+        \bottomrule
+    \end{tabular}
+  \end{table}
+
+\section{Results for specific locations}
+In this section, the distribution of MAPE by station will be considered.
+For each station the primary model MAPE was computed using just the ws$_{15} \geq 10$ data for that station was calculated and the results are shown in Figure \ref{fig:errorMap}.
+The stations with low performance are all near mountains and often in fjords.
+The worst cases are in Vestfirðir and Austfirðir, and several bad stations are e.g.
+in Tröllaskagi and Snæfellsnes, and near Öræfajökull, Esja, Eyjafjallajökull.
+Table \ref{table:more_specific_sites} shows the primary model MAPE together with the constant baseline reference for a few selected stations.
+
+Inland stations seem to have lower error.
+The worst performing station is Seljalandsdalur in Vestfirðir while the best performing station is Garðskagaviti at the tip of the Reykjanes peninsula.
+Evidently rugged landscape makes it difficult to estimate the gust factor accurately.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale = 0.5]{Figures/errorMap.png}
+\caption[MAPE distribution by station]{Mean absolute percentage error (MAPE) at each station, shown as colored circles (dark blue = low error; yellow = high error).
+The lowest station error is about 7\% at Garðskagaviti and the highest about 35\% at Seljalandsdalur.
+The model uses a 10 m/s lower bound based on ws$_{15}$.}
+\label{fig:errorMap}
+\end{figure}
+
+\begin{table}[h]
+    \caption[Primary model MAPE by selected stations]{Constant baseline and primary model MAPE for a few selected stations.
+High MAPE values is seen in fjords near tall mountains in Austfirðir (Fáskrúðsfjörður and Seyðisfjörður), intermediate values at the foothills of mountains (Ingólfsfjall and Kjalarnes), and low values in flatter places (Sandskeið, Þrengsli).}
+    \label{table:more_specific_sites}
+    \centering
+    \begin{tabular}{lcc}
+        \toprule
+        Station name & \multicolumn{2}{c}{MAPE}\\
+         & Baseline & Model\\
+        \midrule
+        Fáskrúðsfjörður & 28.2\% & 21.8\%\\
+        Ingólfsfjall & 30.0\% & 19.6\%\\
+        Kjalarnes & 20.7\% & 13.5\% \\
+        Sandskeið & 13.0\% & 10.2\%\\
+        Seyðisfjörður & 32.1\% & 23.0\%\\
+        Þjórsárdalur & 12.2\% & 11.4\%\\
+        Þrengsli & 13.6\% & 11.3\%\\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+After examining the MAPE at each station individually, it is logical to try accounting for the station (or its location) in the model’s design.
+There are two ways to do this: To include the station location in the training data, and to train on subsets of the data, e.g.
+by station or groups of stations.
+Here the latter approach is considered.
+Results of an investigation for a few stations can be seen in Table \ref{table:specific_sites}.
+For many of the stations, the results are worse than expected, sometimes much worse, calling for further investigation.
+
+\begin{table}[h]
+  \caption[Model result by station]{MAPE for a few selected stations, for a model trained with the whole data set (general training), and for a model trained only with the data of the station itself (site training), with a lower bound of 10 m/s.
+This investigation was done in an earlier phase of the study, when $f$ instead of ws$_{15}$ was being used as the lower bound cut off, and that leads to some data leakage and a somewhat lower MAPE.}
+    \label{table:specific_sites}
+    \centering
+    \begin{tabular}{lccc}
+        \toprule
+        Station name & n & \multicolumn{2}{c}{MAPE}\\
+        & & General training & Site training\\
+        \midrule
+        Akrafjall & 43,000 & 18.6\% & 93.7\%\\
+        Almannaskarð & 4,000 & 12.2\% & 86.7\%\\
+        Ásgarðsfjall & 15,000 & 9.1\% & 9.4\%\\
+        Jökulheimar & 17,000 & 7.7\% & 7.7\%\\
+        Sandbúðir & 19,000 & 6.8\% & 6.4\%\\
+        Stórholt & 35,000 & 7.1\% & 29.2\% \\
+        Þúfuver & 20,000 & 6.4\% & 6.8\%\\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+\section{Results for wind speed intervals}
+In addition to training for individual stations one can also try to train for individual wind speed intervals.
+Table \ref{table:closed_intervals} shows the results of training with the primary parameter set, with and without DEM.
+The table does not show noticeable improvement over the values in the last column in Table \ref{table:results}, in fact the two tables show quite similar results.
+
+\begin{table}[h]
+    \caption[Model result looking at closed wind speed intervals]{The MAPE results for different wind speed limit intervals.
+Here instead of training for all data above a certain threshold of wind speed, training is done only on data between two wind speeds.
+The percentage variance in gust factor as a function of wind speed increases with decreasing wind speed.
+Measured wind speed is used for the cutoff and thus have data leakage.
+These results should thus be somewhat comparable to the last column in Table \ref{table:results}}
+    \label{table:closed_intervals}
+    \centering
+    \begin{tabular}{lcc}
+        \toprule
+        Interval &  \multicolumn{2}{c}{MAPE}\\
+        $[m/s]$ & Without DEM & With DEM \\
+        \midrule
+        $[5, 10[$ & 17.1\% & 16.4\%\\
+        $[10, 15[$ & 14.5\% & 13.0\%\\
+        $[15, 20[$ & 15.0\% & 12.0\%\\
+        $[20, 25[$ & 15.6 \% & 13.1\%\\
+        $[25, 30[$ & 18.4\% & 19.0\%\\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+\section{Feature importance with Shapley values}
+In this section Shapley values (see section \ref{sec:methodology} and \ref{sec:layout}) for the features of the secondary parameter set will be considered.
+Each feature value in each data row will contribute some share in the final prediction of an NN model, that can be computed by shifting the respective value and observe the change in the model output.
+This procedure is illustrated in Figure \ref{fig:ShapleyWaterfall}.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[scale = 0.6]{Figures/shap_plots/waterfall_plot.png}
+  \caption[Shapley contributions for a single observation]{Shapley contributions for a single data row.
+The figure shows an example of the individual values contributing to a Shapley summary plot.
+Each point in a summary plot corresponds to one such instance.
+Several values with the same SHAP value will result in a thicker bulge in the summary.
+In this specific instance the wind direction ($wd_{15}$) has the highest negative contribution and the temperature has the highest positive contribution.
+The last three features contribute nothing.}
+  \label{fig:ShapleyWaterfall}
+\end{figure}
+
+\subsection{Shapley summary plots for all stations}
+
+Figure \ref{fig:ShapleySummary} provides a Shapley summary plot showing the contribution of each feature in the set.
+Computation of these values is expensive so subset of data is used ($\sim$1000 randomly selected data rows).
+There is an obvious outlier in the figure, making it difficult to assess the distribution.
+A second Shapley summary was computed with more data rows ($\sim$10000) and outliers excluded, see the lower panel of the figure.
+Note that it is not feasible to include the DEM data in this analysis, as they contain far too many variables.
+
+Looking specifically at the lower panel in Figure \ref{fig:ShapleySummary}, the station elevation (altitude) is the most influential and the Richardson number's influence is very low, with the other features falling in between.
+For most features the Shapley values are bunch up around 0, while the station elevation, and to some extent ws$_{15}$ have a wider distribution.
+Station altitude/elevation seems to have a wide flat part between –0.1 and 0.1.
+Overall, the values seem to have a longer right tail.
+This is expected as gust factors are positive ($\sim$1.2 to 2).
+Notice also that Ri has virtually no contribution and could safely be removed from the model.
+
+The coloring of the points in the displayed distributions show the corresponding feature values.
+For example, in the lower panel of Figure \ref{figure:ShapleySummary}, high station altitude/elevation corresponds to negative contribution and vice versa, and same for $p_{15}$.
+For ws$_{15}$ and $t_{15}$ the relationship is opposite, with high feature values corresponding to positive contribution.
+When the coloring is more mixed the contribution of the feature is more mixed.
+For example, high twd might have a positve contribution at some locations but negative elsewhere.
+
+\begin{figure}
+  \centering
+  \begin{subfigure}{0.9\textwidth}
+    \includegraphics[width=\textwidth]{Figures/shap_plots/summary_plot.png}
+  \end{subfigure}
+  \vspace{0.5cm}
+  \begin{subfigure}{0.9\textwidth}
+    \includegraphics[width=\textwidth]{Figures/shap_plots/summary_plot_190924_.png}
+  \end{subfigure}
+  \caption[Shapley summary plots]{Shapley summary plots showing feature importance for the secondary model described in Table \ref{table:setsOfParams}, with randomly chosen subsets of the data.
+SHAP values with high absolute value indicate high contribution of the corresponding feature.
+Coloring shows the respective feature value, blue for low and red for high.
+For example high station altitude/elevation corresponds to negative contribution and vice versa, and for ws$_{15}$ the relationship is opposite, with high feature values corresponding to positive contribution.
+It can be seen that generally multiple factors influence the prediction, with the station elevation being highly influential.
+There is one outlier in the upper panel, for the Richardson number, which otherwise has very little influence.
+The upper panel is based on $\sim$1000 data rows.
+The lower panel is based on more data rows ($\sim$10000), and furthermore, outliers have been removed there to make the plot more legible.}
+  \label{fig:ShapleySummary}
+\end{figure}
+
+Figure \ref{fig:ShapleySummary3} in Appendix A shows a Shapley summary plot for all the data rows. However this plot is detrimentally affected outliers.
+
+\subsection{Shapley summary plots for individual stations}
+It can also be informative to look at summary plots for individual stations.
+In this subsection two such examples are considered, for Akrafjall and Keflavíkurflugvöllur.
+It is noteworthy to see that the constant value of a stations altitude produces quite different contributions to the predictions, presumable because of diffent interactions with other features.
+For (very) simple models, such as linear regression without cross terms, this would not happen.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1350.png}
+    \caption[Summary feature importance of a neural network only looking at AWS at Keflavíkurflugvöllur.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at datapoints from Keflavíkurflugvöllur.
+This seems to show the same distribution as previous summary plots.
+Station elevation is influential and Richardson number has no impact.}
+    \label{fig:ShapleySummaryKeflavikurflugvollur}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_35553.png}
+    \caption[Summary feature importance of a neural network only looking at AWS at Almannaskarð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at datapoints from Almannaskarð.
+This seems to show the same distribution as previous summary plots.
+Station elevation is influential and Richardson number has no impact.}
+    \label{fig:ShapleySummaryAlmannaskarð}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_31572.png}
+    \caption[Summary feature importance of a neural network only looking at AWS at Akrafjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at datapoints from Akrafjall.
+This seems to show the same distribution as previous summary plots.
+Station elevation is influential and Richardson number has no impact.}
+    \label{fig:ShapleySummaryAkrafjall}
+\end{figure}
+
+Further examples of Shapley summary plots for individual stations may be found in Appendix A.
+
+It is important to note that Shapley summary plot analysis assumes feature independence, because only one feature at a time is changed \cite{Salih_2024}.
+This might explain why the impact of the Richardson number is so low.
+Both the squared Brunt–Väisälä frequency and the Richardson number are measures of atmospheric stability and partly depend on the same variables, i.e.
+pressure and temperature at different height levels, as seen in Equations (\ref{eqn:Ri}) and (\ref{eqn:N}).
+Shapley tries to assign contribution values for each feature for each observation, but how the contribution should be distributed between the two features is not clear.
+If the Brunt–Väisälä would be excluded from the data, the impact of the Richardson number would likely increase.
+
+Another point to note is that the features are ordered by their impact.
+The station altitude is the most impactful for all the plots except Keflavíkurflugvöllur (Figure \ref{fig:ShapleySummaryKeflavikurflugvollur}), but the ordering of other variables changes.
+Looking at the lower panel of Figure \ref{fig:ShapleySummary}, the wind speed is the second most important feature.
+This is reversed in Figure \ref{fig:ShapleySummaryAlmannaskarð}.
+This is not unexpected as the Almannaskarð station was specifically selected as a station with high MAE between observed and reanalysis wind speed (Table \ref{table:station_mae_distribution}.
+Another station with high MAE is Akrafjall, with a Shapley summary plot in Figure \ref{fig:ShapleySummaryAkrafjall}).
+
+What is interesting is that the reanalysis wind speed is also of low impact at stations like Keflavíkurflugvöllur, as shown in Figure \ref{fig:ShapleySummaryKeflavikurflugvollur} which had the second lowest MAE difference between measured wind speed and reanalysis wind speed (see Table \ref{table:station_mae_distribution}).
+Another such station, in fact the one with the lowest MAE, is Háahlíð in Reykjavík.
+The Shapley summary plot for this station is shown in Appendix A (\ref{fig:ShapleySummaryHaahlid}).

### 2025-05-08 Chapters/Chapter4.tex
Updated and added neural network figure.

--- 2025-05-06 Chapters/Chapter4.tex
@@
-\chapter{Results}
-\label{Chapter4}
-\section{Baseline models}
-The simplest model for the gust factor is to guess the average gust factor for all available observations, i.e.
-use no variables from the training data.
-A slightly better model is obtained by using the wind speed as a training feature, either through regression or NN training.
-The MAPEs for the constructed base models are shown in Table \ref{table:baseline_models}.
-The ws$_{15}$ model was obtained with NN training but regression produces an almost identical result.
+\chapter{Discussion and conclusions}
+\label{Chapter5}
+As shown in Table \ref{table:results}, the neural-network model achieves a significant reduction in mean absolute percentage error (MAPE) compared to a constant baseline.
+Table \ref{table:setsOfParams} further demonstrates an improvement over a simple regression model.
+A statistical test yields $p<0.01$, confirming that these improvements are unlikely to be due to chance.
+The influence of individual features on model performance is illustrated by Shapley plots (Figure \ref{fig:ShapleySummaryKeflavikurflugvollur} and Appendix \ref{appendix:A}).
+Although one might expect stability indicators—such as the Richardson number and the Brunt--Väisälä frequency—to be among the strongest predictors of gust variability, they contributed less than other variables.
+For reanalysis wind speeds above 10~m/s, the network reduces error by about 30\% relative to the regression baseline.
+The largest gains appear in the 10--25~m/s range (Table \ref{table:closed_intervals}).
+At lower speeds, percentage errors inflate even when absolute gust--wind differences remain small; at higher speeds, sample sizes become sparse.
 
-\begin{table}[h]
-  \caption[MAPE for baseline models]{MAPE for several lower limits on the reanalysis 15 m wind speed ($ws_{15}$). Note that the ws$_{15}$ model was only evaluated for the 10 m/s lower bound.}
-    \label{table:baseline_models}
-    \centering
-    \begin{tabular}{lcc}
-        \toprule
-        Wind speed & \multicolumn{2}{c}{MAPE}\\ 
-        limit $[m/s]$ & mean model & ws$_{15}$ model\\
-        \midrule
-        $\geq 0$ & 39.2\%  & \\
-        $\geq 5$ & 28.1\%  & \\
-        $\geq 10$ & 23.9\% & 17.2\%\\
-        $\geq 15$ & 23.2\% & \\
-        $\geq 20$ & 24.7\% & \\
-        $\geq 25$ & 27.7\% & \\
-        \bottomrule
-    \end{tabular}
-\end{table}
+Predictive skill varies significantly from station to station.
+This variability can be partly attributed to discrepancies between reanalysis and observations (Table \ref{table:station_mae_distribution}), differences in local terrain, and the number of measurements per site.
+Training separate models for individual stations did not improve performance (Tables \ref{table:specific_sites} and \ref{table:more_specific_sites}); in fact, some well-sampled stations performed worse when modeled in isolation.
+Future work should investigate whether these anomalies arise from systematic measurement errors, inherent atmospheric variability at those sites, or other factors.
 
-As the average wind speed increases, then the variability in gust as a percentage decreases \cite{mean_gust_HA_HO}.
-Therefore lower MAPE would be expected if the training data is restricted to higher wind speeds.
-This fact explains why several models were created for different lower bounds.
+This study did not include station coordinates (XY) as direct inputs.
+Adding location information could improve site-specific accuracy without losing the benefits of training on the full dataset.
+Similarly, instead of applying a fixed linear interpolation to CARRA and elevation grids, the network could receive raw neighboring grid values and their relative positions, allowing it to learn an optimal interpolation scheme.
 
-\section{Modeling for ws$_{15}$ above 10 m/s}
-As noted in Chapter \ref{Chapter1}, it is more important to predict accurate gust factor when it is windy than when the weather is calm.
-Therefore, at first, the data is restricted to ws$_{15} \geq 10$ m/s.
-The first constructed model incorporates the features temperature, pressure and wind direction at 15 m height.
-This gives a moderately reduced MAPE of 16.7\% compared with 17.2\% for the baseline ws$_{15}$ model.
-When the next pair of features, station altitude and transformed wind direction, is included, the MAPE is reduced considerably, down to 13.8\%.
-The atmospheric stability indicators, Ri and $N^2$, add little to the model performance, contrary to what might have been expected.
-When the atmosphere is unstable, wind gusts from higher up might be expected.
-However, another considerable performance increase is observed when digital elevation model (DEM) information is included, down to a MAPE of 12.0\%.
-This represents over 30\% reduction in the MAPE from the ws$_{15}$ baseline model.
+Finally, treating the data as a time series may offer further gains.
+Here, only instantaneous inputs at forecast time were used, partly because CARRA fields are available every three hours and may not capture rapid stability changes.
+Access to higher-frequency data or the inclusion of temporal history could enable models to exploit evolving atmospheric conditions and further enhance gust-prediction skill.
 
-\begin{table}[h]
-  \caption[Model results for different sets of parameters.]{Models defined by a selection of parameter sets, with increasing complexity. The table shows decreasing MAPE as more variables are added.}
-    \label{table:setsOfParams}
-    \centering
-    \begin{tabular}{lc}
-        \toprule
-        Model variables & MAPE\\
-        \midrule
-        $[\,]$ Baseline & 23.9\%\\
-        $[ws_{15}]$ Baseline & 17.2\%\\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd]$ & 13.8\% \\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ & 13.7\%\\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ + DEM & 12.0\%\\
-        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, \text{altitude}, twd, N^2, Ri]$ & 12.0\% \\
-        \bottomrule
-    \end{tabular}
-\end{table}
-
-\section{Modeling for various ws$_{15}$ lower bounds}
-To assess how model performance depends on ws$_{15}$ lower bound several models were trained using the variable sets in rows 5 and 6 in Table \ref{table:setsOfParams}, i.e.
-$[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$, as well as the same set together with DEM.
-The results can be seen in Table \ref{table:results}.
-For this investigation comparison was only made with the constant baseline model.
-In all cases adding the DEM information is seen to increase the model performance.
-The reason for the difference between the 10 m/s lower bound results seen in Tables \ref{table:setsOfParams} and \ref{table:results} is they were obtained with different training setups.
-
-\begin{table}[h]
-  \caption[Model results for different wind speed limits]{MAPE for each ws$_{15}$ lower bound without and with landscape elevation data (DEM) in a 30° sector into the direction of the reanalysis wind.
-Adding DEM data reduces the error noticably.
-Note that the MAPE is higher for both the lower lower bounds and for the higher ones.}
-    \label{table:results}
-    \centering
-    \begin{tabular}{lccc}
-        \toprule
-        Wind Speed & \multicolumn{3}{c}{MAPE} \\
-        limit $[m/s]$ & Baseline & Without DEM & With DEM \\
-        \midrule
-        $\geq 0$ & 39.2\% & 19.2\% & 18.9\% \\
-        $\geq 5$ & 28.1\% & 15.3\% & 14.9\%\\
-        $\geq 10$ & 23.9\% & 13.3\% & 12.5\%\\
-        $\geq 15$ & 23.2\% & 14.4\% & 11.9\%\\
-        $\geq 20$ & 24.7\% & 15.7\% & 13.3\%\\
-        $\geq 25$ & 27.7\% & 19.4\% & 17.3\%\\
-        \bottomrule
-    \end{tabular}
-  \end{table}
-
-\section{Results for specific locations}
-There are two ways to incorporate location in the model building: To include the station location in the training data, and to train on subsets of the data, e.g.
-by station or groups of stations.
-In this section the latter approach is considered.
-Results of an investigation for a few stations can be seen in Table \ref{table:specific_sites}.
-For many of the stations, the results are worse than expected, calling for further investigation.
-
-\begin{table}[h]
-  \caption[Model result by station]{MAPE for a few selected stations, for a model trained with the whole data set (general training), and for a model trained only with the data of the station itself (site training), with a lower bound of 10 m/s.
-This investigation was done in an earlier phase of the study, when $f$ instead of ws$_{15}$ was being used as the lower bound cut off, and that leads to some data leakage and a somewhat lower MAPE.}
-    \label{table:specific_sites}
-    \centering
-    \begin{tabular}{lccc}
-        \toprule
-        Station name & n & \multicolumn{2}{c}{MAPE}\\
-        & & General training & Site training\\
-        \midrule
-        Akrafjall & 43,000 & 18.6\% & 93.7\%\\
-        Almannaskarð & 4,000 & 12.2\% & 86.7\%\\
-        Ásgarðsfjall & 15,000 & 9.1\% & 9.4\%\\
-        Jökulheimar & 17,000 & 7.7\% & 7.7\%\\
-        Sandbúðir & 19,000 & 6.8\% & 6.4\%\\
-        Stórholt & 35,000 & 7.1\% & 29.2\% \\
-        Þúfuver & 20,000 & 6.4\% & 6.8\%\\
-        \bottomrule
-    \end{tabular}
-\end{table}
-
-% Færum Mynd 4.1 fram fyrir töflu 4.4.   
-% KOMNIR HINGAÐ
-
-Instead of looking at the exact values of MAPE at select stations, a plot of the error distribution can be created.
-This can be seen in Figure (\ref{fig:errorMap}).
-Looking at Figure (\ref{fig:errorMap}), the worst performing stations can be seen at Vestfirðir and around the coastline.
-Stations further inland seem to have lower error.
-The worst performing station, at Seljalandsdalur, is at Vestfirðir while the best performing station Garðskagaviti is at the South-Western tip of Iceland in Reykjanes.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale = 0.5]{Figures/errorMap.png}
-    \caption[MAPE error distribution of stations shown on a map of Iceland.]{The MAPE error of each station in data shown as color gradient circles.
-That is each station is represented by one circle, with the error value represented as a color gradient from dark blue to yellow.
-The lowest error at a single station was around 7\% at Garðskagaviti and the highest around 35\% at Seljalandsdalur.
-It is important to note that the model is trained using a cutoff of 10 m/s and this cutoff point is determined using the reanalysis wind at 15 meters above ground.}
-    \label{fig:errorMap}
-\end{figure}
-
-\begin{table}[h]
-    \caption[Model result looking at closed wind speed intervals]{The MAPE results for different wind speed limit intervals.
-Here instead of training for all data above a certain threshold of wind speed, training is done only on data between two wind speeds.
-The percentage variance in gust factor as a function of wind speed increases with decreasing wind speed.
-Measured wind speed is used for the cutoff and thus have data leakage.
-These results should thus be somewhat comparable to the last column in Table (\ref{table:results})}
-    \label{table:closed_intervals}
-    \centering
-    \begin{tabular}{lcc}
-        \toprule
-        Interval &  \multicolumn{2}{c}{MAPE}\\
-        $[m/s]$ & Without Elevation & With Elevation\\
-        \midrule
-        $[5, 10[$ & 17.1\% & 16.4\%\\
-        $[10, 15[$ & 14.5\% & 13.0\%\\
-        $[15, 20[$ & 15.0\% & 12.0\%\\
-        $[20, 25[$ & 15.6 \% & 13.1\%\\
-        $[25, 30[$ & 18.4\% & 19.0\%\\
-        \bottomrule
-    \end{tabular}
-\end{table}
-
-Table (\ref{table:closed_intervals}) shows no improvement over the values in the last column in Table (\ref{table:results}).
-As previously mentioned, the gust factor decreases with increasing wind speed and thus, training on intervals and lessening this effect might be expected to give better results.
-This does not seem to be the case.
-Some interesting sites to look closer at for drivers might include places such as Ingólfsfjall, Kjalarnes and others.
-These can be seen in Table (\ref{table:more_specific_sites}).
-
-\begin{table}[h]
-    \caption[Model result by stations of interest]{The MAPE results of different stations for several stations of interest, both when training for the specific site and when the stations are a part of the general data.
-For every station the wind speed limit is set at 10 m/s.
-In training for a single station at a time, some site specific information can be gauged.
-This does not mean that the a better result can be reached for that site.
-Factors such as the number of datapoints at given location can significantly impact the result.
-This table uses the measured wind speed to determine the cut off for data points.
-This leads to some data leakage and an increased performance compared to using the reanalysis CARRA speed for cut off.}
-    \label{table:more_specific_sites}
-    \centering
-    \begin{tabular}{lcc}
-        \toprule
-        Station name & \multicolumn{2}{c}{MAPE}\\
-         & Baseline & Model\\
-        \midrule
-        Fáskrúðsfjörður & 28.2\% & 21.8\%\\
-        Ingólfsfjall & 30.0\% & 19.6\%\\
-        Kjalarnes & 20.7\% & 13.5\% \\
-        Sandskeið & 13.0\% & 10.2\%\\
-        Seyðisfjörður & 32.1\% & 23.0\%\\
-        Þjórsárdalur & 12.2\% & 11.4\%\\
-        Þrengsli & 13.6\% & 11.3\%\\
-        \bottomrule
-    \end{tabular}
-\end{table}
-
-\section{Comparison for models with variable parameters}
-
-In Figure (\ref{fig:ShapleySummary}) the contribution of each feature, excluding the elevation points, can be seen for the model in general (a significant subset of data is used).
-Looking at Figure (\ref{fig:ShapleySummary}), there is an outlier.
-Exactly calculating the Shapley values is time intensive, in the subset shown there is an outlier that skews the figure and makes it so that viewing the importance distribution excluding the outlier is difficult.
-For this reason, another shapley summary was created that looked at different, and larger distribution.
-This can be seen in Figure (\ref{fig:ShapleySummary2}).
-The Figures (\ref{fig:ShapleySummary}) and (\ref{fig:ShapleySummary2}) show the summary for a model trained only on the features shown and not on landscape elevation of the surrounding area.
-This is done as the number of points there is too high to display in one figure (70 total points).
-Looking specifically at Figure (\ref{fig:ShapleySummary2}), the station elevation is most influential and the Richardson number's influence is very low.
-Most of the feature values bunch up in the middle, while the station elevation is elongated compared to other features.
-Station elevation seems to have two bunches on either side of 0.
-Overall, the values seem to be skewed to the right of 0.
-This would be expected as the predicted values are expected to be in the range of 1.2-2, or at least always above 1 by definition.
-Finally, for the Shapley values, looking that all the data there are again outliers that skew the data so that spotting general distribution is difficult.
-This can be seen in Figure (\ref{fig:ShapleySummary3}).
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot.png}
-    \caption[Summary feature importance of a neural network.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-We can see that generally multiple factors influence the prediction, with the station elevation being highly influential.
-There is seemingly one outlier for the Richardson number, which usually has very little influence.
-Elevation data is excluded when working with Shapley values, as the contribution of each elevation point is very low and there are very many of them.
-To see their influence on the model output see Table \ref{table:results}.}
-    \label{fig:ShapleySummary}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_.png}
-    \caption[Summary feature importance of a neural network using a larger distribution of data.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-Generally multiple factors influence the prediction, with the station elevation being highly influential.
-Elevation data is excluded when working with Shapley values, as the contribution of each elevation.
-In contrast to Figure (\ref{fig:ShapleySummary}), the distribution doesn't have as extreme outliers.
-This means that more details can be seen in the figure.
-The X-axis shows the influence of feature values on the model.
-The color gradient shows the value of each feature.
-As an example, there is a very red value for station elevation (top line, all the way to the left).
-This means that in this instance, the station elevation contributed around -0.25 to the final output and that the station had an elevation significantly above average.}
-    \label{fig:ShapleySummary2}
-\end{figure}
-
-Further Shapley figures can be seen in Appendix~\ref{appendix:A}.
-In each of these plots, even without the feature labels, the station elevation is easily noticed as the value is constant for a station.
-This is not noteworthy.
-What is noteworthy is the range of impact from this single value.
-For simpler models, this would not happen.
-It is important to note that SHAP assumes feature independence \cite{Salih_2024}.
-This might explain why the impact of the Richardson number is so low.
-Both the squared Brunt–Väisälä frequency and the Richardson number are derived features from reanalysis data.
-They carry with them some extra information over the other features in the dataset.
-This is because both are variables over elevation ranges.
-That is, as seen in Equations (\ref{eqn:Ri}, \ref{eqn:N}), both are dependent on values at lower and upper elevations and try to describe the stability of that range.
-Shapley tries to assign contribution values for each feature for each observation.
-SHAP assumes that the features are independent, but this is not the case.
-It is clearly not the case for the derived variables, but how the contribution should be distributed between the features is not clear.
-Seemingly the SHAP python package is giving all the impact to the Brunt–Väisälä squared frequency and none to Richardson number.
-If the Brunt–Väisälä would be excluded from the data, the impact of the Richardson number would likely increase.
-Another point to note is that the features are ordered by their impact.
-This means that the station elevation is the most impactful for each plot, but the ordering of other variables changes.
-Looking at Figure (\ref{fig:ShapleySummary3}), which shows the Shapley summary plot for all data, the wind speed is the second most important feature.
-This is reversed in Figure (\ref{fig:ShapleySummaryAkrafjall}).
-This is not unexpected as Akrafjall station was specifically selected as the MAE for reanalysis wind speed was very high as can be seen in Table (\ref{fig:ShapleySummaryAkrafjall}), where you will also find the stations whose summary plot is shown in Figures (\ref{fig:ShapleySummaryAlmannaskarð}, \ref{fig:ShapleySummaryAsgarðsfjall}) and these also fall into the category of very high MAE for wind speed.
-What is interesting is that the reanalysis wind speed is also of low impact at stations like Háahlíð and Keflavíkurflugvöllur, as shown in Figures (\ref{fig:ShapleySummaryHaahlid}, \ref{fig:ShapleySummaryKeflavikurflugvollur}).
-These stations had the lowest of MAE for difference between measured wind speed and reanalysis wind speed.
-As the summary plot over all stations (Figure (\ref{fig:ShapleySummary3})) shows that reanalysis wind speed is impactful, might lead to the conclusion that the reanalysis wind speed is not a good predictor at these locations or something else is skewing the data.
-A simpler way to look at feature importance is to create models that are trained on and use to predict different sets of parameters.
-The results of such a comparison can be seen in Table \ref{table:setsOfParams}.
+The power generated by wind turbines increases with the cube of the wind speed \cite{wind_power}.
+The highest wind gusts in Iceland are around 70 m/s.
+Knowing the gust factor with half as much error as before can allow better anticipation and thus spare turbines for high wind gusts.

### 2025-05-08 Chapters/Chapter4.tex
Búinn að uppfæra kafla 4 eftir pósti en á eftir að fara yfir textann. Tók saman kafla 2/3.

@@
 \label{Chapter5}
+
+\section{Main findings}
 As shown in Table \ref{table:results}, the neural-network model achieves a significant reduction in mean absolute percentage error (MAPE) compared to a constant baseline.
@@
 
+\section{Limitations and future work}
 This study did not include station coordinates (XY) as direct inputs.
@@
 
+Many papers cited in this thesis suggest that adding a machine learning model as a backend to more traditional neural networks models could lead to improved performance.
+The results of this study support this.
+
+\section{Possible applications}
+Wind gusts are displayed on road signs in Iceland, if wind gusts exceed \href{https://www.vegagerdin.is/vegagerdin/starfsemi/frettir/tvo-ny-vedur-og-upplysingaskilti-i-notkun#}{expected values}.
+The model could be used to allow predicitons of wind gusts to be displayed.
+In 2023, outside of Reykjavik harbor, a cruise ship nearly \href{https://www.mbl.is/200milur/frettir/2024/04/26/strand_skemmtiferdaskips_kallar_a_erlenda_adstod/}{beached} due to an unexpected wind gust.
+This wind gust of \textasciitilde 26 m/s, pushed the ship into a buoy near Viðey.
+It was only luck that prevented the ship's screws from hitting the buoy.
+The ship was only 10 meters from the shoals of Viðey, where the water is only 0.4 meters deep.
+A better prediction of wind gust could have prevented this incident.
+Similarly, the model could be used to prevent damage to transportation vehicles, that take on a lot of wind, such as trucks and buses, or to prevent damage to buildings.
+
+\section{Conclusions}
+The neural network model presented in this study demonstrates a significant improvement in predicting wind gusts over a simple regression model, particularly for reanalysis wind speeds above 10 m/s.
+The model's performance varies by station, and while some stations benefit from the full dataset, others may require more localized training.
 The power generated by wind turbines increases with the cube of the wind speed \cite{wind_power}.

### 2025-05-09 Chapters/Chapter3.tex
vinna

@@
 \chapter{Results}
-\label{Chapter4}
+\label{Chapter3}
 \section{Baseline models}
@@
 \section{Modeling for various ws$_\textbf{15}$ lower bounds}
-To assess how model performance depends on ws$_{15}$ lower bound several models were trained using the primary variable set with and without DEM.
+To assess how model performance depends on ws$_{15}$ lower bound, several models were trained using the primary variable set, with and without DEM.
 The results can be seen in Table \ref{table:results}.
@@
 In all cases adding the DEM information is seen to increase the model performance.
-The reason for the difference between the 10 m/s lower bound results seen in Tables \ref{table:setsOfParams} and \ref{table:results} is they were obtained with different training setups.
+The reason for the difference between the 10 m/s lower bound results seen in Tables \ref{table:setsOfParams} and \ref{table:results} is that they were obtained with different training setups.
 
@@
 Adding DEM data reduces the error noticably.
-Note that the MAPE is higher for both the lower lower bounds and for the higher ones.}
+Note that the MAPE is worse for both the lower lower bounds and the higher ones, but better for the intermediate ones.}
     \label{table:results}
@@
 The Shapley summary plot for this station is shown in Appendix A (\ref{fig:ShapleySummaryHaahlid}).
+
+\section{CARRA wind gust}
+% In 4.3, add commas in first sentence
+% Later in 4.3, add "that" before "they"
+% Table 4.3, change higher to worse, add: but better for the intermediate ones
+
+The CARRA dataset provides no gust values in its analysis component but does include 10 m wind gust in the forecast.
+With a shortest lead time of one hour and given that a model of the gust factor is to be constructed, it stands to reason to incorporate the one-hour forecast CARRA gust factor, computed as:
+\begin{equation*}
+\text{CARRA gust factor} = \dfrac
+{\text{1 hr forecast 10 m wind gust}}
+{\text{1 hr forecast 10 m wind speed}}
+\end{equation*}
+into the modelling, as a proxy for a reanalysis wind gust.
+
+One issue with the CARRA gust factor is that it is not available for several time points including the period after the middle of 2022, whereas the CARRA reanalysis data used in this study extends through 2023, making it somewhat involved to assess its predictive power.
+Two investigations were carried out.
+First, the reduction in MAPE resulting from the addition of the CARRA gust factor to the baseline ws$_{15}$, secondary, and primary NN models of Section 4.2 was assessed, using only data with CARRA gust available.
+The results of this comparison are shown in Table \ref{table:carra_gust}
+
+\begin{table}[h]
+  \centering
+  \caption[Results adding CARRA gust factor]{Results of adding the CARRA gust factor as a variable in selected models from Sections 4.2 and 4.3.}
+  \label{table:carra_gust}
+  \begin{tabular}{lcc}
+    \toprule
+    & \multicolumn{2}{c}{MAPE\qquad\qquad} \\
+    Model & Unaltered & With CARRA gust factor \\
+    \midrule
+    Baseline ws$_{15}$ model & 17.0\% & 16.7\% \\
+    Secondary model           & 13.3\% & 13.2\% \\
+    Primary model             & 12.5\% & 12.4\% \\
+    \bottomrule
+  \end{tabular}
+\end{table}
+
+The causes of the differences between the numbers in Table \ref{table:carra_gust} and those of Tables 4.2 and 4.3 could be different time period and/or unlike training setup.
+
+Second, three simple linear regression models were constructed involving ws$_{15}$ and the CARRA gust factor, using available data with ws$_{15} \geq 10$ m/s.
+The results are shown in Table \ref{table:carra_gust_regression}.
+
+\begin{table}[h]
+  \centering
+  \caption[Regression MAPE with and without CARRA gust factor]{Mean absolute percentage error (MAPE) for regression models using the CARRA gust factor, ws$_{15}$, and both combined.}
+  \label{table:carra_gust_regression}
+  \begin{tabular}{lc}
+    \toprule
+    Model                         & MAPE   \\
+    \midrule
+    ws$_{15}$                     & 17.03\% \\
+    CARRA–gf                      & 16.81\% \\
+    CARRA–gf + ws$_{15}$          & 16.80\% \\
+    \bottomrule
+  \end{tabular}
+\end{table}
+
+Finally, using the raw CARRA gust factor without any modelling gave MAPE = 18.14\%.
+
+The conclusion that can be drawn from this investigation is that there is a modest information in the CARRA wind gust forecast, potentiially reducing model MAPE by 0.1 to 0.3 percentage points.
+This reduction is however quite small compared with the overall MAPE reduction of $\sim$5 percentage points seen in Table 4.2 when all available variables are added.

### 2025-05-09 MSc.tex
búnir með kafla 4
Updated and added neural network figure.
vinna

@@
 \def\thesisauthor{Brynjar Geir Sigurðsson}				% Thesis authoreiningaraðferðinni
-\def\thesistitle{A Neural Network Approach to Predicting Gust Factors in Complex Landscape} % Title of thesis
+\def\thesistitle{Prediction of Gust Factor in Complex Landscape with Neural Networks} % Title of thesis
 %\def\thesissubtitle{XXSubtitle is rarely usedXX}		% Subtitle of thesis (optional)
@@
 \section*{Abstract}
-In many industries, whether it be windmill farming or transportation, being able to precisely predict the load caused by weather is crucial to prevent harm coming to people and machinery as well as increasing productivity.
-The efficiency of wind turbines in high wind speed conditions is a vital challenge in optimizing renewable energy systems.
-Gust predictions can be instrumental in loss prevention.
-If looking only at wind, the most destructive moments are gusts, by definition.
-To predict these values there are two ways, traditional numerical weather prediction systems and machine learning (or other pattern recognition) methods.
-Computer-aided numerical weather prediction systems originated in the 1950s, while the use of machine learning in numerical weather predictions is much more recent, having gained traction in the last decade.
-The goal of this thesis is to combine these two methods to predict the wind gust factor, defined as the ratio between wind gust and average wind speed.
-This is done by having the input data to the machine learning model be reanalysis variables that were generated by traditional numerical weather prediction systems.
-In addition to these variables (and derived variables), a digital elevation model (DEM) was used.
-A deep neural network was constructed using some set of variables and the results are compared to a baseline as well as each other so as to show the impact each feature has.
-Neural networks can help make wind gust predictions in complex ladscape.
+Whether it is wind farms, traffic accidents, high-voltage towers, or other things affected by wind, accurately predicting wind loads is crucial for reducing damage, improving operational safety, and boosting productivity.
+Wind gusts are multi-second spikes in speed, and wind gust factor is defined as the ratio of the highest three-second gust to the ten-minute mean wind speed.
+It is important to be able to predict both wind and gust factor accurately.
+Historically, numerical weather prediction models based on physical equations have been used to predict such variables.
+In recent years, artificial intelligence has advanced rapidly, and AI-based models have begun to rival conventional model performance.
+This project combines both approaches by using atmospheric model reanalysis output and digital elevation data as inputs to a deep neural network.
+Selected features from the reanalysis fields, terrain elevation, and derived variables are fed into several network architectures, whose performance is then compared against a simple linear regression baseline.
+The results are very promising: this hybrid methodology can substantially improve both instantaneous estimation and short-term forecasting of gust factors, particularly in areas of complex terrain.
 \vfill \vspace*{1cm}
@@
 \section*{Útdráttur}
-Hvort sem um er að ræða vindmyllubúgarða eða stöðuleiki flutningabíla er spágetan um álag vegna vinds mikilvæg til að koma í veg fyrir tjón á iðnviðum, auk þess er hægt að auka framleiðni.
-Hviðuspár geta verið stór þáttur í tjónaminnkun.
-Ef einungis er horft á vind, þá eru hviður hæsti álagspunktur samkvæmt skilgreiningu.
-Sögulega hefur verið notast við hefðbundin spálíkön sem reiða sig á töluleg eðlisfræðileg líkön til að spá fyrir um niðurstöður.
+Hvort sem um er að ræða vindmyllugarða, umferðarslys, háspennumöstur, eða aðra hluti sem vindur getur haft áhrif á er spágeta um álag vegna vindsins mikilvæg, til að minnka eða koma í veg fyrir tjón, auka rekstraröryggi, eða bæta framleiðni.
+Vindhviður eru nokkurra sekúndna aukning á vindi og hviðustuðull mælir hæsta hlutfall milli þriggja sekúndna hviðu og 10 mínútna meðalvinds.
+Góðar hviðuspár eru mikilvægar.
+Sögulega hefur verið notast við hefðbundin spálíkön sem reiða sig á töluleg eðlisfræðileg líkön til að spá fyrir um veður, þar með talið hviðustuðul.
 Á síðustu árum hefur gervigreind þróast mikið og líkön verið þjálfuð til þess að spá fyrir um veður.
 Svo miklar framfarir hafa orðið að sum líkön geta keppt við hefðbundin veðurlíkön.
-Markmið þessa verkefnis er reyna að nýta bæði hefðbundin veðurlíkön og gervigreind til þess að giska á hviðustuðla, þar sem hviðustuðullinn er skilgreindur sem hlutfall hæstu hviðu og meðalvinds.
+Markmið þessa verkefnis er reyna að nýta bæði hefðbundin veðurlíkön og gervigreind til þess að meta hviðustuðla.
 Þetta er gert með því að nota endurgreiningargögn úr hefðbundnu veðurlíkani sem grunngögn í gervigreindarlíkan.
-Einnig eru notað hæðarlíkan til að lýsa umhverfi.
-Djúpt tauganet var búið til með því að velja breytur úr endurgreiningargögnunum ásamt hæðarpunktum og afleiddum breytum.
-Þetta líkan er svo þjálfað og niðurstöður bornar saman við grunnlíkan sem og önnur líkön með öðrum breytum til að skoða áhrif breyta.
-Djúp tauganet geta búið til hjálplegar spár fyrir vindhviður í flóknu landslagi.
+Einnig eru notað hæðarlíkan sem lýsir landslagi.
+Djúpt tauganet er búið til með því að velja breytur úr endurgreiningargögnunum og hæðarlíkaninu og afleiddum breytum.
+Nokkur líkön af þessu tagi eru svo þjálfuð og niðurstöður bornar saman við einfalt grunnlíkan sem fengið er með línulega aðhvarfi.
+Niðurstöðurnar lofa mjög góðu um að þessi aðferðafræði geti bætt bæði samtímamat á hviðustuðli og spár um hann umtalsvert.
+Þetta á ekki síst við í flóknu landslagi.
 \vfill
@@
 \addcontentsline{toc}{chapter}{Acknowledgments}
-Special thanks go to my advisor, Kristján Jónasson.
-He would help me with the actual work of the thesis and sit with go over the code with me.
-Thanks also go to Ólafur Pétur Pálsson, who would read over the thesis with notes and give helpful suggestions.
-Guðrún Nína Petersen, also provided valuable insight whenever questioned about meteoroligcal matters as well as giving notes on the thesis.
+Special thanks go to my advisor, Kristján Jónasson, who helped me with the thesis work and sat with me to go over the code.
+Thanks also to Ólafur Pétur Pálsson who provided detailed notes and offered very helpful suggestions.
+I am also grateful to Guðrún Nína Petersen, who provided valuable insight into meteorological matters and gave constructive feedback on the thesis.
 
@@
 \include{Chapters/Chapter4}
-\include{Chapters/Chapter5}
 

### 2025-05-09 Chapters/Chapter3.tex
Updated pages and figures.

@@
 The coloring of the points in the displayed distributions show the corresponding feature values.
-For example, in the lower panel of Figure \ref{figure:ShapleySummary}, high station altitude/elevation corresponds to negative contribution and vice versa, and same for $p_{15}$.
+For example, in the lower panel of Figure \ref{fig:ShapleySummary}, high station altitude/elevation corresponds to negative contribution and vice versa, and same for $p_{15}$.
 For ws$_{15}$ and $t_{15}$ the relationship is opposite, with high feature values corresponding to positive contribution.
@@
   \centering
-  \begin{subfigure}{0.9\textwidth}
+  \begin{subfigure}{0.85\textwidth}
     \includegraphics[width=\textwidth]{Figures/shap_plots/summary_plot.png}
+    \label{fig:ShapleySummarySmallerSet}
   \end{subfigure}
   \vspace{0.5cm}
-  \begin{subfigure}{0.9\textwidth}
+  \begin{subfigure}{0.85\textwidth}
     \includegraphics[width=\textwidth]{Figures/shap_plots/summary_plot_190924_.png}
+    \label{fig:ShapleySummaryLargerSet}
   \end{subfigure}

### 2025-05-09 Chapters/Chapter2.tex
vinna
Updated pages and figures.

@@
 
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.3]{Figures/storhofdi.jpeg}
+  \caption[AWS at Stórhöfði in Vestmannaeyjar, in south Iceland.]{AWS at Stórhöfði in Vestmannaeyjar, south of Iceland.
+The anemometer can be seen as three pins in the upper-central area of the figure just slightly to the left.}
+  \label{fig:storhofdi}
+\end{figure}
+
 \section{Automatic Weather Station Data}
-
 Automatic weather stations (AWS) have been set up all around the world, including Iceland.
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[scale=0.3]{Figures/storhofdi.jpeg}
-  \caption[AWS at Stórhöfði in Vestmannaeyjar, south of Iceland.]{AWS at Stórhöfði in Vestmannaeyjar, south of Iceland.
-The anemometer can be seen as three pins in the upper-central area of the figure just slightly to the left.}
-  \label{fig:storhofdi}
-\end{figure}
-
 Data from these AWSs were stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases.
@@
     \centering
-    \includegraphics[scale = 0.6]{Figures/station_elevations.png}
+    \includegraphics[width=1.1\textwidth]{Figures/station_elevations.png}
     \caption[Locations of automatic weather stations in Iceland]{Locations of all 412 stations that were looked at in this study.

### 2025-05-09 Chapters/Chapter3.tex
vinna

@@
 
-Inland stations seem to have lower error.
-The worst performing station is Seljalandsdalur in Vestfirðir while the best performing station is Garðskagaviti at the tip of the Reykjanes peninsula.
-Evidently rugged landscape makes it difficult to estimate the gust factor accurately.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale = 0.5]{Figures/errorMap.png}
-\caption[MAPE distribution by station]{Mean absolute percentage error (MAPE) at each station, shown as colored circles (dark blue = low error; yellow = high error).
-The lowest station error is about 7\% at Garðskagaviti and the highest about 35\% at Seljalandsdalur.
-The model uses a 10 m/s lower bound based on ws$_{15}$.}
-\label{fig:errorMap}
-\end{figure}
-
-\begin{table}[h]
+\begin{table}[H]
     \caption[Primary model MAPE by selected stations]{Constant baseline and primary model MAPE for a few selected stations.
@@
 
+Inland stations seem to have lower error.
+The worst performing station is Seljalandsdalur in Vestfirðir while the best performing station is Garðskagaviti at the tip of the Reykjanes peninsula.
+Evidently rugged landscape makes it difficult to estimate the gust factor accurately.
+
 After examining the MAPE at each station individually, it is logical to try accounting for the station (or its location) in the model’s design.
@@
 
-\begin{table}[h]
+\begin{figure}[H]
+  \centering
+  \makebox[\textwidth[c]{
+    \includegraphics[width=1.1\textwidth]{Figures/errorMap.png}
+  }
+\caption[MAPE distribution by station]{Mean absolute percentage error (MAPE) at each station, shown as colored circles (dark blue = low error; yellow = high error).
+The lowest station error is about 7\% at Garðskagaviti and the highest about 35\% at Seljalandsdalur.
+The model uses a 10 m/s lower bound based on ws$_{15}$.}
+\label{fig:errorMap}
+\end{figure}
+
+\begin{table}[H]
   \caption[Model result by station]{MAPE for a few selected stations, for a model trained with the whole data set (general training), and for a model trained only with the data of the station itself (site training), with a lower bound of 10 m/s.
@@
 
-\begin{table}[h]
+\section{Feature importance with Shapley values}\label{sec:shapley}
+In this section Shapley values (see section \ref{sec:methodology} and \ref{sec:layout}) for the features of the secondary parameter set will be considered.
+Each feature value in each data row will contribute some share in the final prediction of an NN model, that can be computed by shifting the respective value and observe the change in the model output.
+This procedure is illustrated in Figure \ref{fig:ShapleyWaterfall}.
+
+\begin{table}
     \caption[Model result looking at closed wind speed intervals]{The MAPE results for different wind speed limit intervals.
@@
 
-\section{Feature importance with Shapley values}
-In this section Shapley values (see section \ref{sec:methodology} and \ref{sec:layout}) for the features of the secondary parameter set will be considered.
-Each feature value in each data row will contribute some share in the final prediction of an NN model, that can be computed by shifting the respective value and observe the change in the model output.
-This procedure is illustrated in Figure \ref{fig:ShapleyWaterfall}.
-
-\begin{figure}[h]
+\begin{figure}
   \centering
@@
 \subsection{Shapley summary plots for all stations}
-
 Figure \ref{fig:ShapleySummary} provides a Shapley summary plot showing the contribution of each feature in the set.

### 2025-05-09 Chapters/Chapter4.tex
vinna

--- 2025-05-08 Chapters/Chapter4.tex
@@
 \label{Chapter5}
-As shown in Table \ref{table:results}, the neural-network model achieves a significant reduction in mean absolute percentage error (MAPE) compared to a constant baseline.
-Table \ref{table:setsOfParams} further demonstrates an improvement over a simple regression model.
-A statistical test yields $p<0.01$, confirming that these improvements are unlikely to be due to chance.
+
+\section{Main findings}
+As shown in Table \ref{table:setsOfParams}, the neural-network model achieves a significant reduction in mean absolute percentage error (MAPE) compared to a constant baseline as well as a simple regression model.
+Table \ref{table:results} further demonstrates this improvement.
+For reanalysis wind speeds above 10~m/s, the network reduces error by about 30\% relative to the regression baseline.
+Table \ref{table:closed_intervals} shows the results of training and model evaluation within closed intervals of the reanalysis wind speed.
+The best results appear in the 10--25~m/s range.
+At lower reanlysis wind speeds, percentage errors inflate even when the differences between wind gust and average wind speed remain small and at both at the higher and the lower speeds the base line MAPE is higher.
+
+Although one might expect staility indicators, such as the Richardson number and the Brunt--Väisälä frequency, to be among the strongest predictors of gust variability, they contributed less than other variables.
 The influence of individual features on model performance is illustrated by Shapley plots (Figure \ref{fig:ShapleySummaryKeflavikurflugvollur} and Appendix \ref{appendix:A}).
-Although one might expect stability indicators—such as the Richardson number and the Brunt--Väisälä frequency—to be among the strongest predictors of gust variability, they contributed less than other variables.
-For reanalysis wind speeds above 10~m/s, the network reduces error by about 30\% relative to the regression baseline.
-The largest gains appear in the 10--25~m/s range (Table \ref{table:closed_intervals}).
-At lower speeds, percentage errors inflate even when absolute gust--wind differences remain small; at higher speeds, sample sizes become sparse.
+According the Shapley analysis, stations altitude, mean wind speed and wind direction are among the more important features for predicting the gust factor, and also, to some extent, the Brunt--Väisälä frequency ($N^2$).
+As explained in Section \ref{sec:shapley} it was infeasible to evaluate the DEM data with Shapley values.
 
@@
 
-This study did not include station coordinates (XY) as direct inputs.
+\section{Limitations and future work}
+This study did not include station coordinates (X, Y) as direct inputs.
 Adding location information could improve site-specific accuracy without losing the benefits of training on the full dataset.
-Similarly, instead of applying a fixed linear interpolation to CARRA and elevation grids, the network could receive raw neighboring grid values and their relative positions, allowing it to learn an optimal interpolation scheme.
+Similarly, instead of applying a fixed linear interpolation to CARRA and DEM grids, the NN training could receive raw neighboring grid values and their relative positions as part of their input, allowing the model to learn an optimal interpolation scheme.
 
@@
 Here, only instantaneous inputs at forecast time were used, partly because CARRA fields are available every three hours and may not capture rapid stability changes.
-Access to higher-frequency data or the inclusion of temporal history could enable models to exploit evolving atmospheric conditions and further enhance gust-prediction skill.
+Access to higher-frequency data or the inclusion of temporal history could enable models to exploit evolving atmospheric conditions and further enhance gust-prediction skills.
+
+Many papers cited in this thesis suggest that adding a machine learning model as a backend to more traditional numerical atmospheric models could lead to improved performance.
+The results of this study support this.
+
+\section{Possible applications}
+Wind gusts at weather stations are displayed on road signs in Iceland, if wind gusts exceed \href{https://www.vegagerdin.is/vegagerdin/starfsemi/frettir/tvo-ny-vedur-og-upplysingaskilti-i-notkun#}{expected values}.
+The model could be used to allow such predicitons of wind gusts to be displayed for every section of the road, not just at the stations.
+In 2023, outside of Reykjavik harbor, a cruise ship nearly \href{https://www.mbl.is/200milur/frettir/2024/04/26/strand_skemmtiferdaskips_kallar_a_erlenda_adstod/}{ran aground} due to an unexpected wind gust.
+This wind gust of \textasciitilde 26 m/s, pushed the ship into a buoy near the island Viðey.
+It was only luck that prevented the ship's screws from hitting the buoy.
+The ship was only 10 meters from the shoals of Viðey, where the water is only 0.4 meters deep.
+A better prediction of wind gust could have prevented this incidence.
+Similarly, the model could be used to prevent damage to transportation vehicles, that take on a lot of wind, such as trucks and buses, or to prevent damage to buildings.
 
@@
 The highest wind gusts in Iceland are around 70 m/s.
-Knowing the gust factor with half as much error as before can allow better anticipation and thus spare turbines for high wind gusts.
+Knowing the gust factor accurately would allow the turbine to be stopped in time, thus avoiding damage.
+
+\section{Conclusions}
+The neural network model presented in this study demonstrates a significant improvement in predicting wind gusts over a simple regression model, particularly for reanalysis wind speeds above 10 m/s.
+The model's performance varies by station, and while some stations benefit from the full dataset, others may require more localized training.

### 2025-05-09 MSc.tex
Updated pages and figures.
vinna

@@
 \def\thesisyear{2025}       					% Year thesis submitted
-\def\thesismonth{June}					% Month thesis submitted, e.g. "December"
+\def\thesismonth{May}					% Month thesis submitted, e.g. "December"
 \def\thesisauthor{Brynjar Geir Sigurðsson}				% Thesis authoreiningaraðferðinni
-\def\thesistitle{Prediction of Gust Factor in Complex Landscape with Neural Networks} % Title of thesis
+\def\thesistitle{Prediction of Gust Factor in \\Complex Landscape with Neural Networks} % Title of thesis
 %\def\thesissubtitle{XXSubtitle is rarely usedXX}		% Subtitle of thesis (optional)
@@
 
-  \newpage
+  %\newpage
 
-  \thispagestyle{empty} \mbox{} % This is the inside page of the cover (cover verso) which remains empty
+  %\thispagestyle{empty} \mbox{} % This is the inside page of the cover (cover verso) which remains empty
 
-  \newpage
+  %\newpage
 
@@
 
-  \newpage
+  %\newpage
 
@@
 Þetta á ekki síst við í flóknu landslagi.
-\vfill
+%\vfill
 
-\newpage % Empty page
+%\newpage % Empty page
 \setkomafont{section}{\FloatBarrier\Large}
@@
 Finally, I would like to thank my parents without whom I would probably never had any want to finish.
-
+\newpage
 \pagenumbering{arabic}

### 2025-05-09 Appendicies/AppendixB.tex
Added appendix B

\chapter*{Appendix B: Github Repositories}
\label{appendix:B}

This appendix provides links to the GitHub repositories associated with this thesis. These repositories contain the \LaTeX{} source files for the thesis and the code used for data handling and modeling.

\section*{Thesis \LaTeX{} Source Code}
The \LaTeX{} source code for this thesis can be found in the following GitHub repository:
\begin{itemize}
    \item \url{https://github.com/BrynjarGeir/lokaverkefni_vel_tex}
\end{itemize}

\section*{Data Handling and Modeling Code}
The code used for data handling and creating models, written in Python (in python files or notebooks), is available in the following GitHub repository:
\begin{itemize}
    \item \url{https://github.com/BrynjarGeir/lokaverkefni_vel}
\end{itemize}

### 2025-05-13 Appendicies/AppendixA.tex
Updated pages and figures.
Updated after comments by GNP

@@
 \label{appendix:A}
+
+In section \ref{sec:shapley}, the Shapley values were used to investigate the feature importance of the neural network.
+Summary plots were created to visualize the feature importance of the model, along with one waterfall plot to show how features for an individual prediction influenced the model prediction.
+More summary plots were created than were displayed in section \ref{sec:shapley}, those plots are displayed here.
 
@@
 
-\begin{figure}
+\begin{figure}[H]
     \centering
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_190924_full_10ms.png}
-    \caption[Summary feature importance of a neural network using entire dataset.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-The distribution seems to be the same as before, discounting the outliers.}
+    \caption[Shapley Summary plot using entire dataset.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.}
     \label{fig:ShapleySummary3}
@@
 
-\begin{figure}
+\begin{figure}[H]
     \centering
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_6745.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Ásgarðsfjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Ásgarðsfjall.
-This seems to show the same distribution as previous summary plots.
+    \caption[Shapley Summary plot at Ásgarðsfjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at data from Ásgarðsfjall.
 Station elevation is influential and Richardson number has no impact.}
@@
 
-\begin{figure}
+\begin{figure}[H]
     \centering
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1470.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Háahlíð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Háahlíð.
-This seems to show the same distribution as previous summary plots.
+    \caption[Shapley Summary plot at Háahlíð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at data from Háahlíð.
 Station elevation is influential and Richardson number has no impact.}

### 2025-05-13 Chapters/Chapter1.tex
vinna
Updated after comments by GNP

@@
 %----------------------------------------------------------------------------------------
-Wind gusts are brief increases in wind speed (lasting seconds) compared to the mean wind speed over intervals of one or several minutes.
+Wind gusts are brief increases in wind speed (lasting seconds) compared to the mean wind speed over a given time interval.
 The gust factor is defined as the peak gust divided by the mean wind speed over a specified time period.
-The peak wind gust is often defined as the highest 3-second rolling-average wind speed measured over a 10-minute period, while the mean wind speed is the average of all measurements in the same interval.
-This thesis uses that definition.
-However, definitions vary: for example, the US uses a 1-minute interval, leading to approximately 14\% higher values \cite{why_wind_gusts}.
+In Iceland the peak wind gust is defined as the highest 3-second and 1-second rolling-average wind speed, by the Icelandic Meterological Office (IMO) and Icelandic Road and Coastal Administration (IRCA) respectively, in a given interval.
+The mean wind speed is the average of all measurements in a given interval.
+This thesis uses these definitions, based on source of data.
+However, definitions vary: for example, the US uses a 1-minute interval, leading to approximately 14\% higher values \parencite{why_wind_gusts}.
 
-The Navier–Stokes equation (\ref{eqn:navierstokes}) shows that changes in wind, both in time and space, depend on the pressure gradient, the oscillating force of the Earth (the Coriolis force), and frictional forces \cite{uncertainties_in_numerical_weather_predictions}.
+The Navier–Stokes Equation (\ref{eqn:navierstokes}) shows that changes in wind, both in time and space, depend on the pressure gradient, the oscillating force of the Earth (the Coriolis force), and frictional forces \parencite{uncertainties_in_numerical_weather_predictions}.
 \begin{equation}
     \label{eqn:navierstokes}
-    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{\text{pressure}} -\overbrace{ f\mathbf{k}\times\mathbf{V}}^{\text{Coriolis}} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{\text{resistance}}
+    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{\text{gradient force}} -\overbrace{ f\mathbf{k}\times\mathbf{V}}^{\text{Coriolis force}} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{\text{frictional force}}
 \end{equation}
 
-Traditionally, numerical weather prediction (NWP) systems are used to forecast and analyze weather patterns \cite{medium_range_3d_weather_forecasting_NN}.
+Traditionally, numerical weather prediction (NWP) systems are used to forecast and analyze weather patterns \parencite{medium_range_3d_weather_forecasting_NN}.
 These models describe the evolution of discretized atmospheric states using partial differential equations grounded in physics.
 Forecasts are typically produced every hour, or at coarser intervals for climate simulations.
-With increasing computational power and efficiency, the trend is to output data more frequently \cite{GNP_vidtal}.
-However, such outputs summarize conditions over each period and may not capture short-term fluctuations well, including variations in wind speed and gusts \cite{canNNBeatNWP}.
+With increasing computational power and efficiency, the trend is to output data more frequently \parencite{GNP_vidtal}.
+However, such outputs summarize conditions over each period and may not capture short-term fluctuations well, including variations in wind speed and gusts \parencite{canNNBeatNWP}.
 
@@
 Accurate gust predictions are crucial, as it is often the peak wind gust that will cause damaging incidents, such as structural failures, traffic accidents and downed power lines.
-Moreover, the prevalence of extreme wind events is expected to increase in the future \cite{nasa_extreme_weather}.
+Moreover, the prevalence of extreme wind events is expected to increase in the future \parencite{nasa_extreme_weather}.
 
@@
 The history of numerical weather prediction dates back to the 1920s, when Lewis Fry Richardson pioneered the field and attempted to produce forecasts.
-His results were flawed due to numerical noise.
+While his methods were reasonable, the results were flawed due to numerical noise.
 The ENIAC, built in 1945, was a general-purpose computer used—among other tasks—for weather prediction.
-These forecasts took 24 hours to compute and predicted 24 hours into the future.
-While a proof of concept, they were not operationally useful \cite{TheENIACForecastsARecreation}.
+The first forecasts took 24 hours to compute and predicted 24 hours into the future.
+While a proof of concept, they were not operationally useful \parencite{TheENIACForecastsARecreation}.
 
@@
 In September 1954, Carl-Gustaf Rossby and his Stockholm-based team produced the first real-time barotropic forecasts.
-The following year, the Joint Numerical Weather Prediction Unit (JNWPU), based in Princeton, New Jersey, released its first 36-hour forecasts at 400, 700, and 900 mb.
-Although these forecasts were inferior to subjective human analyses, they demonstrated feasibility and spurred further development in the field \cite{historyNWP}.
+The following year, the Joint Numerical Weather Prediction Unit (JNWPU), based in Princeton, New Jersey, released its first 36-hour forecasts at 400, 700, and 900 hPa.
+Although these forecasts were inferior to subjective human analyses, they demonstrated feasibility and spurred further development in the field \parencite{historyNWP}.
 Since then, NWP has made tremendous strides in parallel with increases in computational power and efficiency.
@@
 Notable drivers of this wave include advances in computational power (notably parallel processing on graphics processing units, GPUs), the development of neural networks (NNs) for processing massive datasets, and the availability of large online datasets.
-One type of NN is the convolutional neural network (CNN), originally designed for image processing but applicable to any gridded data with spatial structure \cite{canNNBeatNWP}.
+One type of NN is the convolutional neural network (CNN), originally designed for image processing but applicable to any gridded data with spatial structure \parencite{canNNBeatNWP}.
 Since 2018, significant work has applied AI to weather prediction.
@@
 
-In 2018, Düben and Bauer showed that an NN could outperform a simple persistence forecast and compete with very coarse-resolution atmospheric models of similar complexity for short lead times \cite{dueben2018}.
+In 2018, Düben and Bauer showed that an NN could outperform a simple persistence forecast and compete with very coarse-resolution atmospheric models of similar complexity for short lead times \parencite{dueben2018}.
 Also in 2018, Scher developed a deep CNN to emulate a general circulation model (GCM)—a numerical model representing physical processes—by training on GCM output.
-This approach allowed stable emulation of model dynamics for much longer horizons \cite{scher2018}.
+This approach allowed stable emulation of model dynamics for much longer horizons \parencite{scher2018}.
 These papers were proofs of concept rather than production-ready replacements for NWP, but they demonstrated that deep-learning–based models could, with further development, compete with standard models in the field.
@@
 In the last two years, there have been even more developments with the emergence of Large AI Weather forecast Models (LWMs).
-In 2024, Ling et al.\ \cite{SecondRevolution} proposed a standardized definition of LWMs in meteorology, outlining three criteria, referred to as the "Three Large Rules":
+In \cite{SecondRevolution} proposed a standardized definition of LWMs in meteorology, outlining three criteria, referred to as the "Three Large Rules":
 
@@
   \item \textbf{Multiple predictands:} Forecasting at various levels (e.g., pressure or height levels) to provide detailed atmospheric vertical structure and surface conditions.
-  \item \textbf{Scalability and downstream applicability:} Demonstrated, for example, by predicting cyclones even when not explicitly trained on cyclone data (e.g., \href{https://www.youtube.com/watch?v=PD1v5PCJs_o&ab_channel=GregBronevetsky}{GraphCast}) to showcase model versatility \cite{SecondRevolution}.
+  \item \textbf{Scalability and downstream applicability:} Demonstrated, for example, by predicting cyclones even when not explicitly trained on cyclone data (e.g., \href{https://www.youtube.com/watch?v=PD1v5PCJs_o&ab_channel=GregBronevetsky}{GraphCast}) to showcase model versatility.
 \end{enumerate}
@@
 Before 2022, LWMs had been shown to compete with traditional NWP in specific cases and to generate forecasts much faster after training, No model had been shown to be able to completely replace traditional NWP systems.
-In early 2022, Pathak et al.\ \cite{FourCastNet} introduced FourCastNet, which employs an Adaptive Fourier Neural Operator model leveraging transformer architecture instead of convolution.
+In \cite{FourCastNet} introduced FourCastNet, which employs an Adaptive Fourier Neural Operator model leveraging transformer architecture instead of convolution.
 FourCastNet matches the performance of standard forecasting techniques at short lead times for large-scale variables and outperforms them for smaller-scale features.
-It generates a one-week forecast in under two seconds—orders of magnitude faster than conventional physical methods \cite{FourCastNet}.
+It generates a one-week forecast in under two seconds—orders of magnitude faster than conventional physical methods.
 In 2022, several ML-based models demonstrated faster-than-NWP predictions after one-time training, with performance rivaling or exceeding NWP in some cases.
@@
 In 2023, Remi Lam and the GraphCast team at Google released GraphCast, which outperformed the European Centre for Medium-Range Weather Forecasts’ (ECMWF) industry-standard High-Resolution Forecast (HRES).
-GraphCast uses a graph-based representation instead of a regular grid, operating on a global latitude–longitude grid at 0.25° resolution, which introduces nonuniform point spacing near the poles; the graph structure helps mitigate this bias \cite{GraphCast}.
+GraphCast uses a graph-based representation instead of a regular grid, operating on a global latitude–longitude grid at 0.25° resolution, which introduces nonuniform point spacing near the poles; the graph structure helps mitigate this bias \parencite{GraphCast}.
 
 \subsection{Conclusion}
-Substantial progress has been made since 2018 and especially since 2022 \cite{SecondRevolution}.
+Substantial progress has been made since 2018 and especially since 2022 \parencite{SecondRevolution}.
 The evolution from proof-of-concept ML methods to models competitive with—or surpassing—standard NWP has been remarkably rapid.
@@
 The aim of the study is to improve prediction of gust factors that can be applied at any given place in Iceland.
-Currently, gust factor predictions can be obtained directly from NWP output, but one may also create a simple regression model (involving, e.g., wind speed and location) to predict them.
+Currently, gust factor predictions can be obtained directly from NWP output, but one may also create a simple regression model (involving wind speed and location) to predict them.
 To accomplish this, an NN is used as a backend to the NWP.
@@
 
-In 2004, Hálfdán Ágústsson and Haraldur Ólafsson \cite{mean_gust_HA_HO} investigated gust factor variability in complex landscapes.
+In \cite{mean_gust_HA_HO}, gust factor variability in complex landscapes was investigated.
 They used data from automatic weather stations measuring wind at 10~m above ground during 1999–2001.
-They examined how three parameters affected the gust factor: \(d_m\), \(D\), and \(H\)—the wind direction relative to a mountain, the distance to the mountain, and the height of the mountain above the station.
+They examined how three parameters affected the gust factor: $d_m$, $D$, and $H$, the wind direction relative to a mountain, the distance to the mountain, and the height of the mountain above the station.
 Their main results showed that the gust factor is inversely correlated with distance and directly correlated with mountain height.
@@
 
-\subsubsection{Structure of neural networks}
 The general structure of a neural network is shown in Figure \ref{fig:nn_structure}.
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{figures/neural_network.drawio.png}
+    \includegraphics[width=0.8\textwidth]{figures/neural_network.drawio_updated.png}
     \caption[General structure of a fully connected neural network]{General structure of a fully connected neural network.
 The figure shows a network with three input variables and one output variable.
-It also shows one hidden layer (and others are implied).
-This hidden layer has four neurons that each connect to all input neurons and all connect to all neurons in the next hidden layer (not shown).
+It also shows two hidden layers (and others are implied between them).
+These hidden layers have four neurons that each connect to all input neurons and all connect to all neurons in the next hidden layer (not shown).
 This continues for all hidden layers until the output layer, when there is only one neuron.}
@@
 Models were created with different parameter sets to gauge each parameter’s influence.
-Each model produces a single numerical output.
-Every layer also has an activation function; common choices include Rectified Linear Unit (ReLU), Exponential Linear Unit (ELU), and hyperbolic tangent (tanh), defined in Equations (\ref{eqn:relu}), (\ref{eqn:elu}), and (\ref{eqn:tanh}).
+Each model produced a single numerical output.
+Every layer also had an activation function; common choices include Rectified Linear Unit (ReLU, r), Exponential Linear Unit (ELU, f), and hyperbolic tangent (tanh), defined respectively in the following equations.
 
-\begin{equation}
-    \label{eqn:relu}
+\begin{equation*}
     r(x):=\max(0, x)
-\end{equation}
+\end{equation*}
 
-\begin{align}
-    \label{eqn:elu}
+\begin{align*}
     f(x) := x,\ x > 0\\
     f(x) := \alpha (e^x-1),\ x\leq 0,\ \alpha > 0
-\end{align}
+\end{align*}
 
-\begin{equation}
-    \label{eqn:tanh}
+\begin{equation*}
     \operatorname{tanh}(x):=\frac{e^x-e^{-x}}{e^x+e^{-x}}
-\end{equation}
+\end{equation*}
 
@@
 \subsection{Model evaluation}
-To measure the performance of these models, both in training and testing, mean absolute percentage error (MAPE) as defined in Equation (\ref{eqn:mape}) was used.
+To measure the performance of these models, both in training and testing, mean absolute percentage error (MAPE) as defined nn Equation (\ref{eqn:mape})
 \begin{equation}
@@
 \end{equation}
+Where $y_{\text{predict}}$ is the model output.
 This measure was chosen because the target is the gust factor (the wind gust over the average wind).
@@
 \subsection{Model explainability}
-Neural networks are often considered mysterious black boxes \cite{nn_black_box}.
+Neural networks are often considered mysterious black boxes \parencite{nn_black_box}.
 To understand model predictions, explainability methods are used.
-One such method is Shapley values \cite{shapley_information}.
+One such method is Shapley values \parencite{shapley_information}.
 Shapley values are calculated as the average marginal contribution of a feature value across all possible coalitions.
-The contribution for a single feature \(j\) to a prediction \(\hat{f}(X)\) is given by Equation (\ref{eqn:shapley}).
-In Equation (\ref{eqn:shapley}), \(x_j\) is the feature value, \(\beta_j\) its weight, and \(\beta_j E[X_j]\) the mean effect estimate for feature \(j\).
+The contribution for a single feature \(j\) to a prediction \(\hat{f}(X)\) is given by Equation (\ref{eqn:shapley}):
 \begin{equation}
@@
 \end{equation}
+where \(x_j\) is the feature value, \(\beta_j\) its weight, and \(\beta_j E[X_j]\) the mean effect estimate for feature \(j\).
 For any combination of parameters, Shapley values explain the individual prediction by attributing a contribution to each feature.
-Other methods, like ELI5 (Explain Like I’m 5), randomly shuffle a feature and measure the effect on model performance \cite{eli5_information}.
+Other methods, like ELI5 (Explain Like I’m 5), randomly shuffle a feature and measure the effect on model performance \parencite{eli5_information}.
 

### 2025-05-13 Chapters/Chapter2.tex
Updated after comments by GNP

@@
 Data were sourced from several streams.
-The Icelandic Meteorological Office (IMO) provided measurements from weather stations across Iceland, NWP data were downloaded from the Copernicus Arctic Regional Reanalysis dataset (CARRA), and finally a land-elevation model was also provided by the IMO.
+The Icelandic Meteorological Office (IMO) provided measurements from automatic weather stations across Iceland, NWP data were downloaded from the Copernicus Arctic Regional Reanalysis dataset (CARRA), and finally digital elevation model (DEM) was also provided by the IMO.
 
@@
 Data from these AWSs were stored in hourly files, which aggregate the original 10-minute files; measurement errors—unrealistic spikes known as “nails”—have been removed in most cases.
-Each record contains the following information: date and time; station number (convertible to coordinates using another dataset of Icelandic meteorological stations); average wind speed ($f$); wind gust ($f_g$); standard deviation of the wind gust; wind direction ($d$); and standard deviation of the wind direction.
+Each record contains e.g.
+the following information: date and time; station number (convertible to coordinates using another dataset of Icelandic meteorological stations); average wind speed ($f$); wind gust ($f_g$); standard deviation of the wind gust; wind direction ($d$); and standard deviation of the wind direction.
 
@@
     \caption[Locations of automatic weather stations in Iceland]{Locations of all 412 stations that were looked at in this study.
-Most of these were from IMO but over a hundred were from IRCA.
-IMO anemometers are placed at 10 meters above ground, while IRCA ones are placed at 6--7 meters above ground.}
+After filtering out stations with inconsistent data and stations where there were known malfunctions, \nStationsMin stations were left.
+IMO anemometers are placed at 10 meters above ground (red points), while IRCA ones are placed at 6--7 meters above ground (blue points).}
     \label{fig:aws_map}
@@
 \section{CARRA Data}
-CARRA is a high-resolution atmospheric reanalysis produced by the Copernicus Climate Change Service and run by ECMWF.
+CARRA is a high-resolution atmospheric reanalysis produced by the Copernicus Climate Change Service.
 It covers two regions, a west region covering Greenland and Iceland and an east region covering the European Arctic.
@@
 
-The CARRA dataset covers all the IMO observations that fulfill criteria of consistent availability – the oldest observation is from \startDateVedur.
+The CARRA dataset covers all the IMO observations.
+The oldest observation used is from \startDateVedur.
 The CARRA-West region covers a vastly larger area than the area of interest.
@@
 
-The requests to the API were made at each available CARRA hour ([00, 03, 06, 09, 12, 15, 18, 21]) on a grid covering Iceland, for each available observation time.
+The requests to the API were made at each available CARRA analysis hour ([00, 03, 06, 09, 12, 15, 18, 21]) on a grid covering Iceland, for each available observation time.
 The downloaded data were interpolated to get values at the weather stations.
@@
 \section{Elevation data}
-A GeoTIFF file containing a digital elevation model (DEM) for Iceland on a 20~m by 20~m grid was provided by the Icelandic Meteorological Office (IMO).
-The entire country is covered by this file, and its size is approximately 685\,MB.
+A GeoTIFF file containing a DEM for Iceland on a 20~m by 20~m grid was provided by the Icelandic Meteorological Office (IMO).
+The entire country is covered by this file, and its size is approximately 685\,MB \cite{ragnar_geotiff}.
 
@@
 To train the models, these datasets were combined into a single file using the IMO measurements as a reference.
-CARRA data are supplied on a rectangular grid with approximately 2.5~km spacing, while IMO observations are tied to specific station locations.
+CARRA data are supplied on a rectangular grid with 2.5~km spacing, while IMO observations are tied to specific station locations.
 Elevation data are on a 20~m by 20~m grid covering Iceland.
@@
   \includegraphics[scale=0.6]{Figures/data_combination.drawio.png}
-  \caption{Flow chart illustrating the data-combination workflow}
+  \caption[Flow chart illustrating the data-combination workflow.]{Flow chart illustrating the data-combination workflow.
+Data provided by IMO was used to query the CARRA API.
+The CARRA data were then interpolated to each station location as well as the elevation data.
+The data were then combined into a single file.}
   \label{fig:data_preprocessing_flow_chart}
@@
 Elevation values from the GeoTIFF were interpolated in the same manner: the four surrounding grid points were used in a linear interpolation to estimate the elevation at each station location.
-These interpolated values were included in the dataframe, since topography influences both average wind speed and gustiness \cite{GNP_vidtal}.
+These interpolated values were included in the dataframe, since topography influences both average wind speed and gustiness \parencite{GNP_vidtal}.
 
@@
     \toprule
-    $f$ & $n$ & MAE & MAPE \\
+    $f[m/s]$ & $n$ & MAE $[m/s]$ & MAPE \\
     \midrule
@@
 Figure \ref{fig:station_mae_distribution} shows this distribution, and Table \ref{table:station_mae_distribution} lists the five stations with the lowest MAE and the five with the highest MAE.
+The five stations with the lowest MAE are all located in relatively simple landscapes (flat terrain, wide space), while the five with the highest MAE are in complex landscapes (mountains, valleys, and fjords).
+This is mostly echoed in Figure \ref{fig:station_mae_distribution}.
 
@@
   \includegraphics[scale=0.6]{Figures/MAEoverIceland.png}
-  \caption[MAE distribution by station]{Distribution of mean absolute error (MAE) by station for all observed wind speeds.}
+  \caption[MAE distribution by station]{Distribution of mean absolute error (MAE) by station for all observed wind speeds. Image is missing units for MAE, which is in m/s.}
   \label{fig:station_mae_distribution}
@@
     \toprule
-    Station               & $n$     & MAE  \\
+    Station               & $n$     & MAE $[m/s]$ \\
     \midrule
@@
 
-Once data from all three sources have been retrieved and processed—including interpolation—they must be combined and formatted for modeling (training, validation, and testing).
+Once data from all three sources have been retrieved and processed, including interpolation, they must be combined and formatted for modeling.
+This includes splitting data in training, validation, and testing datasets, where the training dataset is used to teach the model, the validation dataset helps tune the model's parameters, and the testing dataset evaluates its performance on unseen data.
 The starting point is a DataFrame containing AWS observations: average wind speed, wind gust, wind direction, station number, and coordinates.
@@
 The DEM elevation data are handled by defining an upwind sector: a range of angles relative to the wind direction $d$ and radial distances from each station.
-Points within this sector are generated as shown in Code Listing \ref{code:sectorElevation}.
+Points within this sector are generated as shown in Code Listing \ref{code:sectorElevation}:
 
@@
     \caption[Example of combined data structure]{Example of the combined data structure of features used for modeling.
-Data include derived variables Richardson number (Ri) and squared Brunt–Väisälä frequency ($N^2$), station altitude (meters above sea level), transformed wind direction (twd), wind speed (ws$_{15}$), wind direction (wd$_{15}$), temperature ($t_{15}$), pressure ($p_{15}$) (CARRA values at 15 m height), and elevation points (from DEM) in a sector pointing upwind.}
+Data include derived variables Richardson number (Ri) and squared Brunt–Väisälä frequency ($N^2$), station elevation (meters above sea level), transformed wind direction (twd), wind speed (ws$_{15}$), wind direction (wd$_{15}$), temperature ($t_{15}$), pressure ($p_{15}$) (CARRA values at 15 m height), and elevation points (from DEM) in a sector pointing upwind.}
     \label{table:trainDataExample}
@@
         \toprule
-        Ri & $N^2$ & \shortstack{station\\altitude} & twd & ws$_{15}$ & wd$_{15}$
+        Ri & $N^2$ & \shortstack{station\\elevation} & twd & ws$_{15}$ & wd$_{15}$
         & $t_{15}$ & $p_{15}$ & $\text{elevation}_0$ & elevation$_1$ & \dots\\
+        $[]$ & $[Hz^2]$ & $[m]$ & $[^{\circ}]$ & $[m/s]$ & $[^{\circ}]$ & $[K]$ & $[hPa]$ & $[m]$ & $[m]$ \\
         \midrule 
-        -1.18 &  26700 & 100 & 1.5 & 10 & 5 & 0 & 100 & 2 & 4 & \dots\\
+        -1.18 &  26700 & 100 & 1.5 & 10 & 5 & 270 & 100 & 2 & 4 & \dots\\
         \dots\\
@@
 
-The transformed wind direction is a proxy showing whether the wind is coming from land or sea.
-It is computed as the angle between the wind direction and a vector from the center of Iceland to the station.
+The transformed wind direction is a proxy showing if the wind is coming from land or sea. It is computed as the angle between the wind direction and a vector from the center of Iceland to the station.
 
-The Richardson number and the Brunt–Väisälä frequency describe atmospheric stability, see Equations (\ref{eqn:Ri}) and (\ref{eqn:N}) \cite{richardson_number_skybrary,brunt_vaisala_freq_eumtrain,mean_gust_HA_HO}.
+The Richardson number and the Brunt–Väisälä frequency describe atmospheric stability, see Equations (\ref{eqn:Ri}) and (\ref{eqn:N}) \parencite{richardson_number_skybrary,brunt_vaisala_freq_eumtrain,mean_gust_HA_HO}.
 These values are calculated using reanalysis data at two different height levels.
-Thus Ri refers to the Richardson number calculated between height levels 15m and 500m.
+Thus Ri refers to the Richardson number calculated between height levels 15 m and 500 m.
 Exactly the same notation is used with the Brunt–Väisälä frequency, except the square is used.
@@
 Here, $g$ is gravitational acceleration; $d\theta$ the potential temperature difference between levels; $dz$ the height difference; $\theta_{\mathrm{ave}}$ the average potential temperature; and $dU$ the wind-speed difference.
-Lower Ri indicates greater turbulence, with typical values between 0.1 and 10, and values below 1 signifying significant turbulence \cite{richardson_number_skybrary}.
-Negative $N^2$ denotes instability, as an air parcel will accelerate away from its original position \cite{brunt_vaisala_freq_eumtrain}.
+Lower Ri indicates greater turbulence, with typical values between 0.1 and 10, and values below 1 signifying significant turbulence \parencite{richardson_number_skybrary}.
+Negative $N^2$ (when potential temperature delta is negative) denotes instability, as an air parcel will accelerate away from its original position \parencite{brunt_vaisala_freq_eumtrain}.
 The potential temperature is computed as:
@@
   \label{eqn:potential_temp}
-  \theta = T\left(\dfrac{P0}{P}\right)^{R/c_p}
+  \theta = T\left(\dfrac{P_0}{P}\right)^{R/c_p}
 \end{equation}
 %
-where $P$ is the pressure and $R = 287$ and $c_p = 1004$ are the specifc gas constant and the specific heat capacity of air.
+where $P, P_0$ is pressure and $R = 287 J/kg\cdot K$ and $c_p = 1004 J/kg\cdot K$ are the specific gas constant and the specific heat capacity of air.
 
-Since Ri and $N^2$ are derived from reanalysis variables, they add limited new information compared to the raw data, but they can reduce input dimensionality and improve model explainability via Shapley-value analysis \cite{shapley_information}.
-Shapley values assess feature importance by evaluating all possible feature subsets—an operation of computational complexity $2^n$.
+Since Ri and $N^2$ are derived from reanalysis variables, they add limited new information compared to the raw data, but they can reduce input dimensionality and improve model explainability via Shapley-value analysis \parencite{shapley_information}.
+Shapley values assess feature importance by evaluating all possible feature subsets, an operation of computational complexity $2^n$.
 In practice, approximations are used but can still be costly for large models.
@@
 
-  \caption{Comparison of observed and reanalysis wind-speed distributions. CARRA values are spatially interpolated by linear weighting of surrounding grid points.}
+  \caption[Comparison of observed and reanalysis wind-speed distributions.]{Comparison of observed and reanalysis wind-speed distributions.
+CARRA values are spatially interpolated by linear weighting of surrounding grid points.}
   \label{fig:obs_carra_wind_speeds}
 \end{figure}
+
+Figure \ref{fig:obs_carra_wind_speeds} shows the distributions of observed and reanalysis wind speeds.
+The observed wind speeds are from the IMO and IRCA stations, while the CARRA wind speeds are interpolated to the station locations.
+The distributions are similar, but the CARRA values tend to be higher.
 
@@
   \includegraphics[width=0.8\textwidth]{Figures/station_heights.png}
-  \caption{Distribution of weather station altitudes above sea level. One outlier near 2000~m was excluded due to limited and inconsistent data.}
+  \caption[Distribution of weather station elevations above sea level.]{Distribution of weather station elevations above sea level.
+One outlier near 2000~m was excluded due to limited and inconsistent data.}
   \label{fig:station_heights}
 \end{figure}
+
+Well over half of the stations are located at elevations below 100~m, with a few at elevations above 1000~m.
+The distribution of station elevations is shown in Figure \ref{fig:station_heights}.
+The outlier at 2000~m was excluded due to limited and inconsistent data.
 
@@
   \includegraphics[width=0.8\textwidth]{Figures/gust_factor_2025.png}
-  \caption{Histogram of gust factors. By definition, gust factor $\ge1$; most values lie between 1.2 and 2. The gust factor declines as wind speed increases.}
+  \caption[Histogram of gust factors.]{Histogram of gust factors.}
   \label{fig:gust_factors}
@@
 
+Figure \ref{fig:gust_factors} shows the distribution of gust factors.
+By definition, gust factors are always greater than or equal to 1.
+The histogram reveals that most gust factors lie between 1.2 and 2.0, with a long tail extending to higher values.
+This distribution highlights the variability in gustiness across different locations and conditions.
+Higher gust factors are typically associated with complex terrain or unstable atmospheric conditions, where turbulence can amplify wind gusts relative to the mean wind speed.
+
 \section{Model architecture}
 \label{sec:model_architecture}
+
+The model architecture is a feedforward neural network, an architecture suitable for regression tasks.
+It is implemented using the \href{https://keras.io/}{Keras} library, which is a high-level API for building and training deep learning models.
+The model consists of several layers: input, output, hidden and dropout layers, along with batch normalization.
+An input layer receives has the same shape as the number of features, while hidden layers have some number of units (defined by the user) and are followed by batch normalization.
+The dropout layer is used to prevent overfitting, by randomly dropping a given fraction of the input units.
+The ouput layer in regression tasks has the same number of outputs as predictants, in this case one, the gust factor.
 
@@
 A grid search was used to find the hyperparameters that minimize the loss.
-Hyperparameters are settings defined before training begins \cite{hyperparameters_definition}.
-They include the number of units per layer, the number of layers, the number of training epochs, batch size, optimizer choice, and regularization penalty.
+Hyperparameters are settings defined before training begins \parencite{hyperparameters_definition}.
+They include the number of units per layer, the number of layers, the number of training epochs (how often all training data is used in training), batch size, optimizer choice, and regularization penalty (imposes cost on optimization function).
 The ranges tested are listed in Table \ref{table:gridSearchHyperparameters}.
@@
     \centering
-    \caption[Hyperparameter search with Hyperband]{Hyperparameter search was performed using the Hyperband algorithm, which begins by sampling configurations at random and then focuses on the most promising ones.
-The table shows the ranges tested and the best-performing combination.
-Unlike random search, Hyperband does not impose a fixed limit on the number of trials.}
+    \caption[Hyperparameter search with Hyperband]{Results of hyperparameter search using Hyperband. The ranges of values tested are shown, along with the selected values.}
     \label{table:gridSearchHyperparameters}
@@
 This determines how often the Hyperband algorithm is run and defaults to 1.
-For each iteration the epochs are distributed between tries (that is each set of hyperparameters) with the total amount of epochs approximately $n_{epochs} = max_{epochs} * log^2(max_{epochs})\approx 10^4$, where $max_{epochs} = 1000$ gives the maximum number of epochs that one set of hyperparameters can be trained for.
-Searching a space generally takes a lot of time but this drastically improves on grid search.
-If each epoch takes around 10 seconds to run then the total search would take around 28 hours on a shared resource.
-This is resource-intensive and cannot be repeated often.
-Another question that remains is whether the ranges given are optimal.
-
-As noted, Hyperband does not impose a fixed limit on the total number of training epochs.
+Hyperband does not impose a fixed limit on the total number of training epochs.
 When using the \href{https://keras.io/api/keras_tuner/tuners/hyperband/}{Hyperband tuner}, the Hyperband iterations parameter controls how many times Hyperband is run and defaults to 1.
@@
 It remains unclear whether the current hyperparameter ranges are optimal.
+This is resource-intensive and cannot be repeated often.
+Another question that remains is whether the ranges given are optimal.
+
+Hyperparameter search was performed using the Hyperband algorithm, which begins by sampling configurations at random and then focuses on the most promising ones.
+Table \ref{table:gridSearchHyperparameters} shows the ranges tested and the best-performing combination.

### 2025-05-13 Chapters/Chapter3.tex
Updated after comments by GNP

@@
 \label{Chapter3}
+This chapter presents the results of modeling efforts aimed at predicting the gust factor under various conditions.
+The analysis begins with baseline models, which serve as simple benchmarks for comparison.
+These models use minimal input features and provide a reference point for evaluating the performance of more complex models.
+
+Subsequently, more sophisticated models are introduced, incorporating additional features such as atmospheric variables, terrain data, and station-specific information.
+These models aim to capture the underlying complexities of the gust factor and improve predictions.
+The results are analyzed in terms of mean absolute percentage error (MAPE), with comparisons of feature sets and modeling approaches.
+
 \section{Baseline models}
-The simplest model for the gust factor is to guess the average gust factor for all available observations, i.e.
+A very simple model for the gust factor is to guess the average gust factor for all available observations, i.e.
 use no variables from the training data.
-A slightly better model is obtained by using the wind speed as a training feature, either through regression or NN training.
+A slightly more complex model is obtained by using the wind speed as a training feature, either through regression or NN training.
 The MAPEs for the constructed base models are shown in Table \ref{table:baseline_models}.
@@
 \begin{table}[h]
-  \caption[MAPE for baseline models]{MAPE for several lower limits on the reanalysis 15 m wind speed ($ws_{15}$). Note that the ws$_{15}$ model was only evaluated for the 10 m/s lower bound.}
+  \caption[MAPE for baseline models]{MAPE for several lower limits on the reanalysis 15 m wind speed (ws$_{15}$). Note that the ws$_{15}$ model was only evaluated for the 10 m/s lower bound.}
     \label{table:baseline_models}
@@
 
-\section{Modeling for ws$_{\textbf{15}}$ above 10 m/s}
+\section{Modeling for ws$_{\textbf{15}}$ above 10 m/s}\label{sec:ws15_10}
 As noted in Chapter \ref{Chapter1}, it is more important to predict accurate gust factor when it is windy than when the weather is calm.
@@
 This gives a moderately reduced MAPE of 16.7\% compared with 17.2\% for the baseline ws$_{15}$ model.
-When the next pair of features, station altitude and transformed wind direction, is included, the MAPE is reduced considerably, down to 13.8\%.
+When the next pair of features, station elevation and transformed wind direction, is included, the MAPE is reduced considerably, down to 13.8\%.
 The atmospheric stability indicators, Ri and $N^2$, add little to the model performance, contrary to what might have been expected.
 When the atmosphere is unstable, wind gusts from higher up might be expected.
-However, another considerable performance increase is observed when digital elevation model (DEM) information is included, down to a MAPE of 12.0\%.
+However, another considerable performance increase is observed when digital elevation model (DEM) information is included, lowering down to MAPE 12.0\%.
 This represents over 30\% reduction in the MAPE from the ws$_{15}$ baseline model.
-Adding all height levels decreases the loss but very little, as seen in the last line of the table.
+Adding all height levels decreases the loss but very little, as seen in the last line of Table \ref{table:setsOfParams}.
 
@@
 \begin{table}[h]
-  \caption[Model results for different sets of parameters.]{Models defined by a selection of parameter sets, with increasing complexity. The table shows decreasing MAPE as more variables are added.}
+  \caption[Model results for different sets of parameters.]{Models defined by a selection of parameter sets, with increasing complexity and the calculated MAPE for each model.}
     \label{table:setsOfParams}
@@
         $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd]$ & 13.8\% \\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$* & 13.7\%\\
-        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ + DEM** & 12.0\%\\
-        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, \text{altitude}, twd, N^2, Ri]$ & 12.0\% \\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{elevation}, twd]$ & 13.8\% \\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{elevation}, twd, N^2, Ri]$* & 13.7\%\\
+        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{elevation}, twd, N^2, Ri]$ + DEM** & 12.0\%\\
+        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, \text{elevation}, twd, N^2, Ri]$ & 12.0\% \\
         \bottomrule
@@
 
-\section{Modeling for various ws$_\textbf{15}$ lower bounds}
+Table \ref{table:setsOfParams} shows decreasing MAPE as more variables are added.
+
+\section{Modeling for various ws$_\textbf{15}$ lower bounds}\label{sec:ws15_lower_bounds}
 To assess how model performance depends on ws$_{15}$ lower bound, several models were trained using the primary variable set, with and without DEM.
 The results can be seen in Table \ref{table:results}.
-For this investigation comparison was only made with the constant baseline model.
+For this investigation comparison was only made with the constant baseline model, that is the mean gust factor for all observations above 10 m/s.
 In all cases adding the DEM information is seen to increase the model performance.
@@
 \begin{table}[h]
-  \caption[Model results for different wind speed limits]{MAPE for each ws$_{15}$ lower bound without and with landscape elevation data (DEM) in a 30° sector into the direction of the reanalysis wind.
-Adding DEM data reduces the error noticably.
-Note that the MAPE is worse for both the lower lower bounds and the higher ones, but better for the intermediate ones.}
+  \caption[Model results for different wind speed limits]{MAPE for each ws$_{15}$ lower bound without and with elevation data (DEM) in a 30° sector into the direction of the reanalysis wind.}
   \label{table:results}
@@
 
+Table \ref{table:results} shows that adding DEM data reduces the error noticably. Note that the MAPE is worse for both the lower lower bounds and the higher ones, but better for the intermediate ones.
+
 \section{Results for specific locations}
 In this section, the distribution of MAPE by station will be considered.
-For each station the primary model MAPE was computed using just the ws$_{15} \geq 10$ data for that station was calculated and the results are shown in Figure \ref{fig:errorMap}.
+For each station the primary model MAPE was computed using just the ws$_{15} \geq 10 m/s$ data for that station and the results are shown in Figure \ref{fig:errorMap}.
 The stations with low performance are all near mountains and often in fjords.
-The worst cases are in Vestfirðir and Austfirðir, and several bad stations are e.g.
+The worst cases are in Vestfirðir and Austfirðir, and several low performing stations are e.g.
 in Tröllaskagi and Snæfellsnes, and near Öræfajökull, Esja, Eyjafjallajökull.
@@
 \begin{table}[H]
-    \caption[Primary model MAPE by selected stations]{Constant baseline and primary model MAPE for a few selected stations.
-High MAPE values is seen in fjords near tall mountains in Austfirðir (Fáskrúðsfjörður and Seyðisfjörður), intermediate values at the foothills of mountains (Ingólfsfjall and Kjalarnes), and low values in flatter places (Sandskeið, Þrengsli).}
+    \caption[Primary model MAPE by selected stations]{Constant baseline and primary model MAPE for a few selected stations.}
     \label{table:more_specific_sites}
@@
         Station name & \multicolumn{2}{c}{MAPE}\\
-         & Baseline & Model\\
+         & Baseline & Primary Model\\
         \midrule
@@
 The worst performing station is Seljalandsdalur in Vestfirðir while the best performing station is Garðskagaviti at the tip of the Reykjanes peninsula.
-Evidently rugged landscape makes it difficult to estimate the gust factor accurately.
+Evidently complex landscape makes it difficult to estimate the gust factor accurately.
+High MAPE values is seen in fjords near tall mountains in Austfirðir (Fáskrúðsfjörður and Seyðisfjörður), intermediate values at the foothills of mountains (Ingólfsfjall and Kjalarnes), and low values in flatter places (Sandskeið, Þrengsli).
 
@@
     \caption[Model result looking at closed wind speed intervals]{The MAPE results for different wind speed limit intervals.
-Here instead of training for all data above a certain threshold of wind speed, training is done only on data between two wind speeds.
-The percentage variance in gust factor as a function of wind speed increases with decreasing wind speed.
-Measured wind speed is used for the cutoff and thus have data leakage.
-These results should thus be somewhat comparable to the last column in Table \ref{table:results}}
+Here instead of training for all data above a certain threshold of wind speed, training is done only on data within a wind speed interval.}
     \label{table:closed_intervals}
@@
 
+The percentage variance in gust factor as a function of wind speed increases with decreasing wind speed.
+Measured wind speed is used for the cutoff and thus have data leakage.
+The results in Table \ref{table:closed_intervals} should thus be somewhat comparable to the last column in Table \ref{table:results}
+
 \begin{figure}
@@
 Computation of these values is expensive so subset of data is used ($\sim$1000 randomly selected data rows).
-There is an obvious outlier in the figure, making it difficult to assess the distribution.
+There is an obvious outlier in the figure for the Richardson number, making it difficult to assess the distribution.
 A second Shapley summary was computed with more data rows ($\sim$10000) and outliers excluded, see the lower panel of the figure.
@@
 
-Looking specifically at the lower panel in Figure \ref{fig:ShapleySummary}, the station elevation (altitude) is the most influential and the Richardson number's influence is very low, with the other features falling in between.
-For most features the Shapley values are bunch up around 0, while the station elevation, and to some extent ws$_{15}$ have a wider distribution.
-Station altitude/elevation seems to have a wide flat part between –0.1 and 0.1.
+Looking specifically at the lower panel in Figure \ref{fig:ShapleySummary}, the station elevation (elevation) is the most influential and the Richardson number's influence is very low, with the other features falling in between.
+For most features the Shapley values are bunched up around 0, while the station elevation, and to some extent ws$_{15}$ have a wider distribution.
+Station elevation/elevation seems to have a wide flat part between –0.1 and 0.1.
 Overall, the values seem to have a longer right tail.
-This is expected as gust factors are positive ($\sim$1.2 to 2).
+This is expected as gust factors are always positive.
 Notice also that Ri has virtually no contribution and could safely be removed from the model.
@@
 The coloring of the points in the displayed distributions show the corresponding feature values.
-For example, in the lower panel of Figure \ref{fig:ShapleySummary}, high station altitude/elevation corresponds to negative contribution and vice versa, and same for $p_{15}$.
+For example, in the lower panel of Figure \ref{fig:ShapleySummary}, high station elevation/elevation corresponds to negative contribution and vice versa, and same for $p_{15}$.
 For ws$_{15}$ and $t_{15}$ the relationship is opposite, with high feature values corresponding to positive contribution.
 When the coloring is more mixed the contribution of the feature is more mixed.
-For example, high twd might have a positve contribution at some locations but negative elsewhere.
+For example, high twd might have a positive contribution at some locations but negative elsewhere.
 
@@
 Coloring shows the respective feature value, blue for low and red for high.
-For example high station altitude/elevation corresponds to negative contribution and vice versa, and for ws$_{15}$ the relationship is opposite, with high feature values corresponding to positive contribution.
-It can be seen that generally multiple factors influence the prediction, with the station elevation being highly influential.
-There is one outlier in the upper panel, for the Richardson number, which otherwise has very little influence.
 The upper panel is based on $\sim$1000 data rows.
-The lower panel is based on more data rows ($\sim$10000), and furthermore, outliers have been removed there to make the plot more legible.}
+The lower panel is based on more data rows ($\sim$10000), outliers have been removed there to make the plot more readable.
+The ordering of the features shows their average absolute contribution.
+The feature displayed at top of plot has the highest contribution to predictions and the one at the bottom has the lowest.}
   \label{fig:ShapleySummary}
@@
 
-Figure \ref{fig:ShapleySummary3} in Appendix A shows a Shapley summary plot for all the data rows. However this plot is detrimentally affected outliers.
+In Shapley summary plots, high station elevation/elevation corresponds to negative contribution and vice versa, and for ws$_{15}$ the relationship is opposite, with high feature values corresponding to positive contribution.
+It can be seen that generally multiple factors influence the prediction, with the station elevation being highly influential.
+There is one outlier in the upper panel, for the Richardson number, which otherwise has very little influence.
+Figure \ref{fig:ShapleySummary3} in Appendix A shows a Shapley summary plot for all the data rows. However this plot is detrimentally affected by outliers.
 
@@
 In this subsection two such examples are considered, for Akrafjall and Keflavíkurflugvöllur.
-It is noteworthy to see that the constant value of a stations altitude produces quite different contributions to the predictions, presumable because of diffent interactions with other features.
+It is noteworthy to see that the constant value of a stations elevation produces quite different contributions to the predictions, presumable because of diffent interactions with other features.
 For (very) simple models, such as linear regression without cross terms, this would not happen.
@@
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_1350.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Keflavíkurflugvöllur.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Keflavíkurflugvöllur.
-This seems to show the same distribution as previous summary plots.
-Station elevation is influential and Richardson number has no impact.}
+    \caption[Shapley Summary plot at Keflavíkurflugvöllur.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at data from Keflavíkurflugvöllur.}
     \label{fig:ShapleySummaryKeflavikurflugvollur}
@@
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_35553.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Almannaskarð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Almannaskarð.
-This seems to show the same distribution as previous summary plots.
+    \caption[Shapley Summary plot at Almannaskarð.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at data from Almannaskarð.
 Station elevation is influential and Richardson number has no impact.}
@@
     \includegraphics[scale = 0.6]{Figures/shap_plots/summary_plot_31572.png}
-    \caption[Summary feature importance of a neural network only looking at AWS at Akrafjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
-This plot only looks at datapoints from Akrafjall.
-This seems to show the same distribution as previous summary plots.
-Station elevation is influential and Richardson number has no impact.}
+    \caption[Shapley Summary plot at Akrafjall.]{Feature importance of a neural network with model architecture as described in Table \ref{table:gridSearchHyperparameters} and data as described in Table \ref{table:trainDataExample}.
+This plot only looks at data from Akrafjall.}
     \label{fig:ShapleySummaryAkrafjall}
@@
 
-Further examples of Shapley summary plots for individual stations may be found in Appendix A.
+These plots seem show the same distribution.
+Station elevation is influential and Richardson number has no impact.Further examples of Shapley summary plots for individual stations may be found in Appendix A.
 
-It is important to note that Shapley summary plot analysis assumes feature independence, because only one feature at a time is changed \cite{Salih_2024}.
+It is important to note that Shapley summary plot analysis assumes feature independence, because only one feature at a time is changed \parencite{Salih_2024}.
 This might explain why the impact of the Richardson number is so low.
@@
 Another point to note is that the features are ordered by their impact.
-The station altitude is the most impactful for all the plots except Keflavíkurflugvöllur (Figure \ref{fig:ShapleySummaryKeflavikurflugvollur}), but the ordering of other variables changes.
+The station elevation is the most impactful for all the plots except Keflavíkurflugvöllur (Figure \ref{fig:ShapleySummaryKeflavikurflugvollur}), but the ordering of other variables changes.
 Looking at the lower panel of Figure \ref{fig:ShapleySummary}, the wind speed is the second most important feature.
 This is reversed in Figure \ref{fig:ShapleySummaryAlmannaskarð}.
-This is not unexpected as the Almannaskarð station was specifically selected as a station with high MAE between observed and reanalysis wind speed (Table \ref{table:station_mae_distribution}.
-Another station with high MAE is Akrafjall, with a Shapley summary plot in Figure \ref{fig:ShapleySummaryAkrafjall}).
+This is not unexpected as the Almannaskarð station was specifically selected as a station with high MAE between observed and reanalysis wind speed (Table \ref{table:station_mae_distribution}).
+Another station with high MAE is Akrafjall, with a Shapley summary plot in Figure \ref{fig:ShapleySummaryAkrafjall}.
 
@@
 First, the reduction in MAPE resulting from the addition of the CARRA gust factor to the baseline ws$_{15}$, secondary, and primary NN models of Section 4.2 was assessed, using only data with CARRA gust available.
-The results of this comparison are shown in Table \ref{table:carra_gust}
+The results of this comparison are shown in Table \ref{table:carra_gust}.
 
@@
   \centering
-  \caption[Results adding CARRA gust factor]{Results of adding the CARRA gust factor as a variable in selected models from Sections 4.2 and 4.3.}
+  \caption[Results adding CARRA gust factor]{Results of adding the CARRA gust factor as a variable in selected models from Sections \ref{sec:ws15_10} and \ref{sec:ws15_lower_bounds}.}
   \label{table:carra_gust}
@@
 
-The causes of the differences between the numbers in Table \ref{table:carra_gust} and those of Tables 4.2 and 4.3 could be different time period and/or unlike training setup.
+The causes of the differences between the numbers in Table \ref{table:carra_gust} and those of Tables \ref{table:results} and \ref{table:setsOfParams} could be different time period and/or unlike training setup.
 
@@
 The conclusion that can be drawn from this investigation is that there is a modest information in the CARRA wind gust forecast, potentiially reducing model MAPE by 0.1 to 0.3 percentage points.
-This reduction is however quite small compared with the overall MAPE reduction of $\sim$5 percentage points seen in Table 4.2 when all available variables are added.
+This reduction is however quite small compared with the overall MAPE reduction of $\sim$5 percentage points seen in Table \ref{table:setsOfParams} when all available variables are added.

### 2025-05-13 Chapters/Chapter4.tex
Updated after comments by GNP

@@
 The best results appear in the 10--25~m/s range.
-At lower reanlysis wind speeds, percentage errors inflate even when the differences between wind gust and average wind speed remain small and at both at the higher and the lower speeds the base line MAPE is higher.
+At lower reanalysis wind speeds, percentage errors inflate even when the differences between wind gust and average wind speed remain small and at both at the higher and the lower speeds the base line MAPE is higher.
 
 Although one might expect staility indicators, such as the Richardson number and the Brunt--Väisälä frequency, to be among the strongest predictors of gust variability, they contributed less than other variables.
-The influence of individual features on model performance is illustrated by Shapley plots (Figure \ref{fig:ShapleySummaryKeflavikurflugvollur} and Appendix \ref{appendix:A}).
-According the Shapley analysis, stations altitude, mean wind speed and wind direction are among the more important features for predicting the gust factor, and also, to some extent, the Brunt--Väisälä frequency ($N^2$).
+The influence of individual features on model performance is illustrated by Shapley plots (Figures \ref{fig:ShapleySummary}, \ref{fig:ShapleySummaryKeflavikurflugvollur}, \ref{fig:ShapleySummaryAlmannaskarð}, \ref{fig:ShapleySummaryAkrafjall} and figures in Appendix \ref{appendix:A}).
+According to the Shapley analysis, stations elevation, mean wind speed and wind direction are among the more important features for predicting the gust factor, and also, to some extent, the Brunt--Väisälä frequency ($N^2$).
 As explained in Section \ref{sec:shapley} it was infeasible to evaluate the DEM data with Shapley values.
@@
 Finally, treating the data as a time series may offer further gains.
-Here, only instantaneous inputs at forecast time were used, partly because CARRA fields are available every three hours and may not capture rapid stability changes.
+Here, only instantaneous inputs at forecast time were used, partly because CARRA analysis fields are available every three hours and may not capture rapid stability changes.
 Access to higher-frequency data or the inclusion of temporal history could enable models to exploit evolving atmospheric conditions and further enhance gust-prediction skills.
@@
 \section{Possible applications}
-Wind gusts at weather stations are displayed on road signs in Iceland, if wind gusts exceed \href{https://www.vegagerdin.is/vegagerdin/starfsemi/frettir/tvo-ny-vedur-og-upplysingaskilti-i-notkun#}{expected values}.
+Wind gusts at weather stations are displayed on road signs in Iceland, when wind gusts exceed \href{https://www.vegagerdin.is/vegagerdin/starfsemi/frettir/tvo-ny-vedur-og-upplysingaskilti-i-notkun#}{certain values}.
 The model could be used to allow such predicitons of wind gusts to be displayed for every section of the road, not just at the stations.
-In 2023, outside of Reykjavik harbor, a cruise ship nearly \href{https://www.mbl.is/200milur/frettir/2024/04/26/strand_skemmtiferdaskips_kallar_a_erlenda_adstod/}{ran aground} due to an unexpected wind gust.
+In 2023, outside of Reykjavik harbor, a cruise ship nearly ran aground due to an unexpected wind gust \parencite{strand_skemmtiferdaskips}.
 This wind gust of \textasciitilde 26 m/s, pushed the ship into a buoy near the island Viðey.

### 2025-05-13 MSc.tex
Updated after comments by GNP

@@
 
-\usepackage{biblatex}  % BibLaTeX used for references. 
+\usepackage[style=apa, backend=biber]{biblatex}  % BibLaTeX used for references. 
 \usepackage{csquotes} % BibLaTex wants to have context sensitive quotes
@@
 Whether it is wind farms, traffic accidents, high-voltage towers, or other things affected by wind, accurately predicting wind loads is crucial for reducing damage, improving operational safety, and boosting productivity.
-Wind gusts are multi-second spikes in speed, and wind gust factor is defined as the ratio of the highest three-second gust to the ten-minute mean wind speed.
+Wind gusts are spikes in wind speed, and wind gust factor is defined as the ratio of the highest gust to the mean wind speed over a given time period.
 It is important to be able to predict both wind and gust factor accurately.
 Historically, numerical weather prediction models based on physical equations have been used to predict such variables.
-In recent years, artificial intelligence has advanced rapidly, and AI-based models have begun to rival conventional model performance.
+In recent years, artificial intelligence (AI) has advanced rapidly, and AI-based models have begun to rival conventional model performance.
 This project combines both approaches by using atmospheric model reanalysis output and digital elevation data as inputs to a deep neural network.
@@
 \section*{Útdráttur}
-Hvort sem um er að ræða vindmyllugarða, umferðarslys, háspennumöstur, eða aðra hluti sem vindur getur haft áhrif á er spágeta um álag vegna vindsins mikilvæg, til að minnka eða koma í veg fyrir tjón, auka rekstraröryggi, eða bæta framleiðni.
-Vindhviður eru nokkurra sekúndna aukning á vindi og hviðustuðull mælir hæsta hlutfall milli þriggja sekúndna hviðu og 10 mínútna meðalvinds.
+Hvort sem um er að ræða vindmyllugarða, umferðarslys, háspennumöstur, eða annað sem vindur getur haft áhrif á er spágeta um álag vegna vinds mikilvæg, til að minnka eða koma í veg fyrir tjón, auka rekstraröryggi, eða bæta framleiðni.
+Vindhviður eru toppar í vindhraða og hviðustuðull er hlutfall milli hæstu vindhviðu og meðalvinds yfir gefið tímabil.
 Góðar hviðuspár eru mikilvægar.
@@
 Djúpt tauganet er búið til með því að velja breytur úr endurgreiningargögnunum og hæðarlíkaninu og afleiddum breytum.
-Nokkur líkön af þessu tagi eru svo þjálfuð og niðurstöður bornar saman við einfalt grunnlíkan sem fengið er með línulega aðhvarfi.
+Nokkur líkön af þessu tagi eru svo þjálfuð og niðurstöður bornar saman við einfalt grunnlíkan sem fengið er með línulegu aðhvarfi.
 Niðurstöðurnar lofa mjög góðu um að þessi aðferðafræði geti bætt bæði samtímamat á hviðustuðli og spár um hann umtalsvert.
@@
   \acro{AWS}{Automatic Weather Stations}
-  \acro{CARRA}{Copernicus Artic Regional ReAnalysis dataset}
+  \acro{CARRA}{Copernicus Artic Regional ReAnalysis}
   \acro{CNN}{Convolutional Neural Networks}
@@
   \acro{GPU}{Graphical Processing Unit}
-  \acro{HRES}{High Resolution forecas}
-  \acro{IMO}{Icelandic Meteoroligcal Office}
+  \acro{GRIB}{GRIdded Binary}
+  \acro{HRES}{High Resolution forecast}
+  \acro{IMO}{Icelandic Meteorological Office}
   \acro{IRCA}{Icelandic Road and Coastal Administration}

### 2025-05-13 Chapters/Chapter2.tex
Updated final image. Think this is done.

@@
   \centering
-  \includegraphics[scale=0.6]{Figures/MAEoverIceland.png}
-  \caption[MAE distribution by station]{Distribution of mean absolute error (MAE) by station for all observed wind speeds. Image is missing units for MAE, which is in m/s.}
+  \includegraphics[scale=0.7]{Figures/mae_by_stations.png}
+  \caption[MAE distribution by station]{Distribution of mean absolute error (MAE) by station for all observed wind speeds.}
   \label{fig:station_mae_distribution}
